### 2019-02-01 popSimManu.tex
added stub for manuscript

\documentclass[12pt,halfline,a4paper]{ouparticle}
% * 
%
% ^.
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{graphicx,rotating}
\usepackage{dsfont}

\newcommand{\Est}[1]{\hat{#1}}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

\begin{document}

\title{PopSim: towards a standardized database of population genetic simulations}

\author{%
\name{The PopSim Consortium}
\address{}
\email{}}

\abstract{The explosion in population genomic data data demands ever increasingly complex modes of analysis. While
the field has met this challenge to date, we have not yet created field-wide standards by which to measure the performance
and accuracy of new tools. Here we describe a new resource, the PopSim database, that attempts to rectify this situation. This
resource contains XXX...}

\date{\today}

\keywords{Population genetics, Simulation, Inference}

\maketitle

\section*{Introduction}

\section*{Methods and Model}

\section*{Results}

\section*{Discussion}

\end{document}

### 2019-11-07 popSimManu.tex
moved text from google doc to latex doc

@@
 \section*{Introduction}
+While population genetics has always used statistical methods to make inferences from data,
+the degree of sophistication of the questions, models, data, and computational approaches
+used have all increased over the past decade. Currently there exist myriad computational methods
+that can infer the histories of populations, the distribution of fitness effects,
+recombination rates, mutational biases, the extent of positive selection from single genomes,
+genotype data, or full genome sequence data (XX some generous citations XX).
+While en masse these methods have increased our understanding of the
+impacts of genetic and evolutionary processes, very little has been done to systematically
+benchmark the quality of inferences gleaned from computational population genetics. Here we
+describe the popSim consortium project, an effort aimed at creating systematic simulation
+standards across a variety of common study organisms that will empower rigorous comparison
+of methods to usher in the next era of genetics data analysis.
 
-\section*{Methods and Model}
+Many population genetic inference methods often make different assumptions as
+well as use different summaries of the genetic variation data, even when
+attempting to address the same biological question. For example, there are at
+least X methods to infer population size changes. The MSMC method allows for the
+population to continuously change in size over time. Other popular methods, like
+Dadi, typically allow for a smaller number of size changes to occur at certain
+points throughout the population's history. Further, the types of data that the
+methods use also differs. MSMC uses 1-4 whole genome sequences and leverages the
+spatial patterns of genetic variation along the sequence. Dadi instead uses the
+site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
+sample, ignoring any information about the correlation structure (linkage
+disequilibrium) among SNPs.
+
+The degree to which modeling assumptions and using different summaries of the
+data can affect the ultimate results has been challenging to assess. Yet, there
+have been several clear examples of different methods yielding fundamentally
+different conclusions. For example, MSMC methods applied to human genomes have
+suggested large ancient (>100 years ago) ancestral population sizes and
+bottlenecks that have not been detected by SFS-based methods (see Beichman et
+al. 2017). These distinct models differ in how they fit various summaries of the
+genetic variation data, suggesting that modeling choices can greatly affect the
+performance of the inference.
+
+Presumably, specific methods will perform better under certain scenarios.
+However, because these conditions are typically unknown, empirical researchers
+lack principled guidelines for deciding which statistical method is best suited
+for to accurately answer their particular question. The need for empirical
+guidance will only increase as researchers continue to generate more genomic
+variation data from non-model taxa and wish to make inferences regarding
+demography and selection.
+
+One way to benchmark different methods for statistical inference in population
+genetics is to apply the methods to simulated datasets where the true model
+parameters are known. Methods that perform well will accurately infer these
+parameters and will also provide reasonable estimates of uncertainty of the
+estimated parameters. Indeed simulations studies are a standard ingredient in
+most papers proposing new methods in population genetics. However, these
+simulation studies typically do not completely compare the performance of
+different statistical methods. As such, they fall short of being able to provide
+practical guidelines to empirical researchers. Standard simulations usually fall
+short because they often focus on a part of the parameter space where the new
+method is likely to perform well. For example, XXXX. Secondly, the degree to
+which different methods are compared can be quite variable.
+
+One reason why simulation studies do not compare the performance of many methods
+is that these simulation studies are challenging to carry out. There are several
+practical impediments to carrying out the ideal benchmarking simulation studies.
+First, while there are many published population genetic models available, their
+parameters are often not easily accessible, often presented in supplemental
+tables. Implementing these parameters into usable commands for simulations can
+be a time-consuming and error prone process. Further, often papers present many
+different models, and different authors may choose to select distinct models,
+making comparison difficult. While improvements in simulation software and ever
+faster computers have partially emialoritated this concern, generating the
+simulated datasets stil can still be a time and computing resource bottleneck.
+
+For these reasons, we have generated a standard resource of simulated datasets
+under a variety of population genetic models models to enable benchmarking of
+new population genetic methods. We term this new resource, the PopSim project.
+It includes…… Can be accessed at XXXX>
 
 \section*{Results}
+\paragraph{The PopSim Resource}
+To address these problems, we developed the stdpopsim package, a community
+maintained library of empirical genome data and population genetics simulation
+models, providing easy access to realistic simulations.  The package organised
+by species, initially consisting of humans, drosophila and arabadopsis. A
+species definition consists of two key elements.  Firstly, the library defines
+some basic information about its genome, with information about chromosome
+lengths, average mutation rates and so on. We also provide access to detailed
+empirical information such as recombination maps, which model empirically
+observed heterogeneity along chromosomes. As such maps are often large and hard
+to find, we provide a mechanism to automatically download (and cache locally) a
+set of known maps. For example, in humans we support the HapMapII (XX cite XX) and
+Decode (XX cite XX) genetic maps. The second key element of a species description
+within stdpopsim is a set of carefully curated population genetic model
+descriptions from the literature, which are the basis for accurate simulations
+of specific historical scenarios.
+
+\paragraph{Figure 1 about here}
+
+Given the genome data and simulation model descriptions defined within the
+library, it is then straightforward to run accurate, standardised simulations
+across a range of organisms. Stdpopsim has a Python API and a user-friendly
+command line interface, allowing users with minimal experience direct access to
+state-of-the-art simulations. Simulations are output in the “tree sequence”
+format (Kelleher, Etheridge, and McVean 2016; Kelleher et al. 2018, 2019), which
+contains complete genealogical information about the simulated samples, is
+extremely compact, and can be processed very efficiently using the tskit library
+(Kelleher, Etheridge, and McVean 2016; Kelleher et al. 2018). Currently,
+stdpopsim uses the  msprime coalescent simulator (Kelleher, Etheridge, and
+McVean 2016) as the simulation engine. We plan to include other simulation
+engines such as SLiM (Haller et al. 2019);(Haller and Messer 2019) to support
+scenarios that cannot be modelled under the coalescent framework.
+
+The stdpopsim command line interface, by default, outputs citation information
+for the models, genetic maps and simulation engines used in any particular run.
+We hope that this will encourage users to appropriately acknowledge the
+resources used in published work, and authors
+of simulation models to contribute to our ongoing community-driven development process.
+
+\subsection*{Single Population demographic models}
+We begin by simulating data under previously published demographic models for
+humans, Drosophila, and Arabidopsis. These models consider the history of only a
+single population at a time,and  were inferred using a variety of different
+methods.
+
+For humans, we implemented a simplified version of the Tennessen et al. (2012)
+model with only the African population modeled (expansion from the ancestral
+population and recent growth).
+
+For Drosophila melanogaster we implemented the model inferred in Sheehan & Song
+(2016) which covered three epochs: ancestral, bottleneck, and re-expansion, each
+with constant within-epoch population size.
+
+For Arabidopsis thaliana, we implemented the model from in Durvasula et al.
+(2017) inferred using MSMC-8. This model includes a continuous change in
+population size over time, rather than pre-sepceciified epochs of different
+population sizes.
+
+\subsection*{Performance of methods to infer demographic history for single
+populations}
+To evaluate and compare the performance of inference analyze methods for single
+population size history, we use Snakemake (cite) to create a pipeline to run the
+comparing inference methods. Concretely, we compared msmc, smc++, and
+stairwayplot (cite all) on simulated genomes sampled from a single population,
+in each of the stdpopsim models described above. The pipeline was generalized to
+run R replicates with C chromosomes for N samples, for a total of R x C
+simulations for each demographic model. After simulation, the pipeline preparess
+input files for each of the respective inference methods by grouping all
+chromosomes, for each sample, into a single file. This step results in three
+input files, for each simulation replicate R, pertaining to each of the
+respective inference methods and derived from the same simulated tree sequences.
+Each of the inference programs are then run in parallel, followed by plotting of
+Ne estimates from each program.
+
+(Conclusion putting everything together for single population models--Kirk will
+fill in after the other parts are here)
+
+\subsection*{Multi Population demographic models}
+Stdpopsim implements three human multi-population demographic models: Gutenkunst
+et al. (2009), Tennesen et al. (2012) , and Browning et al. (2018). Gutenkunst
+et al. modeled the out of Africa event as well as the subsequent divergence of
+the European and Asian populations based on data from the Environmental Genome
+Project (citation needed). Tennessen et al. fit a two population variant of the
+Gutenkunst model which did not include Asian populations, where they focused on
+modeling recent human population growth in European and African populations to
+explain the presence of rare variants in exome sequencing data. Browning et al.
+modeled an admixed American population jointly with its ancestral African,
+European, and Asian populations based on data from the Hispanic Community Health
+Study/Study of Latinos and the African-American and European-American
+populations from Memphis and Pittsburgh (citation needed). Together these models
+contain features found in real data (e.g. bottlenecks, population growth,
+admixture) which are pertinent in the context of method development.
+
+For Drosophila we implemented the model from Li & Stephan (2006) which included
+an ancestral population in Africa, then a population expansion with a population
+split and bottleneck into a European population.
+
+\subsection*{Performance of methods to infer demographic history for multiple
+populations}
+
+Outlining for now since this is not entirely finalized yet Compared performance
+of multi-population inference methods by running simulations under
+multi-population models for humans and Drosophila and inferring simulated
+parameters using dadi and fastsimcoal (and smc++???) For simplicity, conducted
+inference in dadi and FSC using an IM model with constant population sizes and
+bi-directional migration For humans, this meant that, although we simulated
+under the three-population Gutenkunst model, samples were taken only from the
+African and European populations Since inferred model did not match model used
+for simulations, compared inferred parameters to parameters expected based on
+mean coalescence time in simulated models (Fig 3) In general, used default
+settings and did not attempt to optimize dadi or fsc runs Used Snakemake to
+manage pipeline for two population analysis Results from two population pipeline
+demonstrate that the different methods tested were generally able to recover the
+simulated parameters Although we present comparison of results from several
+competing methods, our aim at this stage is not to provide an exhaustive
+evaluation of these methods Our hope is that future work with this pipeline will
+enable more detailed exploration of the strengths and weaknesses of the numerous
+multi-population inference methods that are available to the population genetics
+community
 
@@
 
+\section*{Methods}
+\subsection*{Generating the PopSim resource}
+
+\subsection*{Model QC procedures}
+Each demographic model was implemented independently by two people working from
+the relevant primary sources. Disagreements between complete implementations
+were resolved between the implementers in consultation with the authors of the
+source papers when necessary. A comparison of the two implementations is then
+added as a unit test.Technical details of the QC process can be found in the
+GitHub repository in the developer documentation.
+
+\subsection*{Workflow for analysis of simulated data}
+Snakemake pipeline for 1 pop models
+Snakemake pipeline for 2 pop models
+
+\section*{Extra text}
+[This section describes the stdpopsim Python library. What it’s for, how it’s used and what it can do]
+[First para: running accurate simulations is hard]
+Running realistic population genetic simulations is a complex task today. It is
+straightforward to run simple simulations of population models such as the
+coalescent or Wright-Fisher using (e.g.) msprime or SLiM, but the process of
+instantiating these models to accurately reflect the properties of natural
+populations that are important for a particular study is fraught with
+difficulties. There are typically three important inputs that must be decided:
+the demographic model, recombination and mutation rates. Demographic models for
+a particular species are typically chosen by examining the literature for
+published model specifications, which can consist of dozens of parameters. These
+parameters must be transcribed from the original publication and translated into
+the input format required for the simulator in question. This is a difficult and
+error prone task, and mistakes are common. Recombination and mutation rates are
+simpler to describe and often a single genome-wide estimate is appropriate.
+However, such estimates are updated frequently and there is wide variation in
+the values used for the simulations of the same species. When more fine-grained
+simulations of the local chromosome structure is required, empirical
+recombination and mutation rate maps are used. However, there is no standard
+format for such maps, and finding the maps can be challenging.
+
 \end{document}

### 2019-12-03 popSimManu.tex

--- 2019-02-01 popSimManu.tex
@@
 \newcommand{\Est}[1]{\hat{#1}}
+
 \newcommand{\beginsupplement}{%
@@
      }
+\newcommand{\stopsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{\arabic{figure}}%
+     }
+
+% method names
+
+\newcommand{\stdpopsim}{\texttt{stdpopsim} }
+\newcommand{\dadi}{\texttt{dadi} }
+\newcommand{\MSMC}{\texttt{MSMC} }
+\newcommand{\smcpp}{\texttt{smcpp} }
+\newcommand{\stairwayplot}{\texttt{stairwayplot} }
+\newcommand{\fastsimcoal}{\texttt{fastsimcoal} }
+
+\newcommand{\adk}[1]{\textcolor{red}{ADK: #1}}
+
+\usepackage[colorinlistoftodos]{todonotes}
 
@@
 
-\title{PopSim: towards a standardized database of population genetic simulations}
+\title{\stdpopsim: a community-supported toolkit for population genetic simulation}
 
@@
 
-\abstract{The explosion in population genomic data data demands ever increasingly complex modes of analysis. While
-the field has met this challenge to date, we have not yet created field-wide standards by which to measure the performance
-and accuracy of new tools. Here we describe a new resource, the PopSim database, that attempts to rectify this situation. This
-resource contains XXX...}
+\abstract{The explosion in population genomic data demands ever increasingly complex modes of analysis.
+More and more these analyses depend on sophisticated, often computationally expensive simulation. 
+To date, computational genetics researchers have been often forced to re-implement such simulations
+independently, leading to both wasted time in the duplication of effort and reductions in reproducibility.
+Population genetics, as a field, specifically lacks rigorous standard benchmarks by which new tools for inference might 
+be measured. Here we describe a new resource, the PopSim library, that attempts to rectify this situation. 
+PopSim is a community-driven, open source project, aimed at revolutionizing the ease and reproducibility
+of population genetic simulation for a catalog of published demographic models from both humans and non-model organisms. 
+PopSim assures correctness in our implementations through a mature quality control pipeline that is
+community-based. Here we describe the PopSim resource as a step towards inviting an even broader
+community of developers to participate and demonstrate some rather simple examples of how one
+might use PopSim for comparison of methods of demographic inference.}
 
@@
 \section*{Introduction}
+While population genetics has always used statistical methods to make inferences from data,
+the degree of sophistication of the questions, models, data, and computational approaches
+used have all increased over the past decade. Currently there exist myriad computational methods
+that can infer the histories of populations, the distribution of fitness effects,
+recombination rates, the extent of positive selection in genome sequence data 
+(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
+chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
+kern2018diplos,sugden2018localization}).
+While en masse these methods have increased our understanding of the
+impacts of genetic and evolutionary processes, very little has been done to systematically
+benchmark the quality of inferences gleaned from computational population genetics.
+This is in large part because ``ground-truth'' in population genetics comes from simulation,
+not from empirical observations, and as a field there has been no agreement as to what should
+constitute community standards or best practices. The general modus operandi to date has been to 
+introduce methods which are validated by a set of bespoke simulations produced by individual
+groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
+to decreases in reproducibility and transparency across the entire field.
+Here we describe the PopSim consortium project, a community-driven catalog
+of standardized simulations across a variety of common study organisms
+that will empower the field and usher in the next era of genetics data analysis.
 
-\section*{Methods and Model}
+Benchmarking method performance population genetics is not straightforward.
+Many population genetic inference methods often make different assumptions as
+well as use different summaries of the genetic variation data, even when
+attempting to address the same biological question. For example, there are
+numerous methods to infer population size changes.
+The \MSMC method (\cite{schiffels2014inferring}) allows for the
+population to continuously change in size over time. Other popular methods, like
+\dadi (\cite{gutenkunst2009inferring}), typically allow for a smaller number of size changes to occur at certain
+points throughout the population's history. Further, the types of data that the
+methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
+spatial patterns of genetic variation along the sequence, \dadi  instead uses the
+site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
+sample, ignoring any information about the correlation structure (linkage
+disequilibrium) among SNPs. Thus from the start these methods may not be 
+expected to perform equally well in all scenarios. \adk{I think we are focussed 
+a bit too much on benchmarking currently in the intro. Could be good to add more on efficiency
+and reproducibility}
+
+The degree to which modeling assumptions and using different summaries of the
+data can affect the ultimate results has been challenging to assess. Yet, there
+have been several clear examples of different methods yielding fundamentally
+different conclusions. For example, \MSMC methods applied to human genomes have
+suggested large ancient (>100 years ago) ancestral population sizes and
+bottlenecks that have not been detected by SFS-based methods (see 
+\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
+genetic variation data, suggesting that modeling choices can greatly affect the
+performance of the inference. Further it is reasonable
+to assume generally that specific methods will perform better than others under certain scenarios.
+However, because these conditions are typically unknown, empirical researchers
+lack principled guidelines for deciding which statistical method is best suited
+for to accurately answer their particular question. The need for empirical
+guidance will only increase as researchers continue to generate more genomic
+variation data from non-model taxa and wish to make inferences regarding
+demography and selection.
+
+One way to benchmark different methods for statistical inference in population
+genetics is to apply the methods to simulated datasets where the true model
+parameters are known. Methods that perform well will accurately infer these
+parameters and will also provide reasonable estimates of uncertainty of the
+estimated parameters. Indeed simulations studies are a standard ingredient in
+most papers proposing new methods in population genetics. However, these
+simulation studies are often specifically devised to showcase the performance
+of the novel method and may not adquately compare performance to competing 
+methods. Moreover researchers are generally forced to ``reinvent the wheel''
+with respect to simulations, and are forced to reimplement previously
+inferred scenarios independently. This can lead to errors being incorporated
+during the implementation of the simulations. Lastly, the degree to
+which different methods are compared can be quite variable.
+
+One reason why simulation studies do not compare the performance of many methods
+is that these simulation studies are challenging to carry out. There are several
+practical impediments to carrying out the ideal benchmarking simulation studies.
+First, while there are many published population genetic models available, their
+parameters are often not easily accessible, often presented in supplemental
+tables. Implementing these parameters into usable commands for simulations can
+be a time-consuming and error prone process. Further, often papers present many
+different models, and different authors may choose to select distinct models,
+making comparison difficult. While improvements in simulation software and ever
+faster computers have partially ameliorating this concern, generating the
+simulated datasets stil can still be a time and computing resource bottleneck.
+
+For these reasons, we have generated a standardized, community-driven resource 
+for simulation of published demographic models from a number of popular study systems.
+This resource, which we call PopSim, significantly eases the development
+burden of implementing realistic simulations for population genetic analysis. 
+Briefly, PopSim currently is comprised of a catalog of four organism genomes: humans,
+\emph{Drosophila}, \emph{Arabidopsis}, and \emph{E. coli}. Each of these
+genomes is associated with a physical organization (e.g., a chromosome structure), 
+one or more genetic maps, default population-level parameters (mutation rate,
+generation time) and one or more published demographic histories. Through 
+either a command line interface or a mature Python API, the user can specify what
+organism, map, chromosome(s), and history they are interested in simulating, and are returned
+simulation output from their chosen model. PopSim has been developed using an 
+completely distributed open source model, with strong procedures in place 
+to continue its growth and accuracy. Below we describe PopSim and give an
+example of how benchmarking of methods for demographic inference might proceed.
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=0.5\linewidth]{display_items/Figure1.png}
+\caption{\textbf{Structure of \stdpopsim}. \textbf{A.} The
+hierarchical organization of the PopSim catalog contains all information
+within individual species. Information within the species level is shown here
+for \emph{Homo sapiens} only. We see that \emph{H. sapiens} is associated
+with a default generation time and mutation rate, that it contains within
+it a representation of the physical genome, genetic maps, and as shown a 
+subset of the demographic models currently implemented. \textbf{B.} Caching 
+and automated download of genetic maps. \textbf{C.} Example code to specify
+and simulate models using \stdpopsim via the API (top) or
+command-line interface (bottom). \textbf{D.} Simulation output from 
+\stdpopsim is generated as tree sequence files that can
+be manipulated and analysed with \texttt{tskit}.}
+\label{fig:cartoon}
+\end{center}
+\end{figure}
 
 \section*{Results}
+\paragraph{The \stdpopsim library}
+The central contribution from PopSim currently is \stdpopsim, a community
+maintained library of empirical genome data and population genetics simulation
+models, providing easy access to realistic simulations. Figure \ref{fig:cartoon} shows a graphical
+representation of the structure of \stdpopsim. The package centers 
+on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{Drosophila},
+\emph{Arabidopsis}, \emph{E. coli}, and \emph{Pongo}. A species definition consists
+of two key elements.  Firstly, the library defines
+some basic information about its genome, with information about chromosome
+lengths, average mutation rates and so on. We also provide access to detailed
+empirical information such as recombination maps, which model empirically
+observed heterogeneity along chromosomes. As such maps are often large and hard
+to find, we provide a mechanism to automatically download (and cache locally) a
+set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the HapMapII (\cite{international2007second}) and
+Decode (\cite{kong2010fine}) genetic maps. The second key element of a species description
+within \stdpopsim is a set of carefully curated population genetic model
+descriptions from the literature, which are the basis for accurate simulations
+of specific historical scenarios.
+
+Given the genome data and simulation model descriptions defined within the
+library, it is then straightforward to run accurate, standardized simulations
+across a range of organisms. \stdpopsim has a Python API and a user-friendly
+command line interface (Fig. \ref{fig:cartoon}C), allowing users with minimal experience direct access to
+state-of-the-art simulations. Simulations are output in the “tree sequence”
+format (\cite{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}), which
+contains complete genealogical information about the simulated samples, is
+extremely compact, and can be processed very efficiently using the tskit library
+(\cite{kelleher2016efficient,kelleher2018efficient}; Fig. \ref{fig:cartoon}D). Currently,
+\stdpopsim uses the  \texttt{msprime} coalescent simulator (\cite{kelleher2016efficient})
+as the simulation engine. We plan to include other simulation
+engines to support scenarios that cannot be modelled under the coalescent framework,
+and have already implemented \texttt{SLiM} (\cite{haller2019slim}) as
+an alternative back-end.
+
+The \stdpopsim command line interface, by default, outputs citation information
+for the models, genetic maps and simulation engines used in any particular run.
+We hope that this will encourage users to appropriately acknowledge the
+resources used in published work, and authors
+of simulation models to contribute to our ongoing community-driven development process.
+
+\subsection*{\texttt{stdpopsim} Catalog}
+The current contents of the \texttt{stdpopsim} catalog are shown in Table X. At the time of
+writing the PopSim Consortium has implemented previously published demographic models for
+humans, \emph{Drosophila}, and \emph{Arabidopsis}. These include both
+simpler, single population histories (e.g., in \emph{D. melanogaster} \cite{sheehan2016deep}),
+and much more complex models which include population splitting, migration, and archaic
+admixture (e.g., in humans \cite{ragsdale2019models}).
+
+At the time of writing, more models of human history have been implemented than for
+any other organism. These models include: a simplified version of the \cite{tennessen2012evolution}
+model with only the African population modeled (expansion from the ancestral
+population and recent growth), the three population model of \cite{gutenkunst2009inferring}
+which modeled the out-of-Africa bottleneck as well as the subsequent divergence of
+the European and Asian populations, the \cite{tennessen2012evolution} two population variant of the
+Gutenkunst model which did not include Asian populations, but more explicitly modeled 
+recent recent human population growth, the \cite{browning2018ancestry} admixture model
+for American populations that models ancestral African, European, and Asian population
+components, and the three population out-of-Africa model that includes archaic admixture
+from \cite{ragsdale2019models}. Together these models
+contain features found in real data (e.g. bottlenecks, population growth,
+admixture) which are pertinent in the context of method development.
+
+For non-human genomes we have have implemented two demographic histories for
+\emph{Drosophila melanogaster} and one each from \emph{A. thaliana} and \emph{E. coli}.
+For \emph{D. melanogaster} we have implemented the three epoch model estimated from 
+an African sample from \cite{sheehan2016deep} as well as the out-of-Africa divergence
+and associated bottleneck model of \cite{li2006inferring} which jointly models African
+and European populations. For For \emph{Arabidopsis thaliana}, we implemented the
+model from in \cite{durvasula2017african} inferred using \MSMC. This model includes
+a continuous change in population size over time, rather than pre-specified epochs of different
+population sizes.
+
+\subsection*{Use case: comparing methods of demographic inference}
+As an example of the utility of \stdpopsim we demonstrate how popular
+methods can be simply and fairly compared in the setting of demographic
+inference using our resource. Although we present comparison of results from several
+competing methods, our aim at this stage is not to provide an exhaustive
+evaluation of these methods. Our hope is instead that future work built upon this resource
+will enable more detailed exploration of the strengths and weaknesses of the numerous
+inference methods that are available to the population genetics community.
+
+We start by comparing popular methods for estimating
+population size histories ($N(t)$) of single populations and subsequently
+show simple examples of multi-population inference. 
+To evaluate and compare the performance of inference methods we developed
+\texttt{snakemake} (\cite{koster2012snakemake}) workflows that allow reproducibility 
+in comparing inference methods which are available from \url{https://github.com/popgensims/analysis}
+. The workflows allow efficient computing in multicore environments or on 
+cluster environments.
+
+For single-population population size histories, we compared \MSMC, \smcpp, and
+\stairwayplot (\cite{schiffels2014inferring,terhorst2017robust,liu2015exploring})
+ on simulated genomes sampled from a single population,
+in a number of the \stdpopsim models described above. The pipeline was generalized to
+run $R$ replicates with $C$ chromosomes for $N$ samples, for a total of $R \times C$
+simulations for each demographic model. After simulation, the pipeline prepares
+input files for each of the respective inference methods by grouping all
+chromosomes, for each sample, into a single file. This step results in an
+input file for each simulation replicates $R$, pertaining to each of the
+respective inference methods and derived from the same simulated tree sequences.
+Each of the inference programs are then run in parallel, followed by plotting of
+$N_e$ estimates from each program.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
+\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
+\MSMC with $n=2$ and $n=8$. Data were generated by simulating
+replicate human genomes under the \cite{ragsdale2019models} model and using the genetic map
+inferred in \cite{international2007second}. From top to bottom we show estimates for each
+of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
+for the rate of coalescence in the demographic model.}
+\label{fig:n_t_ragsdale}
+\end{center}
+\end{figure}
+
+In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
+run on whole genome simulations under the \cite{ragsdale2019models} model using the
+human genetic map estimated from \cite{international2007second}. On rows
+we show inferred $N(t)$ for the three extant populations in the model. 
+On columns we show comparisons among the methods (note that we compare two differing
+sample sizes for \MSMC). The solid black lines represent the true effective population
+size as inferred from the inverse coalescent rate defined by the model. 
+Blue lines represent estimates from each of three replicate whole genome simulations. 
+While there is variation among methods and individual population estimates,
+generally spoken the methods are accurate for this model of human history.
+
+\stdpopsim allows us to readily compare these estimates to those drawn from a different
+model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
+$N(t)$ from simulations using the same physical and genetic maps, but a different demographic
+history-- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
+of the methods is capturing relevant parts of the population history, although the
+degree of noise at any given time interval varies. In comparing inferences between the
+models it is interesting to note that $N(t)$ estimates for the CHB and CEU
+simulated populations are generally better across methods than estimates from the YRI
+sample.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
+\caption{\textbf{Comparing estimates of $N(t)$ in \emph{Drosophila}}. Population
+size over time ($N(t)$) estimated from an African population sample. Data were generated by simulating
+replicate \emph{D.melanogaster} genomes under the three-epoch \cite{sheehan2016deep} model 
+with the genetic map inferred in \cite{comeron2012many}. In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
+for the rate of coalescence in the demographic model.}
+\label{fig:n_t_sheehan}
+\end{center}
+\end{figure}
+
+As most population genetic methods development has been focused on human-scale
+data, it is of consequence to ask how such methods might perform in non-human
+genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
+model estimated from an African sample of \emph{Drosophila melanogaster} (\cite{sheehan2016deep}).
+As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under 
+an empirically derived genetic map, and then try to infer back the parameters of model. 
+Accuracy is mixed among methods in this setting, and generally worse than what we 
+observe under simulations of the human genome.
+
+\adk{do we want to include Arabidopsis results too?}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.png}
+\caption{\textbf{A comparison two-population parameter estimates in humans}. Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+A) Data were generated by simulating
+replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
+inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show inferred
+parameters out of the depicted IM model, inferring population sizes, migration rates, and a split
+time between CEU and YRI samples. C) Population size estimates for each population (rows) 
+from \dadi, \texttt{fastsimcoal2}, and \smcpp (columns). In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
+for the rate of coalescence in the demographic model. The population split date, $T_{div}$, is shown as
+an inset panel, with a common X-axis to the background figure.}
+\label{fig:IM_popn_human}
+\end{center}
+\end{figure}
+
+\paragraph*{Multi Population demographic models}
+As \texttt{stdpopsim} implements multi-population demographic models, we also 
+explored parameter estimation of population divergence parameters. In particular
+we performed simulations under multi-population models for humans and Drosophila
+and inferred simulated parameters using \dadi, \fastsimcoal, and \smcpp.
+For simplicity, we conducted inference in \dadi and \fastsimcoal using an IM model
+with constant population sizes and bi-directional migration. For human 
+models with more than two-populations, e.g. \cite{gutenkunst2009inferring},
+this means that we are inferring parameters from a model which itself does
+not match the model from which the data were generated (Figure 
+\ref{fig:IM_popn_human}A and B), however this is common practice in the field
+when applying IM-type models.
+
+In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
+time from each of the methods for samples simulated for the African and European populations
+under the \cite{gutenkunst2009inferring} model. Each of the methods compared has stronger
+and weaker points to their accuracy across parameters. For instance the simple, IM-style
+analyses we have performed do not capture recent population size change in CEU, however
+they do a good job of estimating the quite simple YRI population size history.
+
+Again we can compare between genomes and look at the performance of these methods in
+the context of a two population model from \emph{Drosophila melanogaster}. Figure
+\ref{fig:two_popn_fly} shows parameter estimates from simulations drawn from 
+the model from \cite{li2006inferring} which includes
+an ancestral population in Africa, then a population expansion with a population
+split and bottleneck into a European population. These simulations are generated 
+using the recombination map of \cite{comeron2012many}. Here methods differ dramatically
+in the estimates, with \smcpp yielding quite noisy estimates of population size
+towards the more ancient past and all methods failing to capture the very brief
+population bottleneck in Europe, although the IM-model of course only allows for
+two epochs in the population size history.
+
+Again the goal of these analyses are not to perform thorough benchmarking of 
+the methods at hand under a wide range of simulations, but instead to show how
+the tools that we have built could be used to do so.
 
 \section*{Discussion}
+Here we have described what we view as the first product from the PopSim Consortium:
+the \stdpopsim library. We have founded the Consortium with a number of specific aims
+in mind: standardization of simulation within the computational genetics community,
+increasing reproducibility of simulation based results, community-based development and 
+decision making in guiding best practices in population genetics, and eventual benchmarking
+of inferential methods.
+
+The \stdpopsim library is the first piece of this effort, in that it allows for rigorous
+standardization of complex population genetic simulations. Population genetics, as a field, 
+has yet to coalesce around a set of gold-standards for the crucial task of simulation. 
+Moreover individual research groups are often in the position of having to independently
+reimplement complex, previously published models. \stdpopsim offers solutions to both of
+these issues by creating a quality controlled, easy-to-use, gateway to models and organisms
+that are central to modern research in genetics.
+
+We have built \stdpopsim, and the PopSim Consortium, to be an open, community-developed
+project, meaning that it has excellent potential for future growth. 
+Thus far we have implemented genome representations and genetic maps for the three most
+common study systems in computational genetics: humans, \emph{Drosophila}, and \emph{Arabidopsis}.
+The standard operating procedures that we have outlined in our detailed documentation
+means that new researchers, who are interested in adding their favorite organism/genome
+are free to with very few barriers to entry. Further new demographic models can be added
+to each of the organisms in the catalog through similarly well documented steps. 
+A long-term goal then is that through a mechanism such as \stdpopsim, the community
+can determine (in an open way) what is the correct implementation for a given model, 
+and in turn, what are the models that the community is most interested in.
+
+We have shown in the results above how \stdpopsim can be used for direct comparisons
+of inferential methods on a common set of simulations. While we have not made a concerted
+effort at benchmarking methods yet, we believe that this would be an important effort for
+the PopSim Consortium and population genetics more broadly. The analysis workflows that
+we have released as part of \stdpopsim should provide a cookbook for the community to
+dive in to more rigorous benchmarking than we have done thus far.
+
+\adk{should we mention benchmarking of tree sequence inference here? what else?}
+
+\section*{Methods}
+\subsection*{Generating the PopSim resource}
+
+\subsection*{Model QC procedures}
+Each demographic model was implemented independently by two people working from
+the relevant primary sources. Disagreements between complete implementations
+were resolved between the implementers in consultation with the authors of the
+source papers when necessary. A comparison of the two implementations is then
+added as a unit test.Technical details of the QC process can be found in the
+GitHub repository in the developer documentation.
+
+\subsection*{Workflow for analysis of simulated data}
+Snakemake pipeline for 1 pop models
+Snakemake pipeline for 2 pop models
+
+ In general, used default
+settings and did not attempt to optimize \dadi or \fastsimcoal runs Used Snakemake to
+manage pipeline for two population analysis Results from two population pipeline
+demonstrate that the different methods tested were generally able to recover the
+simulated parameters
+
+% \subsection*{Extra text}
+% [This section describes the stdpopsim Python library. What it’s for, how it’s used and what it can do]
+% [First para: running accurate simulations is hard]
+% Running realistic population genetic simulations is a complex task today. It is
+% straightforward to run simple simulations of population models such as the
+% coalescent or Wright-Fisher using (e.g.) msprime or SLiM, but the process of
+% instantiating these models to accurately reflect the properties of natural
+% populations that are important for a particular study is fraught with
+% difficulties. There are typically three important inputs that must be decided:
+% the demographic model, recombination and mutation rates. Demographic models for
+% a particular species are typically chosen by examining the literature for
+% published model specifications, which can consist of dozens of parameters. These
+% parameters must be transcribed from the original publication and translated into
+% the input format required for the simulator in question. This is a difficult and
+% error prone task, and mistakes are common. Recombination and mutation rates are
+% simpler to describe and often a single genome-wide estimate is appropriate.
+% However, such estimates are updated frequently and there is wide variation in
+% the values used for the simulations of the same species. When more fine-grained
+% simulations of the local chromosome structure is required, empirical
+% recombination and mutation rate maps are used. However, there is no standard
+% format for such maps, and finding the maps can be challenging.
+
+\bibliographystyle{plainnat}
+\bibliography{PopSim.bib}
+
+\pagebreak
+\beginsupplement
+\section*{Supplemental Figures}
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.png}
+\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, \stairwayplot,
+\MSMC with $n=2$ and $n=8$. Data were generated by simulating
+replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
+inferred in \cite{international2007second}. From top to bottom we show estimates for each
+of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
+for the rate of coalescence in the demographic model.}
+\label{fig:n_t_gutenkunst}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}
+\caption{\textbf{A comparison two-population parameter estimates in Drosophila}. Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+Data were generated by simulating
+replicate human genomes under the \cite{li2006inferring} model and using the genetic map
+inferred in \cite{comeron2012many}. See legend of Figure \ref{fig:IM_popn_human} for details.
+In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
+for the rate of coalescence in the demographic model.}
+\label{fig:two_popn_fly}
+\end{center}
+\end{figure}
+
+\stopsupplement
 

### 2019-12-04 src/catalog_table.py
Rough outline for catalog table.

@@ -1,0 +1,33 @@
+"""
+Writes out a Latex description of the stdpopsim catalog to stdout.
+"""
+import stdpopsim
+
+
+def main():
+
+    print("\\begin{tabular}{lllll}")
+    for species_id in ["homsap", "dromel", "aratha", "esccol"]:
+        species = stdpopsim.get_species(species_id)
+        escaped_id = species.id.replace("_", "\\_")
+        species_header = f"{species.name} ({escaped_id})"
+        print("\\hline")
+        print("\\multicolumn{5}{l}{", species_header, "} \\\\")
+        print("\\hline")
+        # print(species.id)
+        for model in species.models:
+            # bibtex_key = "todo"
+            escaped_id = model.id.replace("_", "\\_")
+            print(escaped_id, end="& ")
+            escaped_name = model.name.replace("&", "\\&")
+            print(escaped_name, end="& ")
+            # print(f"\\cite{{{bibtex_key}}} & ", end="")
+            print("(TODO, 2010)", end="& ")
+            print("10s", end="& ")
+            print("1G ", end="")
+            print("\\\\")
+    print("\\end{tabular}{lllll}")
+
+
+if __name__ == "__main__":
+    main()

### 2019-12-05 src/simulation_benchmark.py
Added initial benchmark data.

@@ -1,0 +1,76 @@
+"""
+Run simulations and write timing/memory info to a CSV file.
+"""
+import tempfile
+import sys
+import os
+import subprocess
+
+import pandas as pd
+import stdpopsim
+
+
+def time_cmd(cmd, stdout=sys.stdout):
+    """
+    Runs the specified command line (a list suitable for subprocess.call)
+    and writes the stdout to the specified file object.
+    """
+    if sys.platform == 'darwin':
+        # on OS X, install gtime using `brew install gnu-time`
+        time_cmd = "/usr/local/bin/gtime"
+    else:
+        time_cmd = "/usr/bin/time"
+    full_cmd = [time_cmd, "-f%M %S %U"] + cmd
+    with tempfile.TemporaryFile() as stderr:
+        exit_status = subprocess.call(full_cmd, stderr=stderr, stdout=stdout)
+        stderr.seek(0)
+        if exit_status != 0:
+            raise ValueError(
+                "Error running '{}': status={}:stderr{}".format(
+                    " ".join(cmd), exit_status, stderr.read()))
+
+        split = stderr.readlines()[-1].split()
+        # From the time man page:
+        # M: Maximum resident set size of the process during its lifetime,
+        #    in Kilobytes.
+        # S: Total number of CPU-seconds used by the system on behalf of
+        #    the process (in kernel mode), in seconds.
+        # U: Total number of CPU-seconds that the process used directly
+        #    (in user mode), in seconds.
+        max_memory = int(split[0]) * 1024
+        system_time = float(split[1])
+        user_time = float(split[2])
+    return user_time + system_time, max_memory
+
+
+def main():
+
+    num_samples = 100
+    seed = 42
+
+    data = {"species": [], "model": [], "cpu_time": [], "ram": [], "file_size": []}
+    for species_id in ["homsap", "dromel", "aratha"]:
+        species = stdpopsim.get_species(species_id)
+        # Get the shortest chromosome
+        chrom = sorted(species.genome.chromosomes, key=lambda x: x.length)[0]
+        assert chrom.recombination_rate > 0
+        for model in species.models:
+            with tempfile.NamedTemporaryFile() as out:
+                cmd = (
+                    f"{species_id} -m {model.id} -c {chrom.id} {num_samples} "
+                    f"-s {seed} -o {out.name}")
+                cpu_time, ram = time_cmd(["stdpopsim"] + cmd.split())
+                file_size = os.path.getsize(out.name)
[…]

### 2019-12-05 popSimManu.tex
Added table to manuscript.
Incidental changes are automatic empty space deletes.
Fixed bibliography filename.
Reworked abstract.

@@
 
-\abstract{The explosion in population genomic data demands ever increasingly complex modes of analysis.
-More and more these analyses depend on sophisticated, often computationally expensive simulation. 
-To date, computational genetics researchers have been often forced to re-implement such simulations
-independently, leading to both wasted time in the duplication of effort and reductions in reproducibility.
-Population genetics, as a field, specifically lacks rigorous standard benchmarks by which new tools for inference might 
-be measured. Here we describe a new resource, the PopSim library, that attempts to rectify this situation. 
-PopSim is a community-driven, open source project, aimed at revolutionizing the ease and reproducibility
-of population genetic simulation for a catalog of published demographic models from both humans and non-model organisms. 
-PopSim assures correctness in our implementations through a mature quality control pipeline that is
-community-based. Here we describe the PopSim resource as a step towards inviting an even broader
-community of developers to participate and demonstrate some rather simple examples of how one
-might use PopSim for comparison of methods of demographic inference.}
+% JK: didn't bother with the \texttt stuff for stdpopsim in the abstract as it's nearly
+% always dropped in production anyway.
 
+\abstract{The explosion in population genomic data demands ever more complex
+modes of analysis, and increasingly these analyses depend on sophisticated simulations.
+Recent advances in population genetic simulation have made it possible to
+simulate large and complex models, but describing such models for a particular
+simulation engine remains a difficult and error-prone task.
+Computational genetics researchers currently re-implement simulation models
+independently, leading to duplication of effort and reductions in reproducibility.
+Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
+be measured. Here we describe a new resource, stdpopsim, that attempts to rectify this situation.
+Stdpopsim is a community-driven, open source project, which provides a standard
+catalog of published simulation models from a wide range of organisms,
+and supports multiple simulation engine backends.
+% PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
+Here we describe the stdpopsim resource to encourange an even broader
+community of developers to participate and to demonstrate some examples of how
+it can be used to systematically compare demographic inference methods.}
 \date{\today}
@@
 admixture (e.g., in humans \cite{ragsdale2019models}).
+
+\begin{table}
+\begin{footnotesize}
+\input{model_table.tex}
+\end{footnotesize}
+\caption{\label{tab:catalog}
+Details of the initial set of population models across four species.
+\textbf{NOTES: Column names are ID, name (or description maybe?),
+reference, CPU time, RAM}
+}
+\end{table}
 
@@
 \bibliographystyle{plainnat}
-\bibliography{PopSim.bib}
+\bibliography{popSim.bib}
 

### 2019-12-05 popSimManu.tex
Pass through intro.

--- 2019-12-03 popSimManu.tex
@@
 \newcommand{\stdpopsim}{\texttt{stdpopsim} }
-\newcommand{\dadi}{\texttt{dadi} }
+\newcommand{\dadi}{$\partial a \partial i$ }
 \newcommand{\MSMC}{\texttt{MSMC} }
@@
 to decreases in reproducibility and transparency across the entire field.
-Here we describe the PopSim consortium project, a community-driven catalog
+Here we describe \texttt{stdpopsim}, a community-driven catalog
 of standardized simulations across a variety of common study organisms
@@
 
-Benchmarking method performance population genetics is not straightforward.
-Many population genetic inference methods often make different assumptions as
+Benchmarking the performance of inference methods in population genetics is not straightforward.
+Inference methods often make different assumptions as
 well as use different summaries of the genetic variation data, even when
@@
 The \MSMC method (\cite{schiffels2014inferring}) allows for the
-population to continuously change in size over time. Other popular methods, like
-\dadi (\cite{gutenkunst2009inferring}), typically allow for a smaller number of size changes to occur at certain
+population to continuously change in size over time. Other methods, such as
+\dadi (\cite{gutenkunst2009inferring}), allow for a smaller number of size changes to occur at certain
 points throughout the population's history. Further, the types of data that the
@@
 different conclusions. For example, \MSMC methods applied to human genomes have
-suggested large ancient (>100 years ago) ancestral population sizes and
+% JK is this right? > 100 years doesn't sound very ancient to me.
+suggested large ancient ($>100$ years ago) ancestral population sizes and
 bottlenecks that have not been detected by SFS-based methods (see
@@
 genetic variation data, suggesting that modeling choices can greatly affect the
-performance of the inference. Further it is reasonable
+performance of the inference. Furthermore, it is reasonable
 to assume generally that specific methods will perform better than others under certain scenarios.
@@
 lack principled guidelines for deciding which statistical method is best suited
-for to accurately answer their particular question. The need for empirical
+to accurately answer their particular question. The need for empirical
 guidance will only increase as researchers continue to generate more genomic
@@
 parameters and will also provide reasonable estimates of uncertainty of the
-estimated parameters. Indeed simulations studies are a standard ingredient in
+estimated parameters. Indeed, simulations studies are a standard ingredient in
 most papers proposing new methods in population genetics. However, these
 simulation studies are often specifically devised to showcase the performance
-of the novel method and may not adquately compare performance to competing 
+of the novel method and may not adequately compare performance to competing
 methods. Moreover researchers are generally forced to ``reinvent the wheel''
@@
 First, while there are many published population genetic models available, their
-parameters are often not easily accessible, often presented in supplemental
-tables. Implementing these parameters into usable commands for simulations can
-be a time-consuming and error prone process. Further, often papers present many
+parameters are often not easily accessible, usually presented in supplemental
+tables. Translating these parameters into input for a given simulation engine can
+be a time-consuming and error prone process. In addition, papers often present many
 different models, and different authors may choose to select distinct models,
@@
 faster computers have partially ameliorating this concern, generating the
-simulated datasets stil can still be a time and computing resource bottleneck.
+simulated datasets can still be a time and computing resource bottleneck.
 
 For these reasons, we have generated a standardized, community-driven resource
-for simulation of published demographic models from a number of popular study systems.
-This resource, which we call PopSim, significantly eases the development
-burden of implementing realistic simulations for population genetic analysis. 
-Briefly, PopSim currently is comprised of a catalog of four organism genomes: humans,
-\emph{Drosophila}, \emph{Arabidopsis}, and \emph{E. coli}. Each of these
-genomes is associated with a physical organization (e.g., a chromosome structure), 
-one or more genetic maps, default population-level parameters (mutation rate,
+for simulating  published demographic models from a number of popular study systems.
+This resource, which we call \texttt{stdpopsim}, makes running
+realistic simulations for population genetic analysis a simple matter of
+choosing pre-implemented models from a community maintained catalog.
+The \stdpopsim catalog currently contains four organisms: humans,
+\emph{D.\ melanogaster}, \emph{A. thaliana}, and \emph{E. coli}. For each
+organism the catalog contains details on the physical organization (e.g., chromosome structure)
+of their genomes, one or more genetic maps, default population-level parameters (mutation rate,
 generation time) and one or more published demographic histories. Through
-either a command line interface or a mature Python API, the user can specify what
-organism, map, chromosome(s), and history they are interested in simulating, and are returned
-simulation output from their chosen model. PopSim has been developed using an 
-completely distributed open source model, with strong procedures in place 
-to continue its growth and accuracy. Below we describe PopSim and give an
-example of how benchmarking of methods for demographic inference might proceed.
+either a command line interface or a simple Python API, the user can specify what
+organism, map, chromosome, and demographic history they are interested in simulating, and are returned
+simulation output from their chosen model.
+The software has been developed by the PopSim Consortium using a
+distributed open source model, with strong procedures in place
+to continue its growth and accuracy. Below we describe the resource and give an
+example of how it can be used to benchmark demographic inference methods.
 

### 2019-12-05 popSimManu.tex

@@
 
-\abstract{The explosion in population genomic data demands ever increasingly complex modes of analysis.
-More and more these analyses depend on sophisticated, often computationally expensive simulation.
-To date, computational genetics researchers have been often forced to re-implement such simulations
-independently, leading to both wasted time in the duplication of effort and reductions in reproducibility.
-Population genetics, as a field, specifically lacks rigorous standard benchmarks by which new tools for inference might
-be measured. Here we describe a new resource, the PopSim library, that attempts to rectify this situation.
-PopSim is a community-driven, open source project, aimed at revolutionizing the ease and reproducibility
-of population genetic simulation for a catalog of published demographic models from both humans and non-model organisms.
-PopSim assures correctness in our implementations through a mature quality control pipeline that is
-community-based. Here we describe the PopSim resource as a step towards inviting an even broader
-community of developers to participate and demonstrate some rather simple examples of how one
-might use PopSim for comparison of methods of demographic inference.}
+% JK: didn't bother with the \texttt stuff for stdpopsim in the abstract as it's nearly
+% always dropped in production anyway.
 
+\abstract{The explosion in population genomic data demands ever more complex
+modes of analysis, and increasingly these analyses depend on sophisticated simulations.
+Recent advances in population genetic simulation have made it possible to
+simulate large and complex models, but describing such models for a particular
+simulation engine remains a difficult and error-prone task.
+Computational genetics researchers currently re-implement simulation models
+independently, leading to duplication of effort and reductions in reproducibility.
+Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
+be measured. Here we describe a new resource, stdpopsim, that attempts to rectify this situation.
+Stdpopsim is a community-driven, open source project, which provides a standard
+catalog of published simulation models from a wide range of organisms,
+and supports multiple simulation engine backends.
+% PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
+Here we describe the stdpopsim resource to encourange an even broader
+community of developers to participate and to demonstrate some examples of how
+it can be used to systematically compare demographic inference methods.}
 \date{\today}

### 2019-12-05 popSimManu.tex
shortened the title
KEL edits on abstract & intro

@@
 
-\title{\stdpopsim: a community-supported toolkit for population genetic simulation}
+\title{\stdpopsim: a community toolkit for population genetic simulation}
 
@@
 simulation engine remains a difficult and error-prone task.
-Computational genetics researchers currently re-implement simulation models
-independently, leading to duplication of effort and reductions in reproducibility.
+Computational genomics researchers currently re-implement simulation models
+independently for their own use, leading to duplication of effort and reductions in reproducibility.
 Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
@@
 and supports multiple simulation engine backends.
+%KEL: I think we need another sentence to make it clearer on what exactly stdpopsim is...like what's available?
 % PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
@@
 that can infer the histories of populations, the distribution of fitness effects,
-recombination rates, the extent of positive selection in genome sequence data
+recombination rates, and the extent of positive selection in genome sequence data
 (\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
@@
 constitute community standards or best practices. The general modus operandi to date has been to
-introduce methods which are validated by a set of bespoke simulations produced by individual
+introduce methods which are validated by a set of custom simulations produced by individual
 groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
@@
 % JK is this right? > 100 years doesn't sound very ancient to me.
-suggested large ancient ($>100$ years ago) ancestral population sizes and
+%KEL: Good catch--it should be >100,000 years ago!
+suggested large ancient ($>100,000$ years ago) ancestral population sizes and
 bottlenecks that have not been detected by SFS-based methods (see
@@
 parameters and will also provide reasonable estimates of uncertainty of the
-estimated parameters. Indeed, simulations studies are a standard ingredient in
+estimated parameters. Indeed, simulation studies are a standard ingredient in
 most papers proposing new methods in population genetics. However, these
@@
 parameters are often not easily accessible, usually presented in supplemental
-tables. Translating these parameters into input for a given simulation engine can
-be a time-consuming and error prone process. In addition, papers often present many
+tables, if at all. Translating these parameters into input for a given simulation engine can
+be a time-consuming and error-prone process. In addition, papers often present many
 different models, and different authors may choose to select distinct models,
 making comparison difficult. While improvements in simulation software and ever
-faster computers have partially ameliorating this concern, generating the
+faster computers have partially ameliorated this concern, generating the
 simulated datasets can still be a time and computing resource bottleneck.

### 2019-12-05 popSimManu.tex
added mig rate plots and src

--- 2019-12-05 popSimManu.tex
@@
 \documentclass[12pt,halfline,a4paper]{ouparticle}
-% *
-%
-% ^.
 \usepackage{amsmath,amssymb}
-\usepackage{natbib}
+\usepackage[round]{natbib}
 \usepackage{graphicx,rotating}
 \usepackage{dsfont}
+\usepackage{xspace}
 
@@
 
-\newcommand{\stdpopsim}{\texttt{stdpopsim} }
-\newcommand{\dadi}{$\partial a \partial i$ }
-\newcommand{\MSMC}{\texttt{MSMC} }
-\newcommand{\smcpp}{\texttt{smcpp} }
-\newcommand{\stairwayplot}{\texttt{stairwayplot} }
-\newcommand{\fastsimcoal}{\texttt{fastsimcoal} }
+\newcommand{\stdpopsim}{\texttt{stdpopsim}\xspace}
+\newcommand{\dadi}{$\partial a \partial i$\xspace}
+\newcommand{\MSMC}{\texttt{MSMC}\xspace}
+\newcommand{\smcpp}{\texttt{smcpp}\xspace}
+\newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
+\newcommand{\fastsimcoal}{\texttt{fastsimcoal}\xspace}
 
@@
 
-\usepackage[colorinlistoftodos]{todonotes}
+\usepackage[hidelinks]{hyperref}
 
@@
 
-\title{\stdpopsim: a community-supported toolkit for population genetic simulation}
+\title{A community-maintained standard library of population genetic simulations}
 
@@
 Computational genetics researchers currently re-implement simulation models
-independently, leading to duplication of effort and reductions in reproducibility.
-Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
+independently, leading to duplication of effort and the possibility for error.
+Population genetics, as a field, also lacks standard benchmarks by which new tools for inference might
 be measured. Here we describe a new resource, stdpopsim, that attempts to rectify this situation.
@@
 and supports multiple simulation engine backends.
-% PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
-Here we describe the stdpopsim resource to encourange an even broader
+% PopSim assures correctness in our implementations through a mature, community-based quality control pipeline.
+Here, we encourange an even broader
 community of developers to participate and to demonstrate some examples of how
@@
 recombination rates, the extent of positive selection in genome sequence data
-(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
+\citep{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
 chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
-kern2018diplos,sugden2018localization}).
+kern2018diplos,sugden2018localization}.
 While en masse these methods have increased our understanding of the
@@
 benchmark the quality of inferences gleaned from computational population genetics.
+As large databases of population genetic variation begin to be used to inform public health procedures,
+the accuracy and quality of these inferences is becoming even more important.
 This is in large part because ``ground-truth'' in population genetics comes from simulation,
@@
 to decreases in reproducibility and transparency across the entire field.
-Here we describe \texttt{stdpopsim}, a community-driven catalog
+Here we describe \stdpopsim, a community-driven catalog
 of standardized simulations across a variety of common study organisms
@@
 
-Benchmarking the performance of inference methods in population genetics is not straightforward.
-Inference methods often make different assumptions as
-well as use different summaries of the genetic variation data, even when
-attempting to address the same biological question. For example, there are
-numerous methods to infer population size changes.
-The \MSMC method (\cite{schiffels2014inferring}) allows for the
-population to continuously change in size over time. Other methods, such as
-\dadi (\cite{gutenkunst2009inferring}), allow for a smaller number of size changes to occur at certain
-points throughout the population's history. Further, the types of data that the
-methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
-spatial patterns of genetic variation along the sequence, \dadi  instead uses the
-site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
-sample, ignoring any information about the correlation structure (linkage
-disequilibrium) among SNPs. Thus from the start these methods may not be
-expected to perform equally well in all scenarios. \adk{I think we are focussed
-a bit too much on benchmarking currently in the intro. Could be good to add more on efficiency
-and reproducibility}
-
-The degree to which modeling assumptions and using different summaries of the
+The degree to which modeling assumptions and different summaries of the
 data can affect the ultimate results has been challenging to assess. Yet, there
@@
 suggested large ancient ($>100$ years ago) ancestral population sizes and
-bottlenecks that have not been detected by SFS-based methods (see
-\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
+bottlenecks that have not been detected by SFS-based methods
+\citep[see][]{beichman2017comparison}.
+These distinct models differ in how they fit various summaries of the
 genetic variation data, suggesting that modeling choices can greatly affect the
@@
 most papers proposing new methods in population genetics. However, these
-simulation studies are often specifically devised to showcase the performance
-of the novel method and may not adequately compare performance to competing
-methods. Moreover researchers are generally forced to ``reinvent the wheel''
-with respect to simulations, and are forced to reimplement previously
-inferred scenarios independently. This can lead to errors being incorporated
-during the implementation of the simulations. Lastly, the degree to
-which different methods are compared can be quite variable.
+simulation studies often use simulations of the particular models for which the method was designed,
+which showcases the novel method rather than rigorously comparing performance to competing methods.
+Of course, testing a new method in the situation for which it was designed is an important step,
+but more rigorous testing is necessary for tools that are to be widely used in the field.
+% Moreover researchers are generally forced to reimplement previously
+% inferred scenarios independently, which makes good benchmarking
+% a major additional time investment, and can lead to errors.
+% Lastly, the degree to which different methods are compared can be quite variable.
+Many more researchers in the broader field
+use simulations to perform power and other exploratory analyses.
+The many steps of implementing a moderately complicated population genetics simulation,
+including translating published demographic models to a simulator
+and obtaining genetic maps,
+present a major obstacle, especially to new researchers.
 
-One reason why simulation studies do not compare the performance of many methods
-is that these simulation studies are challenging to carry out. There are several
-practical impediments to carrying out the ideal benchmarking simulation studies.
-First, while there are many published population genetic models available, their
-parameters are often not easily accessible, usually presented in supplemental
-tables. Translating these parameters into input for a given simulation engine can
-be a time-consuming and error prone process. In addition, papers often present many
-different models, and different authors may choose to select distinct models,
-making comparison difficult. While improvements in simulation software and ever
-faster computers have partially ameliorating this concern, generating the
-simulated datasets can still be a time and computing resource bottleneck.
+% One reason why simulation studies do not compare the performance of many methods
+% across a wide range of situations
+% is that simulation studies are challenging to carry out, for several reasons.
+% First, while there are many published population genetic models available, their
+% parameters are often not easily accessible, usually presented in supplemental
+% tables. Translating these parameters into input for a given simulation engine can
+% be a time-consuming and error prone process. In addition, papers often present many
+% different models, and different authors may choose to select distinct models,
+% making comparison difficult. While improvements in simulation software and ever
+% faster computers have partially ameliorated this concern, generating the
+% simulated datasets can still be a time and computing resource bottleneck.
 
@@
 The \stdpopsim catalog currently contains four organisms: humans,
-\emph{D.\ melanogaster}, \emph{A. thaliana}, and \emph{E. coli}. For each
+\emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}. For each
 organism the catalog contains details on the physical organization (e.g., chromosome structure)
@@
 either a command line interface or a simple Python API, the user can specify what
-organism, map, chromosome, and demographic history they are interested in simulating, and are returned
+organism, genetic map, chromosome, and demographic history they are interested in simulating, and are returned
 simulation output from their chosen model.
+
 The software has been developed by the PopSim Consortium using a
 distributed open source model, with strong procedures in place
-to continue its growth and accuracy. Below we describe the resource and give an
+to continue its growth and accuracy.
+Importantly, we have rigorous quality control methods to ensure implemented models are accurate,
+and have documented methods for others to contribute new modules.
+We welcome new collaborators.
+Below we describe the resource and give an
 example of how it can be used to benchmark demographic inference methods.
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{display_items/Figure1.png}
+\includegraphics[width=0.5\linewidth]{display_items/Figure1.pdf}
 \caption{\textbf{Structure of \stdpopsim}. \textbf{A.} The
@@
 to find, we provide a mechanism to automatically download (and cache locally) a
-set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the HapMapII (\cite{international2007second}) and
-Decode (\cite{kong2010fine}) genetic maps. The second key element of a species description
+set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the
+HapMapII~\citep{international2007second} and
+deCODE~\citep{kong2010fine} genetic maps. The second key element of a species description
 within \stdpopsim is a set of carefully curated population genetic model
@@
 state-of-the-art simulations. Simulations are output in the “tree sequence”
-format (\cite{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}), which
+format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
 contains complete genealogical information about the simulated samples, is
 extremely compact, and can be processed very efficiently using the tskit library
-(\cite{kelleher2016efficient,kelleher2018efficient}; Fig. \ref{fig:cartoon}D). Currently,
-\stdpopsim uses the  \texttt{msprime} coalescent simulator (\cite{kelleher2016efficient})
+\citep{kelleher2016efficient,kelleher2018efficient}. Currently,
+\stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
 as the simulation engine. We plan to include other simulation
 engines to support scenarios that cannot be modelled under the coalescent framework,
-and have already implemented \texttt{SLiM} (\cite{haller2019slim}) as
+and have already implemented \texttt{SLiM} \citep{haller2019slim} as
 an alternative back-end.
@@
 writing the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Drosophila}, and \emph{Arabidopsis}. These include both
-simpler, single population histories (e.g., in \emph{D. melanogaster} \cite{sheehan2016deep}),
-and much more complex models which include population splitting, migration, and archaic
-admixture (e.g., in humans \cite{ragsdale2019models}).
+humans, \emph{Drosophila}, and \emph{Arabidopsis}. These range from
+simple, single population histories \cite[e.g.,][]{sheehan2016deep},
+to complex models which include population splitting, migration, and archaic
+admixture \cite[e.g.,][]{ragsdale2019models}.
 
-\begin{table}
+\begin{table}[t]
 \begin{footnotesize}
@@
 and associated bottleneck model of \cite{li2006inferring} which jointly models African
-and European populations. For For \emph{Arabidopsis thaliana}, we implemented the
+and European populations. For \emph{Arabidopsis thaliana}, we implemented the
 model from in \cite{durvasula2017african} inferred using \MSMC. This model includes
@@
 To evaluate and compare the performance of inference methods we developed
-\texttt{snakemake} (\cite{koster2012snakemake}) workflows that allow reproducibility
+\texttt{snakemake} \citep{koster2012snakemake} workflows that allow reproducibility
 in comparing inference methods which are available from \url{https://github.com/popgensims/analysis}
@@
 
-For single-population population size histories, we compared \MSMC, \smcpp, and
-\stairwayplot (\cite{schiffels2014inferring,terhorst2017robust,liu2015exploring})
+For single-population population size histories, we compared
+\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
+\stairwayplot~\citep{liu2015exploring}
  on simulated genomes sampled from a single population,
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
@@
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate human genomes under the \cite{ragsdale2019models} model and using the genetic map
-inferred in \cite{international2007second}. From top to bottom we show estimates for each
+replicate human genomes under the \cite{ragsdale2019models} model and using the
+HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
 of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
@@
 
+% JK here --- I think it would really help if we described what the models
+% were, briefly, rather than saying who inferred them.
 In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
-run on whole genome simulations under the \cite{ragsdale2019models} model using the
-human genetic map estimated from \cite{international2007second}. On rows
+run on whole genome simulations under a sophisticated demographic model
+and empirical genetic map. On rows
 we show inferred $N(t)$ for the three extant populations in the model.
@@
 $N(t)$ from simulations using the same physical and genetic maps, but a different demographic
-history-- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
+history -- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
 of the methods is capturing relevant parts of the population history, although the
-degree of noise at any given time interval varies. In comparing inferences between the
+discrepancy between inferred and simulated values is changes over time. In comparing inferences between the
 models it is interesting to note that $N(t)$ estimates for the CHB and CEU
 simulated populations are generally better across methods than estimates from the YRI
-sample.
+simulated population.
+
+We can even do away with the demographic model entirely, and simulate under a human
+genome architecture and genetic map to see how well methods might to at recovering
+the population history of a constant sized population. We show results of such an
+experiment in Figure \ref{fig:n_t_HomSap_constant}. Again we see that all methods
+are recovering a modern population size within a factor of two of the truth, however
+SMC-based methods, perhaps due to their regularization, tend to learn sinusoidal
+patterns of population size even though no change is present.
 
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in \emph{Drosophila}}. Population
@@
 genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
-model estimated from an African sample of \emph{Drosophila melanogaster} (\cite{sheehan2016deep}).
+model estimated from an African sample of \emph{Drosophila melanogaster} \citep{sheehan2016deep}.
 As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
@@
 observe under simulations of the human genome.
-
-\adk{do we want to include Arabidopsis results too?}
 
@@
 analyses we have performed do not capture recent population size change in CEU, however
-they do a good job of estimating the quite simple YRI population size history.
+they do a good job of estimating the simpler simulated YRI population size history.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Discussion}
+
 Here we have described what we view as the first product from the PopSim Consortium:
@@
 of inferential methods on a common set of simulations. While we have not made a concerted
-effort at benchmarking methods yet, we believe that this would be an important effort for
+effort at benchmarking methods yet, we believe that this would be an important goal for
 the PopSim Consortium and population genetics more broadly. The analysis workflows that
@@
 
-\adk{should we mention benchmarking of tree sequence inference here? what else?}
+In the use case comparisons we have performed,
+we found that the methods tested \emph{do} generally perform well on human genomes,
+up to, for instance, the usual factor-of-two differences in inferred population sizes.
+This will not be surprising to people familiar with the landscape of
+demographic inference methods in population genetics,
+but the nature and scale of uncertainty may be not clear to people new to the field.
+The ability for \stdpopsim to generate data with and without significant features, such
+as the genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
+us to ask directly about the various failure modes of popular methods.
+Moreover the comparison of methods across the various genome organizations, genetic maps,
+and demographic histories of different organisms provides valuable information
+about how methods might perform on non-human systems.
+In our initial comparison we find that inference of population histories from \emph{Drosophila}
+or \emph{Arabidopsis} simulated genomes is a fair bit more difficult from SMC based methods
+than in the human genome. Likely this is consequence of the reduced physical and genetic
+map lengths of these genomes, although we will leave further investigation of this to the future.
+Finally, comparison of results across methods or simulation runs
+gives a strong estimate of inference uncertainty, analogously to parametric bootstrapping --
+although different methods have different strengths and weaknesses,
+strong conclusions should probably not rest entirely on features inferred by only one method.
+
+The broad brush comparisons we have done, while far from exhaustive, are informative,
+and highlight the utility of easy comparison of methods on the same footing.
+The tests here were under best-case scenarios --
+simulations of discrete, randomly mating populations that fit the assumptions of the methods nearly exactly --
+and yet different methods, as well as applications of the same method
+to different simulations of the same model,
+produced somewhat different results.
+
+%\adk{should we mention benchmarking of tree sequence inference here? what else?}
+
+% \paragraph{The problem of comparison} % Moved from the intro
+% Benchmarking the performance of inference methods in population genetics is not straightforward.
+% Different methods often fit different types of model
+% or use different summaries of the genetic variation data, even when
+% attempting to address the same biological question. For example, there are
+% numerous methods to infer population size changes.
+% The \MSMC method \citep{schiffels2014inferring} allows for the
+% population to continuously change in size over time. Other methods, such as
+% \dadi \citep{gutenkunst2009inferring}, allow for a smaller number of size changes to occur at certain
+% points throughout the population's history. The types of data that the
+% methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
+% spatial patterns of genetic variation along the sequence, \dadi  instead uses the
+% site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
+% sample, ignoring any information about the correlation structure (linkage
+% disequilibrium) among SNPs. Thus from the start these methods may not be
+% expected to perform equally well in all scenarios,
+% and indeed, it is not always clear how to compare to the truth --
+% for instance, how well can a method of piecewise constant population sizes
+% describe a period of exponential growth?
+% Or, how should discrete population models be mapped onto real-world (continuous) geography?
+% The answer depends on determining what is most important for the question at hand.
 
@@
 \subsection*{Model QC procedures}
-Each demographic model was implemented independently by two people working from
-the relevant primary sources. Disagreements between complete implementations
-were resolved between the implementers in consultation with the authors of the
-source papers when necessary. A comparison of the two implementations is then
-added as a unit test.Technical details of the QC process can be found in the
-GitHub repository in the developer documentation.
+As a consortium we have agreed to a standardized proceedure for model inclusion
+into \stdpopsim that allows for rigorous quality control. Imagine Developer A
+wants to introduce a new model into \stdpopsim. Developer A implements the
+demographic model for the relevant organism along with clear documentation
+of the model parameters and populations. This model is submitted as a Pull Request,
+where it is evaluated by a reviewer and then included as `preliminary',
+but is not linked to the online documentation nor the command line interface.
+Inclusion of a preliminary model triggers a QC issue, and a second developer,
+Developer B, then independently reimplements the model from the relevant
+primary sources and adds an automatic unit test for equality between the
+QC implementation and the preliminary production model. If the two
+implementations are equal the original model is included in \stdpopsim.
+If not, we move to an arbitration process whereby the A and B first try
+to work out the details of what went wrong. If that fails, the original
+authors of the published model A is trying to implement must be contacted
+to resolve ambiguities. This process can lead to cycles of revision of
+the implementation from Developers A and B but allows the project to assure
+quality. Indeed we have already had a number of interesting cases of arbitration
+to this point. Further details of our QC process can be found in our
+\href{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.
 
 \subsection*{Workflow for analysis of simulated data}
-Snakemake pipeline for 1 pop models
-Snakemake pipeline for 2 pop models
+To demonstrate the utility of \stdpopsim we created \texttt{Snakemake}
+workflows \citep{koster2012snakemake} that perform demographic inference on
+tree sequence output from our package using a few common software packages.
+Our choice of \texttt{Snakemake} allows complete reproducibility of the
+analyses shown and the entire Analysis repository is available from
+\url{https://github.com/popgensims/analysis}.
 
- In general, used default
-settings and did not attempt to optimize \dadi or \fastsimcoal runs Used Snakemake to
-manage pipeline for two population analysis Results from two population pipeline
-demonstrate that the different methods tested were generally able to recover the
-simulated parameters
+We performed demographic analysis for one and two population tasks.
+Our one population task was to infer historical population size over
+time ($N(t)$). This was done using three software packages \stairwayplot
+which uses site frequency spectrum information only \citep{liu2015exploring},
+the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
+sample sizes $n\in (2 , 8)$, and \smcpp~\citep{terhorst2017robust}
+which combines information from the site frequency spectrum with
+recombination information as in SMC-based methods. No attempt
+was made at trying to optimize the analysis from
+any particular software package in present case,
+as our goal was not to benchmark performance of methods but
+instead show how such benchmarking could be easily done using
+the \stdpopsim resource. In this spirit we ran each software package as near
+to default parameters as possibile in each case. For \stairwayplot we
+set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
+``estimate" run mode to infer $N(t)$ with all other parameters set
+to their default values. For \MSMC we used the ``--fixedRecombination"
+option and used the default number of iterations.
+
+For the single population task we ran human, HomSap, simulations
+using a variety of models: OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
+a constant sized generic model, and a two epoch generic model where the
+population size instantaneously decreases from $N=10^4$ to $N=10^3$ 500
+generations before the present. Each HomSap simulation was run
+using the HapmapII\_GRCh37 genetic maps. For \emph{D. melanogaster}
+we estimated $N(t)$ from an African sample simulated under the DroMel,
+African3Epoch\_1S16 model using the Comeron2012\_dm6 map. Finally we ran
+simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18
+model under the Salome2012\_TAIR7. For each model, 3 replicate whole genomes were
+simulated and the population size estimated from those data. In all cases we
+set the sample size of the focal population to $N=50$ chromosomes.
+
+Following simulation, low recombination portions of chromosomes were masked
+from the analysis in a manner that reflects the ``accessible" subset of sites
+used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
+Specifically we masked regions of 1cM or greater in the lowest 5th percentile of the empirical
+distribution of recombination, regions which are nearly uniformly absent for
+empirical analysis.
+
+Our goal for the two population analysis pipeline was to explore two population inference
+on some of the multi-population demographic models implemented in \stdpopsim.
+For HomSap we examined the OutOfAfrica\_3G09 model with the HapmapII\_GRCh37 genetic map,
+and for DroMel we simulated under the OutOfAfrica\_2L06 model with the Comeron2012\_dm6.
+The HomSap model is a three population model (Africa, Europe, and Asia) including post-divergence
+migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
+DroMel model is a two population model (Africa and Europe) with no post-divergence
+migration and constant population sizes (Figure \ref{fig:two_popn_fly}).
+
+To conduct inference on these models, we used three commonly used methods:
+\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
+and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
+generally with default settings and we did not attempt to optimize their performance or fit
+parameter-rich demographic models. As with the single population inference pipeline,
+we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
+our inference pipeline and maintain reproducibility.
+
+For both \dadi and \fastsimcoal2, we fit a two population
+isolation with migration (IM) model with constant population sizes to the simulated
+data from the HomSap and DroMel models. This IM model contains six parameters:
+the ancestral population size, the size of population 1 after the split, the size of
+population 2 after the split, the divergence time, and two migration rate parameters.
+Importantly, this meant that for both species, the
+fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human}C and Sx).
+In the HomSap case, we therefore performed inference solely on the Africa
+and Europe populations, meaning that the Asia population functioned as a ghost
+population that was ignored by our inference. Our motivation for fitting this simple
+IM model was to mimic the typical approach of two population inference on empirical
+data, where the user is not aware of the `true' underlying demography and the inference
+model is often misspecified. To ground-truth our inference approach, we also conducted
+inference on a generic IM model that was identical to the model used for inference.
+
+From HomSap we ran replicate whole genome simulations under the model
+and specified taking 20 samples each from the Europe and Africa populations.
+For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
+For example, both chromosomes 3R and X failed to coalesce within 30 days of starting using $n = 50$.
+Runtimes using the identical sample size were 19, 21, and 12 days for chromosomes 2L, 2R, and 3L,
+respectively. For this reason, we chose to present only data from chromosome 2R for simulations
+under OutOfAfrica\_2L06. For the generic IM simulations, we used the HomSap genome along with the
+HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
+
+Following simulation, we outputted tree sequences and masked low recombination
+regions using the same approach described for the single population pipeline above. We
+converted tree sequences into a two dimensional site frequency spectrum for all
+chromosomes in the appropriate format for \dadi and \fastsimcoal2. For each simulation
+replicate, we performed 10 runs of \dadi and \fastsimcoal2 and checked for convergence.
+Detailed settings for \dadi and \fastsimcoal2 can be found in the Snakefile
+on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
+The parameter estimates from the top run from each simulation replicate
+are shown in Figures \ref{fig:IM_popn_human}C and Sx.
+
+For \smcpp , we converted the tree sequences into vcf format and performed inference
+with default settings. Importantly, \smcpp assumes no migration post-divergence, so
+here again the inference model does not exactly match the simulated model for the
+human model that does include migration. However, because \smcpp allows
+for continuous population size changes, it is better equipped to capture many of the
+more complex aspects of the simulated demographic models (e.g., exponential growth).
+
+To visualize our results, we plotted the inferred population size trajectories
+for each simulation replicate along side the true (census) population sizes
+(Figures \ref{fig:IM_popn_human}C Sx). Here, unlike the single population pipeline,
+we focus on census size rather than the inverse coalescence rate as the `true' population size.
 
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.png}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
@@
 \label{fig:n_t_gutenkunst}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.pdf}
+\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
+\MSMC with $n=2$ and $n=8$. Data were generated by simulating
+replicate human genomes under a constant sized population model with $N=10^4$ and using the
+HapMapII genetic map~\citep{international2007second}. In black we show the true population size history
+of the model.}
+\label{fig:n_t_HomSap_constant}
 \end{center}

### 2019-12-05 src/plot_mig_rates_dromel.py

@@ -1,0 +1,56 @@
+import pandas
+import seaborn as sns
+import matplotlib
+from matplotlib import pyplot as plt
+import numpy as np
+import subprocess
+
+
+
+dadi_infile = "dmel_n50_LSTP_joint_3Mb_no_mask/Results/best_dadi_runs.txt"
+fsc_infile = "dmel_n50_LSTP_joint_3Mb_no_mask/Results/best_fsc_runs.txt"
+simulated_genome_length = 3000000.0
+
+outfile = "dromel_migration_rates.pdf"
+
+
+dadi = pandas.read_csv(dadi_infile, skiprows=0, sep="\t")
+dadi['A-Na'] = dadi['theta']/(4*8.4e-9*simulated_genome_length)
+dadi['A-N2'] = dadi['nu2']*dadi['A-Na']
+dadi['A-N1'] = dadi['nu1']*dadi['A-Na']
+dadi['A-TDIV'] = dadi['T']*dadi['A-Na']*2
+dadi['A-MIG_EU_AF'] = dadi['m1']/(dadi['A-Na']*2)
+dadi['A-MIG_AF_EU'] = dadi['m2']/(dadi['A-Na']*2)
+dadi = dadi[0:3]
+
+fsc = pandas.read_csv(fsc_infile, skiprows=0, sep="\t")
+fsc['A-Na'] = fsc['ANCSIZE']/2
+fsc['A-N1'] = fsc['NPOP2']/2
+fsc['A-N2'] = fsc['NPOP1']/2
+fsc['A-TDIV'] = fsc['TDIV']
+fsc['A-MIG_EU_AF'] =fsc['MIG12']
+fsc['A-MIG_AF_EU'] =fsc['MIG21']
+fsc = fsc[0:3]
+
+
+f, ax = plt.subplots(figsize=(10, 10))
+truth_M = pandas.DataFrame(np.array([[0,0]]), columns=['A-MIG_EU_AF','A-MIG_AF_EU'])
+
+data_M = fsc[['A-MIG_EU_AF','A-MIG_AF_EU']]
+data_M=data_M.append(dadi[['A-MIG_EU_AF','A-MIG_AF_EU']], sort=True)
+data_M=data_M.append(truth_M, sort=True)
+num_sims = dadi.shape[0]
+data_M['method']=['fastsimcoal2']*num_sims+['dadi']*num_sims+['truth']
+data_M=data_M.reset_index(drop=True).reset_index()
[…]

### 2019-12-05 src/plot_mig_rates_homsap.py

@@ -1,0 +1,55 @@
+import pandas
+import seaborn as sns
+import matplotlib
+from matplotlib import pyplot as plt
+import numpy as np
+import subprocess
+
+
+
+dadi_infile = "homo_sapiens_mask_full_genome/Results/best_dadi_runs.txt"
+fsc_infile = "homo_sapiens_mask_full_genome/Results/best_fsc_runs.txt"
+simulated_genome_length = 2681441344.0
+
+outfile = "homsap_migration_rates.pdf"
+
+
+dadi = pandas.read_csv(dadi_infile, skiprows=0, sep="\t")
+dadi['A-Na'] = dadi['theta']/(4*1e-8*simulated_genome_length)
+dadi['A-N2'] = dadi['nu2']*dadi['A-Na']
+dadi['A-N1'] = dadi['nu1']*dadi['A-Na']
+dadi['A-TDIV'] = dadi['T']*dadi['A-Na']*2
+dadi['A-MIG_EU_AF'] = dadi['m1']/(dadi['A-Na']*2)
+dadi['A-MIG_AF_EU'] = dadi['m2']/(dadi['A-Na']*2)
+
+fsc = pandas.read_csv(fsc_infile, skiprows=0, sep="\t")
+fsc['A-Na'] = fsc['ANCSIZE']/2
+fsc['A-N1'] = fsc['NPOP2']/2
+fsc['A-N2'] = fsc['NPOP1']/2
+fsc['A-TDIV'] = fsc['TDIV']
+fsc['A-MIG_EU_AF'] =fsc['MIG12']
+fsc['A-MIG_AF_EU'] =fsc['MIG21']
+
+
+
+f, ax = plt.subplots(figsize=(10, 10))
+truth_M = pandas.DataFrame(np.array([[3e-5,3e-5]]), columns=['A-MIG_EU_AF','A-MIG_AF_EU'])
+
+data_M = fsc[['A-MIG_EU_AF','A-MIG_AF_EU']]
+data_M=data_M.append(dadi[['A-MIG_EU_AF','A-MIG_AF_EU']], sort=True)
+data_M=data_M.append(truth_M, sort=True)
+num_sims = dadi.shape[0]
+data_M['method']=['fastsimcoal2']*num_sims+['dadi']*num_sims+['truth']
+data_M=data_M.reset_index(drop=True).reset_index()
[…]

### 2019-12-05 popSimManu.tex
first pass at two pop methods

--- 2019-12-05 popSimManu.tex
@@
 Snakemake pipeline for 1 pop models
-Snakemake pipeline for 2 pop models
 
- In general, used default
-settings and did not attempt to optimize \dadi or \fastsimcoal runs Used Snakemake to
-manage pipeline for two population analysis Results from two population pipeline
-demonstrate that the different methods tested were generally able to recover the
-simulated parameters
+\subsection*{Workflow for two population analysis}
+Our goal for the two population inference pipeline was to explore two population inference
+on some of the multi-population demographic models implemented in \stdpopsim.
+The models we examined include the human Gutenkunst model (\cite{gutenkunst2009inferring}) as
+well as the \emph{Drosophila} Li & Stephan model (\cite{li2006inferring}). We implemented three
+commonly-used multi-population inference approaches: \dadi (\cite{gutenkunst2009inferring}),
+\fastsimcoal2 (\cite{excoffier2013robust}), and \smcpp (\cite{terhorst2017robust}). Our aim was not
+to exhaustively test the performance of these methods, but rather to explore their behavior
+in two different inference scenarios and demonstrate the potential for this resource to
+eventually be used for more in-depth benchmarking. As a result, these methods were used
+generally with default settings and we did not attempt to optimize their performance or fit
+parameter-rich demographic models. As with the single population inference pipeline,
+we developed a \texttt{snakemake} (\cite{koster2012snakemake}) workflow to manage
+our inference pipeline and maintain reproducibility.
+
+For both \dadi and \fastsimcoal2, our general approach was to fit a two population
+isolation with migration (IM) model with constant population sizes to the simulated
+data from the human and \emph{Drosophila} models. This IM model contains six parameters:
+the ancestral population size, the size of population 1 after split, the size of
+population 2 after split, the divergence time, and two migration rate parameters. Importantly,
+this meant that for both the human and \emph{Drosophila} inference, the fitted model
+did not match the simulated model (Figure \ref{fig:IM_popn_human}C), which, for humans,
+includes three populations with various population size changes as well as exponential
+growth. In the human case, we therefore performed inference solely on the Africa (YRI)
+and Europe (CEU) populations, meaning that the Asia (CHB) population functioned as a ghost
+population that was ignored by our inference. Our motivation for this simple
+approach was to mimic the typical approach of two population inference on empirical
+data, where the user is not aware of the 'true' underlying demography and the inference
+model is often misspecified. To ground-truth our inference approach, we also conducted
+inference on a generic IM model that was identical to the model used for inference.
+% CK: I have a figure of the generic IM results in the works that we can include in the supplement
+
+To perform inference, we began by running simulations under each model, performing
+three replicates to inspect for inter-replicate variability. For humans, we ran
+whole genome simulations (chromosomes 1-22) under the Gutenkunst model (\cite{gutenkunst2009inferring}),
+using the HapMap II recombination map (\cite{international2007second}) and taking 20
+samples each from the CEU and YRI populations. For \emph{Drosophila},
+we simulated 3 Mb of chromosome 2R under the Li & Stephan model (\cite{li2006inferring}),
+using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
+the Africa and Europe populations. Due to computational limitations, we were unable
+to run whole genome simulations for \emph{Drosophila}. For the generic IM simulations,
+we used the human genome (chromosomes 1-22) and HapMap II recombination map (\cite{international2007second})
+and took 20 samples per population.
+% CK: should we say more about the drosophila simulations here?
+
+Following simulation, we outputted tree sequences and masked low recombination
+regions using the same approach described for the single population pipeline above. We
+converted tree sequences into a two dimensional site frequency spectrum for all
+chromosomes in the appropriate format for \dadi and \fastsimcoal2.
+For each simulation replicate, we performed 10 runs of \dadi and \fastsimcoal2 and
+checked for convergence. The parameter estimates from the top run from each simulation
+replicate are shown in Figures \ref{fig:IM_popn_human}C and Sx.
+% CK: how much detail should i go into here? should i include things such as initial
+% parameter values, grid size, priors, etc?
+
+For \smcpp , we converted the tree sequences into vcf format and performed inference
+with default settings. Importantly, \smcpp assumes no migration post-divergence, so
+here again the inference model does not exactly match the simulated model for the
+human Gutenkunst model that does include migration. However, because \smcpp allows
+for continuous population size changes, it is better equipped to capture many of the
+more complex aspect of the simulated demographic models (e.g., exponential growth).
+% CK: would be good to have someone with more smcpp experience to weigh in on
+% these methods (Andy or Deigo?). Are there any settings that need to be mentioned?
+
+To visualize our results, we plotted the inferred population size trajectories
+for each simulation replicate along side the true (census) population sizes
+(Figures \ref{fig:IM_popn_human}C Sx). Here, unlike the single population pipeline,
+ we focus on census size rather than the inverse coalescence rate as the 'true' population size.
 

### 2019-12-05 src/plotResults_3population_4method.R
added plotting code for final figures

@@ -1,0 +1,280 @@
+library(ggplot2)
+library(cowplot)
+
+## define variables
+infile_pop1="/n_t/homo_sapiens_mask_Ragsdale_population1/Results/all_estimated_Ne.txt"
+infile_pop2="/n_t/homo_sapiens_mask_Ragsdale_population2/Results/all_estimated_Ne.txt"
+infile_pop3="/n_t/homo_sapiens_mask_Ragsdale_population3/Results/all_estimated_Ne.txt"
+outfile="/n_t/n_t_plot_Ragsdale.pdf"
+species="Homo sapiens"
+pop1_name="YRI"
+pop2_name="CEU"
+pop3_name="CHB"
+method1_name="SMC++"
+method2_name="stairway plot"
+method3_name="MSMC (n = 2)"
+method4_name="MSMC (n = 8)"
+method1_id="smcpp"
+method2_id="sp"
+method3_id="msmc_2"
+method4_id="msmc_8"
+replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
+truth_color="gray30"
+
+## plot
+cPal_legend<-c(truth_color, replicate_colors)
+
+tab0 <- read.table(file=infile_pop1, h=T ,sep="\t")
+tab1 <- read.table(file=infile_pop2, h=T ,sep="\t")
+tab2 <- read.table(file=infile_pop3, h=T ,sep="\t")
+
+tab0$x[tab0$x==0.0]<-1
+tab1$x[tab1$x==0.0]<-1
+tab2$x[tab2$x==0.0]<-1
+
+xMin = 10
+xMax = max(c(tab0$x,tab1$x,tab2$x))
+y0Min = min(tab0$y)
+y0Max = max(tab0$y)
+y1Min = min(tab1$y)
+y1Max = max(tab1$y)
+y2Min = min(tab2$y)
+y2Max = max(tab2$y)
+
+tab0A = subset(tab0,method == method1_id)
+tab0B = subset(tab0,method == method2_id)
+tab0C = subset(tab0,method == method3_id)
+tab0D = subset(tab0,method == method4_id)
+tab0Z = subset(tab0,method == "coal")
+
+tab1A = subset(tab1,method == method1_id)
+tab1B = subset(tab1,method == method2_id)
+tab1C = subset(tab1,method == method3_id)
+tab1D = subset(tab1,method == method4_id)
+tab1Z = subset(tab1,method == "coal")
+
[…]

### 2019-12-06 popSimManu.tex
added to QC section

--- 2019-12-05 popSimManu.tex
@@
 simulation engine remains a difficult and error-prone task.
-Computational genomics researchers currently re-implement simulation models
-independently for their own use, leading to duplication of effort and reductions in reproducibility.
+Computational genetics researchers currently re-implement simulation models
+independently, leading to duplication of effort and reductions in reproducibility.
 Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
@@
 and supports multiple simulation engine backends.
-%KEL: I think we need another sentence to make it clearer on what exactly stdpopsim is...like what's available?
 % PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
@@
 that can infer the histories of populations, the distribution of fitness effects,
-recombination rates, and the extent of positive selection in genome sequence data
+recombination rates, the extent of positive selection in genome sequence data
 (\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
@@
 constitute community standards or best practices. The general modus operandi to date has been to
-introduce methods which are validated by a set of custom simulations produced by individual
+introduce methods which are validated by a set of bespoke simulations produced by individual
 groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
@@
 disequilibrium) among SNPs. Thus from the start these methods may not be
-expected to perform equally well in all scenarios. \adk{I think we are focussed
-a bit too much on benchmarking currently in the intro. Could be good to add more on efficiency
-and reproducibility}
+expected to perform equally well in all scenarios.
 
@@
 % JK is this right? > 100 years doesn't sound very ancient to me.
-%KEL: Good catch--it should be >100,000 years ago!
-suggested large ancient ($>100,000$ years ago) ancestral population sizes and
+suggested large ancient ($>100$ years ago) ancestral population sizes and
 bottlenecks that have not been detected by SFS-based methods (see
@@
 parameters and will also provide reasonable estimates of uncertainty of the
-estimated parameters. Indeed, simulation studies are a standard ingredient in
+estimated parameters. Indeed, simulations studies are a standard ingredient in
 most papers proposing new methods in population genetics. However, these
@@
 parameters are often not easily accessible, usually presented in supplemental
-tables, if at all. Translating these parameters into input for a given simulation engine can
-be a time-consuming and error-prone process. In addition, papers often present many
+tables. Translating these parameters into input for a given simulation engine can
+be a time-consuming and error prone process. In addition, papers often present many
 different models, and different authors may choose to select distinct models,
 making comparison difficult. While improvements in simulation software and ever
-faster computers have partially ameliorated this concern, generating the
+faster computers have partially ameliorating this concern, generating the
 simulated datasets can still be a time and computing resource bottleneck.
@@
 
-\begin{table}
+\begin{table}[t]
 \begin{footnotesize}
@@
 
-\adk{do we want to include Arabidopsis results too?}
-
 \begin{figure}
@@
 
-\adk{should we mention benchmarking of tree sequence inference here? what else?}
+%\adk{should we mention benchmarking of tree sequence inference here? what else?}
 
@@
 \subsection*{Model QC procedures}
-Each demographic model was implemented independently by two people working from
-the relevant primary sources. Disagreements between complete implementations
-were resolved between the implementers in consultation with the authors of the
-source papers when necessary. A comparison of the two implementations is then
-added as a unit test.Technical details of the QC process can be found in the
-GitHub repository in the developer documentation.
+As a consortium we have agreed to a standardized proceedure for model inclusion
+into \stdpopsim that allows for rigorous quality control. Imagine Developer A
+wants to introduce a new model into \stdpopsim. Developer A implements the
+demographic model for the relevant organism along with a series of required
+unit tests and documentation. This model is submitted as a Pull Request,
+where it is evaluated by a reviewer and then included as `preliminary', 
+B
+but is not linked to the online documentation nor the command line interface.
+Inclusion of a preliminary model triggers a QC issue, and a second developer, 
+Developer B, then independently reimplements the model from the relevant
+primary sources and adds an automatic unit test for equality between the
+QC implementation and the preliminary production model. If the two
+implementations are equal the original model is included in \stdpopsim.
+If not, we move to an arbitration process whereby the A and B first try
+to work out the details of what went wrong. If that fails, the original
+authors of the published model A is trying to implement must be contacted
+to resolve ambiguities. This process can lead to cycles of revision of 
+the implementation from Developers A and B but allows the project to assure
+quality. Indeed we have already had a number of interesting cases of arbitration
+to this point. Further details of our QC process can be found in our
+\href{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.
 

### 2019-12-06 popSimManu.tex
typo fixed

@@
 where it is evaluated by a reviewer and then included as `preliminary', 
-B
 but is not linked to the online documentation nor the command line interface.

### 2019-12-06 popSimManu.tex

@@
 Snakemake pipeline for 1 pop models
-Snakemake pipeline for 2 pop models
 
- In general, used default
-settings and did not attempt to optimize \dadi or \fastsimcoal runs Used Snakemake to
-manage pipeline for two population analysis Results from two population pipeline
-demonstrate that the different methods tested were generally able to recover the
-simulated parameters
+\subsection*{Workflow for two population analysis}
+Our goal for the two population analysis pipeline was to explore two population inference
+on some of the multi-population demographic models implemented in \stdpopsim.
+The models we examined include the human Gutenkunst model (\cite{gutenkunst2009inferring}) as
+well as the \emph{Drosophila} Li & Stephan model (\cite{li2006inferring}). The human
+model is a three population model (Africa, Europe, and Asia) including post-divergence
+migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
+\emph{Drosophila} model is a two population model (Africa and Europe) with no post-divergence
+migration and constant population sizes (Figure Sx).
+
+To conduct inference on these models, we implemented three commonly-used approaches:
+\dadi (\cite{gutenkunst2009inferring}), \fastsimcoal2 (\cite{excoffier2013robust}),
+and \smcpp (\cite{terhorst2017robust}). Our aim was not to exhaustively
+test the performance of these methods, but rather to explore their behavior
+in two different inference scenarios and demonstrate the potential for this resource to
+eventually be used for more in-depth benchmarking. As a result, these methods were used
+generally with default settings and we did not attempt to optimize their performance or fit
+parameter-rich demographic models. As with the single population inference pipeline,
+we developed a \texttt{snakemake} (\cite{koster2012snakemake}) workflow to manage
+our inference pipeline and maintain reproducibility.
+
+For both \dadi and \fastsimcoal2, our general approach was to fit a two population
+isolation with migration (IM) model with constant population sizes to the simulated
+data from the human and \emph{Drosophila} models. This IM model contains six parameters:
+the ancestral population size, the size of population 1 after the split, the size of
+population 2 after the split, the divergence time, and two migration rate parameters.
+Importantly, this meant that for both the human and \emph{Drosophila} inference, the
+fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human}C and Sx).
+In the human case, we therefore performed inference solely on the Africa
+and Europe populations, meaning that the Asia population functioned as a ghost
+population that was ignored by our inference. Our motivation for fitting this simple
+IM model was to mimic the typical approach of two population inference on empirical
+data, where the user is not aware of the 'true' underlying demography and the inference
+model is often misspecified. To ground-truth our inference approach, we also conducted
+inference on a generic IM model that was identical to the model used for inference.
+
+To perform inference, we began by running simulations under each model, running
+three replicates to inspect for inter-replicate variability. For humans, we ran
+whole genome simulations (chromosomes 1-22) under the Gutenkunst model (\cite{gutenkunst2009inferring}),
+using the HapMap II recombination map (\cite{international2007second}) and taking 20
+samples each from the Europe and Africa populations. For \emph{Drosophila},
+we simulated 3 Mb of chromosome 2R under the Li & Stephan model (\cite{li2006inferring}),
+using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
+the Africa and Europe populations. We tried to run whole genome simulations under
+this model but after XXXX amount of time... (JRA fill in details here). For the generic
+IM simulations, we used the human genome (chromosomes 1-22) and HapMap II recombination
+map (\cite{international2007second}) and took 20 samples per population.
+
+Following simulation, we outputted tree sequences and masked low recombination
+regions using the same approach described for the single population pipeline above. We
+converted tree sequences into a two dimensional site frequency spectrum for all
+chromosomes in the appropriate format for \dadi and \fastsimcoal2. For each simulation
+replicate, we performed 10 runs of \dadi and \fastsimcoal2 and checked for convergence.
+Detailed settings for \dadi and \fastsimcoal2 can be found in the Snakefile
+on the popgensims/analysis GitHub repository (https://github.com/popgensims/analysis). The parameter
+estimates from the top run from each simulation replicate are shown in Figures \ref{fig:IM_popn_human}C and Sx.
+
+For \smcpp , we converted the tree sequences into vcf format and performed inference
+with default settings. Importantly, \smcpp assumes no migration post-divergence, so
+here again the inference model does not exactly match the simulated model for the
+human model that does include migration. However, because \smcpp allows
+for continuous population size changes, it is better equipped to capture many of the
+more complex aspects of the simulated demographic models (e.g., exponential growth).
+
+To visualize our results, we plotted the inferred population size trajectories
+for each simulation replicate along side the true (census) population sizes
+(Figures \ref{fig:IM_popn_human}C Sx). Here, unlike the single population pipeline,
+ we focus on census size rather than the inverse coalescence rate as the 'true' population size.
 

### 2019-12-07 popSimManu.tex
cleaning after past PR

@@
 to find, we provide a mechanism to automatically download (and cache locally) a
-set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the HapMapII (\cite{international2007second}) and
+set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the HapMapII 
+\cite{international2007second}) and
 Decode (\cite{kong2010fine}) genetic maps. The second key element of a species description
@@
 The models we examined include the human Gutenkunst model (\cite{gutenkunst2009inferring}) as
-well as the \emph{Drosophila} Li & Stephan model (\cite{li2006inferring}). The human
+well as the \emph{Drosophila} Li and Stephan model (\cite{li2006inferring}). The human
 model is a three population model (Africa, Europe, and Asia) including post-divergence
@@
 \emph{Drosophila} model is a two population model (Africa and Europe) with no post-divergence
-migration and constant population sizes (Figure Sx).
+migration and constant population sizes (Figure \ref{fig:two_popn_fly}).
 
@@
 IM model was to mimic the typical approach of two population inference on empirical
-data, where the user is not aware of the 'true' underlying demography and the inference
+data, where the user is not aware of the `true' underlying demography and the inference
 model is often misspecified. To ground-truth our inference approach, we also conducted
@@
 samples each from the Europe and Africa populations. For \emph{Drosophila},
-we simulated 3 Mb of chromosome 2R under the Li & Stephan model (\cite{li2006inferring}),
+we simulated 3 Mb of chromosome 2R under the Li and Stephan model (\cite{li2006inferring}),
 using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
@@
 Detailed settings for \dadi and \fastsimcoal2 can be found in the Snakefile
-on the popgensims/analysis GitHub repository (https://github.com/popgensims/analysis). The parameter
-estimates from the top run from each simulation replicate are shown in Figures \ref{fig:IM_popn_human}C and Sx.
+on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
+The parameter estimates from the top run from each simulation replicate
+are shown in Figures \ref{fig:IM_popn_human}C and Sx.
 
@@
 (Figures \ref{fig:IM_popn_human}C Sx). Here, unlike the single population pipeline,
- we focus on census size rather than the inverse coalescence rate as the 'true' population size.
+we focus on census size rather than the inverse coalescence rate as the `true' population size.
 

### 2019-12-10 popSimManu.tex
disco para on simulation results

@@
 \documentclass[12pt,halfline,a4paper]{ouparticle}
-% *
-%
-% ^.
 \usepackage{amsmath,amssymb}
@@
 \usepackage{dsfont}
+\usepackage{xspace}
 
@@
 Benchmarking the performance of inference methods in population genetics is not straightforward.
-Inference methods often make different assumptions as
-well as use different summaries of the genetic variation data, even when
+Different methods often fit different types of model
+or use different summaries of the genetic variation data, even when
 attempting to address the same biological question. For example, there are
@@
 \dadi (\cite{gutenkunst2009inferring}), allow for a smaller number of size changes to occur at certain
-points throughout the population's history. Further, the types of data that the
+points throughout the population's history. The types of data that the
 methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
@@
 
-The degree to which modeling assumptions and using different summaries of the
+The degree to which modeling assumptions and different summaries of the
 data can affect the ultimate results has been challenging to assess. Yet, there
@@
 most papers proposing new methods in population genetics. However, these
-simulation studies are often specifically devised to showcase the performance
-of the novel method and may not adequately compare performance to competing
-methods. Moreover researchers are generally forced to ``reinvent the wheel''
-with respect to simulations, and are forced to reimplement previously
-inferred scenarios independently. This can lead to errors being incorporated
-during the implementation of the simulations. Lastly, the degree to
-which different methods are compared can be quite variable.
+simulation studies often use simulations of the particular models for which the method was designed,
+which showcases the novel method rather than rigorously comparing performance to competing methods.
+Of course, testing a new method in the situation for which it was designed is an important step,
+but more rigorous testing is necessary for tools that are to be widely used in the field.
+% Moreover researchers are generally forced to reimplement previously
+% inferred scenarios independently, which makes good benchmarking
+% a major additional time investment, and can lead to errors.
+% Lastly, the degree to which different methods are compared can be quite variable.
 
 One reason why simulation studies do not compare the performance of many methods
-is that these simulation studies are challenging to carry out. There are several
-practical impediments to carrying out the ideal benchmarking simulation studies.
+across a wide range of situations
+is that simulation studies are challenging to carry out, for several reasons.
 First, while there are many published population genetic models available, their
@@
 making comparison difficult. While improvements in simulation software and ever
-faster computers have partially ameliorating this concern, generating the
+faster computers have partially ameliorated this concern, generating the
 simulated datasets can still be a time and computing resource bottleneck.
@@
 either a command line interface or a simple Python API, the user can specify what
-organism, map, chromosome, and demographic history they are interested in simulating, and are returned
+organism, genetic map, chromosome, and demographic history they are interested in simulating, and are returned
 simulation output from their chosen model.
+
 The software has been developed by the PopSim Consortium using a
@@
 \cite{international2007second}) and
-Decode (\cite{kong2010fine}) genetic maps. The second key element of a species description
+deCODE (\cite{kong2010fine}) genetic maps. The second key element of a species description
 within \stdpopsim is a set of carefully curated population genetic model
@@
 dive in to more rigorous benchmarking than we have done thus far.
+
+Although we did not produce an exhaustive comparison of different demographic inference methods,
+our results are informative,
+and highlight the utility of easy comparison of methods on the same footing.
+The tests here were under best-case scenarios --
+simulations of discrete, randomly mating populations that fit the assumptions of the methods nearly exactly --
+and yet different methods, as well as applications of the same method
+to different simulations of the same model,
+produced somewhat different results.
+We found that the methods tested \emph{do} generally perform well,
+up to, for instance, the usual factor-of-two differences in bottleneck population sizes.
+This will not be surprising to people familiar with the landscape of 
+demographic inference methods in population genetics,
+but the nature and scale of uncertainty may be not clear to people new to the field.
+Comparison of results across methods or simulation runs
+gives a strong estimate of inference uncertainty, analogously to parametric bootstrapping --
+although different methods have different strengths and weaknesses,
+strong conclusions should probably not rest entirely on features inferred by only one method.
 

### 2019-12-10 popSimManu.tex

@@
 \cite{international2007second}) and
-deCODE (\cite{kong2010fine}) genetic maps. The second key element of a species description
+Decode (\cite{kong2010fine}) genetic maps. The second key element of a species description
 within \stdpopsim is a set of carefully curated population genetic model
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Discussion}
-
 Here we have described what we view as the first product from the PopSim Consortium:
@@
 dive in to more rigorous benchmarking than we have done thus far.
-
-Although we did not produce an exhaustive comparison of different demographic inference methods,
-our results are informative,
-and highlight the utility of easy comparison of methods on the same footing.
-The tests here were under best-case scenarios --
-simulations of discrete, randomly mating populations that fit the assumptions of the methods nearly exactly --
-and yet different methods, as well as applications of the same method
-to different simulations of the same model,
-produced somewhat different results.
-We found that the methods tested \emph{do} generally perform well,
-up to, for instance, the usual factor-of-two differences in bottleneck population sizes.
-This will not be surprising to people familiar with the landscape of 
-demographic inference methods in population genetics,
-but the nature and scale of uncertainty may be not clear to people new to the field.
-Comparison of results across methods or simulation runs
-gives a strong estimate of inference uncertainty, analogously to parametric bootstrapping --
-although different methods have different strengths and weaknesses,
-strong conclusions should probably not rest entirely on features inferred by only one method.
 

### 2019-12-11 popSimManu.tex

--- 2019-12-10 popSimManu.tex
@@
 to decreases in reproducibility and transparency across the entire field.
-Here we describe \texttt{stdpopsim}, a community-driven catalog
+Here we describe \stdpopsim, a community-driven catalog
 of standardized simulations across a variety of common study organisms
 that will empower the field and usher in the next era of genetics data analysis.
-
-Benchmarking the performance of inference methods in population genetics is not straightforward.
-Different methods often fit different types of model
-or use different summaries of the genetic variation data, even when
-attempting to address the same biological question. For example, there are
-numerous methods to infer population size changes.
-The \MSMC method (\cite{schiffels2014inferring}) allows for the
-population to continuously change in size over time. Other methods, such as
-\dadi (\cite{gutenkunst2009inferring}), allow for a smaller number of size changes to occur at certain
-points throughout the population's history. The types of data that the
-methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
-spatial patterns of genetic variation along the sequence, \dadi  instead uses the
-site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
-sample, ignoring any information about the correlation structure (linkage
-disequilibrium) among SNPs. Thus from the start these methods may not be
-expected to perform equally well in all scenarios.
 
@@
 bottlenecks that have not been detected by SFS-based methods (see
-\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
+\citet{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
 genetic variation data, suggesting that modeling choices can greatly affect the
@@
 % Lastly, the degree to which different methods are compared can be quite variable.
+Many more researchers in the broader field
+use simulations to perform power and other exploratory analyses.
+The many steps of implementing a moderately complicated population genetics simulation,
+including translating published demographic models to a simulator
+and obtaining genetic maps,
+present a major obstacle, especially to new researchers.
 
-One reason why simulation studies do not compare the performance of many methods
-across a wide range of situations
-is that simulation studies are challenging to carry out, for several reasons.
-First, while there are many published population genetic models available, their
-parameters are often not easily accessible, usually presented in supplemental
-tables. Translating these parameters into input for a given simulation engine can
-be a time-consuming and error prone process. In addition, papers often present many
-different models, and different authors may choose to select distinct models,
-making comparison difficult. While improvements in simulation software and ever
-faster computers have partially ameliorated this concern, generating the
-simulated datasets can still be a time and computing resource bottleneck.
+% One reason why simulation studies do not compare the performance of many methods
+% across a wide range of situations
+% is that simulation studies are challenging to carry out, for several reasons.
+% First, while there are many published population genetic models available, their
+% parameters are often not easily accessible, usually presented in supplemental
+% tables. Translating these parameters into input for a given simulation engine can
+% be a time-consuming and error prone process. In addition, papers often present many
+% different models, and different authors may choose to select distinct models,
+% making comparison difficult. While improvements in simulation software and ever
+% faster computers have partially ameliorated this concern, generating the
+% simulated datasets can still be a time and computing resource bottleneck.
 
@@
 distributed open source model, with strong procedures in place
-to continue its growth and accuracy. Below we describe the resource and give an
+to continue its growth and accuracy. 
+Importantly, we have rigorous quality control methods to ensure implemented models are accurate,
+and have documented methods for others to contribute new modules.
+We welcome new collaborators.
+Below we describe the resource and give an
 example of how it can be used to benchmark demographic inference methods.
@@
 
+\paragraph{The problem of comparison} % Moved from the intro
+Benchmarking the performance of inference methods in population genetics is not straightforward.
+Different methods often fit different types of model
+or use different summaries of the genetic variation data, even when
+attempting to address the same biological question. For example, there are
+numerous methods to infer population size changes.
+The \MSMC method \citep{schiffels2014inferring} allows for the
+population to continuously change in size over time. Other methods, such as
+\dadi \citep{gutenkunst2009inferring}, allow for a smaller number of size changes to occur at certain
+points throughout the population's history. The types of data that the
+methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
+spatial patterns of genetic variation along the sequence, \dadi  instead uses the
+site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
+sample, ignoring any information about the correlation structure (linkage
+disequilibrium) among SNPs. Thus from the start these methods may not be
+expected to perform equally well in all scenarios,
+and indeed, it is not always clear how to compare to the truth --
+for instance, how well can a method of piecewise constant population sizes
+describe a period of exponential growth?
+Or, how should discrete population models be mapped onto real-world (continuous) geography?
+The answer depends on determining what is most important for the question at hand.
+
 \section*{Methods}

### 2019-12-16 popSimManu.tex

@@
 \usepackage{amsmath,amssymb}
-\usepackage{natbib}
+\usepackage[round]{natbib}
 \usepackage{graphicx,rotating}
@@
 suggested large ancient ($>100$ years ago) ancestral population sizes and
-bottlenecks that have not been detected by SFS-based methods (see
-\citet{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
+bottlenecks that have not been detected by SFS-based methods
+\citep[see][]{beichman2017comparison}.
+These distinct models differ in how they fit various summaries of the
 genetic variation data, suggesting that modeling choices can greatly affect the
@@
 to find, we provide a mechanism to automatically download (and cache locally) a
-set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the HapMapII
-\cite{international2007second}) and
-deCODE (\cite{kong2010fine}) genetic maps. The second key element of a species description
+set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the
+HapMapII~\citep{international2007second} and
+deCODE~\citep{kong2010fine} genetic maps. The second key element of a species description
 within \stdpopsim is a set of carefully curated population genetic model
@@
 writing the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Drosophila}, and \emph{Arabidopsis}. These include both
-simpler, single population histories (e.g., in \emph{D. melanogaster} \cite{sheehan2016deep}),
-and much more complex models which include population splitting, migration, and archaic
-admixture (e.g., in humans \cite{ragsdale2019models}).
+humans, \emph{Drosophila}, and \emph{Arabidopsis}. These range from
+simple, single population histories \cite[e.g.,][]{sheehan2016deep},
+to complex models which include population splitting, migration, and archaic
+admixture \cite[e.g.,][]{ragsdale2019models}.
 
@@
 $N(t)$ from simulations using the same physical and genetic maps, but a different demographic
-history-- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
+history -- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
 of the methods is capturing relevant parts of the population history, although the
@@
 on some of the multi-population demographic models implemented in \stdpopsim.
-The models we examined include the human Gutenkunst model (\cite{gutenkunst2009inferring}) as
-well as the \emph{Drosophila} Li and Stephan model (\cite{li2006inferring}). The human
+The models we examined include the human Gutenkunst model \citep{gutenkunst2009inferring} as
+well as the \emph{Drosophila} Li and Stephan model \citep{li2006inferring}. The human
 model is a three population model (Africa, Europe, and Asia) including post-divergence
@@
 To conduct inference on these models, we implemented three commonly-used approaches:
-\dadi (\cite{gutenkunst2009inferring}), \fastsimcoal2 (\cite{excoffier2013robust}),
-and \smcpp (\cite{terhorst2017robust}). Our aim was not to exhaustively
+\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
+and \smcpp~\citep{terhorst2017robust}. Our aim was not to exhaustively
 test the performance of these methods, but rather to explore their behavior
@@
 parameter-rich demographic models. As with the single population inference pipeline,
-we developed a \texttt{snakemake} (\cite{koster2012snakemake}) workflow to manage
+we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
 our inference pipeline and maintain reproducibility.
@@
 three replicates to inspect for inter-replicate variability. For humans, we ran
-whole genome simulations (chromosomes 1-22) under the Gutenkunst model (\cite{gutenkunst2009inferring}),
-using the HapMap II recombination map (\cite{international2007second}) and taking 20
+whole genome simulations (chromosomes 1-22) under the Gutenkunst model \citep{gutenkunst2009inferring},
+using the HapMap II recombination map \citep{international2007second} and taking 20
 samples each from the Europe and Africa populations. For \emph{Drosophila},
-we simulated 3 Mb of chromosome 2R under the Li and Stephan model (\cite{li2006inferring}),
+we simulated 3 Mb of chromosome 2R under the Li and Stephan model \citep{li2006inferring},
 using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
@@
 IM simulations, we used the human genome (chromosomes 1-22) and HapMap II recombination
-map (\cite{international2007second}) and took 20 samples per population.
+map \citep{international2007second} and took 20 samples per population.
 

### 2019-12-16 popSimManu.tex
added methods bit on Analysis, edited title
few changes from Chris

@@
 
-\title{\stdpopsim: a community toolkit for population genetic simulation}
+\title{\stdpopsim: a community-maintained standard library of population genetic simulations}
 
@@
 wants to introduce a new model into \stdpopsim. Developer A implements the
-demographic model for the relevant organism along with a series of required
-unit tests and documentation. This model is submitted as a Pull Request,
+demographic model for the relevant organism along with clear documentation
+of the model parameters and populations. This model is submitted as a Pull Request,
 where it is evaluated by a reviewer and then included as `preliminary',
@@
 \subsection*{Workflow for analysis of simulated data}
-Snakemake pipeline for 1 pop models
+To demonstrate the utility of \stdpopsim we created \texttt{Snakemake}
+workflows \citep{koster2012snakemake} that perform demographic inference on
+tree sequence output from our package using a few common software packages. 
+Our choice of \texttt{Snakemake} allows complete reproducibility of the 
+analyses shown and the entire Analysis repository is available from
+\url{https://github.com/popgensims/analysis}.
 
-\subsection*{Workflow for two population analysis}
+We performed demographic analysis for one and two population tasks.
+Our one population task was to infer historical population size over
+time ($N(t)$). This was done using three software packages \stairwayplot
+which uses site frequency spectrum information only \citep{liu2015exploring},
+the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
+sample sizes $n\in (2 , 8)$, and \smcpp~\citep{terhorst2017robust} 
+which combines information from the site frequency spectrum with 
+recombination information as in SMC-based methods. No attempt
+was made at trying to optimize the analysis from
+any particular software package in present case,
+as our goal was not to benchmark performance of methods but
+instead show how such benchmarking could be easily done using
+the \stdpopsim resource. In this spirit we ran each software package as near
+to default parameters as possibile in each case. For \stairwayplot we
+set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
+``estimate" run mode to infer $N(t)$ with all other parameters set
+to their default values. For \MSMC we used the ``--fixedRecombination"
+option and used the default number of iterations.
+
+For the single population task we ran human, HomSap, simulations
+using a variety of models: OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
+a constant sized generic model, and a two epoch generic model where the
+population size instantaneously decreases from $N=10^4$ to $N=10^3$ 500 
+generations before the present. Each HomSap simulation was run
+using the HapmapII\_GRCh37 genetic maps. For \emph{D. melanogaster}
+we estimated $N(t)$ from an African sample simulated under the DroMel, 
+African3Epoch\_1S16 model using the Comeron2012\_dm6 map. Finally we ran
+simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18
+model under the Salome2012\_TAIR7. For each model, 3 replicate whole genomes were
+simulated and the population size estimated from those data. In all cases we
+set the sample size of the focal population to $N=50$ chromosomes.
+
+Following simulation, low recombination portions of chromosomes were masked
+from the analysis in a manner that reflects the ``accessible" subset of sites
+used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
+Specifically we masked regions of 1cM or greater in the lowest 5th percentile of the empirical
+distribution of recombination, regions which are nearly uniformly absent for
+empirical analysis.
+
 Our goal for the two population analysis pipeline was to explore two population inference
 on some of the multi-population demographic models implemented in \stdpopsim.
-The models we examined include the human Gutenkunst model \citep{gutenkunst2009inferring} as
-well as the \emph{Drosophila} Li and Stephan model \citep{li2006inferring}. The human
-model is a three population model (Africa, Europe, and Asia) including post-divergence
+For HomSap we examined the OutOfAfrica_3G09 model with the HapmapII\_GRCh37 genetic map,
+and for DroMel we simulated under the OutOfAfrica_2L06 model with the Comeron2012\_dm6.
+The HomSap model is a three population model (Africa, Europe, and Asia) including post-divergence
 migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
-\emph{Drosophila} model is a two population model (Africa and Europe) with no post-divergence
+DroMel model is a two population model (Africa and Europe) with no post-divergence
 migration and constant population sizes (Figure \ref{fig:two_popn_fly}).
 
-To conduct inference on these models, we implemented three commonly-used approaches:
+To conduct inference on these models, we used three commonly used methods:
 \dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
-and \smcpp~\citep{terhorst2017robust}. Our aim was not to exhaustively
-test the performance of these methods, but rather to explore their behavior
-in two different inference scenarios and demonstrate the potential for this resource to
-eventually be used for more in-depth benchmarking. As a result, these methods were used
+and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
 generally with default settings and we did not attempt to optimize their performance or fit
@@
 
-For both \dadi and \fastsimcoal2, our general approach was to fit a two population
+For both \dadi and \fastsimcoal2, we fit a two population
 isolation with migration (IM) model with constant population sizes to the simulated
-data from the human and \emph{Drosophila} models. This IM model contains six parameters:
+data from the HomSap and DroMel models. This IM model contains six parameters:
 the ancestral population size, the size of population 1 after the split, the size of
 population 2 after the split, the divergence time, and two migration rate parameters.
-Importantly, this meant that for both the human and \emph{Drosophila} inference, the
+Importantly, this meant that for both species, the
 fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human}C and Sx).
-In the human case, we therefore performed inference solely on the Africa
+In the HomSap case, we therefore performed inference solely on the Africa
 and Europe populations, meaning that the Asia population functioned as a ghost
@@
 
-To perform inference, we began by running simulations under each model, running
-three replicates to inspect for inter-replicate variability. For humans, we ran
-whole genome simulations (chromosomes 1-22) under the Gutenkunst model \citep{gutenkunst2009inferring},
-using the HapMap II recombination map \citep{international2007second} and taking 20
-samples each from the Europe and Africa populations. For \emph{Drosophila},
-we simulated 3 Mb of chromosome 2R under the Li and Stephan model \citep{li2006inferring},
-using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
+From HomSap we ran replicate whole genome simulations under the model
+and specified taking 20 samples each from the Europe and Africa populations. 
+For DroMel, runtimes under OutOfAfrica_2L06 were prohibitively slow for
+running whole genome simulations. Instead we simulated 3 Mb of chromosome 2R.
+and take 50 samples from each of
 the Africa and Europe populations. We tried to run whole genome simulations under
 this model but after XXXX amount of time... (JRA fill in details here). For the generic
-IM simulations, we used the human genome (chromosomes 1-22) and HapMap II recombination
-map \citep{international2007second} and took 20 samples per population.
+IM simulations, we used the HomSap genome along with the HapmapII\_GRCh37 genetic map,
+and sampled 20 individuals from each population.
 

### 2019-12-16 popSimManu.tex
cleaned minor typos

@@
 on some of the multi-population demographic models implemented in \stdpopsim.
-For HomSap we examined the OutOfAfrica_3G09 model with the HapmapII\_GRCh37 genetic map,
-and for DroMel we simulated under the OutOfAfrica_2L06 model with the Comeron2012\_dm6.
+For HomSap we examined the OutOfAfrica\_3G09 model with the HapmapII\_GRCh37 genetic map,
+and for DroMel we simulated under the OutOfAfrica\_2L06 model with the Comeron2012\_dm6.
 The HomSap model is a three population model (Africa, Europe, and Asia) including post-divergence
@@
 and specified taking 20 samples each from the Europe and Africa populations. 
-For DroMel, runtimes under OutOfAfrica_2L06 were prohibitively slow for
+For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow for
 running whole genome simulations. Instead we simulated 3 Mb of chromosome 2R.

### 2019-12-17 popSimManu.tex
passed through the discussion

@@
 
-\title{\stdpopsim: a community-maintained standard library of population genetic simulations}
+\title{A community-maintained standard library of population genetic simulations}
 
@@
 
+We can even do away with the demographic model entirely, and simulate under a human
+genome architecture and genetic map to see how well methods might to at recovering
+the population history of a constant sized population. We show results of such an
+experiment in Figure \ref{fig:n_t_HomSap_constant}. Again we see that all methods
+are recovering a modern population size within a factor of two of the truth, however
+SMC-based methods, perhaps due to their regularization, tend to learn sinusoidal 
+patterns of population size even though no change is present.
+
 \begin{figure}
@@
 
-Although we did not produce an exhaustive comparison of different demographic inference methods,
-our results are informative,
+In the use case comparisons we have performed,
+we found that the methods tested \emph{do} generally perform well on human genomes,
+up to, for instance, the usual factor-of-two differences in inferred population sizes.
+This will not be surprising to people familiar with the landscape of
+demographic inference methods in population genetics,
+but the nature and scale of uncertainty may be not clear to people new to the field.
+The ability for \stdpopsim to generate data with and without significant features, such
+as the genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
+us to ask directly about the various failure modes of popular methods. 
+Moreover the comparison of methods across the various genome organizations, genetic maps,
+and demographic histories of different organisms provides valuable information
+about how methods might perform on non-human systems.
+In our initial comparison we find that inference of population histories from \emph{Drosophila}
+or \emph{Arabidopsis} simulated genomes is a fair bit more difficult from SMC based methods
+than in the human genome. Likely this is consequence of the reduced physical and genetic
+map lengths of these genomes, although we will leave further investigation of this to the future.
+Finally, comparison of results across methods or simulation runs
+gives a strong estimate of inference uncertainty, analogously to parametric bootstrapping --
+although different methods have different strengths and weaknesses,
+strong conclusions should probably not rest entirely on features inferred by only one method.
+
+The broad brush comparisons we have done, while far from exhaustive, are informative,
 and highlight the utility of easy comparison of methods on the same footing.
@@
 produced somewhat different results.
-We found that the methods tested \emph{do} generally perform well,
-up to, for instance, the usual factor-of-two differences in bottleneck population sizes.
-This will not be surprising to people familiar with the landscape of
-demographic inference methods in population genetics,
-but the nature and scale of uncertainty may be not clear to people new to the field.
-Comparison of results across methods or simulation runs
-gives a strong estimate of inference uncertainty, analogously to parametric bootstrapping --
-although different methods have different strengths and weaknesses,
-strong conclusions should probably not rest entirely on features inferred by only one method.
 
@@
 
-\paragraph{The problem of comparison} % Moved from the intro
-Benchmarking the performance of inference methods in population genetics is not straightforward.
-Different methods often fit different types of model
-or use different summaries of the genetic variation data, even when
-attempting to address the same biological question. For example, there are
-numerous methods to infer population size changes.
-The \MSMC method \citep{schiffels2014inferring} allows for the
-population to continuously change in size over time. Other methods, such as
-\dadi \citep{gutenkunst2009inferring}, allow for a smaller number of size changes to occur at certain
-points throughout the population's history. The types of data that the
-methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
-spatial patterns of genetic variation along the sequence, \dadi  instead uses the
-site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
-sample, ignoring any information about the correlation structure (linkage
-disequilibrium) among SNPs. Thus from the start these methods may not be
-expected to perform equally well in all scenarios,
-and indeed, it is not always clear how to compare to the truth --
-for instance, how well can a method of piecewise constant population sizes
-describe a period of exponential growth?
-Or, how should discrete population models be mapped onto real-world (continuous) geography?
-The answer depends on determining what is most important for the question at hand.
+% \paragraph{The problem of comparison} % Moved from the intro
+% Benchmarking the performance of inference methods in population genetics is not straightforward.
+% Different methods often fit different types of model
+% or use different summaries of the genetic variation data, even when
+% attempting to address the same biological question. For example, there are
+% numerous methods to infer population size changes.
+% The \MSMC method \citep{schiffels2014inferring} allows for the
+% population to continuously change in size over time. Other methods, such as
+% \dadi \citep{gutenkunst2009inferring}, allow for a smaller number of size changes to occur at certain
+% points throughout the population's history. The types of data that the
+% methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
+% spatial patterns of genetic variation along the sequence, \dadi  instead uses the
+% site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
+% sample, ignoring any information about the correlation structure (linkage
+% disequilibrium) among SNPs. Thus from the start these methods may not be
+% expected to perform equally well in all scenarios,
+% and indeed, it is not always clear how to compare to the truth --
+% for instance, how well can a method of piecewise constant population sizes
+% describe a period of exponential growth?
+% Or, how should discrete population models be mapped onto real-world (continuous) geography?
+% The answer depends on determining what is most important for the question at hand.
 
@@
 \begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.pdf}
+\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
+\MSMC with $n=2$ and $n=8$. Data were generated by simulating
+replicate human genomes under a constant sized population model with $N=10^4$ and using the
+HapMapII genetic map~\citep{international2007second}. In black we show the true population size history
+of the model.}
+\label{fig:n_t_HomSap_constant}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}

### 2019-12-17 popSimManu.tex
added blurb on model limitations

--- 2019-12-16 popSimManu.tex
@@
 
+The models implemented are currently limited to neutral evolution and to populations 
+composed of a finite number of randomly-mating demes. Such models do not reflect 
+the complexities of demographic and selective forces affecting real populations. 
+However, because they are comparatively simple to specify, easy to simulate, 
+and widely use in practice,  they serve as a useful starting point for a 
+comprehensive library of population genetics models.
+
 %\adk{should we mention benchmarking of tree sequence inference here? what else?}

### 2019-12-17 popSimManu.tex
added text about dmel sims to methods
added text about dmel sims to methods

--- 2019-12-17 popSimManu.tex
@@
 analyses we have performed do not capture recent population size change in CEU, however
-they do a good job of estimating the quite simple YRI population size history.
+they do a good job of estimating the more simple simulated YRI population size history.
 
@@
 The tests here were under best-case scenarios --
-simulations of discrete, randomly mating populations that fit the assumptions of the methods nearly exactly --
+simulations of discrete, randomly mating populations that evolve neutrally and 
+fit the assumptions of the methods nearly exactly --
 and yet different methods, as well as applications of the same method
@@
 and specified taking 20 samples each from the Europe and Africa populations. 
-For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow for
-running whole genome simulations. Instead we simulated 3 Mb of chromosome 2R.
-and take 50 samples from each of
-the Africa and Europe populations. We tried to run whole genome simulations under
-this model but after XXXX amount of time... (JRA fill in details here). For the generic
-IM simulations, we used the HomSap genome along with the HapmapII\_GRCh37 genetic map,
-and sampled 20 individuals from each population.
+For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
+For example, both chromosomes 3R and X failed to coalesce within 30 days of starting using $n = 50$. 
+Runtimes using the identical sample size were 19, 21, and 12 days for chromosomes 2L, 2R, and 3L, 
+respectively. For this reason, we chose to present only data from chromosome 2R for simulations
+under OutOfAfrica\_2L06. For the generic IM simulations, we used the HomSap genome along with the 
+HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
 

### 2019-12-17 popSimManu.tex

--- 2019-12-17 popSimManu.tex
@@
 analyses we have performed do not capture recent population size change in CEU, however
-they do a good job of estimating the more simple simulated YRI population size history.
+they do a good job of estimating the quite simple YRI population size history.
 
@@
 and specified taking 20 samples each from the Europe and Africa populations. 
-For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
-For example, both chromosomes 3R and X failed to coalesce within 30 days of starting using $n = 50$. 
-Runtimes using the identical sample size were 19, 21, and 12 days for chromosomes 2L, 2R, and 3L, 
-respectively. For this reason, we chose to present only data from chromosome 2R for simulations
-under OutOfAfrica\_2L06. For the generic IM simulations, we used the HomSap genome along with the 
-HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
+For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow for
+running whole genome simulations. Instead we simulated 3 Mb of chromosome 2R.
+and take 50 samples from each of
+the Africa and Europe populations. We tried to run whole genome simulations under
+this model but after XXXX amount of time... (JRA fill in details here). For the generic
+IM simulations, we used the HomSap genome along with the HapmapII\_GRCh37 genetic map,
+and sampled 20 individuals from each population.
 

### 2019-12-17 popSimManu.tex
consistency: samples vs populations

@@
 models it is interesting to note that $N(t)$ estimates for the CHB and CEU
-simulated populations are generally better across methods than estimates from the YRI
-sample.
+simulated samples are generally better across methods than estimates from the YRI simulations.
 

### 2019-12-17 popSimManu.tex
clarified noise vs discrepancy

@@
 of the methods is capturing relevant parts of the population history, although the
-degree of noise at any given time interval varies. In comparing inferences between the
+mismatch between simulated and inferred values varies over time. In comparing inferences between the
 models it is interesting to note that $N(t)$ estimates for the CHB and CEU

### 2019-12-18 popSimManu.tex

--- 2019-12-17 popSimManu.tex
@@
 of the methods is capturing relevant parts of the population history, although the
-degree of noise at any given time interval varies. In comparing inferences between the
+discrepancy between inferred and simulated values is changes over time. In comparing inferences between the
 models it is interesting to note that $N(t)$ estimates for the CHB and CEU
-simulated samples are generally better across methods than estimates from the YRI simulations.
+simulated populations are generally better across methods than estimates from the YRI
+simulated population.
 
@@
 analyses we have performed do not capture recent population size change in CEU, however
-they do a good job of estimating the more simple simulated YRI population size history.
+they do a good job of estimating the simpler simulated YRI population size history.
 
@@
 The tests here were under best-case scenarios --
-simulations of discrete, randomly mating populations that evolve neutrally and 
-fit the assumptions of the methods nearly exactly --
+simulations of discrete, randomly mating populations that fit the assumptions of the methods nearly exactly --
 and yet different methods, as well as applications of the same method

### 2019-12-18 popSimManu.tex

@@
 \newcommand{\MSMC}{\texttt{MSMC}\xspace}
-\newcommand{\smcpp}{\texttt{smcpp}\xspace}
+\newcommand{\smcpp}{\texttt{smc++}\xspace}
 \newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
@@
 Recent advances in population genetic simulation have made it possible to
-simulate large and complex models, but describing such models for a particular
+simulate large and complex models, but specifying such models for a particular
 simulation engine remains a difficult and error-prone task.
@@
 Population genetics, as a field, also lacks standard benchmarks by which new tools for inference might
-be measured. Here we describe a new resource, stdpopsim, that attempts to rectify this situation.
-Stdpopsim is a community-driven, open source project, which provides a standard
+be measured. Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
+\stdpopsim is a community-driven, open source project, which provides a standard
 catalog of published simulation models from a wide range of organisms,
@@
 % PopSim assures correctness in our implementations through a mature, community-based quality control pipeline.
-Here, we encourange an even broader
-community of developers to participate and to demonstrate some examples of how
-it can be used to systematically compare demographic inference methods.}
+We share some examples demonstrating how \stdpopsim
+can be used to systematically compare demographic inference methods,
+and encourange an even broader
+community of developers to join in the effort to standardize
+population genetic methods.}
 \date{\today}
@@
 Here we describe \stdpopsim, a community-driven catalog
-of standardized simulations across a variety of common study organisms
-that will empower the field and usher in the next era of genetics data analysis.
+of standardized population genetics models spanning a variety of common study organisms.
+This resource will empower the field to rigorously evaluate population genetic software
+and select the best tool for a given task, thereby helping to usher in the next era of population genetic data analysis.
+%DRS: still not in love with this last sentence here, but tried to make it a little less sappy by adding some explanation.
+%ADK: made some edits here
 
-The degree to which modeling assumptions and different summaries of the
+The degree to which modeling assumptions and different choices of summaries of the
 data can affect the ultimate results has been challenging to assess. Yet, there
@@
 \citep[see][]{beichman2017comparison}.
-These distinct models differ in how they fit various summaries of the
-genetic variation data, suggesting that modeling choices can greatly affect the
+These distinct methods differ in how they model, summarize, and optimize fit to
+genetic variation data, suggesting that such design choices can greatly affect the
 performance of the inference. Furthermore, it is reasonable
@@
 most papers proposing new methods in population genetics. However, these
-simulation studies often use simulations of the particular models for which the method was designed,
+studies often use simulations of the particular models for which the method was designed,
 which showcases the novel method rather than rigorously comparing performance to competing methods.
@@
 The many steps of implementing a moderately complicated population genetics simulation,
-including translating published demographic models to a simulator
+including translating published demographic models into input for a simulator
 and obtaining genetic maps,
@@
 generation time) and one or more published demographic histories. Through
-either a command line interface or a simple Python API, the user can specify what
+either a command line interface or a simple Python API, the user can specify which
 organism, genetic map, chromosome, and demographic history they are interested in simulating, and are returned
@@
 distributed open source model, with strong procedures in place
-to continue its growth and accuracy.
+to continue its growth and maintain its accuracy.
 Importantly, we have rigorous quality control methods to ensure implemented models are accurate,
@@
 \paragraph{The \stdpopsim library}
-The central contribution from PopSim currently is \stdpopsim, a community
+The central contribution of the PopSim consortium currently is \stdpopsim, a community
 maintained library of empirical genome data and population genetics simulation
@@
 of two key elements.  Firstly, the library defines
-some basic information about its genome, with information about chromosome
+some basic information about its genome, including information about chromosome
 lengths, average mutation rates and so on. We also provide access to detailed
@@
 to find, we provide a mechanism to automatically download (and cache locally) a
-set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the
+set of previously published maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the
 HapMapII~\citep{international2007second} and
@@
 descriptions from the literature, which are the basis for accurate simulations
-of specific historical scenarios.
+of specific historical scenarios that have influenced present-day patterns of
+genetic polymorphism.
 
@@
 \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
-as the simulation engine. We plan to include other simulation
+as the default simulation engine. We plan to include other simulation
 engines to support scenarios that cannot be modelled under the coalescent framework,
@@
 
-\subsection*{\texttt{stdpopsim} Catalog}
-The current contents of the \texttt{stdpopsim} catalog are shown in Table X. At the time of
+\subsection*{The \stdpopsim Catalog}
+The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}. At the time of
 writing the PopSim Consortium has implemented previously published demographic models for
@@
 Gutenkunst model which did not include Asian populations, but more explicitly modeled
-recent recent human population growth, the \cite{browning2018ancestry} admixture model
+recent rapid human population growth, the \cite{browning2018ancestry} admixture model
 for American populations that models ancestral African, European, and Asian population
-components, and the three population out-of-Africa model that includes archaic admixture
-from \cite{ragsdale2019models}. Together these models
-contain features found in real data (e.g. bottlenecks, population growth,
-admixture) which are pertinent in the context of method development.
+components, and the three-population out-of-Africa model from \cite{ragsdale2019models} that includes archaic admixture
+. Together these models
+contain features believed to have widespread impacts in real data (e.g. bottlenecks, population growth,
+admixture) which are therefore highly pertinent in the context of method development.
 
@@
 \emph{Drosophila melanogaster} and one each from \emph{A. thaliana} and \emph{E. coli}.
-For \emph{D. melanogaster} we have implemented the three epoch model estimated from
-an African sample from \cite{sheehan2016deep} as well as the out-of-Africa divergence
+For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
+an African sample, as well as the out-of-Africa divergence
 and associated bottleneck model of \cite{li2006inferring} which jointly models African
@@
 competing methods, our aim at this stage is not to provide an exhaustive
-evaluation of these methods. Our hope is instead that future work built upon this resource
+evaluation or ranking of these methods. Our hope is instead that future work built upon this resource
 will enable more detailed exploration of the strengths and weaknesses of the numerous
@@
 in a number of the \stdpopsim models described above. The pipeline was generalized to
-run $R$ replicates with $C$ chromosomes for $N$ samples, for a total of $R \times C$
+run $R$ replicates with $C$ chromosomes for $n$ samples, for a total of $R \times C$
 simulations for each demographic model. After simulation, the pipeline prepares
 input files for each of the respective inference methods by grouping all
-chromosomes, for each sample, into a single file. This step results in an
-input file for each simulation replicates $R$, pertaining to each of the
+chromosomes, for each sample, into a single file. For each of the $R$ simulation replicates, this step results in an
+input file for each of the
 respective inference methods and derived from the same simulated tree sequences.
 Each of the inference programs are then run in parallel, followed by plotting of
-$N_e$ estimates from each program.
+$N(t)$ estimates from each program.
 
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
+size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot},
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
-of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
+of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
 $N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
@@
 run on whole genome simulations under a sophisticated demographic model
-and empirical genetic map. On rows
+and empirical genetic map. In each column of this figure
 we show inferred $N(t)$ for the three extant populations in the model.
-On columns we show comparisons among the methods (note that we compare two differing
+In each row we show comparisons among the methods (note that we compare two differing
 sample sizes for \MSMC). The solid black lines represent the true effective population
@@
 Blue lines represent estimates from each of three replicate whole genome simulations.
-While there is variation among methods and individual population estimates,
+While there is variation in accuracy among methods, populations, and individual replicates,
 generally spoken the methods are accurate for this model of human history.
@@
 inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show inferred
-parameters out of the depicted IM model, inferring population sizes, migration rates, and a split
+parameters of the depicted IM model, which include population sizes, migration rates, and a split
 time between CEU and YRI samples. C) Population size estimates for each population (rows)
@@
 $N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model. The population split date, $T_{div}$, is shown as
-an inset panel, with a common X-axis to the background figure.}
+for the rate of coalescence in the demographic model. The population split date, $T_{div}$, is shown at
+the bottom, with a common X-axis to the background figure.}
 \label{fig:IM_popn_human}
@@
 explored parameter estimation of population divergence parameters. In particular
-we performed simulations under multi-population models for humans and Drosophila
+we performed simulations under multi-population models for humans and \emph{Drosophila}
 and inferred simulated parameters using \dadi, \fastsimcoal, and \smcpp.
@@
 analyses we have performed do not capture recent population size change in CEU, however
-they do a good job of estimating the simpler simulated YRI population size history.
+they do a good job of estimating the simpler simulated YRI population size history from this model.
 
-Again we can compare between genomes and look at the performance of these methods in
-the context of a two population model from \emph{Drosophila melanogaster}. Figure
+Again we can compare between species and look at the performance of these methods in
+the context of a two-population model from \emph{Drosophila melanogaster}. Figure
 \ref{fig:two_popn_fly} shows parameter estimates from simulations drawn from
@@
 the methods at hand under a wide range of simulations, but instead to show how
-the tools that we have built could be used to do so.
+the tools that we have built could be used to facilitate such comparisons.
 
@@
 standardization of complex population genetic simulations. Population genetics, as a field,
-has yet to coalesce around a set of gold-standards for the crucial task of simulation.
+has yet to coalesce around a set of gold-standards for the crucial task of method
+evaluation, which in our field hinges on simulation. In contrast, other fields such as
+structural biology (\citep{moult1995large}) and machine learning (\citep{russakovsky2015imagenet}) have a long track record
+of standardized method testing. We hope that our efforts represent the beginning of what
+will prove to be an equally longstanding and invaluable tradition in population genetics.
+
 Moreover individual research groups are often in the position of having to independently
@@
 means that new researchers, who are interested in adding their favorite organism/genome
-are free to with very few barriers to entry. Further new demographic models can be added
+are free to with very few barriers to entry (see Methods). Further new demographic models can be added
 to each of the organisms in the catalog through similarly well documented steps.
@@
 or \emph{Arabidopsis} simulated genomes is a fair bit more difficult from SMC based methods
-than in the human genome. Likely this is consequence of the reduced physical and genetic
+than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
 map lengths of these genomes, although we will leave further investigation of this to the future.
@@
 to different simulations of the same model,
-produced somewhat different results.
+produced somewhat different results. Future efforts will extend the resource described here to
+encompass additional population genetic tasks, such as detecting natural selection, inferring
+recombination rates, etc, in hopes of producing a general-purpose resource for fair and
+standardized evaluation of the ever-increasing arsenal of computational population genetic
+inference tools.
 
@@
 to work out the details of what went wrong. If that fails, the original
-authors of the published model A is trying to implement must be contacted
+authors of the published model must be contacted
 to resolve ambiguities. This process can lead to cycles of revision of
@@
 
-We performed demographic analysis for one and two population tasks.
+We performed demographic analysis for one-and two-population tasks.
 Our one population task was to infer historical population size over
@@
 
-For the single population task we ran human, HomSap, simulations
+For the single population task we ran human (HomSap) simulations
 using a variety of models: OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
-a constant sized generic model, and a two epoch generic model where the
+a constant-sized generic model, and a two-epoch generic model where the
 population size instantaneously decreases from $N=10^4$ to $N=10^3$ 500
@@
 simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18
-model under the Salome2012\_TAIR7. For each model, 3 replicate whole genomes were
+model under the Salome2012\_TAIR7 map. For each model, 3 replicate whole genomes were
 simulated and the population size estimated from those data. In all cases we
@@
 
-Following simulation, low recombination portions of chromosomes were masked
+Following simulation, low-recombination portions of chromosomes were masked
 from the analysis in a manner that reflects the ``accessible" subset of sites
@@
 
-Our goal for the two population analysis pipeline was to explore two population inference
+Our goal for the two-population analysis pipeline was to explore two-population inference
 on some of the multi-population demographic models implemented in \stdpopsim.
 For HomSap we examined the OutOfAfrica\_3G09 model with the HapmapII\_GRCh37 genetic map,
-and for DroMel we simulated under the OutOfAfrica\_2L06 model with the Comeron2012\_dm6.
+and for DroMel we simulated under the OutOfAfrica\_2L06 model with the Comeron2012\_dm6 map.
 The HomSap model is a three population model (Africa, Europe, and Asia) including post-divergence
@@
 For both \dadi and \fastsimcoal2, we fit a two population
-isolation with migration (IM) model with constant population sizes to the simulated
+isolation-with-migration (IM) model with constant population sizes to the simulated
 data from the HomSap and DroMel models. This IM model contains six parameters:
@@
 
-Following simulation, we outputted tree sequences and masked low recombination
+Following simulation, we outputted tree sequences and masked low-recombination
 regions using the same approach described for the single population pipeline above. We
-converted tree sequences into a two dimensional site frequency spectrum for all
+converted tree sequences into a two-dimensional site frequency spectrum for all
 chromosomes in the appropriate format for \dadi and \fastsimcoal2. For each simulation
@@
 \includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}
-\caption{\textbf{A comparison two-population parameter estimates in Drosophila}. Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+\caption{\textbf{A comparison two-population parameter estimates in \emph{Drosophila}}. Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
 Data were generated by simulating

### 2019-12-18 popSimManu.tex
Various edits for clarity and grammar.

@@
 \newcommand{\fastsimcoal}{\texttt{fastsimcoal}\xspace}
+\newcommand{\tskit}{\texttt{tskit}\xspace}
 
@@
 used have all increased over the past decade. Currently there exist myriad computational methods
-that can infer the histories of populations, the distribution of fitness effects,
-recombination rates, the extent of positive selection in genome sequence data
-\citep{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
-chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
-kern2018diplos,sugden2018localization}.
+that can infer the histories of populations 
+\citep{gutenkunst2009inferring,li2011inference,excoffier2013robust,schiffels2014inferring, terhorst2017robust,ragsdale2019models}, 
+the distribution of fitness effects \citep{kim2017inference},
+recombination rates \citep{chan2012genome,lin2013fast,Adrion662247}, 
+the extent of positive selection in genome sequence data
+\citep{eyre2009estimating, alachiotis2012omegaplus,degiorgio2016sweepfinder2,kern2018diplos,sugden2018localization}.
 While en masse these methods have increased our understanding of the
@@
 either a command line interface or a simple Python API, the user can specify which
-organism, genetic map, chromosome, and demographic history they are interested in simulating, and are returned
-simulation output from their chosen model.
+organism, genetic map, chromosome, and demographic history they are interested in simulating, and the
+simulation output from their chosen model is returned.
 
@@
 for \emph{Homo sapiens} only. We see that \emph{H. sapiens} is associated
-with a default generation time and mutation rate, that it contains within
-it a representation of the physical genome, genetic maps, and as shown a
+with a default generation time and mutation rate,
+a representation of the physical genome, and two genetic maps. Also shown is a
 subset of the demographic models currently implemented. \textbf{B.} Caching
@@
 and simulate models using \stdpopsim via the API (top) or
-command-line interface (bottom). \textbf{D.} Simulation output from
-\stdpopsim is generated as tree sequence files that can
-be manipulated and analysed with \texttt{tskit}.}
+command-line interface (bottom). \textbf{D.} Simulation output
+is a tree sequence file that can
+be manipulated and analysed with \tskit.}
 \label{fig:cartoon}
@@
 \paragraph{The \stdpopsim library}
-The central contribution of the PopSim consortium currently is \stdpopsim, a community
-maintained library of empirical genome data and population genetics simulation
+The central contribution of the PopSim consortium currently is \stdpopsim, a
+community-maintained library of empirical genome data and population genetics simulation
 models, providing easy access to realistic simulations. Figure \ref{fig:cartoon} shows a graphical
@@
 of two key elements.  Firstly, the library defines
-some basic information about its genome, including information about chromosome
+some basic information about the species' genome, including information about chromosome
 lengths, average mutation rates and so on. We also provide access to detailed
-empirical information such as recombination maps, which model empirically
-observed heterogeneity along chromosomes. As such maps are often large and hard
+empirical information such as recombination maps, which model empirically-observed
+heterogeneity along chromosomes. As such maps are often large and hard
 to find, we provide a mechanism to automatically download (and cache locally) a
@@
 contains complete genealogical information about the simulated samples, is
-extremely compact, and can be processed very efficiently using the tskit library
+extremely compact, and can be processed very efficiently using the \tskit library
 \citep{kelleher2016efficient,kelleher2018efficient}. Currently,
@@
 any other organism. These models include: a simplified version of the \cite{tennessen2012evolution}
-model with only the African population modeled (expansion from the ancestral
+model with only the African population specified (expansion from the ancestral
 population and recent growth), the three population model of \cite{gutenkunst2009inferring}
-which modeled the out-of-Africa bottleneck as well as the subsequent divergence of
+which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
 the European and Asian populations, the \cite{tennessen2012evolution} two population variant of the
-Gutenkunst model which did not include Asian populations, but more explicitly modeled
+Gutenkunst model which does not include Asian populations, but more explicitly models
 recent rapid human population growth, the \cite{browning2018ancestry} admixture model
-for American populations that models ancestral African, European, and Asian population
-components, and the three-population out-of-Africa model from \cite{ragsdale2019models} that includes archaic admixture
-. Together these models
+for American populations which specifies ancestral African, European, and Asian population
+components, and the three-population out-of-Africa model from \cite{ragsdale2019models}
+which includes archaic admixture.
+Together these models
 contain features believed to have widespread impacts in real data (e.g. bottlenecks, population growth,
-admixture) which are therefore highly pertinent in the context of method development.
+admixture), and are therefore highly pertinent in the context of method development.
 
@@
 and European populations. For \emph{Arabidopsis thaliana}, we implemented the
-model from in \cite{durvasula2017african} inferred using \MSMC. This model includes
+model in \cite{durvasula2017african} inferred using \MSMC. This model includes
 a continuous change in population size over time, rather than pre-specified epochs of different
@@
 \subsection*{Use case: comparing methods of demographic inference}
-As an example of the utility of \stdpopsim we demonstrate how popular
-methods can be simply and fairly compared in the setting of demographic
-inference using our resource. Although we present comparison of results from several
+As an example of the utility of \stdpopsim, we demonstrate how it can be used 
+to simply and fairly compare popular demographic inference methods.
+Although we present comparison of results from several
 competing methods, our aim at this stage is not to provide an exhaustive
@@
 show simple examples of multi-population inference.
-To evaluate and compare the performance of inference methods we developed
-\texttt{snakemake} \citep{koster2012snakemake} workflows that allow reproducibility
-in comparing inference methods which are available from \url{https://github.com/popgensims/analysis}
-. The workflows allow efficient computing in multicore environments or on
-cluster environments.
+To reproducibly evaluate and compare the performance of inference methods, we developed
+\texttt{snakemake} \citep{koster2012snakemake} workflows,
+available from \url{https://github.com/popgensims/analysis}.
+These workflows allow efficient computing in multicore or cluster environments.
 
@@
  on simulated genomes sampled from a single population,
-in a number of the \stdpopsim models described above. The pipeline was generalized to
+in a number of the demographic models described above. The pipeline was generalized to
 run $R$ replicates with $C$ chromosomes for $n$ samples, for a total of $R \times C$
@@
 While there is variation in accuracy among methods, populations, and individual replicates,
-generally spoken the methods are accurate for this model of human history.
+generally the methods are accurate for this model of human history.
 
@@
 $N(t)$ from simulations using the same physical and genetic maps, but a different demographic
-history -- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
+history---that inferred by \cite{gutenkunst2009inferring}. Again we see that each
 of the methods is capturing relevant parts of the population history, although the
-discrepancy between inferred and simulated values is changes over time. In comparing inferences between the
+discrepancy between inferred and simulated values varies with $t$. In comparing inferences between the
 models it is interesting to note that $N(t)$ estimates for the CHB and CEU
@@
 We can even do away with the demographic model entirely, and simulate under a human
-genome architecture and genetic map to see how well methods might to at recovering
+genome architecture and genetic map to see how well methods might do at recovering
 the population history of a constant sized population. We show results of such an
@@
 
-As most population genetic methods development has been focused on human-scale
+As most method development for population genetics has been focused on human-scale
 data, it is of consequence to ask how such methods might perform in non-human
@@
 As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
-an empirically derived genetic map, and then try to infer back the parameters of model.
+an empirically derived genetic map, and then try to infer back the parameters of the model.
 Accuracy is mixed among methods in this setting, and generally worse than what we
@@
 \includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.png}
-\caption{\textbf{A comparison two-population parameter estimates in humans}. Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+\caption{\textbf{Parameters estimated using a multi-population human model}.
+Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
 A) Data were generated by simulating
 replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show inferred
-parameters of the depicted IM model, which include population sizes, migration rates, and a split
+inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show parameters inferred
+by fitting the depicted IM model, which includes population sizes, migration rates, and a split
 time between CEU and YRI samples. C) Population size estimates for each population (rows)
@@
 
-\paragraph*{Multi Population demographic models}
+\paragraph*{Multi-population demographic models}
 As \texttt{stdpopsim} implements multi-population demographic models, we also
@@
 and inferred simulated parameters using \dadi, \fastsimcoal, and \smcpp.
-For simplicity, we conducted inference in \dadi and \fastsimcoal using an IM model
+For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
 with constant population sizes and bi-directional migration. For human
 models with more than two-populations, e.g. \cite{gutenkunst2009inferring},
-this means that we are inferring parameters from a model which itself does
+this means that we are inferring parameters for a model which itself does
 not match the model from which the data were generated (Figure
@@
 In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
-time from each of the methods for samples simulated for the African and European populations
-under the \cite{gutenkunst2009inferring} model. Each of the methods compared has stronger
+time, for each of the inference methods, using samples drawn from African and European populations
+simulated under the \cite{gutenkunst2009inferring} model. Each of the methods compared has stronger
 and weaker points to their accuracy across parameters. For instance the simple, IM-style
@@
 the context of a two-population model from \emph{Drosophila melanogaster}. Figure
-\ref{fig:two_popn_fly} shows parameter estimates from simulations drawn from
-the model from \cite{li2006inferring} which includes
+\ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
+the model of \cite{li2006inferring}, which includes
 an ancestral population in Africa, then a population expansion with a population
@@
 towards the more ancient past and all methods failing to capture the very brief
-population bottleneck in Europe, although the IM-model of course only allows for
+population bottleneck in Europe, although the IM model of course only allows for
 two epochs in the population size history.
@@
 evaluation, which in our field hinges on simulation. In contrast, other fields such as
-structural biology (\citep{moult1995large}) and machine learning (\citep{russakovsky2015imagenet}) have a long track record
+structural biology \citep{moult1995large} and machine learning \citep{russakovsky2015imagenet} have a long track record
 of standardized method testing. We hope that our efforts represent the beginning of what
@@
 means that new researchers, who are interested in adding their favorite organism/genome
-are free to with very few barriers to entry (see Methods). Further new demographic models can be added
+are free to do so with very few barriers to entry (see Methods).
+Further, new demographic models can be added
 to each of the organisms in the catalog through similarly well documented steps.
 A long-term goal then is that through a mechanism such as \stdpopsim, the community
-can determine (in an open way) what is the correct implementation for a given model,
-and in turn, what are the models that the community is most interested in.
+can determine (in an open way) what is the correct implementation for a published demographic model,
+and in turn, what are the models in which the community is most interested.
 
@@
 Moreover the comparison of methods across the various genome organizations, genetic maps,
-and demographic histories of different organisms provides valuable information
+and demographic histories of different organisms, provides valuable information
 about how methods might perform on non-human systems.
-In our initial comparison we find that inference of population histories from \emph{Drosophila}
-or \emph{Arabidopsis} simulated genomes is a fair bit more difficult from SMC based methods
+In our initial comparison we find that inference of population histories from simulated \emph{Drosophila}
+or \emph{Arabidopsis} genomes is a fair bit more difficult from SMC based methods
 than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
@@
 Finally, comparison of results across methods or simulation runs
-gives a strong estimate of inference uncertainty, analogously to parametric bootstrapping --
-although different methods have different strengths and weaknesses,
+gives a strong estimate of inference uncertainty, analogous to parametric
+bootstrapping---although different methods have different strengths and weaknesses,
 strong conclusions should probably not rest entirely on features inferred by only one method.
@@
 and highlight the utility of easy comparison of methods on the same footing.
-The tests here were under best-case scenarios --
-simulations of discrete, randomly mating populations that fit the assumptions of the methods nearly exactly --
-and yet different methods, as well as applications of the same method
+The tests here were under best-case scenarios---simulations of discrete,
+randomly mating populations that fit the assumptions of the methods nearly
+exactly---and yet different methods, as well as applications of the same method
 to different simulations of the same model,
@@
 Our one population task was to infer historical population size over
-time ($N(t)$). This was done using three software packages \stairwayplot
-which uses site frequency spectrum information only \citep{liu2015exploring},
+time ($N(t)$). This was done using three software packages: \stairwayplot
+which uses site frequency spectrum information only \citep{liu2015exploring};
 the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
-sample sizes $n\in (2 , 8)$, and \smcpp~\citep{terhorst2017robust}
-which combines information from the site frequency spectrum with
+sample sizes $n\in (2 , 8)$; and \smcpp~\citep{terhorst2017robust},
+which combines information from the site frequency spectrum, with
 recombination information as in SMC-based methods. No attempt
 was made at trying to optimize the analysis from
-any particular software package in present case,
+any particular software package,
 as our goal was not to benchmark performance of methods but
@@
 the \stdpopsim resource. In this spirit we ran each software package as near
-to default parameters as possibile in each case. For \stairwayplot we
+to default parameters as possibile. For \stairwayplot we
 set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
@@
 
-For the single population task we ran human (HomSap) simulations
+For the single-population task we ran human (HomSap) simulations
 using a variety of models: OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
 a constant-sized generic model, and a two-epoch generic model where the
-population size instantaneously decreases from $N=10^4$ to $N=10^3$ 500
+population size instantaneously decreases from $N=10^4$ to $N=10^3$, 500
 generations before the present. Each HomSap simulation was run
@@
 simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18
-model under the Salome2012\_TAIR7 map. For each model, 3 replicate whole genomes were
+model under the Salome2012\_TAIR7 map. For each model, three replicate whole genomes were
 simulated and the population size estimated from those data. In all cases we
@@
 used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
-Specifically we masked regions of 1cM or greater in the lowest 5th percentile of the empirical
+Specifically we masked regions of 1\,cM or greater in the lowest 5th percentile of the empirical
 distribution of recombination, regions which are nearly uniformly absent for
@@
 
-To conduct inference on these models, we used three commonly used methods:
+To conduct inference on these models, we applied three commonly used methods:
 \dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
@@
 For both \dadi and \fastsimcoal2, we fit a two population
-isolation-with-migration (IM) model with constant population sizes to the simulated
+isolation-with-migration (IM) model, with constant population sizes, to the simulated
 data from the HomSap and DroMel models. This IM model contains six parameters:
@@
 Importantly, this meant that for both species, the
-fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human}C and Sx).
+fitted model did not match the simulated model (Figures \ref{fig:IM_popn_human} and \ref{fig:two_popn_fly}).
 In the HomSap case, we therefore performed inference solely on the Africa
@@
 on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
-The parameter estimates from the top run from each simulation replicate
-are shown in Figures \ref{fig:IM_popn_human}C and Sx.
+The parameter estimates from the top run of each simulation replicate
+are shown in Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C.
 
@@
 with default settings. Importantly, \smcpp assumes no migration post-divergence, so
-here again the inference model does not exactly match the simulated model for the
-human model that does include migration. However, because \smcpp allows
+here again the model fitted during inference does not exactly match the simulated
+human demography that does include migration. However, because \smcpp allows
 for continuous population size changes, it is better equipped to capture many of the
@@
 for each simulation replicate along side the true (census) population sizes
-(Figures \ref{fig:IM_popn_human}C Sx). Here, unlike the single population pipeline,
+(Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C). Here, unlike the single-population pipeline,
 we focus on census size rather than the inverse coalescence rate as the `true' population size.
@@
 \includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.pdf}
-\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
+\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
 size over time ($N(t)$) inferred using 4 different methods, \smcpp, \stairwayplot,
@@
 inferred in \cite{international2007second}. From top to bottom we show estimates for each
-of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
+of the three populations in the model: YRI, CEU, and CHB. In shades of blue we show the estimated
 $N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
@@
 \includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}
-\caption{\textbf{A comparison two-population parameter estimates in \emph{Drosophila}}. Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+\caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
+Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
 Data were generated by simulating
-replicate human genomes under the \cite{li2006inferring} model and using the genetic map
+replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
 inferred in \cite{comeron2012many}. See legend of Figure \ref{fig:IM_popn_human} for details.

### 2019-12-18 popSimManu.tex
minor edits to the intro

@@
 choosing pre-implemented models from a community maintained catalog.
-The \stdpopsim catalog currently contains four organisms: humans, 
+The \stdpopsim catalog currently contains five organisms: humans, orangutans,
 \emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}. For each
@@
 and have documented methods for others to contribute new modules.
-We welcome new collaborators.
+We invite new collaborators to join our community.
 Below we describe the resource and give an

### 2019-12-18 popSimManu.tex
added comment about popsim version and reproducibility. Addresses issue #50 added comment about popsim version and reproducibility. Addresses issue #60
added png version of some figs

@@
 of simulation models to contribute to our ongoing community-driven development process.
+Together with the \stdpopsim version number, this citation information will result in a reasonably well-documented and reproducible simulation workflow.
 
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.pdf}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.pdf}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
 \caption{\textbf{Comparing estimates of $N(t)$ in \emph{Drosophila}}. Population
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.pdf}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.png}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.pdf}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.png}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population

### 2019-12-18 popSimManu.tex
updates to two pop results

@@
 \newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
-\newcommand{\fastsimcoal}{\texttt{fastsimcoal}\xspace}
+\newcommand{\fastsimcoal}{\texttt{fastsimcoal2}\xspace}
 \newcommand{\tskit}{\texttt{tskit}\xspace}
@@
 be measured. Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
-\stdpopsim is a community-driven, open source project, which provides a standard
-catalog of published simulation models from a wide range of organisms,
+\stdpopsim is a community-driven open source project, which provides a standard
+catalog of published simulation models from a wide range of organisms
 and supports multiple simulation engine backends.
@@
 can be used to systematically compare demographic inference methods,
-and encourange an even broader
+and we encourage an even broader
 community of developers to join in the effort to standardize
@@
 recombination rates \citep{chan2012genome,lin2013fast,Adrion662247},
-the extent of positive selection in genome sequence data
+and the extent of positive selection in genome sequence data
 \citep{eyre2009estimating, alachiotis2012omegaplus,degiorgio2016sweepfinder2,kern2018diplos,sugden2018localization}.
@@
 constitute community standards or best practices. The general modus operandi to date has been to
-introduce methods which are validated by a set of bespoke simulations produced by individual
-groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
-to decreases in reproducibility and transparency across the entire field.
+introduce methods that are validated by a set of bespoke simulations produced by individual
+groups. Beyond inefficiency from duplicated effort, such a situation leads
+to decreased reproducibility and transparency across the entire field.
 Here we describe \stdpopsim, a community-driven catalog
@@
 This resource will empower the field to rigorously evaluate population genetic software
-and select the best tool for a given task, thereby helping to usher in the next era of population genetic data analysis.
+and select the best tool for a given task, thereby substantially increasing the reliability of population genetic inferences.
 %DRS: still not in love with this last sentence here, but tried to make it a little less sappy by adding some explanation.
 %ADK: made some edits here
+%RNG: Removed the sap. :-P Feel free to put it back if you like it better.
 
-The degree to which modeling assumptions and different choices of summaries of the
-data can affect the ultimate results has been challenging to assess. Yet, there
+The degree to which modeling assumptions and choices of data summaries can affect the ultimate results of population genetic inference has been challenging to assess. Yet there
 have been several clear examples of different methods yielding fundamentally
-different conclusions. For example, \MSMC methods applied to human genomes have
+different conclusions. For example, Markovian coalescent methods applied to human genomes have
 % JK is this right? > 100 years doesn't sound very ancient to me.
-suggested large ancient ($>100$ years ago) ancestral population sizes and
-bottlenecks that have not been detected by SFS-based methods
+% RNG: Fixed time and removed unexplained acronyms.
+suggested large ancient ($>100,000$ years ago) ancestral population sizes and
+bottlenecks that have not been detected by methods based on allele frequency spectra
 \citep[see][]{beichman2017comparison}.
@@
 realistic simulations for population genetic analysis a simple matter of
-choosing pre-implemented models from a community maintained catalog.
+choosing pre-implemented models from a community-maintained catalog.
 The \stdpopsim catalog currently contains five organisms: humans, orangutans,
@@
 to continue its growth and maintain its accuracy.
-Importantly, we have rigorous quality control methods to ensure implemented models are accurate,
+Importantly, we have rigorous quality control methods to ensure implemented models are accurate
 and have documented methods for others to contribute new modules.
@@
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{Drosophila},
-\emph{Arabidopsis}, \emph{E. coli}, and \emph{Pongo}. A species definition consists
-of two key elements.  Firstly, the library defines
+on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{D.~melanogaster}, \emph{A.~thaliana}, \emph{E.~coli}, and\emph{P.~abelii}.
+% RNG: Abbreviate consistently with introduction
+A species definition consists of two key elements.  Firstly, the library defines
 some basic information about the species' genome, including information about chromosome
@@
 across a range of organisms. \stdpopsim has a Python API and a user-friendly
-command line interface (Fig. \ref{fig:cartoon}C), allowing users with minimal experience direct access to
+command line interface (Fig.~\ref{fig:cartoon}C), allowing users with minimal experience direct access to
 state-of-the-art simulations. Simulations are output in the “tree sequence”
@@
 contains complete genealogical information about the simulated samples, is
-extremely compact, and can be processed very efficiently using the \tskit library
+extremely compact, and can be processed efficiently using the \tskit library
 \citep{kelleher2016efficient,kelleher2018efficient}. Currently,
@@
 engines to support scenarios that cannot be modelled under the coalescent framework,
-and have already implemented \texttt{SLiM} \citep{haller2019slim} as
+and we have already implemented \texttt{SLiM} \citep{haller2019slim} as
 an alternative back-end.
@@
 We hope that this will encourage users to appropriately acknowledge the
-resources used in published work, and authors
+resources used in published work and authors
 of simulation models to contribute to our ongoing community-driven development process.
@@
 The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}. At the time of
-writing the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Drosophila}, and \emph{Arabidopsis}. These range from
+writing, the PopSim Consortium has implemented previously published demographic models for
+humans, \emph{Drosophila}, and \emph{Arabidopsis}.
+% RNG: Pongo added by submission?
+These range from
 simple, single population histories \cite[e.g.,][]{sheehan2016deep},
@@
 contain features believed to have widespread impacts in real data (e.g. bottlenecks, population growth,
-admixture), and are therefore highly pertinent in the context of method development.
+admixture) and are therefore highly pertinent in the context of method development.
 
@@
 Although we present comparison of results from several
-competing methods, our aim at this stage is not to provide an exhaustive
+% RNG: Not competitors... We complement each other ;-)
+methods, our aim at this stage is not to provide an exhaustive
 evaluation or ranking of these methods. Our hope is instead that future work built upon this resource
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot},
+size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot}, and
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
-run on whole genome simulations under a sophisticated demographic model
-and empirical genetic map. In each column of this figure
+run on whole genome simulations under an empirical genetic map and a model of human migration out of Africa that includes archaic admiture \citep{ragsdale2019models}. In each column of this figure
 we show inferred $N(t)$ for the three extant populations in the model.
 In each row we show comparisons among the methods (note that we compare two differing
-sample sizes for \MSMC). The solid black lines represent the true effective population
-size as inferred from the inverse coalescent rate defined by the model.
-Blue lines represent estimates from each of three replicate whole genome simulations.
+sample sizes for \MSMC).
+% RNG: Removed duplicated info from caption.
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 $N(t)$ from simulations using the same physical and genetic maps, but a different demographic
-history---that inferred by \cite{gutenkunst2009inferring}. Again we see that each
+history that does not include archaic admixture \citep{gutenkunst2009inferring}. Again we see that each
 of the methods is capturing relevant parts of the population history, although the
@@
 
-We can even do away with the demographic model entirely, and simulate under a human
+We can even do away with the demographic model entirely and simulate under a human
 genome architecture and genetic map to see how well methods might do at recovering
@@
 
-As most method development for population genetics has been focused on human-scale
+As most method development for population genetics has been focused on human
 data, it is of consequence to ask how such methods might perform in non-human
@@
 an empirically derived genetic map, and then try to infer back the parameters of the model.
-Accuracy is mixed among methods in this setting, and generally worse than what we
+Accuracy is mixed among methods in this setting and generally worse than what we
 observe under simulations of the human genome.
@@
 \begin{center}
-\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.png}
+\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.pdf}
 \caption{\textbf{Parameters estimated using a multi-population human model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
 A) Data were generated by simulating
 replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show parameters inferred
+inferred in \cite{international2007second}. B) For \dadi and \fastsimcoal we show parameters inferred
 by fitting the depicted IM model, which includes population sizes, migration rates, and a split
@@
 from \dadi, \texttt{fastsimcoal2}, and \smcpp (columns). In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model. The population split date, $T_{div}$, is shown at
-the bottom, with a common X-axis to the background figure.}
+% RNG: Updated caption to reflect what is shown.
+$N(t)$ trajectories for each replicate. In black we show the census population size from the simulated model.
+The population split date, $T_{div}$, is shown at
+the bottom, with a common X-axis to the population size panels.}
 \label{fig:IM_popn_human}
@@
 As \texttt{stdpopsim} implements multi-population demographic models, we also
-explored parameter estimation of population divergence parameters. In particular
+explored two population inference with several popular methods. Specifically,
 we performed simulations under multi-population models for humans and \emph{Drosophila}
-and inferred simulated parameters using \dadi, \fastsimcoal, and \smcpp.
+and inferred demographic parameters using \dadi, \fastsimcoal, and \smcpp.
 For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
-with constant population sizes and bi-directional migration. For human
-models with more than two-populations, e.g. \cite{gutenkunst2009inferring},
+with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
+an IM model was to mimic the approach often used on empirical datasets, where a relatively
+simple model is fit that may not reflect the true underlying demography.
+For human models with more than two-populations (e.g. \cite{gutenkunst2009inferring})
 this means that we are inferring parameters for a model which itself does
 not match the model from which the data were generated (Figure
-\ref{fig:IM_popn_human}A and B), however this is common practice in the field
-when applying IM-type models.
+\ref{fig:IM_popn_human}A and B).
 
@@
 time, for each of the inference methods, using samples drawn from African and European populations
-simulated under the \cite{gutenkunst2009inferring} model. Each of the methods compared has stronger
-and weaker points to their accuracy across parameters. For instance the simple, IM-style
-analyses we have performed do not capture recent population size change in CEU, however
-they do a good job of estimating the simpler simulated YRI population size history from this model.
+simulated under the \cite{gutenkunst2009inferring} model. Our results highlight many
+of the strengths and weaknesses of the different types of methods we used.
+For instance, the SFS-based approaches where we fit simple IM models do not capture
+recent exponential growth in the CEU population, however do consistently recover the
+simulated YRI population size history. Moreover, these approaches allow for estimating
+migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
+of divergence times. By contrast, \smcpp is much better at capturing the recent exponential
+growth in the CEU population, though the inferred population sizes are in general more
+noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
+estimates that are consistently underestimated.
 
-Again we can compare between species and look at the performance of these methods in
+Again, we can compare between species and look at the performance of these methods in
 the context of a two-population model from \emph{Drosophila melanogaster}. Figure
@@
 an ancestral population in Africa, then a population expansion with a population
-split and bottleneck into a European population. These simulations are generated
-using the recombination map of \cite{comeron2012many}. Here methods differ dramatically
-in the estimates, with \smcpp yielding quite noisy estimates of population size
-towards the more ancient past and all methods failing to capture the very brief
-population bottleneck in Europe, although the IM model of course only allows for
-two epochs in the population size history.
+split and bottleneck into a European population with no post-divergence migration.
+Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
+parameters, but ignore the brief population bottleneck in Europe. The inferred
+demographic parameters from \smcpp are more noisy, though in some cases are better
+able to capture the bottleneck in the Europe population.
 
-Again the goal of these analyses are not to perform thorough benchmarking of
-the methods at hand under a wide range of simulations, but instead to show how
-the tools that we have built could be used to facilitate such comparisons.
+Although these results do not represent an exhaustive benchmarking,
+we have highlighted some of the strengths and weaknesses of these methods.
+Future work should build on these results and undertake more in-depth comparisons
+under a wider range of simulated demographic models.
 
@@
 standardization of complex population genetic simulations. Population genetics, as a field,
+% RNG: We should have more coalescence puns in the paper! :-P
 has yet to coalesce around a set of gold-standards for the crucial task of method
@@
 Moreover individual research groups are often in the position of having to independently
-reimplement complex, previously published models. \stdpopsim offers solutions to both of
+reimplement complex previously published models. \stdpopsim offers solutions to both of
 these issues by creating a quality controlled, easy-to-use, gateway to models and organisms
@@
 common study systems in computational genetics: humans, \emph{Drosophila}, and \emph{Arabidopsis}.
+% RNG: Needs to be updated for final count.
 The standard operating procedures that we have outlined in our detailed documentation
-means that new researchers, who are interested in adding their favorite organism/genome
-are free to do so with very few barriers to entry (see Methods).
+mean that new researchers, who are interested in adding their favorite organism/genome
+are free to do so with few barriers to entry (see Methods).
 Further, new demographic models can be added
 to each of the organisms in the catalog through similarly well documented steps.
-A long-term goal then is that through a mechanism such as \stdpopsim, the community
+A long-term goal is that the community
 can determine (in an open way) what is the correct implementation for a published demographic model,
@@
 or \emph{Arabidopsis} genomes is a fair bit more difficult from SMC based methods
+% RNG: No Arabidopsis results are presented in this paper
 than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
-map lengths of these genomes, although we will leave further investigation of this to the future.
+map lengths of these genomes, although we leave further investigation of this to the future.
 Finally, comparison of results across methods or simulation runs
@@
 \section*{Methods}
-\subsection*{Generating the PopSim resource}
+%\subsection*{Generating the PopSim resource}
+% RNG: No need to empty subsection.
 
 \subsection*{Model QC procedures}
-As a consortium we have agreed to a standardized proceedure for model inclusion
+As a consortium we have agreed to a standardized procedure for model inclusion
 into \stdpopsim that allows for rigorous quality control. Imagine Developer A
@@
 implementations are equal the original model is included in \stdpopsim.
-If not, we move to an arbitration process whereby the A and B first try
+If not, we move to an arbitration process whereby A and B first try
 to work out the details of what went wrong. If that fails, the original
@@
 Our choice of \texttt{Snakemake} allows complete reproducibility of the
-analyses shown and the entire Analysis repository is available from
+analyses shown, and the entire Analysis repository is available from
 \url{https://github.com/popgensims/analysis}.
@@
 sample sizes $n\in (2 , 8)$; and \smcpp~\citep{terhorst2017robust},
-which combines information from the site frequency spectrum, with
+which combines information from the site frequency spectrum with
 recombination information as in SMC-based methods. No attempt
@@
 we estimated $N(t)$ from an African sample simulated under the DroMel,
-African3Epoch\_1S16 model using the Comeron2012\_dm6 map. Finally we ran
-simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18
-model under the Salome2012\_TAIR7 map. For each model, three replicate whole genomes were
+African3Epoch\_1S16 model using the Comeron2012\_dm6 map.
+% RNG: No results from these simulations are included in the paper
+%Finally we ran simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.
+For each model, three replicate whole genomes were
 simulated and the population size estimated from those data. In all cases we
@@
 To conduct inference on these models, we applied three commonly used methods:
-\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
+\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal~\citep{excoffier2013robust},
 and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
@@
 
-For both \dadi and \fastsimcoal2, we fit a two population
+For both \dadi and \fastsimcoal, we fit a two population
 isolation-with-migration (IM) model, with constant population sizes, to the simulated
@@
 model is often misspecified. To ground-truth our inference approach, we also conducted
-inference on a generic IM model that was identical to the model used for inference.
+inference on a generic IM model that was identical to the model used for inference \ref{fig:generic_IM}.
 
@@
 For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
-For example, both chromosomes 3R and X failed to coalesce within 30 days of starting using $n = 50$.
+% RNG: Removed technical term "coalesce" here, since I don't think that's exactly the issue.
+For example, both chromosomes 3R and X failed to finish within 30 days of starting using $n = 50$.
 Runtimes using the identical sample size were 19, 21, and 12 days for chromosomes 2L, 2R, and 3L,
 respectively. For this reason, we chose to present only data from chromosome 2R for simulations
-under OutOfAfrica\_2L06. For the generic IM simulations, we used the HomSap genome along with the
-HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
+under OutOfAfrica\_2L06.
+% RNG: Not present in manuscript
+% For the generic IM simulations, we used the HomSap genome along with the
+% HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
 
@@
 converted tree sequences into a two-dimensional site frequency spectrum for all
-chromosomes in the appropriate format for \dadi and \fastsimcoal2. For each simulation
-replicate, we performed 10 runs of \dadi and \fastsimcoal2 and checked for convergence.
-Detailed settings for \dadi and \fastsimcoal2 can be found in the Snakefile
+chromosomes in the appropriate format for \dadi and \fastsimcoal. For each simulation
+replicate, we performed 10 runs of \dadi and \fastsimcoal and checked for convergence.
+Detailed settings for \dadi and \fastsimcoal can be found in the Snakefile
 on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
-size over time ($N(t)$) inferred using 4 different methods, \smcpp, \stairwayplot,
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, \stairwayplot, and
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairwayplot},
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.pdf}
 \caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
+Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
 Data were generated by simulating
@@
 In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model.}
+$N(t)$ trajectories for each replicate. In black we show the true population size history
+as given by the census size for the simulated model.}
 \label{fig:two_popn_fly}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/generic_IM.pdf}
+\caption{\textbf{Parameters estimated from a generic IM model} Here we show estimates of $N(t)$
+inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp. Data were generated by simulating
+under a generic IM model with a human genome and \cite{international2007second} genetic map.
+In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the true population size history
+as given by the census size for the simulated model.}
+\label{fig:generic_IM}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/dromel_migration_rates.pdf}
+\caption{\textbf{Migration rate parameters estimated under a two-population \emph{Drosophila} model}.
+Here we show inferred migration rates from \dadi and \fastsimcoal.
+Data were generated by simulating replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
+inferred in \cite{comeron2012many}.}
+\label{fig:dromel_mig_rates}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/homsap_migration_rates.pdf}
+\caption{\textbf{Migration rate estimates for the human Gutenkunst model.}
+Here we show inferred migration rates from \dadi and \fastsimcoal.
+Data were generated by simulating
+replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
+inferred in \cite{international2007second}.}
+\label{fig:homsap_mig_rates}
 \end{center}

### 2019-12-18 popSimManu.tex
Many grammar and other minor fixes. Commented out Methods for simulations not presented. Added IM reference. We need to clarify which simulations are and are not included in the manuscript.

@@
 \newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
-\newcommand{\fastsimcoal}{\texttt{fastsimcoal2}\xspace}
+\newcommand{\fastsimcoal}{\texttt{fastsimcoal}\xspace}
 \newcommand{\tskit}{\texttt{tskit}\xspace}
@@
 \begin{center}
-\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.pdf}
+\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.png}
 \caption{\textbf{Parameters estimated using a multi-population human model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
+Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
 A) Data were generated by simulating
 replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}. B) For \dadi and \fastsimcoal we show parameters inferred
+inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show parameters inferred
 by fitting the depicted IM model, which includes population sizes, migration rates, and a split
@@
 As \texttt{stdpopsim} implements multi-population demographic models, we also
-explored two population inference with several popular methods. Specifically,
+explored parameter estimation of population divergence parameters. In particular
 we performed simulations under multi-population models for humans and \emph{Drosophila}
-and inferred demographic parameters using \dadi, \fastsimcoal, and \smcpp.
+and inferred simulated parameters using \dadi, \fastsimcoal, and \smcpp.
 For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
-with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
-an IM model was to mimic the approach often used on empirical datasets, where a relatively
-simple model is fit that may not reflect the true underlying demography.
-For human models with more than two-populations (e.g. \cite{gutenkunst2009inferring})
+% RNG: Added appropriate Jody Hey citation
+with constant population sizes and bi-directional migration \citep{hey2004im}. For human
+models with more than two-populations, e.g. \cite{gutenkunst2009inferring},
 this means that we are inferring parameters for a model which itself does
 not match the model from which the data were generated (Figure
-\ref{fig:IM_popn_human}A and B).
+\ref{fig:IM_popn_human}A and B), however this is common practice in the field
+when applying IM-type models.
 
@@
 time, for each of the inference methods, using samples drawn from African and European populations
-simulated under the \cite{gutenkunst2009inferring} model. Our results highlight many
-of the strengths and weaknesses of the different types of methods we used.
-For instance, the SFS-based approaches where we fit simple IM models do not capture
-recent exponential growth in the CEU population, however do consistently recover the
-simulated YRI population size history. Moreover, these approaches allow for estimating
-migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
-of divergence times. By contrast, \smcpp is much better at capturing the recent exponential
-growth in the CEU population, though the inferred population sizes are in general more
-noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
-estimates that are consistently underestimated.
+simulated under the \cite{gutenkunst2009inferring} model of human migration out of Africa.
+Each of the methods compared has stronger
+and weaker points to their accuracy across parameters. For instance the simple IM-style
+analyses we have performed do not capture recent population size change in CEU, however
+they do a good job of estimating the simpler simulated YRI population size history from this model.
 
-Again, we can compare between species and look at the performance of these methods in
+Again we can compare between species and look at the performance of these methods in
 the context of a two-population model from \emph{Drosophila melanogaster}. Figure
@@
 an ancestral population in Africa, then a population expansion with a population
-split and bottleneck into a European population with no post-divergence migration.
-Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
-parameters, but ignore the brief population bottleneck in Europe. The inferred
-demographic parameters from \smcpp are more noisy, though in some cases are better
-able to capture the bottleneck in the Europe population.
+split and bottleneck into a European population. These simulations are generated
+using the recombination map of \cite{comeron2012many}. Here methods differ dramatically
+in the estimates, with \smcpp yielding quite noisy estimates of population size
+towards the more ancient past and all methods failing to capture the very brief
+population bottleneck in Europe, although the IM model of course only allows for
+two epochs in the population size history.
 
-Although these results do not represent an exhaustive benchmarking,
-we have highlighted some of the strengths and weaknesses of these methods.
-Future work should build on these results and undertake more in-depth comparisons
-under a wider range of simulated demographic models.
+Again the goal of these analyses are not to perform thorough benchmarking of
+the methods at hand under a wide range of simulations, but instead to show how
+the tools that we have built could be used to facilitate such comparisons.
 
@@
 To conduct inference on these models, we applied three commonly used methods:
-\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal~\citep{excoffier2013robust},
+\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
 and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
@@
 
-For both \dadi and \fastsimcoal, we fit a two population
+For both \dadi and \fastsimcoal2, we fit a two population
 isolation-with-migration (IM) model, with constant population sizes, to the simulated
@@
 data, where the user is not aware of the `true' underlying demography and the inference
-model is often misspecified. To ground-truth our inference approach, we also conducted
-inference on a generic IM model that was identical to the model used for inference \ref{fig:generic_IM}.
+model is often misspecified.
+% RNG: Not in present manuscript.
+%To ground-truth our inference approach, we also conducted inference on a generic IM model that was identical to the model used for inference.
 
@@
 converted tree sequences into a two-dimensional site frequency spectrum for all
-chromosomes in the appropriate format for \dadi and \fastsimcoal. For each simulation
-replicate, we performed 10 runs of \dadi and \fastsimcoal and checked for convergence.
-Detailed settings for \dadi and \fastsimcoal can be found in the Snakefile
+chromosomes in the appropriate format for \dadi and \fastsimcoal2. For each simulation
+replicate, we performed 10 runs of \dadi and \fastsimcoal2 and checked for convergence.
+Detailed settings for \dadi and \fastsimcoal2 can be found in the Snakefile
 on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.pdf}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}
 \caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
+Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
 Data were generated by simulating
@@
 In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history
-as given by the census size for the simulated model.}
+$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
+for the rate of coalescence in the demographic model.}
 \label{fig:two_popn_fly}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/generic_IM.pdf}
-\caption{\textbf{Parameters estimated from a generic IM model} Here we show estimates of $N(t)$
-inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp. Data were generated by simulating
-under a generic IM model with a human genome and \cite{international2007second} genetic map.
-In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history
-as given by the census size for the simulated model.}
-\label{fig:generic_IM}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/dromel_migration_rates.pdf}
-\caption{\textbf{Migration rate parameters estimated under a two-population \emph{Drosophila} model}.
-Here we show inferred migration rates from \dadi and \fastsimcoal.
-Data were generated by simulating replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
-inferred in \cite{comeron2012many}.}
-\label{fig:dromel_mig_rates}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homsap_migration_rates.pdf}
-\caption{\textbf{Migration rate estimates for the human Gutenkunst model.}
-Here we show inferred migration rates from \dadi and \fastsimcoal.
-Data were generated by simulating
-replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}.}
-\label{fig:homsap_mig_rates}
 \end{center}

### 2019-12-18 popSimManu.tex
some reworking of Discussion

--- 2019-12-18 popSimManu.tex
@@
 
-We have built \stdpopsim, and the PopSim Consortium, to be an open, community-developed
-project, meaning that it has excellent potential for future growth.
-Thus far we have implemented genome representations and genetic maps for the three most
-common study systems in computational genetics: humans, \emph{Drosophila}, and \emph{Arabidopsis}.
-The standard operating procedures that we have outlined in our detailed documentation
-means that new researchers, who are interested in adding their favorite organism/genome
-are free to do so with very few barriers to entry (see Methods).
-Further, new demographic models can be added
-to each of the organisms in the catalog through similarly well documented steps.
-A long-term goal then is that through a mechanism such as \stdpopsim, the community
-can determine (in an open way) what is the correct implementation for a published demographic model,
-and in turn, what are the models in which the community is most interested.
+We have illustrated in this paper how \stdpopsim can be used for direct
+comparisons of inferential methods on a common set of simulations. Our
+benchmarking comparisons have been limited so far, but nevertheless, they
+do reveal some broad generalizations about relative performance. For
+example, at the task of estimating $N(t)$ trajectories for single human
+populations, we find that the sequence-based methods (\MSMC and \smcpp)
+perform somewhat better overall---at least for mid-range values of
+$t$---than the site frequency spectrum-based method (\stairwayplot)
+(Figure~\ref{fig:n_t_ragsdale} and Figure~\ref{fig:n_t_gutenkunst}), which
+tends to over-estimate oscillations in size.  By contrast, in the case of
+\emph{D.~melanogaster}, where linkage disequilibrium is considerably
+reduced, \stairwayplot outperforms the sequence-based methods
+(Figure~\ref{fig:n_t_sheehan}). In the two-population setting for human
+(Figure~\ref{fig:IM_popn_human}), most methods do reasonably well at
+reconstructing the YRI history, but struggle with the more complex CEU
+history, in large part because of the restriction of constant sizes per
+population.  An exception is \smcpp, which does not have the same
+restrictions on its inferred history, and as a result does somewhat better
+with the CEU history but tends to overfit the YRI history.  The results for
+the two-population {\em Drosophila} case (Figure~\ref{fig:two_popn_fly})
+are generally similar.
 
-We have shown in the results above how \stdpopsim can be used for direct comparisons
-of inferential methods on a common set of simulations. While we have not made a concerted
-effort at benchmarking methods yet, we believe that this would be an important goal for
-the PopSim Consortium and population genetics more broadly. The analysis workflows that
-we have released as part of \stdpopsim should provide a cookbook for the community to
-dive in to more rigorous benchmarking than we have done thus far.
-
-In the use case comparisons we have performed,
-we found that the methods tested \emph{do} generally perform well on human genomes,
-up to, for instance, the usual factor-of-two differences in inferred population sizes.
-This will not be surprising to people familiar with the landscape of
-demographic inference methods in population genetics,
-but the nature and scale of uncertainty may be not clear to people new to the field.
-The ability for \stdpopsim to generate data with and without significant features, such
+Altogether, these preliminary experiments highlight
+the utility of \stdpopsim for comparing a variety of inference methods on
+the same footing, under a variety of different demographic models.
+In addition, the ability of \stdpopsim to generate data with and without significant features, such
 as the genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
-us to ask directly about the various failure modes of popular methods.
+investigation of the failure modes of popular methods.
 Moreover the comparison of methods across the various genome organizations, genetic maps,
@@
 about how methods might perform on non-human systems.
-In our initial comparison we find that inference of population histories from simulated \emph{Drosophila}
-or \emph{Arabidopsis} genomes is a fair bit more difficult from SMC based methods
-than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
-map lengths of these genomes, although we will leave further investigation of this to the future.
 Finally, comparison of results across methods or simulation runs
-gives a strong estimate of inference uncertainty, analogous to parametric
-bootstrapping---although different methods have different strengths and weaknesses,
-strong conclusions should probably not rest entirely on features inferred by only one method.
+provides an estimate of inference uncertainty, analogous to parametric
+bootstrapping.
 
-The broad brush comparisons we have done, while far from exhaustive, are informative,
-and highlight the utility of easy comparison of methods on the same footing.
-The tests here were under best-case scenarios---simulations of discrete,
-randomly mating populations that fit the assumptions of the methods nearly
-exactly---and yet different methods, as well as applications of the same method
-to different simulations of the same model,
-produced somewhat different results. Future efforts will extend the resource described here to
-encompass additional population genetic tasks, such as detecting natural selection, inferring
-recombination rates, etc, in hopes of producing a general-purpose resource for fair and
-standardized evaluation of the ever-increasing arsenal of computational population genetic
-inference tools.
+\stdpopsim is intended to be a fully open, community-developed project.  
+Our implementations of genome representations and genetic maps for the three most
+common study systems in computational genetics---humans, \emph{Drosophila},
+and \emph{Arabidopsis}---are only intended to be a starting point for
+future development.  
+Researchers are invited to contribute to the resource by adding their
+organisms and demographic models of choice. The \stdpopsim resource is
+accompanied by clearly documented standard operating procedures that are
+intended to minimize barriers to entry for new developers.  In this way, we
+expect the resource to expand and adapt to meet the evolving needs of the
+population genomics community.
 
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.png}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.png}
+\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population

### 2019-12-18 popSimManu.tex
incorporated Kirk's edits

--- 2019-12-18 popSimManu.tex
@@
 \citep{gutenkunst2009inferring,li2011inference,excoffier2013robust,schiffels2014inferring, terhorst2017robust,ragsdale2019models},
-the distribution of fitness effects \citep{kim2017inference},
+the distribution of fitness effects \citep{kim2017inference,tataru2017inference},
 recombination rates \citep{chan2012genome,lin2013fast,Adrion662247},
@@
 the accuracy and quality of these inferences is becoming even more important.
-This is in large part because ``ground-truth'' in population genetics comes from simulation,
+Assessing accuracy in population genetics is challenging in large part because
+ ``ground-truth'' in population genetics comes from simulation,
 not from empirical observations, and as a field there has been no agreement as to what should
@@
 different conclusions. For example, Markovian coalescent methods applied to human genomes have
-% JK is this right? > 100 years doesn't sound very ancient to me.
-% RNG: Fixed time and removed unexplained acronyms.
 suggested large ancient ($>100,000$ years ago) ancestral population sizes and
@@
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five organisms: humans, orangutans,
+The \stdpopsim catalog currently contains five organisms: humans, \emph{P. pygmaeus},
 \emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}. For each
@@
 We invite new collaborators to join our community.
-Below we describe the resource and give an
-example of how it can be used to benchmark demographic inference methods.
+Below we describe the resource and give
+examples of how it can be used to benchmark demographic inference methods.
 
@@
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{D.~melanogaster}, \emph{A.~thaliana}, \emph{E.~coli}, and\emph{P.~abelii}.
+on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{P.~abelii}, 
+\emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}.
 % RNG: Abbreviate consistently with introduction
@@
 of specific historical scenarios that have influenced present-day patterns of
-genetic polymorphism.
+genetic variation.
 
@@
 engines to support scenarios that cannot be modelled under the coalescent framework,
-and we have already implemented \texttt{SLiM} \citep{haller2019slim} as
+and have already implemented \texttt{SLiM} \citep{haller2019slim} as
 an alternative back-end.
@@
 writing, the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Drosophila}, and \emph{Arabidopsis}.
-% RNG: Pongo added by submission?
+humans, \emph{Pongo}, \emph{Drosophila}, and \emph{Arabidopsis}.
 These range from
@@
 Although we present comparison of results from several
-% RNG: Not competitors... We complement each other ;-)
 methods, our aim at this stage is not to provide an exhaustive
@@
 In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
-run on whole genome simulations under an empirical genetic map and a model of human migration out of Africa that includes archaic admiture \citep{ragsdale2019models}. In each column of this figure
+run on whole genome simulations under under a sophisticated demographic model
+and empirical genetic map and a model of human migration out of Africa
+that includes archaic admiture \citep{ragsdale2019models}. In each column of this figure
 we show inferred $N(t)$ for the three extant populations in the model.
@@
 
-We can even do away with the demographic model entirely and simulate under a human
-genome architecture and genetic map to see how well methods might do at recovering
-the population history of a constant sized population. We show results of such an
+To see how well methods might do at recovering the population history of a constant sized population,
+we can simulate under a human genome architecture and genetic map, but without demography.
+We show results of such an
 experiment in Figure \ref{fig:n_t_HomSap_constant}. Again we see that all methods
@@
 As \texttt{stdpopsim} implements multi-population demographic models, we also
-explored two population inference with several popular methods. Specifically,
-we performed simulations under multi-population models for humans and \emph{Drosophila}
-and inferred demographic parameters using \dadi, \fastsimcoal, and \smcpp.
+explored parameter estimation of population divergence parameters. In particular,
+we simulated data under multi-population models for humans and \emph{Drosophila}
+and then inferred  parameters using \dadi, \fastsimcoal, and \smcpp.
 For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
@@
 Here we have described what we view as the first product from the PopSim Consortium:
-the \stdpopsim library. We have founded the Consortium with a number of specific aims
+the \stdpopsim library. We have founded the Consortium with a number of specific goals
 in mind: standardization of simulation within the computational genetics community,

### 2019-12-18 popSimManu.tex
adam + dan comments in a clean PR
massaging

.

@@
 \usepackage[hidelinks]{hyperref}
+\usepackage[blocks]{authblk}
+\renewcommand\Affilfont{\small}
+\makeatletter
+\renewcommand\AB@affilsepx{, \protect\Affilfont}
+\makeatother
+\usepackage[utf8]{inputenc}
 
@@
 \title{A community-maintained standard library of population genetic simulations}
+% First authors
+\author[1,$\star$]{Jeffrey R. Adrion}
+\author[2,$\star$]{Christopher B. Cole}
+\author[3,$\star$]{Noah Dukler}
+\author[1,$\star$]{Jared G. Galloway}
+\author[4,$\star$]{Ariella L. Gladstein}
+\author[5,$\star$]{Graham Gower}
+\author[6,$\star$]{Christopher C. Kyriazis}
+\author[7,$\star$]{Aaron P. Ragsdale}
+\author[8,$\star$]{Georgia Tsambos}
+% Second Authors
+\author[9]{Franz Baumdicker}
+\author[10]{Jedidiah Carlson}
+\author[11]{Reed A. Cartwright}
+\author[12]{Arun Durvasula}
+\author[13]{Bernard Y. Kim}
+\author[14]{Patrick McKenzie}
+\author[15]{Philipp W. Messer}
+\author[16]{Ekaterina Noskova}
+\author[17]{Diego Ortega-Del Vecchyo}
+\author[18]{Fernando Racimo}
+\author[19]{Travis J. Struck}
+%Senior Authors
+\author[7,$\dagger$]{Simon Gravel}
+\author[19,$\dagger$]{Ryan N. Gutenkunst}
+\author[6,$\dagger$]{Kirk E. Lohmeuller}
+\author[1,$\dagger$]{Peter L. Ralph}
+\author[4,$\dagger$]{Daniel R. Schrider}
+\author[3,$\dagger$]{Adam Siepel}
+\author[20,$\dagger$,$\aleph$]{Jerome Kelleher}
+\author[1,$\dagger$,$\aleph$]{Andrew D. Kern}
 
-\author{%
-\name{The PopSim Consortium}
-\address{}
-\email{}}
+\affil[1]{Department of Biology and Institute of Ecology and Evolution, University of Oregon}
+\affil[2]{Wellcome Trust Centre for Human Genetics, University of Oxford}
+\affil[3]{Simons Center for Quantitative Biology, Cold Spring Harbor Laboratory}
+\affil[4]{Department of Genetics, University of North Carolina at Chapel Hill}
+\affil[5]{Lundbeck GeoGenetics Centre, GLOBE Institute, University of Copenhagen}
+\affil[6]{Department of Ecology and Evolutionary Biology, University of California, Los Angeles}
 
+\affil[7]{Department of Human Genetics, McGill University}
+\affil[8]{Melbourne Integrative Genomics, School of Mathematics and Statistics, University of Melbourne}
+\affil[9]{Department of Mathematical Stochastics, University of Freiburg}
+\affil[10]{Department of Genome Sciences, University of Washington}
+\affil[11]{The Biodesign Institute and The School of Life Sciences, Arizona State University}
+\affil[12]{Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles}
+\affil[13]{Department of Biology, Stanford University}
+\affil[14]{Department of Ecology, Evolution, and Environmental Biology, Columbia University}
+\affil[15]{Department of Computational Biology, Cornell University}
+\affil[16]{Computer Technologies Laboratory, ITMO University}
+\affil[17]{International Laboratory for Human Genome Research, National Autonomous University of Mexico}
+\affil[18]{Globe Institute, University of Copenhagen}
+\affil[19]{Department of Molecular and Cellular Biology, University of Arizona}
+\affil[20]{Big Data Institute, Oxford Unviersity}
+\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+\affil[$\aleph$]{Denotes corresponding authors, listed alphabetically}
 % JK: didn't bother with the \texttt stuff for stdpopsim in the abstract as it's nearly
@@
 population genetic methods.}
-\date{\today}
+\date{}
 
@@
 sample sizes for \MSMC).
-% RNG: Removed duplicated info from caption.
+There is no single ``true'' reference for effective population size
+because of model misspecification -- the inference methods are fitting a single population model
+to data simulated from mulitple populations.
+However, many methods work by matching coalescence time distributions,
+and a single-population model with varying population size can match any coalescence time distribution
+(in which case coalescence rate is the inverse of the effective size).
+For this reason, solid black lines do not show historical census sizes,
+but rather inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix).
+Blue lines show estimates from each of three replicate whole genome simulations.
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}. B) For \dadi and \fastsimcoal we show parameters inferred
+inferred in \cite{international2007second}.
+B) For \dadi and \fastsimcoal we show parameters inferred
 by fitting the depicted IM model, which includes population sizes, migration rates, and a split
-time between CEU and YRI samples. C) Population size estimates for each population (rows)
-from \dadi, \texttt{fastsimcoal2}, and \smcpp (columns). In shades of blue we show the estimated
-% RNG: Updated caption to reflect what is shown.
-$N(t)$ trajectories for each replicate. In black we show the census population size from the simulated model.
+time between CEU and YRI samples.
+C) Population size estimates for each population (rows)
+from \dadi, \fastsimcoal, and \smcpp (columns).
+In shades of blue we show $N(t)$ trajectories estimated from each simulation,
+and in black the inverse coalescence rates calculated from the true model (see text).
 The population split date, $T_{div}$, is shown at
@@
 Here we have described what we view as the first product from the PopSim Consortium:
-the \stdpopsim library. We have founded the Consortium with a number of specific goals
-in mind: standardization of simulation within the computational genetics community,
-increasing reproducibility of simulation based results, community-based development and
-decision making in guiding best practices in population genetics, and eventual benchmarking
-of inferential methods.
+the \stdpopsim library. We have founded the Consortium with a number of specific goals in mind:
+standardization of simulation within the computational genetics community,
+increasing reproducibility and ease of use of complex simulation based results,
+community-based development and decision making guiding best practices in population genetics,
+and eventual benchmarking of inferential methods.
 
@@
 standardization of complex population genetic simulations. Population genetics, as a field,
-% RNG: We should have more coalescence puns in the paper! :-P
-has yet to coalesce around a set of gold-standards for the crucial task of method
+has yet to coalesce around a set of standards for the crucial task of method
 evaluation, which in our field hinges on simulation. In contrast, other fields such as
@@
 
-Moreover individual research groups are often in the position of having to independently
-reimplement complex previously published models. \stdpopsim offers solutions to both of
-these issues by creating a quality controlled, easy-to-use, gateway to models and organisms
+Individual research groups are often in the position of having to independently
+reimplement complex, previously published demographic models.
+Additional layers of realism (e.g., recombination maps) makes this especially daunting.
+\stdpopsim aims to lower this barrier
+by creating a quality controlled, easy-to-use, gateway to models and organisms
 that are central to modern research in genetics.
 
-We have built \stdpopsim, and the PopSim Consortium, to be an open, community-developed
-project, meaning that it has excellent potential for future growth.
-Thus far we have implemented genome representations and genetic maps for the three most
-common study systems in computational genetics: humans, \emph{Drosophila}, and \emph{Arabidopsis}.
-% RNG: Needs to be updated for final count.
-The standard operating procedures that we have outlined in our detailed documentation
-mean that new researchers, who are interested in adding their favorite organism/genome
-are free to do so with few barriers to entry (see Methods).
-Further, new demographic models can be added
-to each of the organisms in the catalog through similarly well documented steps.
-A long-term goal is that the community
-can determine (in an open way) what is the correct implementation for a published demographic model,
-and in turn, what are the models in which the community is most interested.
+We have illustrated in this paper how \stdpopsim can be used for direct
+comparisons of inferential methods on a common set of simulations. Our
+benchmarking comparisons have been limited, but nevertheless
+reveal some informative comparisons. For
+example, at the task of estimating $N(t)$ trajectories for simulations of single human
+populations, we find that the sequence-based methods (\MSMC and \smcpp)
+perform somewhat better overall---at least for mid-range values of
+$t$---than the site frequency spectrum-based method (\stairwayplot)
+(Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}), which
+tends to over-estimate the size of oscillations.  By contrast,
+\stairwayplot outperforms the sequence-based methods
+on simulations of \textit{D.~melanogaster} populations,
+in which linkage disequilibrium is considerably reduced (Figure~\ref{fig:n_t_sheehan}).
+In simulations of two human populations
+(Figure~\ref{fig:IM_popn_human}), most methods do reasonably well at
+reconstructing the YRI history, but struggle with the more complex CEU
+history, in large part because of the restriction of constant sizes per
+population.  An exception is \smcpp, which does not have the same
+restrictions on its inferred history, and as a result does somewhat better
+with the CEU history but tends to overfit the YRI history.  The results for
+the two-population \textit{D.~melanogaster} case (Figure~\ref{fig:two_popn_fly})
+are generally similar.
 
-We have shown in the results above how \stdpopsim can be used for direct comparisons
-of inferential methods on a common set of simulations. While we have not made a concerted
-effort at benchmarking methods yet, we believe that this would be an important goal for
-the PopSim Consortium and population genetics more broadly. The analysis workflows that
-we have released as part of \stdpopsim should provide a cookbook for the community to
-dive in to more rigorous benchmarking than we have done thus far.
-
-In the use case comparisons we have performed,
-we found that the methods tested \emph{do} generally perform well on human genomes,
-up to, for instance, the usual factor-of-two differences in inferred population sizes.
-This will not be surprising to people familiar with the landscape of
-demographic inference methods in population genetics,
-but the nature and scale of uncertainty may be not clear to people new to the field.
-The ability for \stdpopsim to generate data with and without significant features, such
-as the genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
-us to ask directly about the various failure modes of popular methods.
+Altogether, these preliminary experiments highlight
+the utility of \stdpopsim for comparing a variety of inference methods on
+the same footing, under a variety of different demographic models.
+In addition, the ability of \stdpopsim to generate data with and without significant features, such
+as a genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
+investigation of the failure modes of popular methods.
 Moreover the comparison of methods across the various genome organizations, genetic maps,
@@
 about how methods might perform on non-human systems.
-In our initial comparison we find that inference of population histories from simulated \emph{Drosophila}
-or \emph{Arabidopsis} genomes is a fair bit more difficult from SMC based methods
-% RNG: No Arabidopsis results are presented in this paper
-than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
-map lengths of these genomes, although we leave further investigation of this to the future.
+
 Finally, comparison of results across methods or simulation runs
-gives a strong estimate of inference uncertainty, analogous to parametric
-bootstrapping---although different methods have different strengths and weaknesses,
-strong conclusions should probably not rest entirely on features inferred by only one method.
+provides an estimate of inference uncertainty, analogous to parametric
+bootstrapping,
+especially since different methods are likely vulnerable to model misspecification
+in different ways.
 
-The broad brush comparisons we have done, while far from exhaustive, are informative,
-and highlight the utility of easy comparison of methods on the same footing.
-The tests here were under best-case scenarios---simulations of discrete,
-randomly mating populations that fit the assumptions of the methods nearly
-exactly---and yet different methods, as well as applications of the same method
-to different simulations of the same model,
-produced somewhat different results. Future efforts will extend the resource described here to
-encompass additional population genetic tasks, such as detecting natural selection, inferring
-recombination rates, etc, in hopes of producing a general-purpose resource for fair and
-standardized evaluation of the ever-increasing arsenal of computational population genetic
-inference tools.
+\stdpopsim is intended to be a fully open, community-developed project.  
+Our implementations of genome representations and genetic maps for the some of
+the most common study systems in computational genetics---humans, \textit{Drosophila},
+and \textit{Arabidopsis} (among others)---are only intended to be a starting point for
+future development.  In addition to other taxa,
+we hope to soon incorporate other common biological processes
+such as selection, gene conversion, and mutational heterogeneity.
+Researchers are invited to contribute to the resource by adding their
+organisms and models of choice. The \stdpopsim resource is
+accompanied by clearly documented standard operating procedures that are
+intended to minimize barriers to entry for new developers.  In this way, we
+expect the resource to expand and adapt to meet the evolving needs of the
+population genomics community.
 
@@
 
+\section*{Acknowledgements}
+We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible.
+Early on in the project we were helped by many people including XXXXXX. 
+CCK and KEL were funded under NIH Award R35GM119856. 
+JRA and ADK were funded under NIH Award R01GM117241.
+
 \bibliographystyle{plainnat}
@@
 
+\appendix
+
+\section*{Calculating coalescence rates}
+
+We compute the coalescence rate of a collection of samples in a given demographic model
+at a particular point back in time
+as the expected number of coalescences happening at that time
+per unit of time and per pair of as-yet-uncoalesced lineages.
+More concretely,
+let $p(t)$ denote the probability that the lineages of a randomly chosen pair of samples
+have not yet coalesced $t$ units of time ago,
+let $p(z, t)$ denote the probability that those lineages have not yet coalesced
+and are furthermore both in location $z$,
+and let $1/(2 N_e(z,t))$ be the rate of coalescence in location $z$ at the time.
+Then, we compute the mean coalescence rate as
+$$  r(t)  = \frac{1}{p(t)} \sum_z \frac{p(z,t)}{2N_e(z,t)} . $$
+This follows because if we have $n$ diploid samples, and hence $\binom{2n}{2}$ lineages,
+the expected number of coalescences in location $z$ between times $t$ and $t+dt$ ago
+$$
+\binom{2n}{2} p(z,t) \frac{ dt }{ 2 N_e(z,t) }, 
+$$
+and the expected number of pairs of uncoalesced lineages at that time is
+$$
+\binom{2n}{2} p(t) .
+$$
+The expression for $r(t)$ is a ratio of these two quantities;
+to obtain it we need to compute $p(t)$ and $p(z,t)$.
+This is relatively straightforward
+using the general theory of Markov chains, and is implemented in \texttt{msprime}.
+
+Note that since these quantities are \emph{per pair of lineages},
+this definition depends on the locations of the samples.
+The coalescence rate also has the intuitive interpretation that 
+it is the average between-lineage coalescence rate,
+averaged over where uncoalesced lineages might be.
+Since the local coalescence rate is the inverse of the population size,
+$1/r(t)$ (as shown for instance in \autoref{fig:n_t_ragsdale})
+is a weighted harmonic mean of the census sizes of the different populations present at that time.
+This is as expected: suppose that we have two populations, one big and one small,
+connected by migration.
+If all our samples are from the big population, 
+the number of recent coalescences should be small, reflecting the large population size,
+while in the long run, the coalescence rate approaches an intermediate rate.
+On the other hand, more recent coalescences are expected
+if all samples are from the small population,
+A method that fits a single, time-varying population size to the data
+might be expected to find a population size trajectory
+to match these time-varying rates of coalescence.
+
+We use the same computations to analytically compute \emph{mean coalescence times}:
+since for any nonnegative random variable $T$, the mean value is
+$\mathbb{E}[T] = \int_0^\infty \mathbb{P}\{T > t\} dt$,
+we can obtain the mean coalescence time as
+$$
+\int_0^\infty p(t) dt ,
+$$
+where $p(t)$ is defined above.
+
 \end{document}

### 2019-12-19 popSimManu.tex
Added some bibliography, minor edits and funding information.

@@
 \author[17]{Diego Ortega-Del Vecchyo}
-\author[18]{Fernando Racimo}
-\author[19]{Travis J. Struck}
+\author[5]{Fernando Racimo}
+\author[18]{Travis J. Struck}
 %Senior Authors
 \author[7,$\dagger$]{Simon Gravel}
-\author[19,$\dagger$]{Ryan N. Gutenkunst}
+\author[18,$\dagger$]{Ryan N. Gutenkunst}
 \author[6,$\dagger$]{Kirk E. Lohmeuller}
@@
 \author[3,$\dagger$]{Adam Siepel}
-\author[20,$\dagger$,$\aleph$]{Jerome Kelleher}
+\author[19,$\dagger$,$\aleph$]{Jerome Kelleher}
 \author[1,$\dagger$,$\aleph$]{Andrew D. Kern}
@@
 \affil[4]{Department of Genetics, University of North Carolina at Chapel Hill}
-\affil[5]{Lundbeck GeoGenetics Centre, GLOBE Institute, University of Copenhagen}
+\affil[5]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
 \affil[6]{Department of Ecology and Evolutionary Biology, University of California, Los Angeles}
@@
 \affil[17]{International Laboratory for Human Genome Research, National Autonomous University of Mexico}
-\affil[18]{Globe Institute, University of Copenhagen}
-\affil[19]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[20]{Big Data Institute, Oxford Unviersity}
+\affil[18]{Department of Molecular and Cellular Biology, University of Arizona}
+\affil[19]{Big Data Institute, Oxford Unviersity}
 \affil[$\star$]{Denotes shared first authorship, listed alphabetically}
@@
 \citep{gutenkunst2009inferring,li2011inference,excoffier2013robust,schiffels2014inferring, terhorst2017robust,ragsdale2019models},
-the distribution of fitness effects \citep{kim2017inference,tataru2017inference},
-recombination rates \citep{chan2012genome,lin2013fast,Adrion662247},
+the distribution of fitness effects \citep{Boyko:2008cr,kim2017inference,tataru2017inference,Fortier703918,Huang2019,Vecchyo770966},
+recombination rates \citep{chan2012genome,lin2013fast,Adrion662247,Barroso2019},
 and the extent of positive selection in genome sequence data
@@
 a continuous change in population size over time, rather than pre-specified epochs of different
-population sizes.
+population sizes. Additionally, we have also implemented two generic models that comprise a piecewise constant size model and an isolation with migration (IM) model.
 
@@
 and then inferred  parameters using \dadi, \fastsimcoal, and \smcpp.
-For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
+For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an IM model
 with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
@@
 about how methods might perform on non-human systems.
-In our initial comparison we find that inference of population histories from simulated \emph{Drosophila}
-or \emph{Arabidopsis} genomes is a fair bit more difficult from SMC based methods
-% RNG: No Arabidopsis results are presented in this paper
-than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
-map lengths of these genomes, although we leave further investigation of this to the future.
+
 Finally, comparison of results across methods or simulation runs
-gives a strong estimate of inference uncertainty, analogous to parametric
-bootstrapping---although different methods have different strengths and weaknesses,
-strong conclusions should probably not rest entirely on features inferred by only one method.
+provides an estimate of inference uncertainty, analogous to parametric
+bootstrapping,
+especially since different methods are likely vulnerable to model misspecification
+in different ways.
 
-The broad brush comparisons we have done, while far from exhaustive, are informative,
-and highlight the utility of easy comparison of methods on the same footing.
-The tests here were under best-case scenarios---simulations of discrete,
-randomly mating populations that fit the assumptions of the methods nearly
-exactly---and yet different methods, as well as applications of the same method
-to different simulations of the same model,
-produced somewhat different results. Future efforts will extend the resource described here to
-encompass additional population genetic tasks, such as detecting natural selection, inferring
-recombination rates, etc, in hopes of producing a general-purpose resource for fair and
-standardized evaluation of the ever-increasing arsenal of computational population genetic
-inference tools.
+\stdpopsim is intended to be a fully open, community-developed project.
+Our implementations of genome representations and genetic maps for the some of
+the most common study systems in computational genetics---humans, \textit{Drosophila},
+and \textit{Arabidopsis} (among others)---are only intended to be a starting point for
+future development.  In addition to other taxa,
+we hope to soon incorporate other common biological processes
+such as selection, gene conversion, and mutational heterogeneity.
+Researchers are invited to contribute to the resource by adding their
+organisms and models of choice. The \stdpopsim resource is
+accompanied by clearly documented standard operating procedures that are
+intended to minimize barriers to entry for new developers.  In this way, we
+expect the resource to expand and adapt to meet the evolving needs of the
+population genomics community.
 
@@
 JRA and ADK were funded under NIH Award R01GM117241.
+FR and GG were supported by a Villum Young Investigator award (project no. 00025300).
+DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
 

### 2019-12-19 popSimManu.tex

@@
 \citep{gutenkunst2009inferring,li2011inference,excoffier2013robust,schiffels2014inferring, terhorst2017robust,ragsdale2019models},
-the distribution of fitness effects \citep{Boyko:2008cr,kim2017inference,tataru2017inference,Fortier703918,Huang2019,Vecchyo770966},
-recombination rates \citep{chan2012genome,lin2013fast,Adrion662247,Barroso2019},
+the distribution of fitness effects \citep{kim2017inference,tataru2017inference},
+recombination rates \citep{chan2012genome,lin2013fast,Adrion662247},
 and the extent of positive selection in genome sequence data
@@
 a continuous change in population size over time, rather than pre-specified epochs of different
-population sizes. Additionally, we have also implemented two generic models that comprise a piecewise constant size model and an isolation with migration (IM) model.
+population sizes.
 
@@
 and then inferred  parameters using \dadi, \fastsimcoal, and \smcpp.
-For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an IM model
+For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
 with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
@@
 FR and GG were supported by a Villum Young Investigator award (project no. 00025300).
-DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
 

### 2019-12-19 popSimManu.tex
Updated model table.

--- 2019-12-19 popSimManu.tex
@@
 \usepackage{xspace}
+\usepackage{booktabs}
+\usepackage{siunitx}
 
@@
 \affil[18]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[19]{Big Data Institute, Oxford Unviersity}
+\affil[19]{Big Data Institute, Li Ka Shing Centre for Health Information and
+Discovery, University of Oxford}
 \affil[$\star$]{Denotes shared first authorship, listed alphabetically}
@@
 
+\renewcommand{\arraystretch}{1.2}
 \begin{table}[t]
+% Make sure the table is centered even though it's too wide
+\makebox[\textwidth][c]{
     \begin{footnotesize}
@@
     \end{footnotesize}
+}
 \caption{\label{tab:catalog}
-Details of the initial set of population models across four species.
-\textbf{NOTES: Column names are ID, name (or description maybe?),
-reference, CPU time, RAM}
+Initial set of demographic models in the Catalog and simple benchmarks.
+For each model we report the CPU time, maximum memory usage and the
+size of the output \tskit file. In each case we simulate 100 samples
+drawn from the first population, for the shortest chromosome of that species
+and a constant chromosome-specific recombination rate.
+The times reported are for a single replicate run on an Intel i5-7600 CPU.
+These are not intended to be rigorous benchmarks, and the computing resources
+required will vary widely depending on sample sizes, chromosome length,
+recombination rates and other factors.
 }
@@
 
-Here we have described what we view as the first product from the PopSim Consortium:
+Here we have described the first major product from the PopSim Consortium:
 the \stdpopsim library. We have founded the Consortium with a number of specific goals in mind:
@@
 
-The \stdpopsim library is the first piece of this effort, in that it allows for rigorous
+The \stdpopsim library allows for rigorous
 standardization of complex population genetic simulations. Population genetics, as a field,
@@
 of standardized method testing. We hope that our efforts represent the beginning of what
-will prove to be an equally longstanding and invaluable tradition in population genetics.
+will prove to be an equally longstanding and valuable tradition in population genetics.
 
-Individual research groups are often in the position of having to independently
-reimplement complex, previously published demographic models.
-Additional layers of realism (e.g., recombination maps) makes this especially daunting.
-\stdpopsim aims to lower this barrier
-by creating a quality controlled, easy-to-use, gateway to models and organisms
-that are central to modern research in genetics.
+% (ACS) I suggest deleting the following paragraph (commented out below).
+% All of the important points are made below (even with some of the same
+% phrases, such as "barriers to entry") and I think it is needlessly
+% redundant.
+
+%Individual research groups are often in the position of having to independently
+%reimplement complex, previously published demographic models.
+%Additional layers of realism (e.g., recombination maps) make this especially daunting.
+%\stdpopsim aims to lower this barrier
+%by creating a quality controlled, easy-to-use, gateway to models and organisms
+%that are central to modern research in genetics.
 
@@
 benchmarking comparisons have been limited, but nevertheless
-reveal some informative comparisons. For
-example, at the task of estimating $N(t)$ trajectories for simulations of single human
+reveal some informative features. For
+example, at the task of estimating $N(t)$ trajectories for simulated human
 populations, we find that the sequence-based methods (\MSMC and \smcpp)
@@
 (Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}), which
-tends to over-estimate the size of oscillations.  By contrast,
+tends to over-estimate the sizes of oscillations.  By contrast,
 \stairwayplot outperforms the sequence-based methods
@@
 JRA and ADK were funded under NIH Award R01GM117241.
-FR and GG were supported by a Villum Young Investigator award (project no. 00025300).
+TJS and RNG were funded under NIH award R01GM127348.
+FR and GG were supported by a Villum Young Investigator award (project no.~00025300).
 DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
+JK is supported by the Robertson Foundation.
 

### 2019-12-19 src/catalog_table.py

@@ -3,40 +3,67 @@ Writes out a Latex description of the stdpopsim catalog to stdout.
 """
 import stdpopsim
 import pandas as pd
-import humanize
+import bibtexparser
+
+
+def get_bibtex_key(bibtex_str, output):
+    bib_database = bibtexparser.loads(bibtex_str)
+    entry = bib_database.entries[0]
+    author = entry["author"].split(",")[0].lower()
+    year = entry["year"]
+    title = entry["title"].split()[0].split("-")[0].lower()
+    key = f"{author}{year}{title}"
+    entry["ID"] = key
+    writer = bibtexparser.bwriter.BibTexWriter()
+    output.write(writer.write(bib_database))
+    return key
 
 
 def main():
 
     df = pd.read_csv("data/benchmark.csv")
     df.set_index(["species", "model"])
+    # Using decimal values here as it's easier than explaining what a real mibibyte is
+    megabyte = 10**6
 
-    print("\\begin{tabular}{lllll}")
-    for species_id in ["homsap", "dromel", "aratha"]:
-        species = stdpopsim.get_species(species_id)
-        escaped_id = species.id.replace("_", "\\_")
-        species_header = f"{species.name} ({escaped_id})"
-        print("\\hline")
-        print("\\multicolumn{5}{l}{", species_header, "} \\\\")
-        print("\\hline")
-        # print(species.id)
-        for model in species.models:
-            row = df[(df.species == species.id) & (df.model == model.id)]
-            cpu_time = float(row.cpu_time)
-            ram = humanize.naturalsize(float(row.ram))
-            # Not currently used.
-            # file_size = humanize.naturalsize(float(row.file_size))
-            # bibtex_key = "todo"
-            escaped_id = model.id.replace("_", "\\_")
-            print(escaped_id, end="& ")
-            escaped_name = model.name.replace("&", "\\&")
-            print(escaped_name, end="& ")
-            # print(f"\\cite{{{bibtex_key}}} & ", end="")
-            print("[x]", end="& ")
-            print("{:.1f}s".format(cpu_time), end="& ")
-            print(ram, end="")
-            print("\\\\")
-    print("\\end{tabular}")
+    with open("model_refs.bib", "w") as model_refs:
+        header = (
+            "\\begin{tabular}{lllS[table-format=3.1]S[table-format=3.1]"
+            "S[table-format=2.1]}")
+        print(header)
+        print("\\toprule")
+        print("& Model ID & Citation & ")
[…]

### 2019-12-19 popSimManu.tex
minor edits to ms

@@
 \citep{eyre2009estimating, alachiotis2012omegaplus,degiorgio2016sweepfinder2,kern2018diplos,sugden2018localization}.
-While en masse these methods have increased our understanding of the
+While these methods have increased our understanding of the
 impacts of genetic and evolutionary processes, very little has been done to systematically
@@
 As large databases of population genetic variation begin to be used to inform public health procedures,
-the accuracy and quality of these inferences is becoming even more important.
+the accuracy and quality of these inferences is becoming ever more important.
+%CK: this first paragraph is super long - might be helpful to break up
+
 Assessing accuracy in population genetics is challenging in large part because
  ``ground-truth'' in population genetics comes from simulation,
-not from empirical observations, and as a field there has been no agreement as to what should
-constitute community standards or best practices. The general modus operandi to date has been to
+not from empirical observations, and as a field there has been no agreement
+on community standards or best practices. The general modus operandi to date has been to
 introduce methods that are validated by a set of bespoke simulations produced by individual
@@
 of standardized population genetics models spanning a variety of common study organisms.
-This resource will empower the field to rigorously evaluate population genetic software
-and select the best tool for a given task, thereby substantially increasing the reliability of population genetic inferences.
+This resource will empower the field to rigorously evaluate population genetic software,
+%and select the best tool for a given task,
+thereby substantially increasing the reliability of population genetic inferences.
 %DRS: still not in love with this last sentence here, but tried to make it a little less sappy by adding some explanation.
@@
 %RNG: Removed the sap. :-P Feel free to put it back if you like it better.
+%JK: not crazy about it either. Making it shorter helps though?
 
-The degree to which modeling assumptions and choices of data summaries can affect the ultimate results of population genetic inference has been challenging to assess. Yet there
+It has been challenging to assess the degree to which modeling assumptions
+and choices of data summaries can affect the ultimate results of population
+genetic inference. Yet there
 have been several clear examples of different methods yielding fundamentally
@@
 parameters are known. Methods that perform well will accurately infer these
-parameters and will also provide reasonable estimates of uncertainty of the
+parameters and will also provide reasonable measures of uncertainty of the
 estimated parameters. Indeed, simulation studies are a standard ingredient in
@@
 Many more researchers in the broader field
-use simulations to perform power and other exploratory analyses.
+use simulations to evaluate statistical power and perform other exploratory analyses.
 The many steps of implementing a moderately complicated population genetics simulation,
@@
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five organisms: humans, \emph{P. pygmaeus},
-\emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}. For each
+The \stdpopsim catalog currently contains five organisms: humans, \emph{Pongo~pygmaeus},
+\emph{Drosophila~melanogaster}, \emph{Arabidopsis~thaliana}, and \emph{Escherichia~coli}. For each
+% CK: i think we should spell out genus at first mention.
 organism the catalog contains details on the physical organization (e.g., chromosome structure)
@@
 generation time) and one or more published demographic histories. Through
-either a command line interface or a simple Python API, the user can specify which
+either a command line interface or a simple Python API, users can specify which
 organism, genetic map, chromosome, and demographic history they are interested in simulating, and the
@@
 distributed open source model, with strong procedures in place
-to continue its growth and maintain its accuracy.
+to continue its growth and maintain its quality.
 Importantly, we have rigorous quality control methods to ensure implemented models are accurate
@@
 is a tree sequence file that can
-be manipulated and analysed with \tskit.}
+be manipulated and analyzed with \tskit.}
 \label{fig:cartoon}
@@
 \paragraph{The \stdpopsim library}
-The central contribution of the PopSim consortium currently is \stdpopsim, a
+The first contribution of the PopSim consortium is \stdpopsim, a
 community-maintained library of empirical genome data and population genetics simulation
-models, providing easy access to realistic simulations. Figure \ref{fig:cartoon} shows a graphical
+models. Figure \ref{fig:cartoon} shows a graphical
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{P.~abelii},
+on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{P.~pygmaeus},
 \emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}.
@@
 lengths, average mutation rates and so on. We also provide access to detailed
-empirical information such as recombination maps, which model empirically-observed
-heterogeneity along chromosomes. As such maps are often large and hard
-to find, we provide a mechanism to automatically download (and cache locally) a
-set of previously published maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the
-HapMapII~\citep{international2007second} and
-deCODE~\citep{kong2010fine} genetic maps. The second key element of a species description
+empirical information such as genetic maps, which model observed
+heterogeneity in recombination rate along chromosomes. As such maps are often large,
+we do not distribute them directly with the software, but make them available
+for download in a standard format. When a simulation using such a map is
+requested by the user, \stdpopsim will transparently download the map
+data into a local cache (\texttt{$\sim$/.local/cache/stdpopsim} by default
+on Unix platforms), where it can be quickly retrieved for subsequent
+simulations.
+In the initial version of \stdpopsim we support
+the HapMapII~\citep{international2007second} and
+deCODE~\citep{kong2010fine} genetic maps in humans;
+the \cite{salome2011recombination} map in \emph{A.~thaliana};
+and the \cite{comeron2012many} map in \emph{D.~melanogaster}.
+The second key element of a species description
 within \stdpopsim is a set of carefully curated population genetic model
@@
 of specific historical scenarios that have influenced present-day patterns of
-genetic variation.
+genetic variation. (See the Methods for a description of the community
+development and quality-control process for these models.)
 
@@
 across a range of organisms. \stdpopsim has a Python API and a user-friendly
-command line interface (Fig.~\ref{fig:cartoon}C), allowing users with minimal experience direct access to
-state-of-the-art simulations. Simulations are output in the “tree sequence”
+command line interface, allowing users with minimal experience direct access to
+state-of-the-art simulations. Simulations are output in the ``tree sequence''
 format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
@@
 engines to support scenarios that cannot be modelled under the coalescent framework,
-and have already implemented \texttt{SLiM} \citep{haller2019slim} as
-an alternative back-end.
+and have already implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} as
+an additional backend.
 
@@
 We hope that this will encourage users to appropriately acknowledge the
-resources used in published work and authors
+resources used in published work, and encourage authors
 of simulation models to contribute to our ongoing community-driven development process.
-Together with the \stdpopsim version number, this citation information will result in a reasonably well-documented and reproducible simulation workflow.
+Together with the \stdpopsim version number and the long-term stable identifiers
+for population models and genetic maps,
+this citation information will result in well-documented and reproducible
+simulation workflows. The individual tree sequence files produced by
+\stdpopsim also contain complete provenance information including the command
+line arguments, operating system environment and versions of key libraries
+used.
 
@@
 for American populations which specifies ancestral African, European, and Asian population
-components, and the three-population out-of-Africa model from \cite{ragsdale2019models}
-which includes archaic admixture.
+components, a three-population out-of-Africa model from \cite{ragsdale2019models}
+which includes archaic admixture, a complex model of ancient eurasian admixture from \cite{kamm2019efficiently},
+and a synthetic model of oscillating population size from \cite{schiffels2014inferring}.
 Together these models
@@
 For non-human genomes we have have implemented two demographic histories for
-\emph{Drosophila melanogaster} and one each from \emph{A. thaliana} and \emph{E. coli}.
+\emph{D. melanogaster}, three from \emph{A. thaliana}, an one each from \emph{P. pygmaeus} and \emph{E. coli}.
 For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 and associated bottleneck model of \cite{li2006inferring} which jointly models African
-and European populations. For \emph{Arabidopsis thaliana}, we implemented the
+and European populations. For \emph{A. thaliana}, we implemented the
 model in \cite{durvasula2017african} inferred using \MSMC. This model includes
 a continuous change in population size over time, rather than pre-specified epochs of different
-population sizes. Additionally, we have also implemented two generic models that comprise a piecewise constant size model and an isolation with migration (IM) model.
+population sizes. We have also implemented a two epoch and a three epoch model estimated from African
+samples of \emph{A. thaliana} in \cite{huber2018gene}. For \emph{P. pygmaeus} we have implemented the two
+species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. pygmaeus}
+and \emph{P. abelii}. For \emph{E. coli} we have a simple, large constant population size model implemented.
+Beyond organism specific models \stdpopsim also includes a generic piecewise constant size model and
+isolation with migration (IM) model which can be used with any genome and genetic map combination.
 
@@
 \texttt{snakemake} \citep{koster2012snakemake} workflows,
-available from \url{https://github.com/popgensims/analysis}.
+available from \url{https://github.com/popsim-consortium/analysis}.
 These workflows allow efficient computing in multicore or cluster environments.
@@
 
-To see how well methods might do at recovering the population history of a constant sized population,
+To see how well methods might do at recovering the population history of a constant-sized population,
 we can simulate under a human genome architecture and genetic map, but without demography.
@@
 genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
-model estimated from an African sample of \emph{Drosophila melanogaster} \citep{sheehan2016deep}.
+model estimated from an African sample of \emph{D. melanogaster} \citep{sheehan2016deep}.
 As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
@@
 not match the model from which the data were generated (Figure
-\ref{fig:IM_popn_human}A and B).
+\ref{fig:IM_popn_human}A and B). However, because the inferred models here better
+match the simulated models relative to the single population case, here we compare
+our inferred population sizes to the census size of the simulated models (black line in
+Figure \ref{fig:IM_popn_human}C).
 
@@
 Again, we can compare between species and look at the performance of these methods in
-the context of a two-population model from \emph{Drosophila melanogaster}. Figure
+the context of a two-population model from \emph{D. melanogaster}. Figure
 \ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
@@
 For the single-population task we ran human (HomSap) simulations
-using a variety of models: OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
+using a variety of models (see Table 1): OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
 a constant-sized generic model, and a two-epoch generic model where the
@@
 For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
-% RNG: Removed technical term "coalesce" here, since I don't think that's exactly the issue.
-For example, both chromosomes 3R and X failed to finish within 30 days of starting using $n = 50$.
-Runtimes using the identical sample size were 19, 21, and 12 days for chromosomes 2L, 2R, and 3L,
-respectively. For this reason, we chose to present only data from chromosome 2R for simulations
+For this reason, we chose to present only data from 3 Mb of chromosome 2R for simulations
 under OutOfAfrica\_2L06.
-% RNG: Not present in manuscript
-% For the generic IM simulations, we used the HomSap genome along with the
-% HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
+For the generic IM simulations, we used the HomSap genome along with the
+HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
 
@@
 
-For \smcpp , we converted the tree sequences into vcf format and performed inference
+For \smcpp, we converted the tree sequences into vcf format and performed inference
 with default settings. Importantly, \smcpp assumes no migration post-divergence, so
@@
 To visualize our results, we plotted the inferred population size trajectories
-for each simulation replicate along side the true (census) population sizes
+for each simulation replicate along side the census population sizes
 (Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C). Here, unlike the single-population pipeline,
@@
 JRA and ADK were funded under NIH Award R01GM117241.
-TJS and RNG were funded under NIH award R01GM127348.
+TJS and RNG were funded under NIH Award R01GM127348.
+ND and AS were supported in part by NIH Awards R01HG010346 and R35GM127070.
 FR and GG were supported by a Villum Young Investigator award (project no.~00025300).

### 2019-12-19 popSimManu.tex
a few more edits to Discussion

--- 2019-12-19 popSimManu.tex
@@
 \usepackage{xspace}
-\usepackage{booktabs}
-\usepackage{siunitx}
 
@@
 \affil[18]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[19]{Big Data Institute, Li Ka Shing Centre for Health Information and
-Discovery, University of Oxford}
+\affil[19]{Big Data Institute, Oxford Unviersity}
 \affil[$\star$]{Denotes shared first authorship, listed alphabetically}
@@
 
-\renewcommand{\arraystretch}{1.2}
 \begin{table}[t]
-% Make sure the table is centered even though it's too wide
-\makebox[\textwidth][c]{
 \begin{footnotesize}
@@
 \end{footnotesize}
-}
 \caption{\label{tab:catalog}
-Initial set of demographic models in the Catalog and simple benchmarks.
-For each model we report the CPU time, maximum memory usage and the
-size of the output \tskit file. In each case we simulate 100 samples
-drawn from the first population, for the shortest chromosome of that species
-and a constant chromosome-specific recombination rate.
-The times reported are for a single replicate run on an Intel i5-7600 CPU.
-These are not intended to be rigorous benchmarks, and the computing resources
-required will vary widely depending on sample sizes, chromosome length,
-recombination rates and other factors.
+Details of the initial set of population models across four species.
+\textbf{NOTES: Column names are ID, name (or description maybe?),
+reference, CPU time, RAM}
 }
@@
 TJS and RNG were funded under NIH award R01GM127348.
-FR and GG were supported by a Villum Young Investigator award (project no.~00025300).
+FR and GG were supported by a Villum Young Investigator award (project no. 00025300).
 DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
-JK is supported by the Robertson Foundation.
 

### 2019-12-19 popSimManu.tex
Funding for JK.

@@
 TJS and RNG were funded under NIH award R01GM127348.
-FR and GG were supported by a Villum Young Investigator award (project no. 00025300).
+FR and GG were supported by a Villum Young Investigator award (project no.~00025300).
 DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
+JK is supported by the Robertson Foundation.
 

### 2019-12-19 popSimManu.tex
Edit pass on intro.

@@
 As large databases of population genetic variation begin to be used to inform public health procedures,
-the accuracy and quality of these inferences is becoming even more important.
+the accuracy and quality of these inferences is becoming ever more important.
 Assessing accuracy in population genetics is challenging in large part because
  ``ground-truth'' in population genetics comes from simulation,
-not from empirical observations, and as a field there has been no agreement as to what should
-constitute community standards or best practices. The general modus operandi to date has been to
+not from empirical observations, and as a field there has been no agreement
+on community standards or best practices. The general modus operandi to date has been to
 introduce methods that are validated by a set of bespoke simulations produced by individual
@@
 %RNG: Removed the sap. :-P Feel free to put it back if you like it better.
+%JK: not crazy about it either. Making it shorter helps though?
 

### 2019-12-19 popSimManu.tex
added model descriptions to results

@@
 For non-human genomes we have have implemented two demographic histories for
-\emph{Drosophila melanogaster} and one each from \emph{A. thaliana} and \emph{E. coli}.
+\emph{Drosophila melanogaster}, three from \emph{A. thaliana}, an one each from \emph{P. abelii} and \emph{E. coli}.
 For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 and associated bottleneck model of \cite{li2006inferring} which jointly models African
-and European populations. For \emph{Arabidopsis thaliana}, we implemented the
+and European populations. For \emph{A. thaliana}, we implemented the
 model in \cite{durvasula2017african} inferred using \MSMC. This model includes
 a continuous change in population size over time, rather than pre-specified epochs of different
-population sizes. Additionally, we have also implemented two generic models that comprise a piecewise constant size model and an isolation with migration (IM) model.
+population sizes. We have also implemented a two epoch and a three epoch model estimated from African
+samples of \emph{A. thaliana} in \cite{huber2018gene}. For \emph{P. abelii} we have implemented the two 
+species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. abelii}
+and \emph{P. pygmaeus}. For \emph{E. coli} we have a simple, large constant population size model implemented.
+Beyond organism specific models \stdpopsim also includes a generic piecewise constant size model and 
+isolation with migration (IM) model which can be used with any genome and genetic map combination.
 

### 2019-12-19 popSimManu.tex
Auto stash before rebase of "popsim-consortium/master"

@@
 
-The degree to which modeling assumptions and choices of data summaries can affect the ultimate results of population genetic inference has been challenging to assess. Yet there
+It has been challenging to assess the degree to which modeling assumptions
+and choices of data summaries can affect the ultimate results of population
+genetic inference. Yet there 
 have been several clear examples of different methods yielding fundamentally

### 2019-12-19 popSimManu.tex

--- 2019-12-19 popSimManu.tex
@@
 \paragraph{The \stdpopsim library}
-The central contribution of the PopSim consortium currently is \stdpopsim, a
+The first contribution of the PopSim consortium is \stdpopsim, a
 community-maintained library of empirical genome data and population genetics simulation
-models, providing easy access to realistic simulations. Figure \ref{fig:cartoon} shows a graphical
+models. Figure \ref{fig:cartoon} shows a graphical
 representation of the structure of \stdpopsim. The package centers

### 2019-12-19 popSimManu.tex
Use Stdpopsim at start of sentence.

@@
 
+\newcommand{\Stdpopsim}{\texttt{Stdpopsim}\xspace}
 \newcommand{\stdpopsim}{\texttt{stdpopsim}\xspace}
@@
 be measured. Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
-\stdpopsim is a community-driven open source project, which provides a standard
+\Stdpopsim is a community-driven open source project, which provides a standard
 catalog of published simulation models from a wide range of organisms
@@
 the accuracy and quality of these inferences is becoming ever more important.
+%CK: this first paragraph is super long - might be helpful to break up
+
 Assessing accuracy in population genetics is challenging in large part because
@@
 for simulating published demographic models from a number of popular study systems.
-This resource, which we call \texttt{stdpopsim}, makes running
+This resource, which we call \stdpopsim, makes running
 realistic simulations for population genetic analysis a simple matter of
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five organisms: humans, \emph{P. pygmaeus},
-\emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}. For each
+The \stdpopsim catalog currently contains five organisms: humans, \emph{Pongo~pygmaeus},
+\emph{Drosophila~melanogaster}, \emph{Arabidopsis~thaliana}, and \emph{Escherichia~coli}. For each
+% CK: i think we should spell out genus at first mention.
 organism the catalog contains details on the physical organization (e.g., chromosome structure)
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{display_items/Figure1.pdf}
+\includegraphics[width=0.7\linewidth]{display_items/Figure1.pdf}
 \caption{\textbf{Structure of \stdpopsim}. \textbf{A.} The
-hierarchical organization of the PopSim catalog contains all information
-within individual species. Information within the species level is shown here
-for \emph{Homo sapiens} only. We see that \emph{H. sapiens} is associated
-with a default generation time and mutation rate,
-a representation of the physical genome, and two genetic maps. Also shown is a
-subset of the demographic models currently implemented. \textbf{B.} Caching
-and automated download of genetic maps. \textbf{C.} Example code to specify
-and simulate models using \stdpopsim via the API (top) or
-command-line interface (bottom). \textbf{D.} Simulation output
-is a tree sequence file that can
-be manipulated and analysed with \tskit.}
+hierarchical organization of the \stdpopsim catalog contains all model simulation information
+within individual species (expanded information shown here for \emph{Homo sapiens} only). 
+Each species is associated with a representation of the physical genome, and one or more genetic maps and demographic models.
+Dotted lines indicate that only a subset of these categories is shown. \textbf{B.} Example code to specify
+and simulate models using \stdpopsim via the API (top) or\textbf{C.} through a command line interface.}
 \label{fig:cartoon}
@@
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{P.~abelii},
+on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{P.~pygmaeus},
 \emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}.
@@
 library, it is then straightforward to run accurate, standardized simulations
-across a range of organisms. \stdpopsim has a Python API and a user-friendly
+across a range of organisms. \Stdpopsim has a Python API and a user-friendly
 command line interface, allowing users with minimal experience direct access to
@@
 For non-human genomes we have have implemented two demographic histories for
-\emph{Drosophila melanogaster}, three from \emph{A. thaliana}, an one each from \emph{P. abelii} and \emph{E. coli}.
+\emph{D. melanogaster}, three from \emph{A. thaliana}, an one each from \emph{P. pygmaeus} and \emph{E. coli}.
 For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 population sizes. We have also implemented a two epoch and a three epoch model estimated from African
-samples of \emph{A. thaliana} in \cite{huber2018gene}. For \emph{P. abelii} we have implemented the two 
-species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. abelii}
-and \emph{P. pygmaeus}. For \emph{E. coli} we have a simple, large constant population size model implemented.
+samples of \emph{A. thaliana} in \cite{huber2018gene}. For \emph{P. pygmaeus} we have implemented the two
+species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. pygmaeus}
+and \emph{P. abelii}. For \emph{E. coli} we have a simple, large constant population size model implemented.
 Beyond organism specific models \stdpopsim also includes a generic piecewise constant size model and
@@
 
-\stdpopsim allows us to readily compare these estimates to those drawn from a different
+\Stdpopsim allows us to readily compare these estimates to those drawn from a different
 model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
@@
 about how methods might perform on non-human systems.
-
 Finally, comparison of results across methods or simulation runs
@@
 
-\stdpopsim is intended to be a fully open, community-developed project.
+\Stdpopsim is intended to be a fully open, community-developed project.
 Our implementations of genome representations and genetic maps for the some of
@@
 TJS and RNG were funded under NIH Award R01GM127348.
+ALG and DRS were funded under NIH award R00HG008696
 ND and AS were supported in part by NIH Awards R01HG010346 and R35GM127070.

### 2019-12-19 popSimManu.tex
added acknowledgments. https://github.com/popsim-consortium/manuscript/issues/84

@@
 
-\newcommand{\Stdpopsim}{\texttt{Stdpopsim}\xspace}
 \newcommand{\stdpopsim}{\texttt{stdpopsim}\xspace}
@@
 be measured. Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
-\Stdpopsim is a community-driven open source project, which provides a standard
+\stdpopsim is a community-driven open source project, which provides a standard
 catalog of published simulation models from a wide range of organisms
@@
 for simulating published demographic models from a number of popular study systems.
-This resource, which we call \stdpopsim, makes running
+This resource, which we call \texttt{stdpopsim}, makes running
 realistic simulations for population genetic analysis a simple matter of
@@
 \begin{center}
-\includegraphics[width=0.7\linewidth]{display_items/Figure1.pdf}
+\includegraphics[width=0.5\linewidth]{display_items/Figure1.pdf}
 \caption{\textbf{Structure of \stdpopsim}. \textbf{A.} The
-hierarchical organization of the \stdpopsim catalog contains all model simulation information
-within individual species (expanded information shown here for \emph{Homo sapiens} only). 
-Each species is associated with a representation of the physical genome, and one or more genetic maps and demographic models.
-Dotted lines indicate that only a subset of these categories is shown. \textbf{B.} Example code to specify
-and simulate models using \stdpopsim via the API (top) or\textbf{C.} through a command line interface.}
+hierarchical organization of the PopSim catalog contains all information
+within individual species. Information within the species level is shown here
+for \emph{Homo sapiens} only. We see that \emph{H. sapiens} is associated
+with a default generation time and mutation rate,
+a representation of the physical genome, and two genetic maps. Also shown is a
+subset of the demographic models currently implemented. \textbf{B.} Caching
+and automated download of genetic maps. \textbf{C.} Example code to specify
+and simulate models using \stdpopsim via the API (top) or
+command-line interface (bottom). \textbf{D.} Simulation output
+is a tree sequence file that can
+be manipulated and analyzed with \tskit.}
 \label{fig:cartoon}
@@
 library, it is then straightforward to run accurate, standardized simulations
-across a range of organisms. \Stdpopsim has a Python API and a user-friendly
+across a range of organisms. \stdpopsim has a Python API and a user-friendly
 command line interface, allowing users with minimal experience direct access to
@@
 
-\Stdpopsim allows us to readily compare these estimates to those drawn from a different
+\stdpopsim allows us to readily compare these estimates to those drawn from a different
 model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
@@
 about how methods might perform on non-human systems.
+
 Finally, comparison of results across methods or simulation runs
@@
 
-\Stdpopsim is intended to be a fully open, community-developed project.
+\stdpopsim is intended to be a fully open, community-developed project.
 Our implementations of genome representations and genetic maps for the some of

### 2019-12-19 popSimManu.tex
text edits and figure updates

@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{display_items/Figure1.pdf}
+\includegraphics[width=0.7\linewidth]{display_items/Figure1.pdf}
 \caption{\textbf{Structure of \stdpopsim}. \textbf{A.} The
-hierarchical organization of the PopSim catalog contains all information
-within individual species. Information within the species level is shown here
-for \emph{Homo sapiens} only. We see that \emph{H. sapiens} is associated
-with a default generation time and mutation rate,
-a representation of the physical genome, and two genetic maps. Also shown is a
-subset of the demographic models currently implemented. \textbf{B.} Caching
-and automated download of genetic maps. \textbf{C.} Example code to specify
-and simulate models using \stdpopsim via the API (top) or
-command-line interface (bottom). \textbf{D.} Simulation output
-is a tree sequence file that can
-be manipulated and analyzed with \tskit.}
+hierarchical organization of the \stdpopsim catalog contains all model simulation information
+within individual species (expanded information shown here for \emph{Homo sapiens} only). 
+Each species is associated with a representation of the physical genome, and one or more genetic maps and demographic models.
+Dotted lines indicate that only a subset of these categories is shown. \textbf{B.} Example code to specify
+and simulate models using \stdpopsim via the API (top) or\textbf{C.} through a command line interface.}
 \label{fig:cartoon}
@@
 TJS and RNG were funded under NIH Award R01GM127348.
-ALG and DRS were funded under NIH award R00HG008696
 ND and AS were supported in part by NIH Awards R01HG010346 and R35GM127070.

### 2019-12-19 popSimManu.tex
added comments and minor edits

--- 2019-12-19 popSimManu.tex
@@
 \usepackage{booktabs}
-\usepackage{siunitx}
+\usepackage[binary-units]{siunitx}
+\usepackage[section]{placeins}
 
@@
 \newcommand{\smcpp}{\texttt{smc++}\xspace}
-\newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
+\newcommand{\stairwayplot}{\texttt{stairway plot}\xspace}
 \newcommand{\fastsimcoal}{\texttt{fastsimcoal2}\xspace}
@@
 
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+\rhead{\texttt{stdpopsim}}
+\lhead{PopSim Consortium}
+\fancyfoot[C]{\thepage}
+
 \begin{document}
 
-\title{A community-maintained standard library of population genetic simulations}
+\title{A community-maintained standard library of population genetic models}
 % First authors
@@
 
-Assessing accuracy in population genetics is challenging in large part because
- ``ground-truth'' in population genetics comes from simulation,
-not from empirical observations, and as a field there has been no agreement
-on community standards or best practices. The general modus operandi to date has been to
-introduce methods that are validated by a set of bespoke simulations produced by individual
-groups. Beyond inefficiency from duplicated effort, such a situation leads
-to decreased reproducibility and transparency across the entire field.
-Here we describe \stdpopsim, a community-driven catalog
-of standardized population genetics models spanning a variety of common study organisms.
-This resource will empower the field to rigorously evaluate population genetic software,
-%and select the best tool for a given task,
-thereby substantially increasing the reliability of population genetic inferences.
-%DRS: still not in love with this last sentence here, but tried to make it a little less sappy by adding some explanation.
-%ADK: made some edits here
-%RNG: Removed the sap. :-P Feel free to put it back if you like it better.
-%JK: not crazy about it either. Making it shorter helps though?
+Assessing the accuracy of inference methods for population genetics is
+challenging in large part because the ``ground-truth'' in question
+generally comes not from direct empirical observations, as the relevant
+historical processes can rarely be observed, but instead from simulations.
+Population genetic simulations are therefore critically important to the
+field, yet there has been no systematic attempt to establish community
+standards or best practices for executing them. Instead, the general modus
+operandi to date has been for individual groups to validate their own
+methods using bespoke simulations. 
+Often these studies focus more on showcasing a novel method than on
+Moreover, this approach results in a great deal of duplicated effort across research
+It is also a barrier to entry to the field, because new researchers can
+struggle with 
+the many steps involved
+in implementing a state-of-the-art population genetics simulation,
+including identifying appropriate demographic models from the literature, 
+translating them into input for a simulator, obtaining appropriate genetic
+maps, and choosing appropriate values for key population genetic parameters.
+groups, and contributes to decreased reproducibility and transparency.
+rigorously comparing it with competing methods.
 
-It has been challenging to assess the degree to which modeling assumptions
-and choices of data summaries can affect the ultimate results of population
-genetic inference. Yet there
-have been several clear examples of different methods yielding fundamentally
+A related issue is that it has been challenging to assess the degree to which modeling assumptions
+and choices of data summaries can affect population
+genetic inferences. Yet there
+are clear examples of different methods yielding fundamentally
 different conclusions. For example, Markovian coalescent methods applied to human genomes have
@@
 genetic variation data, suggesting that such design choices can greatly affect the
-performance of the inference. Furthermore, it is reasonable
-to assume generally that specific methods will perform better than others under certain scenarios.
-However, because these conditions are typically unknown, empirical researchers
-lack principled guidelines for deciding which statistical method is best suited
-to accurately answer their particular question. The need for empirical
-guidance will only increase as researchers continue to generate more genomic
-variation data from non-model taxa and wish to make inferences regarding
-demography and selection.
-
-One way to benchmark different methods for statistical inference in population
-genetics is to apply the methods to simulated datasets where the true model
-parameters are known. Methods that perform well will accurately infer these
-parameters and will also provide reasonable measures of uncertainty of the
-estimated parameters. Indeed, simulation studies are a standard ingredient in
-most papers proposing new methods in population genetics. However, these
-studies often use simulations of the particular models for which the method was designed,
-which showcases the novel method rather than rigorously comparing performance to competing methods.
-Of course, testing a new method in the situation for which it was designed is an important step,
-but more rigorous testing is necessary for tools that are to be widely used in the field.
-% Moreover researchers are generally forced to reimplement previously
-% inferred scenarios independently, which makes good benchmarking
-% a major additional time investment, and can lead to errors.
-% Lastly, the degree to which different methods are compared can be quite variable.
-Many more researchers in the broader field
-use simulations to evaluate statistical power and perform other exploratory analyses.
-The many steps of implementing a moderately complicated population genetics simulation,
-including translating published demographic models into input for a simulator
-and obtaining genetic maps,
-present a major obstacle, especially to new researchers.
-
-% One reason why simulation studies do not compare the performance of many methods
-% across a wide range of situations
-% is that simulation studies are challenging to carry out, for several reasons.
-% First, while there are many published population genetic models available, their
-% parameters are often not easily accessible, usually presented in supplemental
-% tables. Translating these parameters into input for a given simulation engine can
-% be a time-consuming and error prone process. In addition, papers often present many
-% different models, and different authors may choose to select distinct models,
-% making comparison difficult. While improvements in simulation software and ever
-% faster computers have partially ameliorated this concern, generating the
-% simulated datasets can still be a time and computing resource bottleneck.
+performance of the inference. Furthermore, some methods are likely to
+perform better than others under certain scenarios, but 
+researchers lack principled guidelines for selecting the best method for addressing 
+their particular questions. The need for empirical
+guidance will only increase as researchers 
+seek to apply population genetic methods to a growing collection of non-model taxa.
 
@@
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five organisms: humans, \emph{Pongo~pygmaeus},
-\emph{Drosophila~melanogaster}, \emph{Arabidopsis~thaliana}, and \emph{Escherichia~coli}. For each
-% CK: i think we should spell out genus at first mention.
+The \stdpopsim catalog currently contains three organisms: humans,
+\emph{Drosophila~melanogaster}, and \emph{Arabidopsis~thaliana}. For each
 organism the catalog contains details on the physical organization (e.g., chromosome structure)
@@
 simulation output from their chosen model is returned.
+In this way, \stdpopsim will enable rigorous evaluation of population genetic software,
+and contribute to increased reliability of population genetic inferences.
 
-The software has been developed by the PopSim Consortium using a
+The \stdpopsim library has been developed by the PopSim Consortium using a
 distributed open source model, with strong procedures in place
@@
 \includegraphics[width=0.7\linewidth]{display_items/Figure1.pdf}
-\caption{\textbf{Structure of \stdpopsim}. \textbf{A.} The
+    \caption{\textbf{Structure of \stdpopsim}. \textbf{(A)} The
 hierarchical organization of the \stdpopsim catalog contains all model simulation information
-within individual species (expanded information shown here for \emph{Homo sapiens} only). 
+within individual species (expanded information shown here for \emph{H. sapiens} only). 
 Each species is associated with a representation of the physical genome, and one or more genetic maps and demographic models.
-Dotted lines indicate that only a subset of these categories is shown. \textbf{B.} Example code to specify
-and simulate models using \stdpopsim via the API (top) or\textbf{C.} through a command line interface.}
+Dotted lines indicate that only a subset of these categories is shown. \textbf{(B)} Example code to specify
+and simulate models using \stdpopsim via the API or \textbf{(C)} through a command line interface.}
 \label{fig:cartoon}
@@
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{P.~pygmaeus},
-\emph{D.~melanogaster}, \emph{A.~thaliana}, and \emph{E.~coli}.
+on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans,
+\emph{D.~melanogaster}, \emph{A.~thaliana}.
 % RNG: Abbreviate consistently with introduction
@@
 \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
-as the default simulation engine. We plan to include other simulation
-engines to support scenarios that cannot be modelled under the coalescent framework,
-and have already implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} as
-an additional backend.
+as the default simulation engine. 
+We have implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim}
+as an alternative backend,
+to allow simulation of processes that cannot be modeled under the coalescent.
 
@@
 
-\subsection*{The \stdpopsim Catalog}
+\subsection*{The Species Catalog}
+The central feature of \stdpopsim is the species catalog, a systematic organization
+of the key quantitative data needed to simulate a given species. These include a description
+of the assembly, information about mutation rate, recombination rate(s), and generation time
+in addition to a series of demographic models that are specific to that organism.
+
 The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}. At the time of
 writing, the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Pongo}, \emph{Drosophila}, and \emph{Arabidopsis}.
+humans, \emph{Drosophila}, and \emph{Arabidopsis}, 
+and we have models in development for \emph{Pongo abelii} and \emph{Escherichia coli}.
 These range from
@@
 
-At the time of writing, more models of human history have been implemented than for
-any other organism. These models include: a simplified version of the \cite{tennessen2012evolution}
+Currently, more models of human history have been implemented than for
+any other organism (see Table \ref{tab:catalog}). 
+These models include: a simplified version of the \cite{tennessen2012evolution}
 model with only the African population specified (expansion from the ancestral
-population and recent growth), the three population model of \cite{gutenkunst2009inferring}
+population and recent growth; Africa\_1T12), the three-population model of \cite{gutenkunst2009inferring}
 which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
-the European and Asian populations, the \cite{tennessen2012evolution} two population variant of the
+the European and Asian populations (OutOfAfrica\_3G09), the \cite{tennessen2012evolution} two-population variant of the
 Gutenkunst model which does not include Asian populations, but more explicitly models
-recent rapid human population growth, the \cite{browning2018ancestry} admixture model
+recent rapid human population growth (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
 for American populations which specifies ancestral African, European, and Asian population
-components, a three-population out-of-Africa model from \cite{ragsdale2019models}
-which includes archaic admixture, a complex model of ancient eurasian admixture from \cite{kamm2019efficiently},
-and a synthetic model of oscillating population size from \cite{schiffels2014inferring}.
+components (AmericanAdmixture\_4B11), a three-population out-of-Africa model from \cite{ragsdale2019models}
+which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19),
+a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19),
+and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).
 Together these models
@@
 For non-human genomes we have have implemented two demographic histories for
-\emph{D. melanogaster}, three from \emph{A. thaliana}, an one each from \emph{P. pygmaeus} and \emph{E. coli}.
+\emph{D. melanogaster}, three from \emph{A. thaliana}, and have under development
+model for \emph{P. abelii} and \emph{E. coli}.
 For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
-an African sample, as well as the out-of-Africa divergence
+an African sample (African3Epoch\_1S16), as well as the out-of-Africa divergence
 and associated bottleneck model of \cite{li2006inferring} which jointly models African
-and European populations. For \emph{A. thaliana}, we implemented the
+and European populations (OutOfAfrica\_2L06). For \emph{A. thaliana}, we implemented the
 model in \cite{durvasula2017african} inferred using \MSMC. This model includes
 a continuous change in population size over time, rather than pre-specified epochs of different
-population sizes. We have also implemented a two epoch and a three epoch model estimated from African
-samples of \emph{A. thaliana} in \cite{huber2018gene}. For \emph{P. pygmaeus} we have implemented the two
-species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. pygmaeus}
-and \emph{P. abelii}. For \emph{E. coli} we have a simple, large constant population size model implemented.
+population sizes (SouthMiddleAtlas\_1D17). We have also implemented a two-epoch and a three-epoch model estimated from African
+samples of \emph{A. thaliana} in \cite{huber2018gene} (African2Epoch\_1H18 and African3Epoch\_1H18). 
+%For \emph{P. pygmaeus} we have implemented the two
+%species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. pygmaeus}
+%and \emph{P. abelii}. 
+%For \emph{E. coli} we have a simple, large constant population size model implemented.
 Beyond organism specific models \stdpopsim also includes a generic piecewise constant size model and
@@
 
+We note that we have put a good deal of effort into coming up with a standardized naming
+conventions for species, genetic maps, and demographic models that will enable long term stability
+of unique identifiers used throughout \stdpopsim. This is an essential feature to guarantee reproducibility.
 \subsection*{Use case: comparing methods of demographic inference}
@@
  on simulated genomes sampled from a single population,
-in a number of the demographic models described above. The pipeline was generalized to
+in a number of the demographic models described above. The workflow was generalized to
 run $R$ replicates with $C$ chromosomes for $n$ samples, for a total of $R \times C$
-simulations for each demographic model. After simulation, the pipeline prepares
+simulations for each demographic model. After simulation, the workfow prepares
 input files for each of the respective inference methods by grouping all
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot}, and
+size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate human genomes under the \cite{ragsdale2019models} model and using the
-HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
+replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model and using the
+HapMapII\_GRCh37 genetic map~\citep{international2007second}. From top to bottom we show estimates for each
 of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
@@
 In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
-run on whole genome simulations under under a sophisticated demographic model
-and empirical genetic map and a model of human migration out of Africa
-that includes archaic admiture \citep{ragsdale2019models}. In each column of this figure
+run on whole genome simulations under under OutOfAfricaArchaicAdmixture\_5R19,
+a model of human migration out of Africa that includes archaic admixture 
+\citep{ragsdale2019models}, along with an empirical genetic map. In each column of this figure
 we show inferred $N(t)$ for the three extant populations in the model.
@@
 sample sizes for \MSMC).
+% ALG: I still think this sentence is confusing.
 There is no single ``true'' reference for effective population size
-because of model misspecification -- the inference methods are fitting a single population model
+because of model misspecification---the inference methods are fitting a single population model
 to data simulated from mulitple populations.
@@
 
-\Stdpopsim allows us to readily compare these estimates to those drawn from a different
+\stdpopsim allows us to readily compare these estimates to those drawn from a different
 model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
-$N(t)$ from simulations using the same physical and genetic maps, but a different demographic
-history that does not include archaic admixture \citep{gutenkunst2009inferring}. Again we see that each
+$N(t)$ from simulations using the same physical and genetic maps, but from the OutOfAfrica\_3G09
+demographic model that does not include archaic admixture. Again we see that each
 of the methods is capturing relevant parts of the population history, although the
@@
 experiment in Figure \ref{fig:n_t_HomSap_constant}. Again we see that all methods
-are recovering a modern population size within a factor of two of the truth, however
+recover a modern population size within a factor of two of the truth, however
 SMC-based methods, perhaps due to their regularization, tend to learn sinusoidal
@@
 size over time ($N(t)$) estimated from an African population sample. Data were generated by simulating
-replicate \emph{D.melanogaster} genomes under the three-epoch \cite{sheehan2016deep} model
+replicate \emph{D.melanogaster} genomes under the African3Epoch\_1S16 model
 with the genetic map inferred in \cite{comeron2012many}. In shades of blue we show the estimated
@@
 data, it is of consequence to ask how such methods might perform in non-human
-genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
-model estimated from an African sample of \emph{D. melanogaster} \citep{sheehan2016deep}.
+genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from the African3Epoch\_1S16
+model, originally estimated from an African sample of \emph{D. melanogaster} \citep{sheehan2016deep}.
 As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
@@
 \caption{\textbf{Parameters estimated using a multi-population human model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
-A) Data were generated by simulating
-replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}.
-B) For \dadi and \fastsimcoal we show parameters inferred
+Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp. \textbf{(A)} Data were generated by simulating
+replicate human genomes under the OutOfAfrica\_3G09 model and using the HapMapII\_GRCh37 genetic map
+inferred in \cite{international2007second}. \textbf{(B)} For \dadi and \fastsimcoal we show parameters inferred
 by fitting the depicted IM model, which includes population sizes, migration rates, and a split
-time between CEU and YRI samples.
-C) Population size estimates for each population (rows)
+time between CEU and YRI samples. \textbf{(C)} Population size estimates for each population (rows)
 from \dadi, \fastsimcoal, and \smcpp (columns).
@@
 time, for each of the inference methods, using samples drawn from African and European populations
-simulated under the \cite{gutenkunst2009inferring} model. Our results highlight many
+simulated under the OutOfAfrica\_3G09 model. Our results highlight many
 of the strengths and weaknesses of the different types of methods we used.
 For instance, the SFS-based approaches where we fit simple IM models do not capture
-recent exponential growth in the CEU population, however do consistently recover the
+recent exponential growth in the CEU population, but do consistently recover the
 simulated YRI population size history. Moreover, these approaches allow for estimating
 migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
-of divergence times. By contrast, \smcpp is much better at capturing the recent exponential
-growth in the CEU population, though the inferred population sizes are in general more
-noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
-estimates that are consistently underestimated.
+of divergence times. However, these migration rate estimates are somewhat biased likely
+due to model misspecification (Figure \ref{fig:homsap_mig_rates}).
+By contrast, \smcpp is much better at capturing the recent exponential
+growth in the CEU population, though the inferred population sizes are in generally noisier. 
+In addition, the assumption of no migration by \smcpp leads to divergence time
+estimates that are consistently underestimated (Figure \ref{fig:IM_popn_human}C).
 
@@
 \ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
-the model of \cite{li2006inferring}, which includes
+the OutOfAfrica\_2L06 model, which includes
 an ancestral population in Africa, then a population expansion with a population
@@
 Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
-parameters, but ignore the brief population bottleneck in Europe. The inferred
-demographic parameters from \smcpp are more noisy, though in some cases are better
-able to capture the bottleneck in the Europe population.
+parameters, but ignore the brief population bottleneck in Europe. In addition, \dadi
+and \fastsimcoal both do reasonably well at correctly inferring the absence of
+migration (Figure \ref{fig:dromel_mig_rates}). By contrast, the inferred
+demographic parameters from \smcpp are more noisy, though in some cases better
+capture the short bottleneck in the Europe population.
 
@@
 the \stdpopsim library. We have founded the Consortium with a number of specific goals in mind:
-standardization of simulation within the computational genetics community,
+standardization of simulation within the population genetics community,
 increasing reproducibility and ease of use of complex simulation based results,
@@
 % redundant.
+% ALG: I'm not sure about that. I don't really see that it's covered below.
 
@@
 (Figure~\ref{fig:IM_popn_human}), most methods do reasonably well at
-reconstructing the YRI history, but struggle with the more complex CEU
+reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
 history, in large part because of the restriction of constant sizes per
-population.  An exception is \smcpp, which does not have the same
+population.  
+An exception is \smcpp, which does not have the same
 restrictions on its inferred history, and as a result does somewhat better
@@
 %\adk{should we mention benchmarking of tree sequence inference here? what else?}
+% ALG: Mention in what way?
+
+% ALG: I think the paragraph below has unnecessary detail that we don't need at this point.
 
@@
 demographic model for the relevant organism along with clear documentation
-of the model parameters and populations. This model is submitted as a Pull Request,
+of the model parameters and populations. 
+% ALG: I’m not sure what documentation of populations means. 
+This model is submitted on GitHub as a Pull Request,
 where it is evaluated by a reviewer and then included as `preliminary',
 but is not linked to the online documentation nor the command line interface.
-Inclusion of a preliminary model triggers a QC issue, and a second developer,
+Inclusion of a preliminary model 
+% ALG: automatically?
+triggers a QC issue, and a second developer,
 Developer B, then independently reimplements the model from the relevant
 primary sources and adds an automatic unit test for equality between the
-QC implementation and the preliminary production model. If the two
-implementations are equal the original model is included in \stdpopsim.
+QC implementation and the preliminary production model. 
+If the two implementations are equal the original model is included in \stdpopsim.
 If not, we move to an arbitration process whereby A and B first try
@@
 the \stdpopsim resource. In this spirit we ran each software package as near
-to default parameters as possibile. For \stairwayplot we
+to default parameters as possible. For \stairwayplot we
 set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
@@
 used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
-Specifically we masked regions of 1\,cM or greater in the lowest 5th percentile of the empirical
+Specifically we masked regions of \SI{1}{cM} or greater in the lowest 5th percentile of the empirical
 distribution of recombination, regions which are nearly uniformly absent for
@@
 
-From HomSap we ran replicate whole genome simulations under the model
-and specified taking 20 samples each from the Europe and Africa populations.
-For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
-For this reason, we chose to present only data from 3 Mb of chromosome 2R for simulations
-under OutOfAfrica\_2L06.
+For HomSap we ran replicate whole genome simulations under the OutOfAfrica\_3G09 model
+with the HapmapII\_GRCh37 genetic map, taking 20 samples each from the Europe and Africa populations.
+For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes
+with the Comeron2012\_dm6 map due to the large effective population sizes leading to
+high effective recombination rates. For this reason, we chose to present only data
+from \SI{3}{\mega\byte} of chromosome 2R for simulations under OutOfAfrica\_2L06.
 For the generic IM simulations, we used the HomSap genome along with the
-HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
+HapmapII\_GRCh37 genetic map and sampled 20 individuals from each population.
 
@@
 
+% ALG: Is this also in the Snakefile?
 For \smcpp, we converted the tree sequences into vcf format and performed inference
@@
 we focus on census size rather than the inverse coalescence rate as the `true' population size.
+% ALG: If you have Snakefile for the whole workflow, I would just say that at the end here.
 
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible.
@@
 TJS and RNG were funded under NIH Award R01GM127348.
-ALG and DRS were funded under NIH award R00HG008696
+ALG and DRS were funded under NIH award R00HG008696.
 ND and AS were supported in part by NIH Awards R01HG010346 and R35GM127070.
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairwayplot},
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairway plot},
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating

### 2019-12-19 popSimManu.tex

--- 2019-12-19 popSimManu.tex
@@
 
-\title{A community-maintained standard library of population genetic simulations}
+\title{A community-maintained standard library of population genetic models}
 % First authors
@@
 for simulating published demographic models from a number of popular study systems.
-This resource, which we call \texttt{stdpopsim}, makes running
+This resource, which we call \stdpopsim, makes running
 realistic simulations for population genetic analysis a simple matter of
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five organisms: humans, \emph{Pongo~pygmaeus},
-\emph{Drosophila~melanogaster}, \emph{Arabidopsis~thaliana}, and \emph{Escherichia~coli}. For each
-% CK: i think we should spell out genus at first mention.
+The \stdpopsim catalog currently contains three organisms: humans,
+\emph{Drosophila~melanogaster}, and \emph{Arabidopsis~thaliana}. For each
 organism the catalog contains details on the physical organization (e.g., chromosome structure)
@@
 writing, the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Pongo}, \emph{Drosophila}, and \emph{Arabidopsis}.
+humans, \emph{Drosophila}, and \emph{Arabidopsis}, 
+and we have models in development for \emph{Pongo abelii} and \emph{Escherichia coli}.
 These range from
@@
 For non-human genomes we have have implemented two demographic histories for
-\emph{D. melanogaster}, three from \emph{A. thaliana}, an one each from \emph{P. pygmaeus} and \emph{E. coli}.
+\emph{D. melanogaster}, three from \emph{A. thaliana}, and have under development
+model for \emph{P. abelii} and \emph{E. coli}.
 For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 population sizes. We have also implemented a two epoch and a three epoch model estimated from African
-samples of \emph{A. thaliana} in \cite{huber2018gene}. For \emph{P. pygmaeus} we have implemented the two
-species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. pygmaeus}
-and \emph{P. abelii}. For \emph{E. coli} we have a simple, large constant population size model implemented.
+samples of \emph{A. thaliana} in \cite{huber2018gene}. 
+%For \emph{P. pygmaeus} we have implemented the two
+%species divergence model of \cite{locke2011comparative}, which models isolation with migration between \emph{P. pygmaeus}
+%and \emph{P. abelii}. 
+%For \emph{E. coli} we have a simple, large constant population size model implemented.
 Beyond organism specific models \stdpopsim also includes a generic piecewise constant size model and

### 2019-12-19 popSimManu.tex

@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot}, and
+size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 \caption{\textbf{Parameters estimated using a multi-population human model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
-A) Data were generated by simulating
+Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp. \textbf{(A)} Data were generated by simulating
 replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}.
-B) For \dadi and \fastsimcoal we show parameters inferred
+inferred in \cite{international2007second}. \textbf{(B)} For \dadi and \fastsimcoal we show parameters inferred
 by fitting the depicted IM model, which includes population sizes, migration rates, and a split
-time between CEU and YRI samples.
-C) Population size estimates for each population (rows)
+time between CEU and YRI samples. \textbf{(C)} Population size estimates for each population (rows)
 from \dadi, \fastsimcoal, and \smcpp (columns).

### 2019-12-19 popSimManu.tex

@@
 For instance, the SFS-based approaches where we fit simple IM models do not capture
-recent exponential growth in the CEU population, however do consistently recover the
+recent exponential growth in the CEU population, but do consistently recover the
 simulated YRI population size history. Moreover, these approaches allow for estimating
 migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
-of divergence times. By contrast, \smcpp is much better at capturing the recent exponential
+of divergence times. However, these migration rate estimates are somewhat biased likely
+due to model misspecification (Figure \ref{fig:homsap_mig_rates}).
+By contrast, \smcpp is much better at capturing the recent exponential
 growth in the CEU population, though the inferred population sizes are in general more
 noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
-estimates that are consistently underestimated.
+estimates that are consistently underestimated (Figure \ref{fig:IM_popn_human}C).
 
@@
 
-From HomSap we ran replicate whole genome simulations under the model
-and specified taking 20 samples each from the Europe and Africa populations.
-For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes.
-For this reason, we chose to present only data from 3 Mb of chromosome 2R for simulations
-under OutOfAfrica\_2L06.
+From HomSap we ran replicate whole genome simulations under the OutOfAfrica\_3G09 model
+with the HapmapII\_GRCh37 genetic map, taking 20 samples each from the Europe and Africa populations.
+For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes
+with the Comeron2012\_dm6 map due to the large effective population sizes leading to
+high effective recombination rates. For this reason, we chose to present only data
+from 3 Mb of chromosome 2R for simulations under OutOfAfrica\_2L06.
 For the generic IM simulations, we used the HomSap genome along with the
-HapmapII\_GRCh37 genetic map, and sampled 20 individuals from each population.
+HapmapII\_GRCh37 genetic map and sampled 20 individuals from each population.
 

### 2019-12-20 popSimManu.tex

@@
 
-\subsection*{The \stdpopsim Catalog}
+\subsection*{The Species Catalog}
+The central feature of \stdpopsim is the species catalog, a systematic organization
+of the key quantitative data needed to simulate a given species. These include a description
+of the assembly, information about mutation rate, recombination rate(s), and generation time
+in addition to a series of demographic models that are specific to that organism.
+
 The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}. At the time of
@@
 At the time of writing, more models of human history have been implemented than for
-any other organism. These models include: a simplified version of the \cite{tennessen2012evolution}
+any other organism (see Table \ref{tab:catalog}). 
+These models include: a simplified version of the \cite{tennessen2012evolution}
 model with only the African population specified (expansion from the ancestral
-population and recent growth), the three population model of \cite{gutenkunst2009inferring}
+population and recent growth; Africa\_1T12), the three-population model of \cite{gutenkunst2009inferring}
 which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
-the European and Asian populations, the \cite{tennessen2012evolution} two population variant of the
+the European and Asian populations (OutOfAfrica\_3G09), the \cite{tennessen2012evolution} two-population variant of the
 Gutenkunst model which does not include Asian populations, but more explicitly models
-recent rapid human population growth, the \cite{browning2018ancestry} admixture model
+recent rapid human population growth (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
 for American populations which specifies ancestral African, European, and Asian population
-components, a three-population out-of-Africa model from \cite{ragsdale2019models}
-which includes archaic admixture, a complex model of ancient eurasian admixture from \cite{kamm2019efficiently},
-and a synthetic model of oscillating population size from \cite{schiffels2014inferring}.
+components (AmericanAdmixture\_4B11), a three-population out-of-Africa model from \cite{ragsdale2019models}
+which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19),
+a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19),
+and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).
 Together these models
@@
 For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
-an African sample, as well as the out-of-Africa divergence
+an African sample (African3Epoch\_1S16), as well as the out-of-Africa divergence
 and associated bottleneck model of \cite{li2006inferring} which jointly models African
-and European populations. For \emph{A. thaliana}, we implemented the
+and European populations (OutOfAfrica\_2L06). For \emph{A. thaliana}, we implemented the
 model in \cite{durvasula2017african} inferred using \MSMC. This model includes
 a continuous change in population size over time, rather than pre-specified epochs of different
-population sizes. We have also implemented a two epoch and a three epoch model estimated from African
-samples of \emph{A. thaliana} in \cite{huber2018gene}. 
+population sizes (SouthMiddleAtlas\_1D17). We have also implemented a two-epoch and a three-epoch model estimated from African
+samples of \emph{A. thaliana} in \cite{huber2018gene} (African2Epoch\_1H18 and African3Epoch\_1H18). 
 %For \emph{P. pygmaeus} we have implemented the two

### 2019-12-20 popSimManu.tex
some reworking of intro

@@
 
-Assessing accuracy in population genetics is challenging in large part because
- ``ground-truth'' in population genetics comes from simulation,
-not from empirical observations, and as a field there has been no agreement
-on community standards or best practices. The general modus operandi to date has been to
-introduce methods that are validated by a set of bespoke simulations produced by individual
-groups. Beyond inefficiency from duplicated effort, such a situation leads
-to decreased reproducibility and transparency across the entire field.
-Here we describe \stdpopsim, a community-driven catalog
-of standardized population genetics models spanning a variety of common study organisms.
-This resource will empower the field to rigorously evaluate population genetic software,
-%and select the best tool for a given task,
-thereby substantially increasing the reliability of population genetic inferences.
-%DRS: still not in love with this last sentence here, but tried to make it a little less sappy by adding some explanation.
-%ADK: made some edits here
-%RNG: Removed the sap. :-P Feel free to put it back if you like it better.
-%JK: not crazy about it either. Making it shorter helps though?
+Assessing the accuracy of inference methods for population genetics is
+challenging in large part because the ``ground-truth'' in question
+generally comes not from direct empirical observations, as the relevant
+historical processes can rarely be observed, but instead from simulations.
+Population genetic simulations are therefore critically important to the
+field, yet there has been no systematic attempt to establish community
+standards or best practices for executing them. Instead, the general modus
+operandi to date has been for individual groups to validate their own
+methods using bespoke simulations. 
+Often these studies focus more on showcasing a novel method than on
+rigorously comparing it with competing methods.
+Moreover, this approach results in a great deal of duplicated effort across research
+groups, and contributes to decreased reproducibility and transparency.
+It is also a barrier to entry to the field, because new researchers can
+struggle with 
+the many steps involved
+in implementing a state-of-the-art population genetics simulation,
+including identifying appropriate demographic models from the literature, 
+translating them into input for a simulator, obtaining appropriate genetic
+maps, and choosing appropriate values for key population genetic parameters.
 
-It has been challenging to assess the degree to which modeling assumptions
-and choices of data summaries can affect the ultimate results of population
-genetic inference. Yet there
-have been several clear examples of different methods yielding fundamentally
+A related issue is that it has been challenging to assess the degree to which modeling assumptions
+and choices of data summaries can affect population
+genetic inferences. Yet there
+are clear examples of different methods yielding fundamentally
 different conclusions. For example, Markovian coalescent methods applied to human genomes have
@@
 realistic simulations for population genetic analysis a simple matter of
-choosing pre-implemented models from a community-maintained catalog.
+choosing pre-implemented models from a community-maintained catalog, 
 The \stdpopsim catalog currently contains three organisms: humans,
@@
 simulation output from their chosen model is returned.
+In this way, \stdpopsim will enable rigorous evaluation of population genetic software,
+and contribute to increased reliability of population genetic inferences.
 
-The software has been developed by the PopSim Consortium using a
+The \stdpopsim software has been developed by the PopSim Consortium using a
 distributed open source model, with strong procedures in place

### 2019-12-20 popSimManu.tex
some reworking of intro

@@
 \usepackage[utf8]{inputenc}
+
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+\rhead{\texttt{stdpopsim}}
+\lhead{PopSim Consortium}
+\fancyfoot[C]{\thepage}
 
@@
 
-The \stdpopsim software has been developed by the PopSim Consortium using a
+The \stdpopsim library has been developed by the PopSim Consortium using a
 distributed open source model, with strong procedures in place
@@
 as the default simulation engine. We plan to include other simulation
-engines to support scenarios that cannot be modelled under the coalescent framework,
+engines to support scenarios that cannot be modeled under the coalescent framework,
 and have already implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} as
@@
 
-At the time of writing, more models of human history have been implemented than for
+Currently, more models of human history have been implemented than for
 any other organism (see Table \ref{tab:catalog}). 
@@
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate human genomes under the \cite{ragsdale2019models} model and using the
-HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
+replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model and using the
+HapMapII\_GRCh37 genetic map~\citep{international2007second}. From top to bottom we show estimates for each
 of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
@@
 There is no single ``true'' reference for effective population size
-because of model misspecification -- the inference methods are fitting a single population model
+because of model misspecification---the inference methods are fitting a single population model
 to data simulated from mulitple populations.
@@
 Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp. \textbf{(A)} Data were generated by simulating
-replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
+replicate human genomes under the OutOfAfrica\_3G09 model and using the HapMapII\_GRCh37 genetic map
 inferred in \cite{international2007second}. \textbf{(B)} For \dadi and \fastsimcoal we show parameters inferred
@@
 time, for each of the inference methods, using samples drawn from African and European populations
-simulated under the \cite{gutenkunst2009inferring} model. Our results highlight many
+simulated under the OutOfAfrica\_3G09 model. Our results highlight many
 of the strengths and weaknesses of the different types of methods we used.
@@
 high effective recombination rates. For this reason, we chose to present only data
-from 3 Mb of chromosome 2R for simulations under OutOfAfrica\_2L06.
+from \SI{3}{\mega\byte} of chromosome 2R for simulations under OutOfAfrica\_2L06.
 For the generic IM simulations, we used the HomSap genome along with the

### 2019-12-20 popSimManu.tex
some reworking of intro
fixed typo

software->library

--- 2019-12-20 popSimManu.tex
@@
 realistic simulations for population genetic analysis a simple matter of
-choosing pre-implemented models from a community-maintained catalog, 
+choosing pre-implemented models from a community-maintained catalog.
 The \stdpopsim catalog currently contains three organisms: humans,

### 2019-12-20 popSimManu.tex
fixed typo

@@
 
-The \stdpopsim library has been developed by the PopSim Consortium using a
+The \stdpopsim software has been developed by the PopSim Consortium using a
 distributed open source model, with strong procedures in place

### 2019-12-20 popSimManu.tex
Typos and minor formatting.

--- 2019-12-20 popSimManu.tex
@@
 and empirical genetic map and a model of human migration out of Africa
-that includes archaic admiture \citep{ragsdale2019models}. In each column of this figure
+that includes archaic admixture \citep{ragsdale2019models}. In each column of this figure
 we show inferred $N(t)$ for the three extant populations in the model.

### 2019-12-20 popSimManu.tex
minor edits -- typos, grammar, clarity

--- 2019-12-20 popSimManu.tex
@@
 \emph{Drosophila~melanogaster}, and \emph{Arabidopsis~thaliana}. For each
-organism the catalog contains details on the physical organization (e.g., chromosome structure)
-of their genomes, one or more genetic maps, default population-level parameters (mutation rate,
+organism, the catalog contains details on the physical organization (e.g., chromosome structure)
+of its genome, one or more genetic maps, default population-level parameters (mutation rate,
 generation time) and one or more published demographic histories. Through
@@
 \section*{Results}
-\paragraph{The \stdpopsim library}
+\paragraph{The \stdpopsim library.}
 The first contribution of the PopSim consortium is \stdpopsim, a
@@
 A species definition consists of two key elements.  Firstly, the library defines
-some basic information about the species' genome, including information about chromosome
-lengths, average mutation rates and so on. We also provide access to detailed
+some basic information about each species' genome, including information about chromosome
+lengths, average mutation rates, and generation times. We also provide access to detailed
 empirical information such as genetic maps, which model observed
@@
 
-Currently, more models of human history have been implemented than for
-any other organism (see Table \ref{tab:catalog}). 
+Currently, {\em Homo sapiens} has the largest number of population models in
+\stdpopsim (see Table \ref{tab:catalog}).  
 These models include: a simplified version of the \cite{tennessen2012evolution}
@@
 an African sample (African3Epoch\_1S16), as well as the out-of-Africa divergence
-and associated bottleneck model of \cite{li2006inferring} which jointly models African
+and associated bottleneck model of \cite{li2006inferring}, which jointly models African
 and European populations (OutOfAfrica\_2L06). For \emph{A. thaliana}, we implemented the
@@
 %For \emph{E. coli} we have a simple, large constant population size model implemented.
-Beyond organism specific models \stdpopsim also includes a generic piecewise constant size model and
-isolation with migration (IM) model which can be used with any genome and genetic map combination.
+Beyond organism specific models \stdpopsim also includes a generic piecewise-constant-size model and
+an isolation-with-migration (IM) model, both of which can be used with any genome and genetic-map combination.
 
@@
 to simply and fairly compare popular demographic inference methods.
-Although we present comparison of results from several
+Although we present comparisons of results from several
 methods, our aim at this stage is not to provide an exhaustive
@@
 will enable more detailed exploration of the strengths and weaknesses of the numerous
-inference methods that are available to the population genetics community.
+inference methods that are available to the population genetics community
+(see Discussion).
 
@@
 $N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model.}
+for the rate of coalescence in the demographic model (see Appendix).}
 \label{fig:n_t_ragsdale}
@@
 % were, briefly, rather than saying who inferred them.
-In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
-run on whole genome simulations under under OutOfAfricaArchaicAdmixture\_5R19,
+In Figure \ref{fig:n_t_ragsdale}, we present results from our $N(t)$ analysis pipeline
+run on whole genome simulations under OutOfAfricaArchaicAdmixture\_5R19,
 a model of human migration out of Africa that includes archaic admixture 
@@
 because of model misspecification---the inference methods are fitting a single population model
-to data simulated from mulitple populations.
+to data simulated from multiple populations.  
 However, many methods work by matching coalescence time distributions,
@@
 (in which case coalescence rate is the inverse of the effective size).
-For this reason, solid black lines do not show historical census sizes,
+For this reason, we used as our ``ground-truth'' (solid black lines) 
+not historical census sizes,
 but rather inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix).
@@
 While there is variation in accuracy among methods, populations, and individual replicates,
-generally the methods are accurate for this model of human history.
+the methods are generally accurate for this model of human history.
 
-\stdpopsim allows us to readily compare these estimates to those drawn from a different
+\stdpopsim allows us to readily compare relative performance on this benchmark to
+that based on a different
 model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
@@
 
-\paragraph*{Multi-population demographic models}
+\paragraph*{Multi-population demographic models.}
 As \texttt{stdpopsim} implements multi-population demographic models, we also
@@
 For human models with more than two-populations (e.g. \cite{gutenkunst2009inferring})
-this means that we are inferring parameters for a model which itself does
+this means that we are inferring parameters for a model that does
 not match the model from which the data were generated (Figure

### 2019-12-20 popSimManu.tex
edits until the Methods, including ones from #112
finished edits

fixed supp fig referencing, closes #109

number of samples; closes #128

what is top; closes #129

responses to edits; closes #124

url fix

@@
         \renewcommand{\thetable}{S\arabic{table}}%
+        \renewcommand{\theHtable}{S\thetable}
         \setcounter{figure}{0}
         \renewcommand{\thefigure}{S\arabic{figure}}%
+        \renewcommand{\theHfigure}{S\thefigure}
      }
@@
         \renewcommand{\thetable}{\arabic{table}}%
+        \renewcommand{\theHtable}{X\thetable}
         \setcounter{figure}{0}
         \renewcommand{\thefigure}{\arabic{figure}}%
+        \renewcommand{\theHfigure}{X\thefigure}
      }
@@
 be measured. Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
-\Stdpopsim is a community-driven open source project, which provides a standard
+\Stdpopsim is a community-driven open source project, which provides easy access to a standard
 catalog of published simulation models from a wide range of organisms
@@
 and we encourage an even broader
-community of developers to join in the effort to standardize
-population genetic methods.}
+community of developers to contribute to this growing resource.
+}
 \date{}
 
-\keywords{Population genetics, Simulation, Inference}
+\keywords{Population genetics, Simulation, Inference, Reproducibility}
 
@@
 the degree of sophistication of the questions, models, data, and computational approaches
-used have all increased over the past decade. Currently there exist myriad computational methods
+used have all increased over the past two decades. Currently there exist myriad computational methods
 that can infer the histories of populations
@@
 the accuracy and quality of these inferences is becoming ever more important.
-%CK: this first paragraph is super long - might be helpful to break up
 
@@
 rigorously comparing it with competing methods.
-Moreover, this approach results in a great deal of duplicated effort across research
-groups, and contributes to decreased reproducibility and transparency.
+Moreover, this situation results in a great deal of duplicated effort,
+and contributes to decreased reproducibility and transparency across the entire field.
 It is also a barrier to entry to the field, because new researchers can
@@
 suggested large ancient ($>100,000$ years ago) ancestral population sizes and
-bottlenecks that have not been detected by methods based on allele frequency spectra
+bottlenecks that have not been detected by other methods based on allele frequency spectra
 \citep[see][]{beichman2017comparison}.
@@
 simulation output from their chosen model is returned.
-In this way, \stdpopsim will enable rigorous evaluation of population genetic software,
+In this way, \stdpopsim will lower the barrier to high-quality simulation for exploratory analyses,
+enable rigorous evaluation of population genetic software,
 and contribute to increased reliability of population genetic inferences.
@@
 hierarchical organization of the \stdpopsim catalog contains all model simulation information
-within individual species (expanded information shown here for \emph{H. sapiens} only). 
+within individual species (expanded information shown here for \emph{H.~sapiens} only). 
 Each species is associated with a representation of the physical genome, and one or more genetic maps and demographic models.
-Dotted lines indicate that only a subset of these categories is shown. \textbf{(B)} Example code to specify
-and simulate models using \stdpopsim via the API or \textbf{(C)} through a command line interface.}
+Dotted lines indicate that only a subset of these categories is shown. 
+At right we show example code to specify
+and simulate models using \textbf{(B)} the python API or \textbf{(C)} the command line interface.
+    }
 \label{fig:cartoon}
@@
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans,
-\emph{D.~melanogaster}, \emph{A.~thaliana}.
-% RNG: Abbreviate consistently with introduction
+on a catalog of species (Fig.~\ref{fig:cartoon}A), initially consisting of humans,
+\textit{D.~melanogaster}, and \textit{A.~thaliana}.
 A species definition consists of two key elements.  Firstly, the library defines
@@
 requested by the user, \stdpopsim will transparently download the map
-data into a local cache (\texttt{$\sim$/.local/cache/stdpopsim} by default
-on Unix platforms), where it can be quickly retrieved for subsequent
-simulations.
+data into a local cache,
+% (\texttt{$\sim$/.local/cache/stdpopsim} by default on Unix platforms)
+where it can be quickly retrieved for subsequent simulations.
 In the initial version of \stdpopsim we support
 the HapMapII~\citep{international2007second} and
-deCODE~\citep{kong2010fine} genetic maps in humans;
-the \cite{salome2011recombination} map in \emph{A.~thaliana};
-and the \cite{comeron2012many} map in \emph{D.~melanogaster}.
+deCODE~\citep{kong2010fine} genetic maps for humans;
+the \cite{salome2011recombination} map for \emph{A.~thaliana};
+and the \cite{comeron2012many} map for \emph{D.~melanogaster}.
+Adding futher maps to the library is trivial.
 The second key element of a species description
 within \stdpopsim is a set of carefully curated population genetic model
-descriptions from the literature, which are the basis for accurate simulations
-of specific historical scenarios that have influenced present-day patterns of
+descriptions from the literature, which allow simulation under
+specific historical scenarios that have been fit to present-day patterns of
 genetic variation. (See the Methods for a description of the community
@@
 \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
-as the default simulation engine. We plan to include other simulation
-engines to support scenarios that cannot be modeled under the coalescent framework,
-and have already implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} as
-an additional backend.
+as the default simulation engine.
+We have implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} 
+as an alternative backend,
+to allow simulation of processes that cannot be modeled under the coalescent.
 
@@
 resources used in published work, and encourage authors
-of simulation models to contribute to our ongoing community-driven development process.
+publishing demographic models to contribute to our ongoing community-driven development process.
 Together with the \stdpopsim version number and the long-term stable identifiers
@@
 
-The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}. At the time of
-writing, the PopSim Consortium has implemented previously published demographic models for
-humans, \emph{Drosophila}, and \emph{Arabidopsis}, 
-and we have models in development for \emph{Pongo abelii} and \emph{Escherichia coli}.
+The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}.
 These range from
@@
 admixture \cite[e.g.,][]{ragsdale2019models}.
+In addition to those models shown, at time or writing
+the PopSim Consortium has models in development for \textit{Pongo abelii} and \textit{Escherichia coli}.
 
@@
 and a constant chromosome-specific recombination rate.
-The times reported are for a single replicate run on an Intel i5-7600 CPU.
-These are not intended to be rigorous benchmarks, and the computing resources
-required will vary widely depending on sample sizes, chromosome length,
+The times reported are for a single run on an Intel i5-7600 CPU.
+Computing resources required will vary widely depending on sample sizes, chromosome length,
 recombination rates and other factors.
@@
 the European and Asian populations (OutOfAfrica\_3G09), the \cite{tennessen2012evolution} two-population variant of the
-Gutenkunst model which does not include Asian populations, but more explicitly models
+\citeauthor{gutenkunst2009inferring} model which does not include Asian populations, but more explicitly models
 recent rapid human population growth (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
@@
 Together these models
-contain features believed to have widespread impacts in real data (e.g. bottlenecks, population growth,
+contain features believed to have widespread impacts in real data (e.g., bottlenecks, population growth,
 admixture) and are therefore highly pertinent in the context of method development.
 
-For non-human genomes we have have implemented two demographic histories for
-\emph{D. melanogaster}, three from \emph{A. thaliana}, and have under development
-model for \emph{P. abelii} and \emph{E. coli}.
-For \emph{D. melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
+For other species we have have implemented two demographic histories for
+\textit{D.~melanogaster}, three from \textit{A.~thaliana}, and have under development
+model for \textit{P.~abelii} and \textit{E.~coli}.
+For \textit{D.~melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
 an African sample (African3Epoch\_1S16), as well as the out-of-Africa divergence
@@
 %For \emph{E. coli} we have a simple, large constant population size model implemented.
-Beyond organism specific models \stdpopsim also includes a generic piecewise-constant-size model and
-an isolation-with-migration (IM) model, both of which can be used with any genome and genetic-map combination.
+In addition to organism-specific models, \stdpopsim also includes a generic piecewise constant size model and
+isolation with migration (IM) model which can be used with any genome and genetic map.
 
-We note that we have put a good deal of effort into coming up with a standardized naming
+To guarantee reproducibility,
+we have standardized naming
 conventions for species, genetic maps, and demographic models that will enable long term stability
-of unique identifiers used throughout \stdpopsim. This is an essential feature to guarantee reproducibility.
+of unique identifiers used throughout \stdpopsim,
+as described in our documentation
+(\url{https://stdpopsim.readthedocs.io/en/latest/development.html#naming-conventions}).
+
 \subsection*{Use case: comparing methods of demographic inference}
+
 As an example of the utility of \stdpopsim, we demonstrate how it can be used
-to simply and fairly compare popular demographic inference methods.
-Although we present comparisons of results from several
+to easily and fairly compare popular demographic inference methods.
+Although we present comparison of results from several
 methods, our aim at this stage is not to provide an exhaustive
@@
 To reproducibly evaluate and compare the performance of inference methods, we developed
-\texttt{snakemake} \citep{koster2012snakemake} workflows,
-available from \url{https://github.com/popsim-consortium/analysis}.
-These workflows allow efficient computing in multicore or cluster environments.
+workflows using \texttt{snakemake} \citep{koster2012snakemake}
+that are available from \url{https://github.com/popsim-consortium/analysis},
+that allow efficient computing in multicore or cluster environments.
 
@@
  on simulated genomes sampled from a single population,
-in a number of the demographic models described above. The pipeline was generalized to
-run $R$ replicates with $C$ chromosomes for $n$ samples, for a total of $R \times C$
-simulations for each demographic model. After simulation, the pipeline prepares
-input files for each of the respective inference methods by grouping all
+in a number of the demographic models described above.
+Our workflow generates $R$ replicates of $C$ chromosomes, producing $n$ samples in each of a total of $R \times C$
+simulations for each demographic model. After simulation, the workflow prepares
+input files for each inference method by grouping all
 chromosomes, for each sample, into a single file. For each of the $R$ simulation replicates, this step results in an
@@
 of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model (see Appendix).}
+$N(t)$ trajectories for each replicate.
+As a proxy for the truth, in black we show inverse coalescence rates
+as calculated from the true demographic model (see text).
+}
 \label{fig:n_t_ragsdale}
@@
 
-% JK here --- I think it would really help if we described what the models
-% were, briefly, rather than saying who inferred them.
-In Figure \ref{fig:n_t_ragsdale}, we present results from our $N(t)$ analysis pipeline
-run on whole genome simulations under OutOfAfricaArchaicAdmixture\_5R19,
+Figure \ref{fig:n_t_ragsdale} presents the results 
+from simulations under OutOfAfricaArchaicAdmixture\_5R19,
 a model of human migration out of Africa that includes archaic admixture 
 \citep{ragsdale2019models}, along with an empirical genetic map. In each column of this figure
-we show inferred $N(t)$ for the three extant populations in the model.
-In each row we show comparisons among the methods (note that we compare two differing
-sample sizes for \MSMC).
+we show $N(t)$ inferred from samples taken from each of the three extant populations in the model.
+In each row we show comparisons among the methods (including two sample sizes for \MSMC).
+Blue lines show estimates from each of three replicate whole genome simulations.
 There is no single ``true'' reference for effective population size
@@
 but rather inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix).
-Blue lines show estimates from each of three replicate whole genome simulations.
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 of the methods is capturing relevant parts of the population history, although the
-discrepancy between inferred and simulated values varies with $t$. In comparing inferences between the
+accuracy varies across time. In comparing inferences between the
 models it is interesting to note that $N(t)$ estimates for the CHB and CEU
@@
 
-To see how well methods might do at recovering the population history of a constant-sized population,
-we can simulate under a human genome architecture and genetic map, but without demography.
+We can also
+see how well methods might do at recovering the population history of a constant-sized population,
+with human genome architecture and genetic map.
 We show results of such an
-experiment in Figure \ref{fig:n_t_HomSap_constant}. Again we see that all methods
-are recovering a modern population size within a factor of two of the truth, however
-SMC-based methods, perhaps due to their regularization, tend to learn sinusoidal
-patterns of population size even though no change is present.
+experiment in Figure \ref{fig:n_t_HomSap_constant}.
+All methods recover population size within a factor of two of the truth, however
+SMC-based methods, perhaps due to their regularization, tend to infer sinusoidal
+patterns of population size even though no such change is present.
 
@@
 \includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
-\caption{\textbf{Comparing estimates of $N(t)$ in \emph{Drosophila}}. Population
+\caption{\textbf{Comparing estimates of $N(t)$ in \textit{Drosophila}}. Population
 size over time ($N(t)$) estimated from an African population sample. Data were generated by simulating
-replicate \emph{D.melanogaster} genomes under the African3Epoch\_1S16 model
-with the genetic map inferred in \cite{comeron2012many}. In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model.}
+replicate \textit{D.~melanogaster} genomes under the African3Epoch\_1S16 model
+with the genetic map of \cite{comeron2012many}. In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. 
+As a proxy for the truth, in black we show inverse coalescence rates
+as calculated from the true demographic model (see text).
+}
 \label{fig:n_t_sheehan}
@@
 genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from the African3Epoch\_1S16
-model, originally estimated from an African sample of \emph{D. melanogaster} \citep{sheehan2016deep}.
-As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
+model, originally estimated from an African sample of \textit{D.~melanogaster} \citep{sheehan2016deep}.
+As with humans, we use \stdpopsim to simulate replicate \textit{D.~melanogaster} genomes under
 an empirically derived genetic map, and then try to infer back the parameters of the model.
@@
 In shades of blue we show $N(t)$ trajectories estimated from each simulation,
-and in black the inverse coalescence rates calculated from the true model (see text).
-The population split date, $T_{div}$, is shown at
+and in black census sizes for the respective population.
+The population split date, $T_{DIV}$, is shown at
 the bottom, with a common X-axis to the population size panels.}
@@
 \paragraph*{Multi-population demographic models.}
-As \texttt{stdpopsim} implements multi-population demographic models, we also
+As \stdpopsim implements multi-population demographic models, we also
 explored parameter estimation of population divergence parameters. In particular,
-we simulated data under multi-population models for humans and \emph{Drosophila}
+we simulated data under multi-population models for humans and \textit{D.~melanogaster}
 and then inferred  parameters using \dadi, \fastsimcoal, and \smcpp.
-For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an IM model
+For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
 with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
@@
 simple model is fit that may not reflect the true underlying demography.
-For human models with more than two-populations (e.g. \cite{gutenkunst2009inferring})
+For human models with more than two populations (e.g., \cite{gutenkunst2009inferring})
 this means that we are inferring parameters for a model that does
-not match the model from which the data were generated (Figure
+not match the model from which the data were generated (Figures
 \ref{fig:IM_popn_human}A and B). However, because the inferred models here better
-match the simulated models relative to the single population case, here we compare
-our inferred population sizes to the census size of the simulated models (black line in
+match the simulated models than in the single population case, here we compare
+our inferred population sizes directly to the census size of the simulated models (black line in
 Figure \ref{fig:IM_popn_human}C).
@@
 By contrast, \smcpp is much better at capturing the recent exponential
-growth in the CEU population, though the inferred population sizes are in general more
-noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
+growth in the CEU population, though the inferred population sizes are generally 
+noisier. In addition, the assumption of no migration by \smcpp leads to divergence time
 estimates that are consistently underestimated (Figure \ref{fig:IM_popn_human}C).
@@
 Again, we can compare between species and look at the performance of these methods in
-the context of a two-population model from \emph{D. melanogaster}. Figure
+on a two-population model of \textit{D.~melanogaster}. Figure
 \ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
@@
 split and bottleneck into a European population with no post-divergence migration.
-Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
-parameters, but ignore the brief population bottleneck in Europe. In addition, \dadi
+Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
+but ignore the brief population bottleneck in Europe. In addition, \dadi
 and \fastsimcoal both do reasonably well at correctly inferring the absence of
@@
 the \stdpopsim library. We have founded the Consortium with a number of specific goals in mind:
-standardization of simulation within the computational genetics community,
-increasing reproducibility and ease of use of complex simulation based results,
+standardization of simulation within the population genetics community,
+increased reproducibility and ease of use of complex simulations,
 community-based development and decision making guiding best practices in population genetics,
-and eventual benchmarking of inferential methods.
+and benchmarking of inference methods.
 
@@
 will prove to be an equally longstanding and valuable tradition in population genetics.
-
-% (ACS) I suggest deleting the following paragraph (commented out below).
-% All of the important points are made below (even with some of the same
-% phrases, such as "barriers to entry") and I think it is needlessly
-% redundant.
 
@@
 (Figure~\ref{fig:IM_popn_human}), most methods do reasonably well at
-reconstructing the YRI history, but struggle with the more complex CEU
-history, in large part because of the restriction of constant sizes per
-population.  An exception is \smcpp, which does not have the same
+reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
+history, in large part because of the restriction of constant population sizes.
+An exception is \smcpp, which does not have the same
 restrictions on its inferred history, and as a result does somewhat better
 with the CEU history but tends to overfit the YRI history.  The results for
-the two-population \textit{D.~melanogaster} case (Figure~\ref{fig:two_popn_fly})
+the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly})
 are generally similar.
@@
 future development.  In addition to other taxa,
-we hope to soon incorporate other common biological processes
+we plan to incorporate other common biological processes
 such as selection, gene conversion, and mutational heterogeneity.
@@
 population genomics community.
-
-%\adk{should we mention benchmarking of tree sequence inference here? what else?}
 
@@
 
-\subsection*{Model QC procedures}
+\subsection*{Model quality control}
 As a consortium we have agreed to a standardized procedure for model inclusion
@@
 demographic model for the relevant organism along with clear documentation
-of the model parameters and populations. This model is submitted as a Pull Request,
+of the model parameters and populations. This model is submitted as a ``pull request'',
 where it is evaluated by a reviewer and then included as `preliminary',
 but is not linked to the online documentation nor the command line interface.
-Inclusion of a preliminary model triggers a QC issue, and a second developer,
+Developer A submits a quality control (QC) issue, after which a second developer,
 Developer B, then independently reimplements the model from the relevant
@@
 QC implementation and the preliminary production model. If the two
-implementations are equal the original model is included in \stdpopsim.
+implementations are equivalent, the original model is included in \stdpopsim.
 If not, we move to an arbitration process whereby A and B first try
@@
 authors of the published model must be contacted
-to resolve ambiguities. This process can lead to cycles of revision of
-the implementation from Developers A and B but allows the project to assure
-quality. Indeed we have already had a number of interesting cases of arbitration
-to this point. Further details of our QC process can be found in our
-\href{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.
+to resolve ambiguities. 
+% This process can lead to cycles of revision of the implementation from Developers A and B but allows the project to assure quality.
+% Indeed we have already had a number of interesting cases of arbitration to this point. 
+Further details of our QC process can be found in our
+\url{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.
 
@@
 Our choice of \texttt{Snakemake} allows complete reproducibility of the
-analyses shown, and the entire Analysis repository is available from
-\url{https://github.com/popgensims/analysis}.
+analyses shown, and all code is available from \url{https://github.com/popsim-consortium/analysis}.
 
-We performed demographic analysis for one-and two-population tasks.
-Our one population task was to infer historical population size over
-time ($N(t)$). This was done using three software packages: \stairwayplot
+We performed two types of demographic inference.
+Our first task was to infer effective population size over time ($N(t)$).
+This was done using three software packages: \stairwayplot,
 which uses site frequency spectrum information only \citep{liu2015exploring};
-the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
-sample sizes $n\in (2 , 8)$; and \smcpp~\citep{terhorst2017robust},
+\MSMC~\citep{schiffels2014inferring},
+which is based on the sequentially Markovian coalescent (SMC),
+run with two different sample sizes ($n = 2 , 8$);
+and \smcpp~\citep{terhorst2017robust},
 which combines information from the site frequency spectrum with
@@
 to default parameters as possible. For \stairwayplot we
-set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
+set the parameters ``numRuns=1'' and ``dimFactor=5000''. For \smcpp we used the
 ``estimate" run mode to infer $N(t)$ with all other parameters set
@@
 a constant-sized generic model, and a two-epoch generic model where the
-population size instantaneously decreases from $N=10^4$ to $N=10^3$, 500
-generations before the present. Each HomSap simulation was run
-using the HapmapII\_GRCh37 genetic maps. For \emph{D. melanogaster}
+population size instantaneously decreased from $N=10^4$ to $N=10^3$
+five hundred generations before the present. Each HomSap simulation was run
+using the HapmapII\_GRCh37 genetic map. For \textit{D.~melanogaster}
 we estimated $N(t)$ from an African sample simulated under the DroMel,
@@
 % RNG: No results from these simulations are included in the paper
-%Finally we ran simulations of \emph{A. thaliana} genomes using the AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.
-For each model, three replicate whole genomes were
-simulated and the population size estimated from those data. In all cases we
-set the sample size of the focal population to $N=50$ chromosomes.
+%Finally we ran simulations of \textit{A.~thaliana} genomes using the AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.
+For each model, we estimated population size histories
+separately from each of three replicate simulations of whole genome datasets.
+In all cases we simulated 50 sampled chromosomes in the focal population.
 
@@
 used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
-Specifically we masked regions of \SI{1}{cM} or greater in the lowest 5th percentile of the empirical
+Specifically we masked all regions of \SI{1}{cM} or greater in the lowest 5th percentile of the empirical
 distribution of recombination, regions which are nearly uniformly absent for
@@
 
-Our goal for the two-population analysis pipeline was to explore two-population inference
-on some of the multi-population demographic models implemented in \stdpopsim.
-For HomSap we examined the OutOfAfrica\_3G09 model with the HapmapII\_GRCh37 genetic map,
-and for DroMel we simulated under the OutOfAfrica\_2L06 model with the Comeron2012\_dm6 map.
+Our second task was to explore inference with two-population models
+using some of the multi-population demographic models implemented in \stdpopsim.
+For HomSap we used the OutOfAfrica\_3G09 model with the HapmapII\_GRCh37 genetic map,
+and for DroMel we used the OutOfAfrica\_2L06 model with the Comeron2012\_dm6 map.
 The HomSap model is a three population model (Africa, Europe, and Asia) including post-divergence
@@
 generally with default settings and we did not attempt to optimize their performance or fit
-parameter-rich demographic models. As with the single population inference pipeline,
-we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
-our inference pipeline and maintain reproducibility.
+parameter-rich demographic models.
 
 For both \dadi and \fastsimcoal, we fit a two population
-isolation-with-migration (IM) model, with constant population sizes, to the simulated
-data from the HomSap and DroMel models. This IM model contains six parameters:
-the ancestral population size, the size of population 1 after the split, the size of
-population 2 after the split, the divergence time, and two migration rate parameters.
+isolation-with-migration (IM) model with constant population sizes.
+This IM model contains six parameters:
+the ancestral population size, the sizes of each population 1 after the split,
+the divergence time, and two migration rate parameters.
 Importantly, this meant that for both species, the
@@
 In the HomSap case, we therefore performed inference solely on the Africa
-and Europe populations, meaning that the Asia population functioned as a ghost
+and Europe populations, meaning that the Asia population functioned as a ``ghost''
 population that was ignored by our inference. Our motivation for fitting this simple
@@
 
-From HomSap we ran replicate whole genome simulations under the OutOfAfrica\_3G09 model
-with the HapmapII\_GRCh37 genetic map, taking 20 samples each from the Europe and Africa populations.
-For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes
-with the Comeron2012\_dm6 map due to the large effective population sizes leading to
-high effective recombination rates. For this reason, we chose to present only data
-from \SI{3}{\mega\byte} of chromosome 2R for simulations under OutOfAfrica\_2L06.
+From HomSap simulations we took 20 whole genome samples each from the Europe and Africa populations
+from each replicate.
+Runtimes of DroMel simulations were prohibitively slow when simulating whole genomes
+with the Comeron2012\_dm6 map due to large effective population sizes leading to
+high effective recombination rates. For this reason, we present only data
+from 20 samples of a \SI{3}{\mega\byte} region of chromosome 2R from simulations under OutOfAfrica\_2L06.
 For the generic IM simulations, we used the HomSap genome along with the
@@
 
-Following simulation, we outputted tree sequences and masked low-recombination
-regions using the same approach described for the single population pipeline above. We
+Following simulation, we output tree sequences and masked low-recombination
+regions using the same approach described for the single population workflow above. We
 converted tree sequences into a two-dimensional site frequency spectrum for all
@@
 Detailed settings for \dadi and \fastsimcoal can be found in the Snakefile
-on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
-The parameter estimates from the top run of each simulation replicate
+on our git repository (\url{https://github.com/popsim-consortium/analysis}).
+Estimates from the highest log-likelihood (out of 10 runs) for each simulation replicate
 are shown in Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C.
 
-For \smcpp, we converted the tree sequences into vcf format and performed inference
-with default settings. Importantly, \smcpp assumes no migration post-divergence, so
-here again the model fitted during inference does not exactly match the simulated
-human demography that does include migration. However, because \smcpp allows
+For \smcpp, we converted the tree sequences into VCF format and performed inference
+with default settings. Importantly, \smcpp assumes no migration post-divergence,
+deviating from the simulated model.  However, because \smcpp allows
 for continuous population size changes, it is better equipped to capture many of the
@@
 To visualize our results, we plotted the inferred population size trajectories
-for each simulation replicate along side the census population sizes
-(Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C). Here, unlike the single-population pipeline,
-we focus on census size rather than the inverse coalescence rate as the `true' population size.
+for each simulation replicate alongside census population sizes
+(Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C).
+Here, unlike the single-population workflow,
+we compare to census size rather than the inverse coalescence rate as the `true' population size.
 
@@
 \section*{Acknowledgments}
-We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible.
-Early on in the project we were helped by many people including XXXXXX.
+We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible,
+and the Simons Center for Quantitative Biology at Cold Spring Harbor Laboratory 
+for sponsoring the first workshop.
+Early on in the project we were encouraged by many people including Patrick Phillips, Richard Durbin, Philipp Messer, and Dmitri Petrov.
 CCK and KEL were funded under NIH Award R35GM119856.

### 2019-12-20 popSimManu.tex

--- 2019-12-20 popSimManu.tex
@@
 %For \emph{E. coli} we have a simple, large constant population size model implemented.
-Beyond organism specific models \stdpopsim also includes a generic piecewise-constant-size model and
+Beyond organism-specific models \stdpopsim also includes a generic piecewise-constant-size model and
 an isolation-with-migration (IM) model, both of which can be used with any genome and genetic-map combination.

### 2019-12-20 popSimManu.tex
fixed file name for figure
Changed display_items/homo_sapiens_two_popn_comp.pdf
to display_items/homo_sapiens_two_pop_comp.pdf to avoid compilation error

--- 2019-12-20 popSimManu.tex
@@
 \fancyfoot[C]{\thepage}
+
+\usepackage{etoolbox}
+\pretocmd{\abstractname}{\newpage}{}{}
 
@@
 genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from the African3Epoch\_1S16
-model, originally estimated from an African sample of \textit{D.~melanogaster} \citep{sheehan2016deep}.
-As with humans, we use \stdpopsim to simulate replicate \textit{D.~melanogaster} genomes under
-an empirically derived genetic map, and then try to infer back the parameters of the model.
+model, originally estimated from an African sample of \textit{D.~melanogaster} \citep{sheehan2016deep},
+and Figure \ref{fig:n_t_AraTha} shows estimates from simulations of \textit{A.~thaliana}
+under the African2Epoch\_1H18 model originally inferred by \cite{huber2018gene}.
+In both cases, as with humans, we use \stdpopsim to simulate replicate genomes using an empirically derived genetic map,
+and try to infer back parameters of the simulation model.
 Accuracy is mixed among methods in this setting and generally worse than what we
-observe under simulations of the human genome.
+observe for simulations of the human genome.
 
@@
 \begin{center}
-\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.pdf}
+\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_pop_comp.pdf}
 \caption{\textbf{Parameters estimated using a multi-population human model}.
@@
 \stairwayplot outperforms the sequence-based methods
-on simulations of \textit{D.~melanogaster} populations,
-in which linkage disequilibrium is considerably reduced (Figure~\ref{fig:n_t_sheehan}).
+on simulations of \textit{D.~melanogaster} or \textit{A.~thaliana} populations,
+in which linkage disequilibrium is reduced (Figures~\ref{fig:n_t_sheehan} \& \ref{fig:n_t_AraTha}).
 In simulations of two human populations
@@
 African3Epoch\_1S16 model using the Comeron2012\_dm6 map.
-% RNG: No results from these simulations are included in the paper
-%Finally we ran simulations of \textit{A.~thaliana} genomes using the AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.
-For each model, we estimated population size histories
-separately from each of three replicate simulations of whole genome datasets.
-In all cases we simulated 50 sampled chromosomes in the focal population.
+Finally we ran simulations of \emph{A. thaliana} genomes using the 
+AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.
+For each model, three replicate whole genomes were
+simulated and the population size estimated from those data. In all cases we
+set the sample size of the focal population to $N=50$ chromosomes.
 
@@
 for sponsoring the first workshop.
-Early on in the project we were encouraged by many people including Patrick Phillips, Richard Durbin, Philipp Messer, and Dmitri Petrov.
+Early on in the project we were encouraged by many people including Patrick Phillips, Richard Durbin,
+Dmitri Petrov, and Sohini Ramachandran.
 CCK and KEL were funded under NIH Award R35GM119856.
@@
 \label{fig:n_t_HomSap_constant}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/AraTha_African2Epoch_1H18.png}
+\caption{\textbf{Comparing estimates of $N(t)$ in \textit{A.~thaliana}}. Here we show estimates of population
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairway plot},
+\MSMC with $n=2$ and $n=8$. Data were generated by simulating
+replicate \textit{A.~thaliana} genomes under the African2Epoch\_1H18 model and using
+the genetic map of \cite{salome2011recombination}. In black we show the true population size history
+of the model.}
+\label{fig:n_t_AraTha}
 \end{center}

### 2019-12-20 popSimManu.tex
capitalized stdpopsim when starting a sentence

@@
 
-\stdpopsim allows us to readily compare relative performance on this benchmark to
+\Stdpopsim allows us to readily compare relative performance on this benchmark to
 that based on a different

### 2019-12-20 popSimManu.tex
final touches and added resource availability

@@
 
-For other species we have have implemented two demographic histories for
-\textit{D.~melanogaster}, three from \textit{A.~thaliana}, and have under development
-model for \textit{P.~abelii} and \textit{E.~coli}.
+Beyond humans we have have implemented two demographic histories for
+\textit{D.~melanogaster}, three from \textit{A.~thaliana}, and are currently developing
+models for \textit{P.~abelii} and \textit{E.~coli}.
 For \textit{D.~melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 has yet to coalesce around a set of standards for the crucial task of method
-evaluation, which in our field hinges on simulation. In contrast, other fields such as
+evaluation, which in our discipline hinges on simulation. In contrast, other fields such as
 structural biology \citep{moult1995large} and machine learning \citep{russakovsky2015imagenet} have a long track record
@@
 high effective recombination rates. For this reason, we present only data
-from 20 samples of a \SI{3}{\mega\byte} region of chromosome 2R from simulations under OutOfAfrica\_2L06.
+from 50 samples of a \SI{3}{\mega\byte} region of chromosome 2R from simulations under OutOfAfrica\_2L06.
 For the generic IM simulations, we used the HomSap genome along with the
@@
 we compare to census size rather than the inverse coalescence rate as the `true' population size.
+
+\section*{Resource availability}
+The version 0.1 release of \stdpopsim is available for download on Github: 
+\url{https://github.com/popsim-consortium/stdpopsim/releases}. 
+Documentation for the project can be found here: 
+\url{https://stdpopsim.readthedocs.io/en/latest/}.
 

### 2019-12-20 popSimManu.tex
added figure scripts, minor edits

@@
 
-Again, we can compare between species and look at the performance of these methods in
-on a two-population model of \textit{D.~melanogaster}. Figure
+Again, we can compare between species and look at the performance of these methods 
+for a two-population model of \textit{D.~melanogaster}. Figure
 \ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
@@
 % Indeed we have already had a number of interesting cases of arbitration to this point. 
-Further details of our QC process can be found in our
-\url{https://stdpopsim.readthedocs.io/en/latest/development.html#}{developer documentation}.
+Further details of our QC process can be found in our developer documentation 
+(\url{https://stdpopsim.readthedocs.io/en/latest/development.html#}).
 
@@
 African3Epoch\_1S16 model using the Comeron2012\_dm6 map.
-Finally we ran simulations of \emph{A. thaliana} genomes using the 
+Finally, we ran simulations of \emph{A. thaliana} genomes using the 
 AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.

### 2019-12-20 src/plotResults_3population_4method.R

@@ -1,280 +1,0 @@
-library(ggplot2)
-library(cowplot)
-
-## define variables
-infile_pop1="/n_t/homo_sapiens_mask_Ragsdale_population1/Results/all_estimated_Ne.txt"
-infile_pop2="/n_t/homo_sapiens_mask_Ragsdale_population2/Results/all_estimated_Ne.txt"
-infile_pop3="/n_t/homo_sapiens_mask_Ragsdale_population3/Results/all_estimated_Ne.txt"
-outfile="/n_t/n_t_plot_Ragsdale.pdf"
-species="Homo sapiens"
-pop1_name="YRI"
-pop2_name="CEU"
-pop3_name="CHB"
-method1_name="SMC++"
-method2_name="stairway plot"
-method3_name="MSMC (n = 2)"
-method4_name="MSMC (n = 8)"
-method1_id="smcpp"
-method2_id="sp"
-method3_id="msmc_2"
-method4_id="msmc_8"
-replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
-truth_color="gray30"
-
-## plot
-cPal_legend<-c(truth_color, replicate_colors)
-
-tab0 <- read.table(file=infile_pop1, h=T ,sep="\t")
-tab1 <- read.table(file=infile_pop2, h=T ,sep="\t")
-tab2 <- read.table(file=infile_pop3, h=T ,sep="\t")
-
-tab0$x[tab0$x==0.0]<-1
-tab1$x[tab1$x==0.0]<-1
-tab2$x[tab2$x==0.0]<-1
-
-xMin = 10
-xMax = max(c(tab0$x,tab1$x,tab2$x))
-y0Min = min(tab0$y)
-y0Max = max(tab0$y)
-y1Min = min(tab1$y)
-y1Max = max(tab1$y)
-y2Min = min(tab2$y)
-y2Max = max(tab2$y)
-
-tab0A = subset(tab0,method == method1_id)
-tab0B = subset(tab0,method == method2_id)
-tab0C = subset(tab0,method == method3_id)
-tab0D = subset(tab0,method == method4_id)
-tab0Z = subset(tab0,method == "coal")
-
-tab1A = subset(tab1,method == method1_id)
-tab1B = subset(tab1,method == method2_id)
-tab1C = subset(tab1,method == method3_id)
-tab1D = subset(tab1,method == method4_id)
-tab1Z = subset(tab1,method == "coal")
-
[…]

### 2019-12-20 src/plot_AraTha_African2Epoch_1H18.R

@@ -1,0 +1,112 @@
+library(ggplot2)
+library(cowplot)
+
+## define variables
+infile="../data/AraTha_African2Epoch_1H18.txt"
+outfile="../display_items/AraTha_African2Epoch_1H18.pdf"
+species="Arabidopsis thaliana"
+pop0_name="Africa"
+method1_name="SMC++"
+method2_name="stairway plot"
+method3_name="MSMC (n = 2)"
+method4_name="MSMC (n = 8)"
+method1_id="smcpp"
+method2_id="sp"
+method3_id="msmc_2"
+method4_id="msmc_8"
+truth_color="gray30"
+replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
+
+## plot
+cPal_legend<-c(truth_color, replicate_colors)
+
+tab0 <- read.table(file=infile, h=T ,sep="\t")
+
+tab0$x[tab0$x==0.0]<-1
+
+xMin = 10
+xMax = max(tab0$x)
+yMin = min(tab0$y)
+yMax = max(tab0$y)
+
+tab0A = subset(tab0,method == method1_id)
+tab0B = subset(tab0,method == method2_id)
+tab0C = subset(tab0,method == method3_id)
+tab0D = subset(tab0,method == method4_id)
+tab0Z = subset(tab0,method == "coal")
+
+speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
+
+p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
+  coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
+  geom_step(data=tab0Z,aes(x=x, y=y),size=2.5,alpha=1.0, col=truth_color) +
+  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
+  scale_x_continuous(trans='log10') +
+  scale_y_continuous(trans='log10') +
+  annotation_logticks(sides="b") +
+  annotation_logticks(sides="l") +
+  scale_color_manual(values=replicate_colors) +
+  guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
+  theme_classic(base_size = 18) +
+  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
+  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
+
+
[…]

### 2019-12-20 src/plot_homo_sapiens_constant.R

@@ -1,0 +1,112 @@
+library(ggplot2)
+library(cowplot)
+
+## define variables
+infile="../data/homo_sapiens_constant.txt"
+outfile="../display_items/homo_sapiens_constant.pdf"
+species="Homo sapiens"
+pop0_name="constant size"
+method1_name="SMC++"
+method2_name="stairway plot"
+method3_name="MSMC (n = 2)"
+method4_name="MSMC (n = 8)"
+method1_id="smcpp"
+method2_id="sp"
+method3_id="msmc_2"
+method4_id="msmc_8"
+truth_color="gray30"
+replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
+
+## plot
+cPal_legend<-c(truth_color, replicate_colors)
+
+tab0 <- read.table(file=infile, h=T ,sep="\t")
+
+tab0$x[tab0$x==0.0]<-1
+
+xMin = 10
+xMax = max(tab0$x)
+yMin = min(tab0$y)
+yMax = max(tab0$y)
+
+tab0A = subset(tab0,method == method1_id)
+tab0B = subset(tab0,method == method2_id)
+tab0C = subset(tab0,method == method3_id)
+tab0D = subset(tab0,method == method4_id)
+tab0Z = subset(tab0,method == "coal")
+
+speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
+
+p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
+  coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
+  geom_step(data=tab0Z,aes(x=x, y=y),size=2.5,alpha=1.0, col=truth_color) +
+  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
+  scale_x_continuous(trans='log10') +
+  scale_y_continuous(trans='log10') +
+  annotation_logticks(sides="b") +
+  annotation_logticks(sides="l") +
+  scale_color_manual(values=replicate_colors) +
+  guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
+  theme_classic(base_size = 18) +
+  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
+  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
+
+
[…]

### 2019-12-22 popSimManu.tex

@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/n_t_dag.pdf}
+\caption{\textbf{Workflow for our N(t) inference methods comparison.}
+Here we show single replicate for two chromosomes, chr22 and chrX, simulated under the HomSap 
+OutOfAfrica\_3G09 demographic model, with a HapmapII\_GRCh37 genetic map.
+Note that resulting data for input to all inference methods \smcpp, \MSMC, and \stairwayplot,
+come from the same set of simulations.}
+\label{fig:n_t_dag}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/two_pop_dag.pdf}
+\caption{\textbf{Workflow for our two population N(t) inference methods comparison}
+Here we show single replicate for one chromosome, chr22, simulated under the HomSap 
+OutOfAfrica\_3G09 demographic model, with a HapmapII\_GRCh37 recombination map.
+Note that resulting data for input to all inference methods \dadi, \smcpp,
+and \fastsimcoal,
+derive from the same set of simulations.}
+\label{fig:two_pop_dag}
+\end{center}
+\end{figure}
+
 \stopsupplement

### 2020-01-21 popSimManu.tex
typo in Kirk's name!!
minor fixes for first full submission to eLife

@@
 \author[12]{Arun Durvasula}
-\author[13]{Bernard Y. Kim}
-\author[14]{Patrick McKenzie}
-\author[15]{Philipp W. Messer}
-\author[16]{Ekaterina Noskova}
-\author[17]{Diego Ortega-Del Vecchyo}
+\author[13]{Ilan Gronau}
+\author[14]{Bernard Y. Kim}
+\author[15]{Patrick McKenzie}
+\author[16]{Philipp W. Messer}
+\author[17]{Ekaterina Noskova}
+\author[18]{Diego Ortega-Del Vecchyo}
 \author[5]{Fernando Racimo}
-\author[18]{Travis J. Struck}
+\author[19]{Travis J. Struck}
 %Senior Authors
 \author[7,$\dagger$]{Simon Gravel}
-\author[18,$\dagger$]{Ryan N. Gutenkunst}
-\author[6,$\dagger$]{Kirk E. Lohmeuller}
+\author[19,$\dagger$]{Ryan N. Gutenkunst}
+\author[6,$\dagger$]{Kirk E. Lohmueller}
 \author[1,$\dagger$]{Peter L. Ralph}
@@
 \author[3,$\dagger$]{Adam Siepel}
-\author[19,$\dagger$,$\aleph$]{Jerome Kelleher}
+\author[20,$\dagger$,$\aleph$]{Jerome Kelleher}
 \author[1,$\dagger$,$\aleph$]{Andrew D. Kern}
@@
 \affil[12]{Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles}
-\affil[13]{Department of Biology, Stanford University}
-\affil[14]{Department of Ecology, Evolution, and Environmental Biology, Columbia University}
-\affil[15]{Department of Computational Biology, Cornell University}
-\affil[16]{Computer Technologies Laboratory, ITMO University}
-\affil[17]{International Laboratory for Human Genome Research, National Autonomous University of Mexico}
-\affil[18]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[19]{Big Data Institute, Li Ka Shing Centre for Health Information and
+\affil[13]{The Efi Arazi School of Computer Science, The Herzliya Interdisciplinary Center, Israel}
+\affil[14]{Department of Biology, Stanford University}
+\affil[15]{Department of Ecology, Evolution, and Environmental Biology, Columbia University}
+\affil[16]{Department of Computational Biology, Cornell University}
+\affil[17]{Computer Technologies Laboratory, ITMO University}
+\affil[18]{International Laboratory for Human Genome Research, National Autonomous University of Mexico}
+\affil[19]{Department of Molecular and Cellular Biology, University of Arizona}
+\affil[20]{Big Data Institute, Li Ka Shing Centre for Health Information and
 Discovery, University of Oxford}

### 2020-03-17 response-to-reviewers.md
Add outline for response to reviewers.
Closes #151

Dear Graham,

[Introductory text]

Please see below a point-by-point response to the comments.

----

> # Editors comments
> 1) Ensuring that the language about the models & parameters conveys the correct sense of uncertainty. Obviously the reviewers and the authors know that these just represent best guesses at the moment, but as this platform catches on we don't want these numbers & models to be seen as gospel. 
>

FIXME: An example response just to illustrate how to respond to comments in this Markdown file. 
This will definitely need to be removed before we send the actual response!

> 2) We would like to see the SLIM integration demonstrated with an application that MsPrime doesn't cover. This could be, for example, a figure of the average value some summary statistics surrounding a selective under the human demographic model. The reviewers didn't want this to be a lot of work, e.g. it would be outside of the scope of the current paper to demonstrate the power of a range of selection methods. However, we felt that a simple demonstration of the functionality would substantially increase the scenarios easily available and imaginable by readers of the paper. 
> 
> 3) We felt that the discussion would benefit from a discussion of the details not yet incorporated into the platform, e.g. sex-specific recombination, gene conversion, assembly error, structural variants etc. Such an addition might spur future development, and also give the authors an opportunity to discuss general ways forward on these issues. 
> 
> 4) We would be interested to see more of a discussion of the strengths and weakness of the various demographic inference methods. 
> 
> Beyond testing methods, another major use axis that I can personally see for this platform is for empirical researchers to see whether patterns they see in their data are consistent with known demographic models. For example, demonstrating whether the frequency spectrum in larger samples lines up with previously inferred models. Having all of these models implemented in a central place will significantly lower the barrier of entry of empirical researchers into rigorous simulation frameworks. 
> 
> Overall we are very excited to see this big move forward and welcome the authors' careful work in this important area. 
>

-----

# Reviewer 1

> 
> The manuscript describes a community effort to standardize population genetic simulations and it presents an example of the resource's utility for method's testing. This development answers a long-standing need in the population genetics community for greater standardization and ease-of-implementation of simulation protocols. As such, I expect it will be a well-used resource and it represents an important advance for the field's practices. From the test implementation included in the paper, it was remarkable to see how highly variable (and often poor) the performance of the demographic inference methods was, and those results add to the scientific contribution of this manuscript. 
> 
> Some of the specific features that are nice about the work are the incorporation of inferred genetic maps where possible, the automated output of citation information, the use of long-term stable identifiers for models and genetic maps, and the outputting of provenance information in the tree sequence files. 
> 
> 
> ## Major comments
> 
> 1. Emphasize parameters are current best-guesses: 
> a. I am most worried about the fragility of the models over time and the misperception one might have that these models are "accurate" for a species. Genome assemblies change, recombination rates improve, mutation rates change, etc. I would like to see that uncertainty reflected more in the language used in the paper so that it's clear the catalog is a collection of inferred parameters that are subject to change over time. This is a subtle and cosmetic, but important I think. For example, page 4 "the library defines basic information about each species genome, including information about chromosome lengths, average mutation rates, and generation times." Yet, this is information we don't have - each of these parameters is actively under revision/discussion for even the best studied species. It would be great to hammer home these are all inferred values. (Also see page 3 sentence on "details on =the= physical organization of the genome"; the same comment applies, and an improvement might be something like "details on the physical organization of the latest reference genome assembled for the species"). 
> b. On a related note, there is wording on page 4, regarding ensuring "implemented models are accurate" - I think what is meant is that the "implemented models are faithful to the source publications from which they derive". As the second half of this paper makes clear, because of errors in inference, many of these models will not be accurate, in the sense of representing the true history of the species. 
> 2. It would be wonderful to have a comments section for the models that would be either curated by a set of editors or crowd-sourced. I say this because overtime, models will proliferate, and some will come to be regarded as out-of-date. One can imagine those approaching the field afresh will be overwhelmed by the possible selections and possible implement models that become outdated. If the goal is standardization, how do we communicate that standards change? A comment system (or even star-rating system?) may be wise to implement now. Assuming it can be layered on top of the existing structure, it may be enough for this publication to note this as a future challenge that needs development/addressing. 
> 3. In terms of the maturity of the examples developed for the initial release, I would have liked to see at least one simulation model with a selective sweep, one with background selection, and one with spatial stepping-stone structure. Each of these would be helpful test cases to implement to be sure that the existing catalog framework has the breadth/flexibility necessary to accommodate future use cases. I do not think this is a requirement for publication, but it would add great value to this initial release of the resource. 
> 4. Page 15. The approach of masking "low-recombination" portions of the chromosomes seems like an incomplete/indirect attempt to model the inherent limitations of sequencing to an "accessible" genome. 
> a. Shouldn't the approach instead be to drop "low complexity" regions (e.g., as defined by an excessive number of "N"'s in the reference, low mapability scores, or via tools like RepeatMasker?). This part of the pipeline seems open to refinement. 
> b. Are the "masks" a separate configuration file for the simulations? It seems that it would be preferable for them to be separate from the recombination rate files - right now it reads as if the mask applied is a function of the genetic map file, but this seems too inflexible for users who prefer an alternative approach to masking. 
> 
> 
> ## Minor comments
> 
> 1. Is the provenance information only provided if msprime is used to produce tree sequence files? Is a similar functionality in place with the SLiM back-end? 
> 
> 2. Little detail is given on the system of identifiers/versioning for models, maps, etc. 
> 
> 3. In the abstract, it's left vague what the nature of the resource is. 
> 
> a. The resource is described as an open source project and a catalog, but the fact that it's a web repository that contains a set of configuration files might be helpful to say explicitly. 
> b. Also, while I wouldn't normally include url to software in an abstract, in this special case it seems like it could be very helpful to readers who want to jump right to the resource quickly? 
> 
> 4. Page 3: "bespoke simulations": 
> 
> a. "bespoke" struck me as a distracting word choice, given its contemporary associations. Upon checking a dictionary, I see how it is an appropriate and precise word choice, but maybe it's best to instead simply say: "custom" or "one-off". 
> b. To elaborate on the point you're making here, you could also share how the general modus operandi often includes the pasting of long unwieldy command-line calls for generating simulations into methods sections. Many of these command-line values are buried across the literature, often in supplementary methods. Additional virtues of your catalog is having: 1) a central repository for such models, and 2) modular configuration files (e.g. I can keep the same demographic model, but change the recombination map, without having to re-write, or learn, the innards of a complex command-line for ms/msprime) 
> 
> 5. Page 5, "accurate, standardized simulations": related to the point above about "accurate" models - I wonder if a different word should be used here, or even omit "accurate". The word doesn't add much value over simply saying the models are standardized - the standardization includes that the model implementation is quality controlled to match the original source publication, which I think is what is meant by accuracy here? 
> 
> 6. Page 6. Typo in "time or writing". Change to "time of writing" 
> 
> 7. Page 8. I like the approach of showing inverse coalescent rates, but it would be instructive to practitioners and useful for teaching, if in all figures you showed both the census sizes (e.g. with light grey dashed lines) and inverse rates (e.g. with dark grey solid lines as now in Fig 2). 
> 
> 8. Figure 4. 
> a. Confusing how in panels A and B time moves left to right, and in panel C time moves right to left. 
> b. Also, the figure would be improved by showing with vertical lines the timing of the CEU/YRI divergence in the true model. (similar comments hold for Figures S4 and S5) 
> 
> 9. Page 16 "checked for convergence" : Wording - sounds like checking for convergence between the methods - I think you mean convergence by each method to an optimum in their respective objective functions. 
> 
> 10. Page 31. This is all very nice. To save readers a step, could you include the command-line flag for how to have msprime output the coalescent rate function? 
>

# Reviewer #2

> The authors in this manuscript describe the implementation of a publicly curated, open source simulation package called stdpopsim - equipped with commonly utilized population genetic demographic models in humans, Drosophila melanogaster, and Arabidopsis thaliana. I am in awe of what these authors have achieved, in terms of benchmarking these standard models in an effort to avoid duplication of effort, and possibility of erroneous inference. The package is currently equipped with several "in-built" models that allow the simulation of trees with msprime and SLiM. The authors explain the application of these models by simulating and benchmarking estimates of Ne under a couple of scenarios. The manuscript is also well written, and use popular tools like dadi and smc++ to estimate and benchmark the simulations under a variety of models. Across all simulated scenarios (except under more complex models), the simulations seem relatively accurate. 
> 
> Despite a little hiccup with python version mismatch prior to me successfully installing stdpopsim, I was able to successfully get the tutorials running within minutes after. I have one recommendation however for the tutorials - it would definitely help if the CLI versus python tutorials were kept separate. I found it a little confusing since they are all listed on the same page (https://stdpopsim.readthedocs.io/en/latest/tutorial.html). The simulations, testing models, calculating divergences, plotting ran without a hitch, and I am impressed and excited to play around with more models in coming days. Having also developed similar libraries/pipelines, I have also found it extremely useful for developers to provide some more detailed documentation/tutorials via Jupyter notebooks, or some similar platform. I did however notice that the authors have provided their analyses as Snakemake files in the interest of replication. I did not replicate their analyses, but I trust that the documentation for these analyses are detailed enough to aid readers/users in establishing similar analysis pipelines for stdpopsim simulated data. 
> 
> I thoroughly enjoyed reading this manuscript, and learning of all the new features that have been written into stdpopsim, and believe that this will be an invaluable contribution to the population genomics community. 
> 
> ## Minor comments: 
> 
> The authors mention the implementation of the SLiM package into stdpopsim, but I was unable to find any details of this either in the manuscript or on the readthedocs page. If this was indeed implemented, I would also like to see the authors add some tutorials/results of analyses based on SLiM simulated data. 
> 
> "Adding further maps to the library is trivial" - discuss - since the audience for this is the broader pop-gen community, perhaps pointing them to the detailed development environment on readthedocs would be helpful. 
> 
> Fig S6: The figure captions/labels are acronyms, which aren't explained. Would help readers to understand what this plot is estimating. 
>

# Reviewer #3

> 
> ## Summary 
> 
> 
> The authors present a standardized framework for creating reproducible population genetic simulations. This resource will allow researchers to create models for new species/scenarios, and easily compare methods on the same dataset. The authors are correct that the current state of benchmarking in population genetics causes inconsistency, duplicated effort, and confusion about the performance of different methods. The authors highlight that creating a realistic and meaningful simulation study is a barrier to entering this field, and I absolutely agree. I will be passing this resource on to my students and look forward to using it myself. Stdpopsim is a crucial step forward. The manuscript itself is well-written and describes an extensive comparison of demography methods in a variety of species/scenarios. I have a few comments and some typos, etc. 
> 
> 
> ## Comments 
> 
> 1) The authors mention that SLiM can be used as an alternative backend, which would presumably allow for simulations with selection. Although I don't think an extensive comparison of selection methods is necessary for this paper, it would be ideal if the authors can give some idea of how this would work (example command line, etc). There are also a myriad of methods for detecting/quantifying selection, and these simulations are not consistent either. 
> 
> 2) I like the inclusion of the "zigzag" history, as well as generic piecewise constant models and IM models (page 7). I wonder if these could be included in a separate section (not organism specific) in the documentation and software (and then in Table 1). Right now the zigzag model is under humans in the catalog. 
> 
> 3) On page 8 the authors set up notation for the number of replicates (R), number of chromosomes (C), and sample size (n), but don't seem to use it afterward (or use it inconsistently). It would be helpful if all the figure captions and main text included this notation (I am guessing the number of replicates is 3 based on the images, but this should be clarified). The authors use N in the Methods (i.e. page 15) to refer to population size (which makes sense), but then also say "In all cases we set the sample size of the focal population to N = 50 chromosomes." For MSMC, the sample size was set to n=2,8 which suggests haploid samples, but the "Calculating coalescence rates" section says that n is the diploid sample size. 
> 
> 4) "Calculating coalescence rates" section needs a read through. Reword first sentence and add some citations (especially regarding computing p(t) and p(z,t)). It was unclear to me how the "mean coalescence times" were used (the rate was used to compute the ground truth over time). This section is also referred to as the Appendix in the main text. 
> 
> 
> ## Minor Comments 
> 
> Typos, Wording, etc 
> 
> For a future draft, line numbers would be very helpful. 
> 
> -Abstract: "possibility for error" -> inconsistency 
> 
> -Page 5: trivial -> straightforward 
> 
> -"tree sequence": make it clear that users can easily obtain traditional variant matrices from this format, in case they are unfamiliar
> 
> -Page 6: "time or writing" -> of 
> 
> -Page 8: "While there is variation in accuracy among methods, populations, and individual replicates, the methods are generally accurate for this model of human history." I'm not sure I agree here (although perhaps relative to other models this statement is true). Discuss a bit more. 
> 
> -Page 8, paragraph "Stdpopsim allows...": reword this first sentence 
> 
> -Figure 3 caption: wrong citation, should be Sheehan and Song 
> 
> -Page 10, last sentence before "Multi-population demographic models" section: clarify what "this setting" is (both species I assume) 
> 
> -Figure 4 captain: "or smc++" -> "and smc++" (some other figure captions have this too), I would also say "$x$-axis" 
> 
> -Page 12, first paragraph: include a reference where people could learn more about the effective population size vs. census size, random-mating models, etc 
> 
> -Page 12 (and elsewhere): "By contrast" -> "In contrast"(?) I think both are technically correct though 
> 
> -Page 12: sentence beginning with "Figure S4 shows..." is long, perhaps breakup or rephrase ("bottleneck into a European population" sounds odd to me.) 
> 
> -Page 12: "Europe population" -> "European population" 
> 
> -Page 12: S7, S4, S6 are mentioned in a different order than they appear in the supplement (I understand why but maybe revisit this part) 
> 
> -Page 13: "overfit the YRI history" do you mean overfit in a technical sense? 
> 
> -Page 14: "Method quality control" section. Could you say a bit more about how Developer B would be identified? Someone in the consortium or someone else in the community? What kind of leadership structures are in place? (Maybe the goal is to make this a bit more distributed?) And just to be clear, Developer B would not have access to Developer A's implementation of the model while they were independently implementing it, right? 
> 
> -Page 16, line 6: "population 1 after"? 
> 
> -Page 16: S5 -> Figure S5 
> 
> -Page 17: "with making" -> "for making"? 
> 
> -Figure S2 caption: move "and" to before MSMC (in a few other figure captions too) 
> 
> -Figure S6: explain x-axis "MIG_AF_EU" and "MIG_EU_AF". I assume these are directional migration rates 
> 
> -Figure S9: unreadable when printed (not sure it adds a lot) 
>

### 2020-03-30 response-to-reviewers.md

@@
 I did not replicate their analyses, but I trust that the documentation for these analyses are detailed enough to aid readers/users in establishing similar analysis pipelines for stdpopsim simulated data.
+
+We have added in a basic `stdpopsim` API and CLI example in a Jupyter Notebook that can be accessed and used interactively via Binder and encourage users to try out the tutorials there.
+We have linked to the Binder in the README on the GitHub.
+See https://mybinder.org/v2/gh/popsim-consortium/stdpopsim/master?filepath=stdpopsim_example.ipynb
+
 > 

### 2020-04-01 response-to-reviewers.md
Response to request for dynamic catalog.

@@
 Assuming it can be layered on top of the existing structure, it may be enough for this publication to note this as a future challenge that needs development/addressing.
+
+This is an excellent idea and would be a great resource. Unfortunately, implementing something sufficiently dynamic 
+with good tools for curation and moderation of crowd-sourced input is a major task. The current automated rendering of the 
+catalog HTML from source was quite challenging and it is fundamentally static. Having dynamic user input would require
+a different approach which we don't have sufficient developer bandwidth for. However, this could be the basis of a 
+something like a Google Summer of Code student project.
+
 > 3.

### 2020-04-01 response-to-reviewers.md
Expand discussion of strengths and weaknesses of inference methods.

--- 2020-03-30 response-to-reviewers.md
@@
 > 4) We would be interested to see more of a discussion of the strengths and weakness of the various demographic inference methods.
-> 
+
+We have added an additional paragraph to the discussion exploring some of the additional factors that may affect choice of inference method, beyond results like those presented here.
+These include data required, type of model, and software engineering.
+A comprehensive investigation of these issues is beyond the scope of the current manuscript, with focuses on the resource itself, rather than its application.
+
 > Beyond testing methods, another major use axis that I can personally see for this platform is for empirical researchers to see whether patterns they see in their data are consistent with known demographic models.

### 2020-04-01 popSimManu.tex
Expand discussion of strengths and weaknesses of inference methods.

@@
 restrictions on its inferred history, and as a result does somewhat better
-with the CEU history but tends to overfit the YRI history.  The results for
+with the CEU history but tends to overfit the YRI history.
+The results for
 the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly})
 are generally similar.
+In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, differing only in how they calculate model expectations and optimize parameters.
+
+All methods for inferring demographic history have strengths and weaknesses (as recently reviewed by \citet{beichman2018review}.)
+We compared inferences from simulated whole genome data, but many factors affect choice of methodology.
+Markovian coalescent methods (\MSMC and \smcpp) require long contiguous stretches of sequence data.
+By contrast, frequency spectrum methods (\stairwayplot, \dadi, and \fastsimcoal) can use reduced-representation sequencing data, such as RADseq \citep{andrews2016radseq}.
+\dadi and \fastsimcoal require a pre-specified parametric model, unlike \MSMC, \smcpp, and \stairwayplot.
+Using a parametric approach yields less noisy results, but a model that is too simple may not capture important demographic events (Figures~\ref{fig:IM_popn_human} and~\ref{fig:two_popn_fly}).
+From a software engineering perspective, methods also differ in their ease of installation and use.
+We hope our workflows will assist in the application of all the methods we have considered.
 

### 2020-04-08 popSimManu.tex
minor changes to two pop figs and captions

--- 2020-01-21 popSimManu.tex
@@
 The population split date, $T_{DIV}$, is shown at
-the bottom, with a common X-axis to the population size panels.}
+the bottom, with a common $x$-axis to the population size panels.}
 \label{fig:IM_popn_human}
@@
 Data were generated by simulating replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
-inferred in \cite{comeron2012many}.}
+inferred in \cite{comeron2012many}.
+Directional migration from Europe to Africa is represented as MIG_AF_EU and
+migration from Africa to Europe is represented as MIG_EU_AF.}
 \label{fig:dromel_mig_rates}
@@
 replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}.}
+inferred in \cite{international2007second}.
+Directional migration from Europe to Africa is represented as MIG_AF_EU and
+migration from Africa to Europe is represented as MIG_EU_AF.}
 \label{fig:homsap_mig_rates}

### 2020-04-12 response-to-reviewers.md
minor changes to two pop figs and captions
Typo fix

@@
 These include data required, type of model, and software engineering.
-A comprehensive investigation of these issues is beyond the scope of the current manuscript, with focuses on the resource itself, rather than its application.
+A comprehensive investigation of these issues is beyond the scope of the current manuscript, which focuses on the resource itself, rather than its application.
 

### 2020-04-14 response-to-reviewers.md
responses notes from call

@@
 
-FIXME: An example response just to illustrate how to respond to comments in this Markdown file. 
-This will definitely need to be removed before we send the actual response!
+Issue #163 (ckyriazis)
 
@@
 However, we felt that a simple demonstration of the functionality would substantially increase the scenarios easily available and imaginable by readers of the paper.
-> 
+
+Issue #150 (jadrion) will add the figures we need and some explanation;
+we need to explain here (and maybe in the text) that selection is beyond the scope of this paper,
+and instead we are focusing in the paper on validating scaling.
+This will also be covered by issue #164.
+
 > 3) We felt that the discussion would benefit from a discussion of the details not yet incorporated into the platform, e.g.
@@
 Such an addition might spur future development, and also give the authors an opportunity to discuss general ways forward on these issues.
-> 
+
+Issue #164. (petrelharp)
+
 > 4) We would be interested to see more of a discussion of the strengths and weakness of the various demographic inference methods.
@@
 Having all of these models implemented in a central place will significantly lower the barrier of entry of empirical researchers into rigorous simulation frameworks.
-> 
+
+TODO: discussion other applications
+petrelharp
+
 > Overall we are very excited to see this big move forward and welcome the authors' careful work in this important area. 
@@
 As the second half of this paper makes clear, because of errors in inference, many of these models will not be accurate, in the sense of representing the true history of the species.
+
+Issue #163.
+
 > 2.
@@
 Assuming it can be layered on top of the existing structure, it may be enough for this publication to note this as a future challenge that needs development/addressing.
+
+Response summary:
+Thank you for this forward-thinking comment.
+Good idea but how would you actually do it?
+We will have to let this emerge, and can't solve the problem until we actually have it.
+
 > 3.
@@
 I do not think this is a requirement for publication, but it would add great value to this initial release of the resource.
+
+We're not going to do this either;
+Issue #164.
+Emphasize importance of selection in the discussion.
+
 > 4.
@@
 This part of the pipeline seems open to refinement.
+
+Maybe so, but beyond the scope?
+Issue #166 (andrewkern)
+
 > b.
@@
 It seems that it would be preferable for them to be separate from the recombination rate files - right now it reads as if the mask applied is a function of the genetic map file, but this seems too inflexible for users who prefer an alternative approach to masking.
+
+Each software package requires its own kind of masking - eg smc++ requires a bedfile.
+Issue #166
+
+Issue minor comments
+ckyriazis and others
+
 > 
@@
 I can keep the same demographic model, but change the recombination map, without having to re-write, or learn, the innards of a complex command-line for ms/msprime)
+
+Maybe make this wording more descriptive and precise.
+
 > 
@@
 with dark grey solid lines as now in Fig 2).
+
+Add census size ("simulated pop size as taken from the published model"?) to Fig 2, 3,
+Also S1, S2, S3.
+Explain in the text (maybe caption), and also explain why not to Fig 4.
+Issue #160 (jradrion)
+
 > 
@@
 > b. Also, the figure would be improved by showing with vertical lines the timing of the CEU/YRI divergence in the true model. (similar comments hold for Figures S4 and S5)
+
+Done in PR #161, thanks ckyriazis!
+
 > 
@@
 > I thoroughly enjoyed reading this manuscript, and learning of all the new features that have been written into stdpopsim, and believe that this will be an invaluable contribution to the population genomics community.
+
+Thank YOU!
+
 > 
@@
 >
+
 > Fig S6: The figure captions/labels are acronyms, which aren't explained. Would help readers to understand what this plot is estimating. 
+We added a sentence to the caption better explaining this figure.
+
 >
+
+Done in PR #161.
 
@@
 There are also a myriad of methods for detecting/quantifying selection, and these simulations are not consistent either.
+
+Issue #150 and #164, as above.
+
 > 
@@
 Right now the zigzag model is under humans in the catalog.
+
+Summary of response:
+This is done, in the documentation now.
+But note that zigzag is in human because the parameterization is human-ish.
+
 > 
@@
 page 15) to refer to population size (which makes sense), but then also say "In all cases we set the sample size of the focal population to N = 50 chromosomes." For MSMC, the sample size was set to n=2,8 which suggests haploid samples, but the "Calculating coalescence rates" section says that n is the diploid sample size.
+
+Issue #168 (petrelharp)
+
 > 
@@
 This section is also referred to as the Appendix in the main text.
+
+Issue #169 (petrelharp)
+
 > 
@@
 > -Figure 4 captain: "or smc++" -> "and smc++" (some other figure captions have this too), I would also say "$x$-axis" 
+We made the suggested change.
 > 
@@
 (Maybe the goal is to make this a bit more distributed?) And just to be clear, Developer B would not have access to Developer A's implementation of the model while they were independently implementing it, right?
+
+(B would have access - it's not blinded - they just aren't supposed to look at it)
+
 > 
@@
 > -Figure S6: explain x-axis "MIG_AF_EU" and "MIG_EU_AF". I assume these are directional migration rates
+
+We added a sentence to the caption better explaining this figure.
+
 > 
@@
 >
+
+Remove this figure.

### 2020-04-15 response-to-reviewers.md
various edits throughout ms

@@
 
-Issue #163 (ckyriazis)
+We fully agree that the language in the original manuscript was not careful enough in conveying uncertainty about the models and parameters we discuss.
+To address this issue, we have made several edits throughout the manuscript trying to better convey this uncertainty.
+See our response to reviewer 1 below for examples.
 
@@
 (Also see page 3 sentence on "details on =the= physical organization of the genome"; the same comment applies, and an improvement might be something like "details on the physical organization of the latest reference genome assembled for the species").
+
+We thank the reviewer for bringing up this important point and fully agree that the language in the manuscript needs to better emphasize the uncertainty in our current parameter estimates.
+To address this, we have made several changes throughout the manuscript, including the passages cited above:
+
+On page 3, it now reads:
+"For each organism, the catalog contains information on our current understanding of the physical organization (e.g., chromosome structure)
+of its genome, one or more inferred genetic maps, default population-level parameters (mutation rate and generation time estimates)
+and one or more published demographic models.. These available models
+and parameters are meant to represent the current standards in the field, however our
+intention is that this resource will evolve as new results become available"
+
+On page 4, it now reads:
+"Firstly, the library defines
+some basic information on our current understanding of each
+species' genome, including information about chromosome
+lengths, average mutation rate estimates, and generation times. We also provide access to detailed
+empirical information such as inferred genetic maps, which model observed
+heterogeneity in recombination rate along chromosomes."
+
 > b.
@@
 
-Issue #163.
+We made the suggested change:
+"Importantly, we have rigorous quality control methods to ensure that models
+are correctly implemented with respect to their original publication
+and have documented methods for others to contribute new modules."
 
@@
 
-Issue minor comments
-ckyriazis and others
-
 >
@@
 > 1. Is the provenance information only provided if msprime is used to produce tree sequence files? Is a similar functionality in place with the SLiM back-end?
+
+FIXME
+
 >
 > 2. Little detail is given on the system of identifiers/versioning for models, maps, etc.
+
+FIXME
 >
 > 3. In the abstract, it's left vague what the nature of the resource is.
+
+FIXME
 >
@@
 > b. Also, while I wouldn't normally include url to software in an abstract, in this special case it seems like it could be very helpful to readers who want to jump right to the resource quickly?
+
+FIXME
+
 >
@@
 
-Maybe make this wording more descriptive and precise.
+FIXME: Maybe make this wording more descriptive and precise.
 
@@
 The word doesn't add much value over simply saying the models are standardized - the standardization includes that the model implementation is quality controlled to match the original source publication, which I think is what is meant by accuracy here?
+
+We agree that "accurate" could be a misleading term here so have removed it from the manuscript.
+
 >
 > 6. Page 6. Typo in "time or writing". Change to "time of writing"
+
+We corrected this typo.
+
 >
@@
 
-Add census size ("simulated pop size as taken from the published model"?) to Fig 2, 3,
-Also S1, S2, S3.
-Explain in the text (maybe caption), and also explain why not to Fig 4.
+We agree that adding both census size and inverse coalescence rates to these plots would instructive.
+Thus, we have added census size to all of the single population analysis plots (Figs 2, 3, S1, S2, and S3) in addition to ICR.
+However, after some discussion, we have decided that adding ICR to the two population analysis plots (Figs 4, S5, and S7, which currently only have census size) might not be appropriate since these multi-population analyses do to some extent account for the effects of migration that are ignored in the single-population case, and should therefore be able to recover the true simulated population sizes.
+
+Finally, we note that we have decided to instead describe the true simulated population sizes as the "simulated population size" rather than "census size" as we realized the use of "census" here could be misleading (see below for further discussion).
+
+FIXME: add the census sizes to Figs 2, 3, S1, S2, and S3 and edit figure captions
 Issue #160 (jradrion)
@@
 
-Done in PR #161, thanks ckyriazis!
+We agree that the orientation of this figure is confusing, and so have changed it as suggested.
+In addition, we had previously plotted divergence times here as vertical lines, however we felt that plotting it this way made the figure hard to read since there was a great deal of overlap between these lines and the N_t lines on the figure, which was was especially difficult to read when including multiple replicates.
+Thus, we prefer to keep the lower divergence times panel for improved readability.
 
@@
 Page 16 "checked for convergence" : Wording - sounds like checking for convergence between the methods - I think you mean convergence by each method to an optimum in their respective objective functions.
+
+We changed the wording here as follows:
+
+"For each simulation
+replicate, we performed 10 runs of dadi and fastsimcoal, checking to ensure that
+each method reached convergence."
+
 >
@@
 >
+
+FIXME: Peter could you add this?
 
@@
 > Fig S6: The figure captions/labels are acronyms, which aren't explained. Would help readers to understand what this plot is estimating.
+
 We added a sentence to the caption better explaining this figure.
-
->
-
-Done in PR #161.
 
@@
 > For a future draft, line numbers would be very helpful.
+
+We made the suggested change.
+
 >
 > -Abstract: "possibility for error" -> inconsistency
+
+We made the suggested change.
 >
 > -Page 5: trivial -> straightforward
+
+We made the suggested change.
+
 >
 > -"tree sequence": make it clear that users can easily obtain traditional variant matrices from this format, in case they are unfamiliar
+
+We added the following statement:  
+"The tree sequence format could also be converted
+to other formats (e.g., vcf) by the user if desired. "
 >
 > -Page 6: "time or writing" -> of
+
+We corrected this typo.
+
 >
@@
 Discuss a bit more.
+
+FIXME: could someone add more discussion on this?
+
 >
 > -Page 8, paragraph "Stdpopsim allows...": reword this first sentence
+
+We changed the sentence as follows:  
+"Using stdpopsim, we can readily compare performance on this benchmark to
+that based on a different model of human history. "
+
 >
 > -Figure 3 caption: wrong citation, should be Sheehan and Song
+
+FIXME: I think the citation is correct here (for the genetic map), maybe the issue is that the demographic model isnt actually cited? Should we add that in?
+
 >
 > -Page 10, last sentence before "Multi-population demographic models" section: clarify what "this setting" is (both species I assume)
+
+We made the following change:  
+"Accuracy is mixed among methods when doing inference on simulated data from these D. melanogaster
+and A. thaliana models, and generally worse than what we
+observe for simulations of the human genome."
+
 >
 > -Figure 4 captain: "or smc++" -> "and smc++" (some other figure captions have this too), I would also say "$x$-axis"
+
+We made the suggested changes.
+>
+> -Page 12, first paragraph: include a reference where people could learn more about the effective population size vs. census size, random-mating models, etc
+
+We thank the reviewer for this suggestion.
+However, we have realized that our use of "census size" here to describe the "true" simulated population size was a bit misleading, as our intention was not to characterize these as actual "census" (as opposed to "effective") population sizes.
+To avoid this confusion, we have changed the manuscript to now describe them as the "simulated population size".
+For example, this section now reads:
+
+"However, because model misspecification
+here is less severe  compared to the single-population case, we compare
+our inferred population sizes only to the simulated population sizes (black line in
+Figure 4C) and not the inverse coalescence rates."
+
+As a result, we dont necessarily feel that referring to the census vs effective population size literature would be appropriate here.
+
+>
+> -Page 12 (and elsewhere): "By contrast" -> "In contrast"(?) I think both are technically correct though
+
+We changed all instances of "by contrast" to "in contrast".
+>
+> -Page 12: sentence beginning with "Figure S4 shows..." is long, perhaps breakup or rephrase ("bottleneck into a European population" sounds odd to me.)
+
+We changed this section to the following:  
+"Figure S4 shows inference results using data simulated under
+the OutOfAfrica_2L06 model. This model includes an ancestral population in
+Africa from which a European population splits off following a bottleneck, with no
+post-divergence gene flow between the African and European population (Figure S4A)."
+>
+> -Page 12: "Europe population" -> "European population"
+
 We made the suggested change.
 >
-> -Page 12, first paragraph: include a reference where people could learn more about the effective population size vs. census size, random-mating models, etc 
-> 
-> -Page 12 (and elsewhere): "By contrast" -> "In contrast"(?) I think both are technically correct though 
-> 
-> -Page 12: sentence beginning with "Figure S4 shows..." is long, perhaps breakup or rephrase ("bottleneck into a European population" sounds odd to me.) 
-> 
-> -Page 12: "Europe population" -> "European population" 
-> 
 > -Page 12: S7, S4, S6 are mentioned in a different order than they appear in the supplement (I understand why but maybe revisit this part)
+
+We reordered the figures to make sure they were referenced in the same order they appear in the supplement.
+
 >
 > -Page 13: "overfit the YRI history" do you mean overfit in a technical sense?
+
+FIXME: Could someone weigh in on what was originally meant here?
+
 >
@@
 > -Page 16, line 6: "population 1 after"?
+
+The "1" here was a typo - we removed it.
+
 >
 > -Page 16: S5 -> Figure S5
+
+We corrected this mistake.
+
 >
 > -Page 17: "with making" -> "for making"?
+
+We corrected this mistake.
+
 >
 > -Figure S2 caption: move "and" to before MSMC (in a few other figure captions too)
+
+We made the suggested change.
+
 >
@@
 
-Remove this figure.
+We removed this figure.

### 2020-04-15 popSimManu.tex
various edits throughout ms

@@
 \pretocmd{\abstractname}{\newpage}{}{}
+
+\usepackage{lineno}
+\linenumbers
 
@@
 Computational genetics researchers currently re-implement simulation models
-independently, leading to duplication of effort and the possibility for error.
+independently, leading to duplication of effort and inconsistency.
 Population genetics, as a field, also lacks standard benchmarks by which new tools for inference might
@@
 \emph{Drosophila~melanogaster}, and \emph{Arabidopsis~thaliana}. For each
-organism, the catalog contains details on the physical organization (e.g., chromosome structure)
-of its genome, one or more genetic maps, default population-level parameters (mutation rate,
-generation time) and one or more published demographic histories. Through
+organism, the catalog contains information on our current understanding of
+ the physical organization (e.g., chromosome structure)
+of its genome, one or more inferred genetic maps, default population-level parameters (mutation rate and generation time estimates) 
+and one or more published demographic models. These available models
+and parameters are meant to represent the current standards in the field, however our 
+intention is that this resource will evolve as new results become available. Through
 either a command line interface or a simple Python API, users can specify which
@@
 to continue its growth and maintain its quality.
-Importantly, we have rigorous quality control methods to ensure implemented models are accurate
+Importantly, we have rigorous quality control methods to ensure that models
+are correctly implemented with respect to their original publication 
 and have documented methods for others to contribute new modules.
 We invite new collaborators to join our community.
+% CK: perhaps we could add a bit more info here on how to contribute demographic models and a link to the tutorial when its available?
 Below we describe the resource and give
@@
 A species definition consists of two key elements.  Firstly, the library defines
-some basic information about each species' genome, including information about chromosome
-lengths, average mutation rates, and generation times. We also provide access to detailed
-empirical information such as genetic maps, which model observed
+some basic information on our current understanding of each 
+species' genome, including information about chromosome
+lengths, average mutation rate estimates, and generation times. We also provide access to detailed
+empirical information such as inferred genetic maps, which model observed
 heterogeneity in recombination rate along chromosomes. As such maps are often large,
@@
 and the \cite{comeron2012many} map for \emph{D.~melanogaster}.
-Adding futher maps to the library is trivial.
+Adding further maps to the library is straightforward.
 The second key element of a species description
@@
 specific historical scenarios that have been fit to present-day patterns of
-genetic variation. (See the Methods for a description of the community
+genetic variation. Our aim is that this catalog will grow as we continue
+to refine our understanding of different species' demographic histories
+and new models are published. (See the Methods for a description of the community
 development and quality-control process for these models.)
@@
 Given the genome data and simulation model descriptions defined within the
-library, it is then straightforward to run accurate, standardized simulations
+library, it is then straightforward to run standardized simulations
 across a range of organisms. \Stdpopsim has a Python API and a user-friendly
@@
 extremely compact, and can be processed efficiently using the \tskit library
-\citep{kelleher2016efficient,kelleher2018efficient}. Currently,
+\citep{kelleher2016efficient,kelleher2018efficient}. The tree sequence format could also be converted
+to other formats (e.g., vcf) by the user if desired. Currently,
 \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
@@
 of the key quantitative data needed to simulate a given species. These include a description
-of the assembly, information about mutation rate, recombination rate(s), and generation time
-in addition to a series of demographic models that are specific to that organism.
+of the assembly, information about mutation rate estimates, inferred recombination rate(s), and generation time
+in addition to a series of demographic models that have been estimated for that organism.
 
@@
 admixture \cite[e.g.,][]{ragsdale2019models}.
-In addition to those models shown, at time or writing
-the PopSim Consortium has models in development for \textit{Pongo abelii} and \textit{Escherichia coli}.
+In addition to those models shown, at time of writing
+the PopSim Consortium has models in development for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli}.
 
@@
 
-Beyond humans we have have implemented two demographic histories for
+Beyond humans we have have implemented two demographic models for
 \textit{D.~melanogaster}, three from \textit{A.~thaliana}, and are currently developing
-models for \textit{P.~abelii} and \textit{E.~coli}.
+models for \textit{P.~abelii} , \textit{Canis familiaris}, and \textit{E.~coli}.
 For \textit{D.~melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 we have standardized naming
-conventions for species, genetic maps, and demographic models that will enable long term stability
+conventions for species, genetic maps, and demographic models that will enable long-term stability
 of unique identifiers used throughout \stdpopsim,
@@
 In each row we show comparisons among the methods (including two sample sizes for \MSMC).
-Blue lines show estimates from each of three replicate whole genome simulations.
-There is no single ``true'' reference for effective population size
-because of model misspecification---the inference methods are fitting a single population model
-to data simulated from multiple populations.
-However, many methods work by matching coalescence time distributions,
-and a single-population model with varying population size can match any coalescence time distribution
-(in which case coalescence rate is the inverse of the effective size).
-For this reason, we used as our ``ground-truth'' (solid black lines)
-not historical census sizes,
-but rather inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix).
+Blue lines show estimates from each of three replicate whole genome simulations and (FIXME: what color?)
+lines show the simulated population sizes. When fitting single-population models to data simulated for multiple
+populations with migration, however, we do not necessarily expect the inference methods to recover the 
+``true'' simulated population size due to model misspecification. Instead, many methods work by matching
+ coalescence time distributions, and a single-population model with varying population size can match any 
+ coalescence time distribution (in which case coalescence rate is the inverse of the effective size).
+For this reason, we also plot the inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix)
+as another benchmark of the ``truth'' in the single-population inference setting. 
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 
-\Stdpopsim allows us to readily compare relative performance on this benchmark to
-that based on a different
-model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
+Using \stdpopsim, we can readily compare performance on this benchmark to
+that based on a different model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
 $N(t)$ from simulations using the same physical and genetic maps, but from the OutOfAfrica\_3G09
@@
 under the African2Epoch\_1H18 model originally inferred by \cite{huber2018gene}.
-In both cases, as with humans, we use \stdpopsim to simulate replicate genomes using an empirically derived genetic map,
+In both cases, as with humans, we use \stdpopsim to simulate replicate genomes using an empirically-derived genetic map,
 and try to infer back parameters of the simulation model.
-Accuracy is mixed among methods in this setting and generally worse than what we
+Accuracy is mixed among methods when doing inference on simulated data from these \textit{D.~melanogaster}
+and \textit{A.~thaliana} models, and generally worse than what we
 observe for simulations of the human genome.
@@
 \caption{\textbf{Parameters estimated using a multi-population human model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp. \textbf{(A)} Data were generated by simulating
+Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, and \smcpp. \textbf{(A)} Data were generated by simulating
 replicate human genomes under the OutOfAfrica\_3G09 model and using the HapMapII\_GRCh37 genetic map
@@
 In shades of blue we show $N(t)$ trajectories estimated from each simulation,
-and in black census sizes for the respective population.
+and in black simulated population sizes for the respective population.
 The population split date, $T_{DIV}$, is shown at
@@
 not match the model from which the data were generated (Figures
-\ref{fig:IM_popn_human}A and B). However, because the inferred models here better
-match the simulated models than in the single population case, here we compare
-our inferred population sizes directly to the census size of the simulated models (black line in
-Figure \ref{fig:IM_popn_human}C).
+\ref{fig:IM_popn_human}A and B). However, because model misspecification 
+here is less severe  compared to the single-population case, we compare
+our inferred population sizes only to the simulated population sizes (black line in
+Figure \ref{fig:IM_popn_human}C) and not the inverse coalescence rates.
 
@@
 due to model misspecification (Figure \ref{fig:homsap_mig_rates}).
-By contrast, \smcpp is much better at capturing the recent exponential
+In contrast, \smcpp is much better at capturing the recent exponential
 growth in the CEU population, though the inferred population sizes are generally
@@
 for a two-population model of \textit{D.~melanogaster}. Figure
-\ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
-the OutOfAfrica\_2L06 model, which includes
-an ancestral population in Africa, then a population expansion with a population
-split and bottleneck into a European population with no post-divergence migration.
+\ref{fig:two_popn_fly} shows inference results using data simulated under 
+the OutOfAfrica\_2L06 model. This model includes an ancestral population in
+Africa from which a European population splits off following a bottleneck, with no 
+post-divergence gene flow between the African and European population (Figure \ref{fig:two_popn_fly}A).
 Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
@@
 and \fastsimcoal both do reasonably well at correctly inferring the absence of
-migration (Figure \ref{fig:dromel_mig_rates}). By contrast, the inferred
+migration (Figure \ref{fig:dromel_mig_rates}). In contrast, the inferred
 demographic parameters from \smcpp are more noisy, though in some cases better
-capture the short bottleneck in the Europe population.
+capture the short bottleneck in the European population.
 
@@
 (Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}), which
-tends to over-estimate the sizes of oscillations.  By contrast,
+tends to over-estimate the sizes of oscillations.  In contrast,
 \stairwayplot outperforms the sequence-based methods
@@
 
-All methods for inferring demographic history have strengths and weaknesses (as recently reviewed by \citet{beichman2018review}.)
+All methods for inferring demographic history have strengths and weaknesses (as recently reviewed by \citep{beichman2018review}.)
 We compared inferences from simulated whole genome data, but many factors affect choice of methodology.
 Markovian coalescent methods (\MSMC and \smcpp) require long contiguous stretches of sequence data.
-By contrast, frequency spectrum methods (\stairwayplot, \dadi, and \fastsimcoal) can use reduced-representation sequencing data, such as RADseq \citep{andrews2016radseq}.
+In contrast, frequency spectrum methods (\stairwayplot, \dadi, and \fastsimcoal) can use reduced-representation sequencing data, such as RADseq \citep{andrews2016radseq}.
 \dadi and \fastsimcoal require a pre-specified parametric model, unlike \MSMC, \smcpp, and \stairwayplot.
@@
 workflows \citep{koster2012snakemake} that perform demographic inference on
-tree sequence output from our package using a few common software packages.
+tree sequence output from our package using a few common software packages (see Figure \ref{fig:n_t_dag} for an example workflow).
 Our choice of \texttt{Snakemake} allows complete reproducibility of the
@@
 This IM model contains six parameters:
-the ancestral population size, the sizes of each population 1 after the split,
+the ancestral population size, the sizes of each population after the split,
 the divergence time, and two migration rate parameters.
@@
 model is often misspecified. To ground-truth our inference approach, we also conducted
-inference on a generic IM model that was identical to the model used for inference \ref{fig:generic_IM}.
+inference on a generic IM model that was identical to the model used for inference (Figure \ref{fig:generic_IM}).
 
@@
 chromosomes in the appropriate format for \dadi and \fastsimcoal. For each simulation
-replicate, we performed 10 runs of \dadi and \fastsimcoal and checked for convergence.
+replicate, we performed 10 runs of \dadi and \fastsimcoal, checking to ensure that 
+each method reached convergence.
+
+checked for convergence.
 Detailed settings for \dadi and \fastsimcoal can be found in the Snakefile
@@
 To visualize our results, we plotted the inferred population size trajectories
-for each simulation replicate alongside census population sizes
+for each simulation replicate alongside the simulated population sizes
 (Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C).
 Here, unlike the single-population workflow,
-we compare to census size rather than the inverse coalescence rate as the `true' population size.
+we compare our inferred population sizes only to the simulated population
+sizes and not the inverse coalescence rates.
 
@@
 \section*{Acknowledgments}
-We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible,
+We thank the Probabilistic Modeling in Genomics conference organizers for making this collaboration possible,
 and the Simons Center for Quantitative Biology at Cold Spring Harbor Laboratory
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairway plot},
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairway plot}, and
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 \caption{\textbf{Comparing estimates of $N(t)$ in \textit{A.~thaliana}}. Here we show estimates of population
-size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairway plot},
+size over time ($N(t)$) inferred using 4 different methods, \smcpp, and \texttt{stairway plot}, and
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
@@
 \label{fig:n_t_AraTha}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.pdf}
-\caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
-Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
-Data were generated by simulating
-replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
-inferred in \cite{comeron2012many}. See legend of Figure \ref{fig:IM_popn_human} for details.
-In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history
-as given by the census size for the simulated model.}
-\label{fig:two_popn_fly}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/generic_IM.pdf}
-\caption{\textbf{Parameters estimated from a generic IM model} Here we show estimates of $N(t)$
-inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp. Data were generated by simulating
-under a generic IM model with a human genome and \cite{international2007second} genetic map.
-In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history
-as given by the census size for the simulated model.}
-\label{fig:generic_IM}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/dromel_migration_rates.pdf}
-\caption{\textbf{Migration rate parameters estimated under a two-population \emph{Drosophila} model}.
-Here we show inferred migration rates from \dadi and \fastsimcoal.
-Data were generated by simulating replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
-inferred in \cite{comeron2012many}.
-Directional migration from Europe to Africa is represented as MIG_AF_EU and
-migration from Africa to Europe is represented as MIG_EU_AF.}
-\label{fig:dromel_mig_rates}
 \end{center}
@@
 \begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.pdf}
+\caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
+Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, and \smcpp.
+Data were generated by simulating
+replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
+inferred in \cite{comeron2012many}. See legend of Figure \ref{fig:IM_popn_human} for details.
+In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the simulated population sizes.}
+\label{fig:two_popn_fly}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/dromel_migration_rates.pdf}
+\caption{\textbf{Migration rate parameters estimated under a two-population \emph{Drosophila} model}.
+Here we show inferred migration rates from \dadi and \fastsimcoal.
+Data were generated by simulating replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
+inferred in \cite{comeron2012many}.
+Directional migration from Europe to Africa is represented as MIG_AF_EU and
+migration from Africa to Europe is represented as MIG_EU_AF.}
+\label{fig:dromel_mig_rates}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[width=0.8\linewidth]{display_items/n_t_dag.pdf}
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/two_pop_dag.pdf}
-\caption{\textbf{Workflow for our two population N(t) inference methods comparison}
-Here we show single replicate for one chromosome, chr22, simulated under the HomSap
-OutOfAfrica\_3G09 demographic model, with a HapmapII\_GRCh37 recombination map.
-Note that resulting data for input to all inference methods \dadi, \smcpp,
-and \fastsimcoal,
-derive from the same set of simulations.}
-\label{fig:two_pop_dag}
+\includegraphics[width=0.8\linewidth]{display_items/generic_IM.pdf}
+\caption{\textbf{Parameters estimated from a generic IM model} Here we show estimates of $N(t)$
+inferred using \dadi, \texttt{fastsimcoal2}, and \smcpp. Data were generated by simulating
+under a generic IM model with a human genome and \cite{international2007second} genetic map.
+In shades of blue we show the estimated
+$N(t)$ trajectories for each replicate. In black we show the simulated population sizes.}
+\label{fig:generic_IM}
 \end{center}

### 2020-04-15 response-to-reviewers.md
edits from Peter

@@
 of its genome, one or more inferred genetic maps, default population-level parameters (mutation rate and generation time estimates)
-and one or more published demographic models.. These available models
+and one or more published demographic models. These available models
 and parameters are meant to represent the current standards in the field, however our

### 2020-04-15 popSimManu.tex
edits from Peter

@@
 and have documented methods for others to contribute new modules.
-We invite new collaborators to join our community.
-% CK: perhaps we could add a bit more info here on how to contribute demographic models and a link to the tutorial when its available?
+We invite new collaborators to join our community (https://stdpopsim.readthedocs.io/en/latest/development.html).
 Below we describe the resource and give
@@
 specific historical scenarios that have been fit to present-day patterns of
-genetic variation. Our aim is that this catalog will grow as we continue
-to refine our understanding of different species' demographic histories
-and new models are published. (See the Methods for a description of the community
+genetic variation (See the Methods for a description of the community
 development and quality-control process for these models.)
@@
 
-All methods for inferring demographic history have strengths and weaknesses (as recently reviewed by \citep{beichman2018review}.)
+All methods for inferring demographic history have strengths and weaknesses (as recently reviewed by \citet{beichman2018review}.)
 We compared inferences from simulated whole genome data, but many factors affect choice of methodology.
@@
 
-checked for convergence.
 Detailed settings for \dadi and \fastsimcoal can be found in the Snakefile
@@
 inferred in \cite{international2007second}.
-Directional migration from Europe to Africa is represented as MIG_AF_EU and
-migration from Africa to Europe is represented as MIG_EU_AF.}
+Directional migration from Europe to Africa is represented as $MIG\_AF\_EU$ and
+migration from Africa to Europe is represented as $MIG\_EU\_AF$.
+Note that the $x$-axis coordinates are arbitrary.}
 \label{fig:homsap_mig_rates}
@@
 inferred in \cite{comeron2012many}.
-Directional migration from Europe to Africa is represented as MIG_AF_EU and
-migration from Africa to Europe is represented as MIG_EU_AF.}
+Directional migration from Europe to Africa is represented as $MIG\_AF\_EU$ and
+migration from Africa to Europe is represented as $MIG\_EU\_AF$.
+Note that the $x$-axis coordinates are arbitrary.}
 \label{fig:dromel_mig_rates}

### 2020-04-16 popSimManu.tex

@@
 to continue its growth and maintain its quality.
-Importantly, we have rigorous quality control methods to ensure that models
-are correctly implemented with respect to their original publication 
+Importantly, we have rigorous quality control methods to ensure that we have
+correctly implemented the models as described in their original publication
 and have documented methods for others to contribute new modules.

### 2020-04-18 popSimManu.tex
edits and many responses

@@
 \author[6,$\dagger$]{Kirk E. Lohmueller}
-\author[1,$\dagger$]{Peter L. Ralph}
+\author[1,20,$\dagger$]{Peter L. Ralph}
 \author[4,$\dagger$]{Daniel R. Schrider}
 \author[3,$\dagger$]{Adam Siepel}
-\author[20,$\dagger$,$\aleph$]{Jerome Kelleher}
+\author[21,$\dagger$,$\aleph$]{Jerome Kelleher}
 \author[1,$\dagger$,$\aleph$]{Andrew D. Kern}
@@
 \affil[19]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[20]{Big Data Institute, Li Ka Shing Centre for Health Information and
+\affil[20]{Department of Mathematics, University of Oregon}
+\affil[21]{Big Data Institute, Li Ka Shing Centre for Health Information and
 Discovery, University of Oxford}
@@
 \affil[$\aleph$]{Denotes corresponding authors, listed alphabetically}
-% JK: didn't bother with the \texttt stuff for stdpopsim in the abstract as it's nearly
-% always dropped in production anyway.
-
-\abstract{The explosion in population genomic data demands ever more complex
+
+\abstract{
+The explosion in population genomic data demands ever more complex
 modes of analysis, and increasingly these analyses depend on sophisticated simulations.
@@
 simulation engine remains a difficult and error-prone task.
-Computational genetics researchers currently re-implement simulation models
-independently, leading to duplication of effort and inconsistency.
-Population genetics, as a field, also lacks standard benchmarks by which new tools for inference might
-be measured. Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
-\Stdpopsim is a community-driven open source project, which provides easy access to a standard
-catalog of published simulation models from a wide range of organisms
+Computational genetics researchers currently re-implement simulation models independently,
+leading to inconsistency and duplication of effort.
+This presents a major barrier to empirical researchers seeking
+to use simulations for power analyses of upcoming studies
+or sanity checks on existing genomic data.
+Population genetics, as a field,
+also lacks standard benchmarks by which new tools for inference might be measured.
+Here we describe a new resource, \stdpopsim, that attempts to rectify this situation.
+\Stdpopsim is a community-driven open source project, which provides easy access to a growing
+catalog of published simulation models from a range of organisms
 and supports multiple simulation engine backends.
-% PopSim assures correctness in our implementations through a mature, community-based quality control pipeline.
+This resource is available as a well-documented python library
+with a simple command-line interface.
 We share some examples demonstrating how \stdpopsim
@@
 \citep{eyre2009estimating, alachiotis2012omegaplus,degiorgio2016sweepfinder2,kern2018diplos,sugden2018localization}.
-While these methods have increased our understanding of the
-impacts of genetic and evolutionary processes, very little has been done to systematically
-benchmark the quality of inferences gleaned from computational population genetics.
+While these methods have undoubtedly increased our understanding of
+genetic and evolutionary processes, very little has been done to systematically
+benchmark the quality of these inferences.
+% or their robustness to deviations from the assumptions.
 As large databases of population genetic variation begin to be used to inform public health procedures,
@@
 field, yet there has been no systematic attempt to establish community
-standards or best practices for executing them. Instead, the general modus
+standards or best practices for executing them.
+Instead, the general modus
 operandi to date has been for individual groups to validate their own
-methods using bespoke simulations.
+methods using simulations coded from scratch
+(perhaps provided as long command-line calls in supplementary material).
 Often these studies focus more on showcasing a novel method than on
@@
 It is also a barrier to entry to the field, because new researchers can
-struggle with
-the many steps involved
+struggle with the many steps involved
 in implementing a state-of-the-art population genetics simulation,
@@
 A related issue is that it has been challenging to assess the degree to which modeling assumptions
-and choices of data summaries can affect population
-genetic inferences. Yet there
-are clear examples of different methods yielding fundamentally
+and choices of data summaries can affect population genetic inferences.
+However, there are clear examples of different methods yielding fundamentally
 different conclusions. For example, Markovian coalescent methods applied to human genomes have
@@
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains three organisms: humans,
-\emph{Drosophila~melanogaster}, and \emph{Arabidopsis~thaliana}. For each
-organism, the catalog contains information on our current understanding of
- the physical organization (e.g., chromosome structure)
-of its genome, one or more inferred genetic maps, default population-level parameters (mutation rate and generation time estimates) 
-and one or more published demographic models. These available models
-and parameters are meant to represent the current standards in the field, however our 
-intention is that this resource will evolve as new results become available. Through
-either a command line interface or a simple Python API, users can specify which
+The \stdpopsim catalog currently contains five organisms: humans,
+\textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
+\textit{Escherichia coli}, and \textit{Canis familiaris}.
+For each organism, the catalog contains information on our current understanding of
+the physical organization (e.g., chromosome structure) of its genome,
+one or more inferred genetic maps (except for \textit{E.~coli}),
+default population-level parameters (mutation rate and generation time estimates).
+The catalog also contains numerous published demographic models for the first three species.
+These available models and parameters are meant to represent the field's current understanding,
+and we intend for this resource to evolve as new results become available.
+Through either a command line interface or a simple Python API, users can specify which
 organism, genetic map, chromosome, and demographic history they are interested in simulating, and the
@@
 and have documented methods for others to contribute new modules.
-We invite new collaborators to join our community (https://stdpopsim.readthedocs.io/en/latest/development.html).
+We invite new collaborators to join our community:
+those interested should visit our developer documentation at
+\url{https://stdpopsim.readthedocs.io/en/latest/development.html}.
 Below we describe the resource and give
@@
 hierarchical organization of the \stdpopsim catalog contains all model simulation information
-within individual species (expanded information shown here for \emph{H.~sapiens} only).
+within individual species (expanded information shown here for \textit{H.~sapiens} only).
 Each species is associated with a representation of the physical genome, and one or more genetic maps and demographic models.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Results}
+
 \paragraph{The \stdpopsim library.}
@@
 \textit{D.~melanogaster}, and \textit{A.~thaliana}.
-A species definition consists of two key elements.  Firstly, the library defines
-some basic information on our current understanding of each 
+A species definition consists of two key elements.
+Firstly, the library defines some basic information about our current understanding of each 
 species' genome, including information about chromosome
-lengths, average mutation rate estimates, and generation times. We also provide access to detailed
-empirical information such as inferred genetic maps, which model observed
-heterogeneity in recombination rate along chromosomes. As such maps are often large,
-we do not distribute them directly with the software, but make them available
-for download in a standard format. When a simulation using such a map is
-requested by the user, \stdpopsim will transparently download the map
-data into a local cache,
+lengths, average mutation rate estimates, and generation times.
+We also provide access to detailed empirical information such as inferred genetic maps,
+which model observed heterogeneity in recombination rate along chromosomes.
+Such maps are often large,
+so we do not distribute them directly with the software, but make them available
+for download in a standard format.
+When a simulation using such a map is requested by the user,
+\stdpopsim will transparently download the map data into a local cache,
 % (\texttt{$\sim$/.local/cache/stdpopsim} by default on Unix platforms)
@@
 deCODE~\citep{kong2010fine} genetic maps for humans;
-the \cite{salome2011recombination} map for \emph{A.~thaliana};
-and the \cite{comeron2012many} map for \emph{D.~melanogaster}.
+the \cite{salome2011recombination} map for \textit{A.~thaliana};
+the \cite{comeron2012many} map for \textit{D.~melanogaster},
+and the \cite{campbell2016pedigree} map for \textit{C.~familiaris}.
 Adding further maps to the library is straightforward.
@@
 extremely compact, and can be processed efficiently using the \tskit library
-\citep{kelleher2016efficient,kelleher2018efficient}. The tree sequence format could also be converted
-to other formats (e.g., vcf) by the user if desired. Currently,
-\stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
+\citep{kelleher2016efficient,kelleher2018efficient}.
+The tree sequence format could also be converted
+to other formats (e.g., VCF) by the user if desired.
+Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
 as the default simulation engine.
@@
 of the key quantitative data needed to simulate a given species. These include a description
-of the assembly, information about mutation rate estimates, inferred recombination rate(s), and generation time
+of the genome assembly, information about mutation rate estimates, inferred recombination rate(s), and generation time
 in addition to a series of demographic models that have been estimated for that organism.
 
-The current contents of the \stdpopsim catalog are shown in \autoref{tab:catalog}.
+The current demographic models in the \stdpopsim catalog are shown in \autoref{tab:catalog}.
 These range from
@@
 In addition to those models shown, at time of writing
-the PopSim Consortium has models in development for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli}.
+the PopSim Consortium has demographic models in development for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli}.
 
@@
 For each model we report the CPU time, maximum memory usage and the
-size of the output \tskit file. In each case we simulate 100 samples
+size of the output \tskit file, as simulated using the \texttt{msprime} simulation engine.
+In each case we simulate 100 samples
 drawn from the first population, for the shortest chromosome of that species
@@
 
-Beyond humans we have have implemented two demographic models for
-\textit{D.~melanogaster}, three from \textit{A.~thaliana}, and are currently developing
-models for \textit{P.~abelii} , \textit{Canis familiaris}, and \textit{E.~coli}.
 For \textit{D.~melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
@@
 and associated bottleneck model of \cite{li2006inferring}, which jointly models African
-and European populations (OutOfAfrica\_2L06). For \emph{A. thaliana}, we implemented the
+and European populations (OutOfAfrica\_2L06). For \textit{A.~thaliana}, we implemented the
 model in \cite{durvasula2017african} inferred using \MSMC. This model includes
@@
 population sizes (SouthMiddleAtlas\_1D17). We have also implemented a two-epoch and a three-epoch model estimated from African
-samples of \emph{A. thaliana} in \cite{huber2018gene} (African2Epoch\_1H18 and African3Epoch\_1H18).
+samples of \textit{A.~thaliana} in \cite{huber2018gene} (African2Epoch\_1H18 and African3Epoch\_1H18).
 %For \emph{P. pygmaeus} we have implemented the two
@@
 We start by comparing popular methods for estimating
-population size histories ($N(t)$) of single populations and subsequently
+population size histories of single populations and subsequently
 show simple examples of multi-population inference.
 To reproducibly evaluate and compare the performance of inference methods, we developed
-workflows using \texttt{snakemake} \citep{koster2012snakemake}
-that are available from \url{https://github.com/popsim-consortium/analysis},
+workflows using \texttt{snakemake} \citep{koster2012snakemake},
+available from \url{https://github.com/popsim-consortium/analysis},
 that allow efficient computing in multicore or cluster environments.
@@
 in a number of the demographic models described above.
-Our workflow generates $R$ replicates of $C$ chromosomes, producing $n$ samples in each of a total of $R \times C$
-simulations for each demographic model. After simulation, the workflow prepares
-input files for each inference method by grouping all
-chromosomes, for each sample, into a single file. For each of the $R$ simulation replicates, this step results in an
-input file for each of the
-respective inference methods and derived from the same simulated tree sequences.
+Our workflow generates $R$ replicates of $C$ chromosomes,
+producing $n$ samples in each of a total of $R \times C$ simulations for each demographic model.
+After simulation,
+the workflow prepares input files for each inference method
+by grouping all $n \times R \times C$ simulated chromosomes
+into a single file.
+Each file is then converted into an input file appropriate for each inference method
+(so, all inference methods run on the same simulation replicates).
 Each of the inference programs are then run in parallel, followed by plotting of
-$N(t)$ estimates from each program.
+estimates of population size history from each program.
 
@@
 size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
-\MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model and using the
+\MSMC with $n=2$ and $n=8$ samples. Data were generated by simulating
+replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model \citep{ragsdale2019models} and using the
 HapMapII\_GRCh37 genetic map~\citep{international2007second}. From top to bottom we show estimates for each
@@
 $N(t)$ trajectories for each replicate.
-As a proxy for the truth, in black we show inverse coalescence rates
-as calculated from the true demographic model (see text).
+As a proxy for the ``truth'', in black we show inverse coalescence rates
+as calculated from the demographic model used for simulation (see text).
 }
@@
 \citep{ragsdale2019models}, along with an empirical genetic map. In each column of this figure
-we show $N(t)$ inferred from samples taken from each of the three extant populations in the model.
+we show the inferred population size history (denoted $N(t)$)
+from samples taken from each of the three extant populations in the model.
 In each row we show comparisons among the methods (including two sample sizes for \MSMC).
 Blue lines show estimates from each of three replicate whole genome simulations and (FIXME: what color?)
-lines show the simulated population sizes. When fitting single-population models to data simulated for multiple
+lines show the simulated population sizes.
+When fitting single-population models to data simulated for multiple
 populations with migration, however, we do not necessarily expect the inference methods to recover the 
-``true'' simulated population size due to model misspecification. Instead, many methods work by matching
- coalescence time distributions, and a single-population model with varying population size can match any 
- coalescence time distribution (in which case coalescence rate is the inverse of the effective size).
-For this reason, we also plot the inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix)
+``true'' simulated population size due to model misspecification.
+Instead, many methods work by matching coalescence time distributions,
+and a single-population model with varying population size can match any coalescence time distribution
+(in which case coalescence rate is the inverse of the effective size).
+For this reason,
+we also plot the inverse coalescence rates calculated analytically in \texttt{msprime}
 as another benchmark of the ``truth'' in the single-population inference setting. 
+See the Appendix for a precise definition and description of the computation.
 While there is variation in accuracy among methods, populations, and individual replicates,
-the methods are generally accurate for this model of human history.
+the methods generally infer back a good estimate
+of the true effective population sizes of the simulations --
+inferred values are mostly within a factor of two of the truth,
+and many of the methods infer a bottleneck at the correct time.
 
@@
 experiment in Figure \ref{fig:n_t_HomSap_constant}.
-All methods recover population size within a factor of two of the truth, however
+All methods recover population size within a factor of two of the simulated values, however
 SMC-based methods, perhaps due to their regularization, tend to infer sinusoidal
@@
 size over time ($N(t)$) estimated from an African population sample. Data were generated by simulating
-replicate \textit{D.~melanogaster} genomes under the African3Epoch\_1S16 model
+replicate \textit{D.~melanogaster} genomes under the African3Epoch\_1S16 model \citep{sheehan2016deep}
 with the genetic map of \cite{comeron2012many}. In shades of blue we show the estimated
 $N(t)$ trajectories for each replicate.
-As a proxy for the truth, in black we show inverse coalescence rates
-as calculated from the true demographic model (see text).
+As a proxy for the ``truth'', in black we show inverse coalescence rates
+as calculated from the demographic model used for simulation (see text).
 }
@@
 For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
-with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
-an IM model was to mimic the approach often used on empirical datasets, where a relatively
-simple model is fit that may not reflect the true underlying demography.
-For human models with more than two populations (e.g., \cite{gutenkunst2009inferring})
+with constant population sizes and bi-directional migration \citep{hey2004im}.
+Our motivation for fitting this simple
+IM model was to mimic the typical approach of two population inference on empirical
+data, where the user is not aware of the `true' underlying demography and the inference
+model is often misspecified.
+For human models with more than two populations \citep[e.g.,][]{gutenkunst2009inferring}
 this means that we are inferring parameters for a model that does
@@
 migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
-of divergence times. However, these migration rate estimates are somewhat biased likely
-due to model misspecification (Figure \ref{fig:homsap_mig_rates}).
+of divergence times. However, these migration rate estimates are somewhat biased,
+likely due to model misspecification.
 In contrast, \smcpp is much better at capturing the recent exponential
-growth in the CEU population, though the inferred population sizes are generally
-noisier. In addition, the assumption of no migration by \smcpp leads to divergence time
-estimates that are consistently underestimated (Figure \ref{fig:IM_popn_human}C).
+growth in the CEU population, though the inferred population sizes are generally noisier.
+In addition, \smcpp consistently underestimates divergence times
+because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).
 
@@
 
-%Individual research groups are often in the position of having to independently
-%reimplement complex, previously published demographic models.
-%Additional layers of realism (e.g., recombination maps) make this especially daunting.
-%\stdpopsim aims to lower this barrier
-%by creating a quality controlled, easy-to-use, gateway to models and organisms
-%that are central to modern research in genetics.
-
+Besides being a resource for developers of computational methods,
+we aim for \stdpopsim to be a resource for empirical researchers using genomic data.
+For instance, \stdpopsim could be used in power analyses to determine adequate sample sizes,
+or in sanity checks, to see if observed data
+(e.g., levels of divergence or the allele frequency spectrum)
+are roughly consistent with the hypothesized scenario.
+Currently, many studies would benefit from simulation-based checks such as these,
+but this presents a major barrier,
+since individual research groups must reimplement complex, previously published demographic models,
+a task made especially daunting by additional layers of realism (e.g., recombination maps).
+
+\paragraph{Benchmarking population size inference}
 We have illustrated in this paper how \stdpopsim can be used for direct
@@
 benchmarking comparisons have been limited, but nevertheless
-reveal some informative features. For
-example, at the task of estimating $N(t)$ trajectories for simulated human
+reveal some informative features.
+For example, at the task of estimating population size histories for simulated human
 populations, we find that the sequence-based methods (\MSMC and \smcpp)
-perform somewhat better overall---at least for mid-range values of
-$t$---than the site frequency spectrum-based method (\stairwayplot)
-(Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}), which
-tends to over-estimate the sizes of oscillations.  In contrast,
-\stairwayplot outperforms the sequence-based methods
+perform somewhat better overall---at least for moderate times in the past---than
+the site frequency spectrum-based method (\stairwayplot),
+which tends to over-estimate the sizes of oscillations.
+(Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}).
+In contrast, \stairwayplot outperforms the sequence-based methods
 on simulations of \textit{D.~melanogaster} or \textit{A.~thaliana} populations,
@@
 reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
-history, in large part because of the restriction of constant population sizes.
+history, in large part because the methods assume constant population sizes.
 An exception is \smcpp, which does not have the same
 restrictions on its inferred history, and as a result does somewhat better
-with the CEU history but tends to overfit the YRI history.
+with the CEU history but tends to overcomplicate the simple YRI history.
 The results for
@@
 
-All methods for inferring demographic history have strengths and weaknesses (as recently reviewed by \citet{beichman2018review}.)
+All methods for inferring demographic history have strengths and weaknesses \citep[as recently reviewed by][]{beichman2018review}.
 We compared inferences from simulated whole genome data, but many factors affect choice of methodology.
@@
 Finally, comparison of results across methods or simulation runs
-provides an estimate of inference uncertainty, analogous to parametric
-bootstrapping,
-especially since different methods are likely vulnerable to model misspecification
-in different ways.
-
+provides an estimate of inference uncertainty, analogous to parametric bootstrapping,
+especially when different methods are vulnerable to model misspecification in different ways.
+
+\paragraph{Next steps}
 \Stdpopsim is intended to be a fully open, community-developed project.
@@
 and \textit{Arabidopsis} (among others)---are only intended to be a starting point for
-future development.  In addition to other taxa,
-we plan to incorporate other common biological processes
-such as selection, gene conversion, and mutational heterogeneity.
+future development.  
 Researchers are invited to contribute to the resource by adding their
@@
 population genomics community.
+
+One of our goals is to engage research communities studying other taxa,
+so as to expand the resource to many more species.
+Although we have included demographic models and recombination maps,
+there are many biological processes that we do not model.
+Some of these additions that we are enthusiastic to add are:
+selection (including distributions of fitness effects, maps of functional elements, and single sweeps),
+gene conversion, mutation models (rate heterogeneity),
+ancient samples,
+more realistic demography (overlapping generations, separate sexes, mortality/fecundity schedules),
+geographic population structure,
+and downstream aspects of data quality (genotyping and mapping error).
+These plans would not be possible without recent important work on SLiM \citep{haller2019slim},
+as well as current work on \texttt{msprime}.
+Some other important processes are more challenging to model with current simulation software,
+such as
+structural variation,
+changing recombination maps over time,
+transposable elements,
+and context-dependent mutation.
+
+% note: this is addressing reviewer comments about model uncertainty
+In particular, although the included demographic models represent the current state of the art,
+we anticipate the set of available models to expand
+as new methods and new modeling frameworks are developed.
+For instance, the current models all describe a small set of discrete, randomly mating populations,
+which are likely good approximations for deep-time population history,
+but may be less useful for methods describing dynamics of contemporary populations.
+\Stdpopsim's framework is sufficiently general that more realistic population models
+will be easily incorporated, once they are published.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Methods}
@@
 Developer A submits a quality control (QC) issue, after which a second developer,
-Developer B, then independently reimplements the model from the relevant
+Developer B (perhaps found by requesting review from the broader Consortium),
+then independently reimplements the model from the relevant
 primary sources and adds an automatic unit test for equality between the
-QC implementation and the preliminary production model. If the two
-implementations are equivalent, the original model is included in \stdpopsim.
+QC implementation and the preliminary production model.
+If the two implementations are equivalent, the original model is included in \stdpopsim.
 If not, we move to an arbitration process whereby A and B first try
@@
 We performed two types of demographic inference.
-Our first task was to infer effective population size over time ($N(t)$).
+Our first task was to infer effective population size over time (denoted $N(t)$).
 This was done using three software packages: \stairwayplot,
@@
 to default parameters as possible. For \stairwayplot we
-set the parameters ``numRuns=1'' and ``dimFactor=5000''. For \smcpp we used the
-``estimate" run mode to infer $N(t)$ with all other parameters set
-to their default values. For \MSMC we used the ``--fixedRecombination"
+set the parameters \texttt{numRuns=1} and \texttt{dimFactor=5000}.
+For \smcpp we used the ``estimate" run mode to infer $N(t)$ with all other parameters set
+to their default values. For \MSMC we used the \texttt{--fixedRecombination}
 option and used the default number of iterations.
@@
 population size instantaneously decreased from $N=10^4$ to $N=10^3$
-five hundred generations before the present. Each HomSap simulation was run
-using the HapmapII\_GRCh37 genetic map. For \textit{D.~melanogaster}
+five hundred generations before the present. Each simulation
+used the HapmapII\_GRCh37 genetic map. For \textit{D.~melanogaster}
 we estimated $N(t)$ from an African sample simulated under the DroMel,
 African3Epoch\_1S16 model using the Comeron2012\_dm6 map.
-Finally, we ran simulations of \emph{A. thaliana} genomes using the
+Finally, we ran simulations of \textit{A.~thaliana} genomes using the
 AraTha African2Epoch\_1H18 model under the Salome2012\_TAIR7 map.
@@
 and Europe populations, meaning that the Asia population functioned as a ``ghost''
-population that was ignored by our inference. Our motivation for fitting this simple
-IM model was to mimic the typical approach of two population inference on empirical
-data, where the user is not aware of the `true' underlying demography and the inference
-model is often misspecified. To ground-truth our inference approach, we also conducted
+population that was ignored by our inference.
+To ground-truth our inference approach, we also conducted
 inference on a generic IM model that was identical to the model used for inference (Figure \ref{fig:generic_IM}).
@@
 converted tree sequences into a two-dimensional site frequency spectrum for all
-chromosomes in the appropriate format for \dadi and \fastsimcoal. For each simulation
+chromosomes in the appropriate format for \dadi and \fastsimcoal.
+For each simulation
 replicate, we performed 10 runs of \dadi and \fastsimcoal, checking to ensure that 
@@
 of the three populations in the model: YRI, CEU, and CHB. In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
-for the rate of coalescence in the demographic model.}
+$N(t)$ trajectories for each replicate.
+As a proxy for the ``truth'', in black we show inverse coalescence rates
+as calculated from the demographic model used for simulation (see text).
+}
 \label{fig:n_t_gutenkunst}
@@
 replicate human genomes under a constant sized population model with $N=10^4$ and using the
-HapMapII genetic map~\citep{international2007second}. In black we show the true population size history
-of the model.}
+HapMapII genetic map~\citep{international2007second}.
+As a proxy for the ``truth'', in black we show inverse coalescence rates
+as calculated from the demographic model used for simulation (see text).
+}
 \label{fig:n_t_HomSap_constant}
@@
 replicate \textit{A.~thaliana} genomes under the African2Epoch\_1H18 model and using
-the genetic map of \cite{salome2011recombination}. In black we show the true population size history
-of the model.}
+the genetic map of \cite{salome2011recombination}.
+As a proxy for the ``truth'', in black we show inverse coalescence rates
+as calculated from the demographic model used for simulation (see text).
+}
 \label{fig:n_t_AraTha}
@@
 
-\section*{Calculating coalescence rates}
+\section*{Appendix: Calculating coalescence rates}
+
+In population genetics, the ``effective population size''
+of a population model with constant (census) size
+is often defined to be the number of diploids in a Wright-Fisher population
+that would have the same coalescence rate (or, equivalently, genetic drift)
+as the population in question \citep[reviewed in][]{crow1988inbreeding}.
+One reason the concept is useful
+is because theory predicts that genetic data from distinct populations
+with the same effective population size will look similar in many ways:
+for instance, their mean coalescence times will be the same.
+Conversely, this implies that effective population size
+should be easier to infer from genomic data than aspects of population demography
+that do not affect effective population size.
+An analogous observation holds for populations of changing size,
+if we define the "coalescence rate" of a given demographic model
+at a particular point back in time
+to be the rate of coalescence of remaining lineages
+and define the "coalescence effective size" at that time,
+denoted $N_e(t)$, so that the coalescence rate at time $t$ in the past
+is $1/(2N_e(t))$.
+With these definitions, any two models with the same effective population size trajectory ($N_e(t)$)
+will have the same \emph{distribution} of coalescence times.
+For this reason,
+we might guess that if we apply an inference method
+that assumes a Wright-Fisher population with changing size through time
+to a different population model,
+the inferred demographic history will match the ``effective population size history''
+defined in this way.
+These observations and the following calculations are standard in coalescent theory
+\citep[see e.g.,][]{wakeley2005coalescent},
+but they are provided here for completeness.
 
@@
 and are furthermore both in location $z$,
-and let $1/(2 N_e(z,t))$ be the rate of coalescence in location $z$ at the time.
+and let $N_e(z,t)$ be the (effective) diploid population size in location $z$ at the time,
+so that $1/(2 N_e(z,t))$ is the rate of coalescence there.
 Then, we compute the mean coalescence rate as
 $$  r(t)  = \frac{1}{p(t)} \sum_z \frac{p(z,t)}{2N_e(z,t)} . $$
-This follows because if we have $n$ diploid samples, and hence $\binom{2n}{2}$ lineages,
-the expected number of coalescences in location $z$ between times $t$ and $t+dt$ ago
+This follows because if we have $m$ diploid samples, and hence $\binom{2m}{2}$ lineages,
+the expected number of coalescences in location $z$ between times $t$ and $t+dt$ ago is
 $$
-\binom{2n}{2} p(z,t) \frac{ dt }{ 2 N_e(z,t) },
+\binom{2m}{2} p(z,t) \frac{ dt }{ 2 N_e(z,t) },
 $$
@@
 $$
-\binom{2n}{2} p(t) .
+\binom{2m}{2} p(t) .
 $$
@@
 This is relatively straightforward
-using the general theory of Markov chains, and is implemented in \texttt{msprime}.
+using the general theory of Markov chains \citep[e.g,.][]{kemeny2012denumerable},
+and is implemented in \texttt{msprime}.
 
@@
 
+The coalescence rate trajectories can be computed from a model in \texttt{msprime}
+using the \texttt{coalescence\_rate\_trajectory} method of the Demography Debugger class,
+which can be obtained from a \stdpopsim model using the
+\texttt{model.get\_demography\_debugger()} method.
+
 \end{document}

### 2020-04-20 popSimManu.tex
removed "ancient samples"

@@
 gene conversion, mutation models (rate heterogeneity),
-ancient samples,
 more realistic demography (overlapping generations, separate sexes, mortality/fecundity schedules),

### 2020-04-21 response-to-reviewers.md

@@
 
-----
+# Editors comments
 
-> # Editors comments
 > 1) Ensuring that the language about the models & parameters conveys the correct sense of uncertainty.
 Obviously the reviewers and the authors know that these just represent best guesses at the moment, but as this platform catches on we don't want these numbers & models to be seen as gospel.
->
 
 We fully agree that the language in the original manuscript was not careful enough in conveying uncertainty about the models and parameters we discuss.
-To address this issue, we have made several edits throughout the manuscript trying to better convey this uncertainty.
-See our response to reviewer 1 below for examples.
+To address this, we have made edited the text throughout the manuscript trying to better convey this uncertainty.
+See our response to Reviewer 1 below for examples.
+We have also added a bit more to the discussion (below "next steps") on this point.
 
-> 2) We would like to see the SLIM integration demonstrated with an application that MsPrime doesn't cover.
+> 2) We would like to see the SLIM integration demonstrated with an application that msprime doesn't cover.
 This could be, for example, a figure of the average value some summary statistics surrounding a selective under the human demographic model.
-The reviewers didn't want this to be a lot of work, e.g.
-it would be outside of the scope of the current paper to demonstrate the power of a range of selection methods.
+The reviewers didn't want this to be a lot of work, e.g., it would be outside of the scope of the current paper to demonstrate the power of a range of selection methods.
 However, we felt that a simple demonstration of the functionality would substantially increase the scenarios easily available and imaginable by readers of the paper.
@@
 
-Issue #164. (petrelharp)
+Good idea -
+we have expanded the section of the discussion ("next steps")
+with more detail on our future plans.
+We're trying to walk the line between being forward-looking and promising too much.
 
@@
 
-TODO: discussion other applications
-petrelharp
+This is an important point - we have (naturally) focused on methods benchmarking,
+but our impact could well be greater if stdpopsim becomes widely used among empirical researchers
+(for whom on average running realistic simulations presents a greater barrier than to developers of computational methods).
+We've added a paragraph about this to the Discussion
+and additional words to the Abstract.
 
 > Overall we are very excited to see this big move forward and welcome the authors' careful work in this important area.
->
+
+Thanks for the kind words!
+We are enthusiastic about it as well.
 
@@
 
->
 > The manuscript describes a community effort to standardize population genetic simulations and it presents an example of the resource's utility for method's testing.
@@
 > Some of the specific features that are nice about the work are the incorporation of inferred genetic maps where possible, the automated output of citation information, the use of long-term stable identifiers for models and genetic maps, and the outputting of provenance information in the tree sequence files.
->
->
-> ## Major comments
->
+
+Thank you!
+
+## Major comments
+
 > 1. Emphasize parameters are current best-guesses:
@@
 
-We thank the reviewer for bringing up this important point and fully agree that the language in the manuscript needs to better emphasize the uncertainty in our current parameter estimates.
+We thank the reviewer for bringing up this important point
+and fully agree that the language in the manuscript needs to better emphasize the uncertainty in our current parameter estimates.
 To address this, we have made several changes throughout the manuscript, including the passages cited above:
 
-On page 3, it now reads:
-"For each organism, the catalog contains information on our current understanding of the physical organization (e.g., chromosome structure)
-of its genome, one or more inferred genetic maps, default population-level parameters (mutation rate and generation time estimates)
-and one or more published demographic models. These available models
-and parameters are meant to represent the current standards in the field, however our
-intention is that this resource will evolve as new results become available"
+The first quoted passage now reads:
+"For each organism, the catalog contains information on our current understanding of
+the physical organization (e.g., chromosome structure) of its genome,
+one or more inferred genetic maps (except for \textit{E.~coli}),
+default population-level parameters (mutation rate and generation time estimates).
+The catalog also contains numerous published demographic models for the first three species.
+These available models and parameters are meant to represent the field's current understanding,
+and we intend for this resource to evolve as new results become available."
 
-On page 4, it now reads:
-"Firstly, the library defines
-some basic information on our current understanding of each
+The second passage now reads:
+" Firstly, the library defines some basic information about our current understanding of each 
 species' genome, including information about chromosome
-lengths, average mutation rate estimates, and generation times. We also provide access to detailed
-empirical information such as inferred genetic maps, which model observed
-heterogeneity in recombination rate along chromosomes."
+lengths, average mutation rate estimates, and generation times.
+We also provide access to detailed empirical information such as inferred genetic maps,
+which model observed heterogeneity in recombination rate along chromosomes."
+
+(The reviewer's suggestion is good, but we think that "current understanding"
+conveys roughly the same meaning as "the latest reference genome" in less words,
+and without making promises about genome assemblies.)
 
@@
 We made the suggested change:
-"Importantly, we have rigorous quality control methods to ensure that models
-are correctly implemented with respect to their original publication
+"Importantly, we have rigorous quality control methods to ensure that we have
+correctly implemented the models as described in their original publication
 and have documented methods for others to contribute new modules."
@@
 It would be wonderful to have a comments section for the models that would be either curated by a set of editors or crowd-sourced.
-I say this because overtime, models will proliferate, and some will come to be regarded as out-of-date.
+I say this because over time, models will proliferate, and some will come to be regarded as out-of-date.
 One can imagine those approaching the field afresh will be overwhelmed by the possible selections and possible implement models that become outdated.
@@
 
-Response summary:
 Thank you for this forward-thinking comment.
-Good idea but how would you actually do it?
-We will have to let this emerge, and can't solve the problem until we actually have it.
+We considered various ways to do this
+(e.g., by enabling the wiki associated with the github repository:
+\url{https://github.com/popsim-consortium/stdpopsim/issues/415}),
+but ended up deciding that (a) it's not yet clear what the right way to do this is,
+and (b) it'll be significant work to do right that we aren't prepared to do at the moment.
+Since we haven't encountered the problem of out of date models yet,
+we feel happy about postponing this until the problem arises.
 
@@
 
-We're not going to do this either;
-Issue #164.
-Emphasize importance of selection in the discussion.
+We would like to see this also!
+But, we felt that the amount of work it took to get to this point
+made this a good point to release the publication.
+Furthermore, discussing selection would distract from the current relatively simple focus
+on inference of demographic history.
+We definitely intend to include selection (and expand the available generic models) in future work,
+however, as we now discuss in greater detail in the Discussion ("next steps").
 
@@
 
->
->
-> ## Minor comments
->
+## Minor comments
+
 > 1. Is the provenance information only provided if msprime is used to produce tree sequence files? Is a similar functionality in place with the SLiM back-end?
 
-FIXME
+SLiM also records sufficient information to reconstruct the simulation in the tree sequence provenance
+(including the software version, random seed, and a copy of the simulation script).
 
->
 > 2. Little detail is given on the system of identifiers/versioning for models, maps, etc.
 
-FIXME
->
+True, but the details are rather pedantic, and we provide a link
+to our documentation where we describe the system in detail.
+
 > 3. In the abstract, it's left vague what the nature of the resource is.
-
-FIXME
 >
@@
 
-FIXME
+Good point - we've clarified the nature of the resource,
+but haven't included the URL since googling "stdpopsim" features the correct page as the top hit.
 
->
 > 4. Page 3: "bespoke simulations":
@@
 
-FIXME: Maybe make this wording more descriptive and precise.
+We've replaced "bespoke" with "coded from scratch", which we think is more descriptive,
+and added a parenthetical note about command-line calls in supplementary material (which is a good point).
+We hope that the other points are made elsewhere.
 
->
 > 5.
@@
 
->
 > 6. Page 6. Typo in "time or writing". Change to "time of writing"
 
-We corrected this typo.
+Thanks for spotting this typo! We corrected it.
 
->
 > 7.
@@
 
->
 > 8. Figure 4.
@@
 We agree that the orientation of this figure is confusing, and so have changed it as suggested.
-In addition, we had previously plotted divergence times here as vertical lines, however we felt that plotting it this way made the figure hard to read since there was a great deal of overlap between these lines and the N_t lines on the figure, which was was especially difficult to read when including multiple replicates.
+In addition, we had previously plotted divergence times here as vertical lines, however we felt that plotting it this way made the figure hard to read since there was a great deal of overlap between these lines and the N(t) lines on the figure, which was was especially difficult to read when including multiple replicates.
 Thus, we prefer to keep the lower divergence times panel for improved readability.
 
->
 > 9.
@@
 
-We changed the wording here as follows:
+Good point - we changed the wording here as follows:
 
@@
 
->
-> 10. Page 31. This is all very nice. To save readers a step, could you include the command-line flag for how to have msprime output the coalescent rate function?
->
+> 10. Page 31. This is all very nice. To save readers a step, could you include the command-line flag for how to have msprime output the coalescence rate function?
 
-FIXME: Peter could you add this?
+Thanks!
+There is not a command-line flag for obtaining the coalescence rates,
+but this is easily obtainable in python.
+We've added a paragraph to the supplement explaining what methods to use.
 
@@
 Across all simulated scenarios (except under more complex models), the simulations seem relatively accurate.
->
+
+Thanks for the encouragement!
+
 > Despite a little hiccup with python version mismatch prior to me successfully installing stdpopsim, I was able to successfully get the tutorials running within minutes after.
@@
 
-We have added in a basic `stdpopsim` API and CLI example in a Jupyter Notebook that can be accessed and used interactively via Binder and encourage users to try out the tutorials there.
-We have linked to the Binder in the README on the GitHub.
-See https://mybinder.org/v2/gh/popsim-consortium/stdpopsim/master?filepath=stdpopsim_example.ipynb
+Good idea -
+we have reorganized the Tutorials and added to them,
+and have also have added in a basic `stdpopsim` API and CLI example in a Jupyter Notebook that can be accessed and used interactively via Binder and encourage users to try out the tutorials there.
+We have linked to the Binder in the README on the GitHub. See https://mybinder.org/v2/gh/popsim-consortium/stdpopsim/master?filepath=stdpopsim_example.ipynb
 
->
 > I thoroughly enjoyed reading this manuscript, and learning of all the new features that have been written into stdpopsim, and believe that this will be an invaluable contribution to the population genomics community.
 
-Thank YOU!
+Thank YOU! We hope you are right.
 
->
-> ## Minor comments:
->
+## Minor comments:
+
 > The authors mention the implementation of the SLiM package into stdpopsim, but I was unable to find any details of this either in the manuscript or on the readthedocs page.
 If this was indeed implemented, I would also like to see the authors add some tutorials/results of analyses based on SLiM simulated data.
->
+
+FIXME add description of what we added to the manuscript here.
+
 > "Adding further maps to the library is trivial" - discuss - since the audience for this is the broader pop-gen community, perhaps pointing them to the detailed development environment on readthedocs would be helpful.
->
+
+We've changed "trivial" to "straightforward", which is more accurate.
+Unfortunately, we don't have documentation written for this --
+currently, the suggestion would be to look at how existing maps work and copy that (which should in fact be straightforward) --
+but it's on our list of things to do to document this.
 
@@
 
->
-> ## Summary
->
->
+## Summary
+
 > The authors present a standardized framework for creating reproducible population genetic simulations.
@@
 I have a few comments and some typos, etc.
->
->
-> ## Comments
->
+
+Thanks! We hope it is useful to you and your students.
+
+## Comments
+
 > 1) The authors mention that SLiM can be used as an alternative backend, which would presumably allow for simulations with selection.
@@
 
+FIXME: fill this in.
 Issue #150 and #164, as above.
 
->
 > 2) I like the inclusion of the "zigzag" history, as well as generic piecewise constant models and IM models (page 7).
@@
 
-Summary of response:
-This is done, in the documentation now.
-But note that zigzag is in human because the parameterization is human-ish.
+FIXME: maybe we should do this?
+
+Note that zigzag is in human because the parameterization is human-ish.
 
@@
 
-Issue #168 (petrelharp)
+We see why this is confusing.
+We introduced this notation (R, C, and n) to make it completely clear
+in this paragraph what exactly was being simulated,
+but we feel that continuing to use this notation elsewhere would actually obscure things,
+since we don't do any calculations with these quantities.
+We have changed the "n" in the Supplement to an "m".
 
->
-> 4) "Calculating coalescence rates" section needs a read through.
-Reword first sentence and add some citations (especially regarding computing p(t) and p(z,t)).
-It was unclear to me how the "mean coalescence times" were used (the rate was used to compute the ground truth over time).
-This section is also referred to as the Appendix in the main text.
+> 4) "Calculating coalescence rates" section needs a read through. Reword first sentence and add some citations (especially regarding computing p(t) and p(z,t)).
+> It was unclear to me how the "mean coalescence times" were used (the rate was used to compute the ground truth over time). This section is also referred to as the Appendix in the main text.
 
-Issue #169 (petrelharp)
+We've expanded the first sentence substantially, and added a few more citations.
+We apologize that we don't know of a paper to cite that does precisely the same calculations
+(but don't doubt that such a paper exists),
+and instead refer only vaguely to "general theory of Markov chains" (but now with a citation).
+The section is now explicitly labeled "Appendix".
+Hopefully this is more clear now.
 
->
->
 > ## Minor Comments
->
-> Typos, Wording, etc
->
+
 > For a future draft, line numbers would be very helpful.
+
+Done, and sorry we omitted this the first time around.
+
+> -Abstract: "possibility for error" -> inconsistency
 
@@
 
->
-> -Abstract: "possibility for error" -> inconsistency
-
-We made the suggested change.
->
 > -Page 5: trivial -> straightforward
 
-We made the suggested change.
+Good point: this was probably inaccurate. We made the suggested change.
 
->
 > -"tree sequence": make it clear that users can easily obtain traditional variant matrices from this format, in case they are unfamiliar
@@
 "The tree sequence format could also be converted
-to other formats (e.g., vcf) by the user if desired. "
->
+to other formats (e.g., VCF) by the user if desired."
+
 > -Page 6: "time or writing" -> of
 
-We corrected this typo.
+Whoops: we corrected this typo.
 
->
 > -Page 8: "While there is variation in accuracy among methods, populations, and individual replicates, the methods are generally accurate for this model of human history." I'm not sure I agree here (although perhaps relative to other models this statement is true).
@@
 
-FIXME: could someone add more discussion on this?
+"Accurately" is subjective -- we're sure the reviewer can imagine much worse inferences! --
+but in any case we've rephrased this bit to say
+"While there is variation in accuracy among methods, populations, and individual replicates,
+the methods generally infer back a good estimate
+of the true effective population sizes of the simulations --
+inferred values are mostly within a factor of two of the truth,
+and many of the methods infer a bottleneck at the correct time."
 
->
 > -Page 8, paragraph "Stdpopsim allows...": reword this first sentence
@@
 
->
 > -Figure 3 caption: wrong citation, should be Sheehan and Song
 
-FIXME: I think the citation is correct here (for the genetic map), maybe the issue is that the demographic model isnt actually cited? Should we add that in?
+The citation was for the genetic map, but we've added the citation to Sheehan and Song for the model (also in the caption of Figure 2).
 
->
 > -Page 10, last sentence before "Multi-population demographic models" section: clarify what "this setting" is (both species I assume)
@@
 
->
 > -Figure 4 captain: "or smc++" -> "and smc++" (some other figure captions have this too), I would also say "$x$-axis"
@@
 We made the suggested changes.
->
+
 > -Page 12, first paragraph: include a reference where people could learn more about the effective population size vs. census size, random-mating models, etc
@@
 We thank the reviewer for this suggestion.
-However, we have realized that our use of "census size" here to describe the "true" simulated population size was a bit misleading, as our intention was not to characterize these as actual "census" (as opposed to "effective") population sizes.
-To avoid this confusion, we have changed the manuscript to now describe them as the "simulated population size".
-For example, this section now reads:
+However, we have realized that our use of "census size" here to describe the "true" simulated population size was a bit misleading,
+as our intention was not to characterize these as actual "census" (as opposed to "effective") population sizes.
+To avoid this confusion, we have changed the manuscript to now describe them as the "simulated population size". For example, this section now reads:
 
@@
 
-As a result, we dont necessarily feel that referring to the census vs effective population size literature would be appropriate here.
+We have, however, added more discussion of this point (with a citation to a review)
+to the start of the Appendix.
 
->
 > -Page 12 (and elsewhere): "By contrast" -> "In contrast"(?) I think both are technically correct though
@@
 We changed all instances of "by contrast" to "in contrast".
->
+
 > -Page 12: sentence beginning with "Figure S4 shows..." is long, perhaps breakup or rephrase ("bottleneck into a European population" sounds odd to me.)
@@
 post-divergence gene flow between the African and European population (Figure S4A)."
->
+
 > -Page 12: "Europe population" -> "European population"
@@
 We made the suggested change.
->
+
 > -Page 12: S7, S4, S6 are mentioned in a different order than they appear in the supplement (I understand why but maybe revisit this part)
@@
 
->
 > -Page 13: "overfit the YRI history" do you mean overfit in a technical sense?
 
-FIXME: Could someone weigh in on what was originally meant here?
+Really, we just mean that the inferred history has more population size changes
+than the truth did;
+we've change this to "overcomplicate the simple YRI history".
 
->
 > -Page 14: "Method quality control" section.
@@
 
-(B would have access - it's not blinded - they just aren't supposed to look at it)
+We've clarified this to say that "Developer B (perhaps found by requesting review from the broader Consortium)".
+The process is not blinded in any way -- developer B *could* look at developer A's implementation --
+but we don't think this will be a problem in practice.
 
->
 > -Page 16, line 6: "population 1 after"?
@@
 
-We corrected this mistake.
+Thanks! We corrected this.
 
->
 > -Figure S2 caption: move "and" to before MSMC (in a few other figure captions too)
@@
 
->
 > -Figure S6: explain x-axis "MIG_AF_EU" and "MIG_EU_AF". I assume these are directional migration rates
@@
 
->
 > -Figure S9: unreadable when printed (not sure it adds a lot)
->
 

### 2020-04-23 src/plot_AraTha_African2Epoch_1H18.R
added new figs edited old figs

@@ -33,36 +33,36 @@ tab0A = subset(tab0,method == method1_id)
 tab0B = subset(tab0,method == method2_id)
 tab0C = subset(tab0,method == method3_id)
 tab0D = subset(tab0,method == method4_id)
+tab0Y = subset(tab0,method == "census")
 tab0Z = subset(tab0,method == "coal")
 
 speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
-
 p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
-  geom_step(data=tab0Z,aes(x=x, y=y),size=2.5,alpha=1.0, col=truth_color) +
-  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
+  geom_step(data=tab0Z,aes(x=x, y=y),size=3.3,alpha=1.0, col=truth_color) +
+  geom_step(size=2, alpha=0.9, show.legend = FALSE) +
   scale_x_continuous(trans='log10') +
   scale_y_continuous(trans='log10') +
   annotation_logticks(sides="b") +
   annotation_logticks(sides="l") +
   scale_color_manual(values=replicate_colors) +
   guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
-  theme_classic(base_size = 18) +
+  theme_classic(base_size = 24) +
   labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
   theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
 
 
 p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
-  geom_step(data=tab0Z,aes(x=x, y=y),size=2.5,alpha=1.0, col=truth_color) +
-  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
+  geom_step(data=tab0Z,aes(x=x, y=y),size=3.3,alpha=1.0, col=truth_color) +
+  geom_step(size=2, alpha=0.9, show.legend = FALSE) +
   scale_x_continuous(trans='log10') +
   scale_y_continuous(trans='log10') +
   annotation_logticks(sides="b") +
[…]

### 2020-04-23 src/plot_HomSap_OutOfAfricaArchaicAdmixture_5R19_census_vs_IRC.R

@@ -1,0 +1,91 @@
+library(ggplot2)
+library(cowplot)
+
+## define variables
+infile_pop0="../data/HomSap_OutOfAfricaArchaicAdmixture_5R19_pop0_sizes.txt"
+infile_pop1="../data/HomSap_OutOfAfricaArchaicAdmixture_5R19_pop1_sizes.txt"
+infile_pop2="../data/HomSap_OutOfAfricaArchaicAdmixture_5R19_pop2_sizes.txt"
+outfile="../display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19_census_vs_IRC.pdf"
+species="Homo sapiens"
+pop0_name="(YRI)"
+pop1_name="(CEU)"
+pop2_name="(CHB)"
+IRC="#bd0026"
+census="black"
+
+
+## plot
+tab0 <- read.table(file=infile_pop0, h=T ,sep="\t")
+tab1 <- read.table(file=infile_pop1, h=T ,sep="\t")
+tab2 <- read.table(file=infile_pop2, h=T ,sep="\t")
+
+
+tab0Y = subset(tab0,method == "census")
+tab0Z = subset(tab0,method == "coal")
+
+
+tab1Y = subset(tab1,method == "census")
+tab1Z = subset(tab1,method == "coal")
+
+
+tab2Y = subset(tab2,method == "census")
+tab2Z = subset(tab2,method == "coal")
+
+
+xMin = 10
+xMax = max(c(tab0$x,tab1$x,tab2$x))
+yMin = min(tab0Y$y, tab1Y$y, tab2Y$y)
+yMax = max(tab0Y$y, tab1Y$y, tab2Y$y)
+
+
+speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
+p0A<-ggplot() +
+  coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
+  geom_step(data=tab0Y,aes(x=x, y=y),size=3,alpha=0.9, col=census, linetype=1) +
+  geom_step(data=tab0Z,aes(x=x, y=y),size=3,alpha=0.7, col=IRC) +
+  scale_x_continuous(trans='log10') +
+  scale_y_continuous(trans='log10') +
+  annotation_logticks(sides="b") +
+  annotation_logticks(sides="l") +
+  theme_classic(base_size = 24) +
+  labs(x="", y="", title="", subtitle=pop0_name, colour="", fill="")+
[…]

### 2020-04-23 src/plot_HomSap_OutOfAfrica_3G09.R

@@ -1,0 +1,283 @@
+library(ggplot2)
+library(cowplot)
+
+## define variables
+infile_pop0="../data/HomSap_OutOfAfrica_3G09_pop0_sizes.txt"
+infile_pop1="../data/HomSap_OutOfAfrica_3G09_pop1_sizes.txt"
+infile_pop2="../data/HomSap_OutOfAfrica_3G09_pop2_sizes.txt"
+outfile="../display_items/HomSap_OutOfAfrica_3G09.pdf"
+species="Homo sapiens"
+pop0_name="YRI"
+pop1_name="CEU"
+pop2_name="CHB"
+method1_name="SMC++"
+method2_name="stairway plot"
+method3_name="MSMC (n = 2)"
+method4_name="MSMC (n = 8)"
+method1_id="smcpp"
+method2_id="sp"
+method3_id="msmc_2"
+method4_id="msmc_8"
+truth_color="gray30"
+replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
+
+## plot
+cPal_legend<-c(truth_color, replicate_colors)
+
+tab0 <- read.table(file=infile_pop0, h=T ,sep="\t")
+tab1 <- read.table(file=infile_pop1, h=T ,sep="\t")
+tab2 <- read.table(file=infile_pop2, h=T ,sep="\t")
+
+tab0$x[tab0$x==0.0]<-1
+tab1$x[tab1$x==0.0]<-1
+tab2$x[tab2$x==0.0]<-1
+
+xMin = 10
+xMax = max(c(tab0$x,tab1$x,tab2$x))
+y0Min = min(tab0$y)
+y0Max = max(tab0$y)
+y1Min = min(tab1$y)
+y1Max = max(tab1$y)
+y2Min = min(tab2$y)
+y2Max = max(tab2$y)
+
+tab0A = subset(tab0,method == method1_id)
+tab0B = subset(tab0,method == method2_id)
+tab0C = subset(tab0,method == method3_id)
+tab0D = subset(tab0,method == method4_id)
+tab0Y = subset(tab0,method == "census")
+tab0Z = subset(tab0,method == "coal")
+
+tab1A = subset(tab1,method == method1_id)
+tab1B = subset(tab1,method == method2_id)
+tab1C = subset(tab1,method == method3_id)
+tab1D = subset(tab1,method == method4_id)
+tab1Y = subset(tab1,method == "census")
+tab1Z = subset(tab1,method == "coal")
+
[…]

### 2020-04-23 src/plot_HomSap_constant_size.R

@@ -1,0 +1,111 @@
+library(ggplot2)
+library(cowplot)
+
+## define variables
+infile="../data/HomSap_constant_size.txt"
+outfile="../display_items/HomSap_constant_size.pdf"
+species="Homo sapiens"
+pop0_name="Constant size"
+method1_name="SMC++"
+method2_name="stairway plot"
+method3_name="MSMC (n = 2)"
+method4_name="MSMC (n = 8)"
+method1_id="smcpp"
+method2_id="sp"
+method3_id="msmc_2"
+method4_id="msmc_8"
+truth_color="gray30"
+replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
+
+## plot
+cPal_legend<-c(truth_color, replicate_colors)
+
+tab0 <- read.table(file=infile, h=T ,sep="\t")
+
+tab0$x[tab0$x==0.0]<-1
+
+xMin = 10
+xMax = max(tab0$x)
+yMin = min(tab0$y)
+yMax = max(tab0$y)
+
+tab0A = subset(tab0,method == method1_id)
+tab0B = subset(tab0,method == method2_id)
+tab0C = subset(tab0,method == method3_id)
+tab0D = subset(tab0,method == method4_id)
+#tab0Y = subset(tab0,method == "census")
+tab0Z = subset(tab0,method == "coal")
+
+speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
+p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
+  coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
+  geom_step(data=tab0Z,aes(x=x, y=y),size=3.3,alpha=1.0, col=truth_color) +
+  geom_step(size=2, alpha=0.9, show.legend = FALSE) +
+  scale_x_continuous(trans='log10') +
+  scale_y_continuous(trans='log10') +
+  annotation_logticks(sides="b") +
+  annotation_logticks(sides="l") +
+  scale_color_manual(values=replicate_colors) +
+  guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
+  theme_classic(base_size = 24) +
+  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
+  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
+
+
[…]

### 2020-04-23 src/plot_d_mel_Sheehan_mask2.R

@@ -1,112 +1,0 @@
-library(ggplot2)
-library(cowplot)
-
-## define variables
-infile="../data/d_mel_Sheehan_mask2.txt"
-outfile="../display_items/d_mel_Sheehan_mask2.pdf"
-species="Drosophila melanogaster"
-pop0_name="Africa"
-method1_name="SMC++"
-method2_name="stairway plot"
-method3_name="MSMC (n = 2)"
-method4_name="MSMC (n = 8)"
-method1_id="smcpp"
-method2_id="sp"
-method3_id="msmc_2"
-method4_id="msmc_8"
-truth_color="gray30"
-replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
-
-## plot
-cPal_legend<-c(truth_color, replicate_colors)
-
-tab0 <- read.table(file=infile, h=T ,sep="\t")
-
-tab0$x[tab0$x==0.0]<-1
-
-xMin = 25
-xMax = max(tab0$x)
-yMin = min(tab0$y)
-yMax = max(tab0$y)
-
-tab0A = subset(tab0,method == method1_id)
-tab0B = subset(tab0,method == method2_id)
-tab0C = subset(tab0,method == method3_id)
-tab0D = subset(tab0,method == method4_id)
-tab0Z = subset(tab0,method == "coal")
-
-speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
-
-p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
-  coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
-  geom_step(data=tab0Z,aes(x=x, y=y),size=2.5,alpha=1.0, col=truth_color) +
-  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
-  scale_x_continuous(trans='log10') +
-  scale_y_continuous(trans='log10') +
-  annotation_logticks(sides="b") +
-  annotation_logticks(sides="l") +
-  scale_color_manual(values=replicate_colors) +
-  guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
-  theme_classic(base_size = 18) +
-  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
-  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
-
-
[…]

### 2020-04-23 src/plot_homo_sapiens_constant.R

@@ -1,112 +1,0 @@
-library(ggplot2)
-library(cowplot)
-
-## define variables
-infile="../data/homo_sapiens_constant.txt"
-outfile="../display_items/homo_sapiens_constant.pdf"
-species="Homo sapiens"
-pop0_name="constant size"
-method1_name="SMC++"
-method2_name="stairway plot"
-method3_name="MSMC (n = 2)"
-method4_name="MSMC (n = 8)"
-method1_id="smcpp"
-method2_id="sp"
-method3_id="msmc_2"
-method4_id="msmc_8"
-truth_color="gray30"
-replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
-
-## plot
-cPal_legend<-c(truth_color, replicate_colors)
-
-tab0 <- read.table(file=infile, h=T ,sep="\t")
-
-tab0$x[tab0$x==0.0]<-1
-
-xMin = 10
-xMax = max(tab0$x)
-yMin = min(tab0$y)
-yMax = max(tab0$y)
-
-tab0A = subset(tab0,method == method1_id)
-tab0B = subset(tab0,method == method2_id)
-tab0C = subset(tab0,method == method3_id)
-tab0D = subset(tab0,method == method4_id)
-tab0Z = subset(tab0,method == "coal")
-
-speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
-
-p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
-  coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
-  geom_step(data=tab0Z,aes(x=x, y=y),size=2.5,alpha=1.0, col=truth_color) +
-  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
-  scale_x_continuous(trans='log10') +
-  scale_y_continuous(trans='log10') +
-  annotation_logticks(sides="b") +
-  annotation_logticks(sides="l") +
-  scale_color_manual(values=replicate_colors) +
-  guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
-  theme_classic(base_size = 18) +
-  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
-  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
-
-
[…]

### 2020-04-23 src/plot_homo_sapiens_mask_Ragsdale.R

@@ -1,280 +1,0 @@
-library(ggplot2)
-library(cowplot)
-
-## define variables
-infile_pop0="../data/homo_sapiens_mask_Ragsdale_pop0.txt"
-infile_pop1="../data/homo_sapiens_mask_Ragsdale_pop1.txt"
-infile_pop2="../data/homo_sapiens_mask_Ragsdale_pop2.txt"
-outfile="../display_items/homo_sapiens_mask_Ragsdale.pdf"
-species="Homo sapiens"
-pop0_name="YRI"
-pop1_name="CEU"
-pop2_name="CHB"
-method1_name="SMC++"
-method2_name="stairway plot"
-method3_name="MSMC (n = 2)"
-method4_name="MSMC (n = 8)"
-method1_id="smcpp"
-method2_id="sp"
-method3_id="msmc_2"
-method4_id="msmc_8"
-truth_color="gray30"
-replicate_colors<-c("#66c2a4","#41b6c4","#225ea8")
-
-## plot
-cPal_legend<-c(truth_color, replicate_colors)
-
-tab0 <- read.table(file=infile_pop0, h=T ,sep="\t")
-tab1 <- read.table(file=infile_pop1, h=T ,sep="\t")
-tab2 <- read.table(file=infile_pop2, h=T ,sep="\t")
-
-tab0$x[tab0$x==0.0]<-1
-tab1$x[tab1$x==0.0]<-1
-tab2$x[tab2$x==0.0]<-1
-
-xMin = 10
-xMax = max(c(tab0$x,tab1$x,tab2$x))
-y0Min = min(tab0$y)
-y0Max = max(tab0$y)
-y1Min = min(tab1$y)
-y1Max = max(tab1$y)
-y2Min = min(tab2$y)
-y2Max = max(tab2$y)
-
-tab0A = subset(tab0,method == method1_id)
-tab0B = subset(tab0,method == method2_id)
-tab0C = subset(tab0,method == method3_id)
-tab0D = subset(tab0,method == method4_id)
-tab0Z = subset(tab0,method == "coal")
-
-tab1A = subset(tab1,method == method1_id)
-tab1B = subset(tab1,method == method2_id)
-tab1C = subset(tab1,method == method3_id)
-tab1D = subset(tab1,method == method4_id)
-tab1Z = subset(tab1,method == "coal")
-
-tab2A = subset(tab2,method == method1_id)
-tab2B = subset(tab2,method == method2_id)
[…]

### 2020-04-27 popSimManu.tex
adding acknowledgement to 2014 meeting
John Novembre reminded me and Dan that we should probably acknowledge SimBank. I agree and so have added that edit to this PR.

@@
 as an alternative backend,
-to allow simulation of processes that cannot be modeled under the coalescent.
+to allow simulation of processes that cannot be modeled under the coalescent. We validated our
+implementation of the \texttt{SLiM} engine by comparing estimates of several population
+genetic summary statistics for simulations generated by both engines 
+(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
 
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
+\includegraphics[width=0.8\linewidth]{display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
@@
 In each row we show comparisons among the methods (including two sample sizes for \MSMC).
-Blue lines show estimates from each of three replicate whole genome simulations and (FIXME: what color?)
-lines show the simulated population sizes.
+Blue lines show estimates from each of three replicate whole genome simulations. 
 When fitting single-population models to data simulated for multiple
@@
 For this reason,
-we also plot the inverse coalescence rates calculated analytically in \texttt{msprime}
-as another benchmark of the ``truth'' in the single-population inference setting. 
-See the Appendix for a precise definition and description of the computation.
+we plot the inverse coalescence rates (black lines) calculated analytically in \texttt{msprime}
+as a benchmark of the ``truth'' in the single-population inference setting. However, 
+in this case inverse coalescence rates are very close to census population sizes 
+(Figure \ref{fig:census_vs_IRC}).
+See the Appendix for a precise definition and description of the inverse coalescence rate computation.
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
+\includegraphics[width=0.8\linewidth]{display_items/DroMel_African3Epoch_1S16.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in \textit{Drosophila}}. Population
@@
 Dmitri Petrov, and Sohini Ramachandran. 
+In addition we would like to thank NESCENT and Matt Hahn, Victoria Sork, and Michael Whitlock for organizing a 2014
+catalysis meeting in which many of the goals of this effort were first laid out. 
 CCK and KEL were funded under NIH Award R35GM119856.
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.png}
+\includegraphics[width=0.8\linewidth]{display_items/HomSap_AE9K19_chr22_map_AI.pdf}
+\caption{\textbf{Validating the \texttt{SLiM} engine backend under a genetic map}.
+Here we validate our integration of the \texttt{SLiM} \citep{haller2019tree,haller2019slim} engine backend. 
+We show quantile-quantile plots between \texttt{SLiM} and \texttt{msprime} engines for three 
+population genetic summary statistics: \textit{r}\textsuperscript{2}, Tajima's $\pi$, and Tajima's D. Additionally, we 
+show runtimes for generating each simulation replicate. Data were generated by simulating 100 replicates 
+of human chromosome 22 under the AncientEurasia\_9K19 model \citep{kamm2019efficiently} and using the
+HapMapII\_GRCh37 genetic map \citep{international2007second}. From top to bottom we show results 
+using population-size rescaling for three scaling factors: Q=1, Q=10, and Q=100.
+}
+\label{fig:slim_val_map}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/HomSap_AE9K19_chr22_nomap_AI.pdf}
+\caption{\textbf{Validating the \texttt{SLiM} engine backend under uniform recombination}.
+Here we validate our integration of the \texttt{SLiM} \citep{haller2019tree,haller2019slim} engine backend. 
+We show quantile-quantile plots between \texttt{SLiM} and \texttt{msprime} engines for three 
+population genetic summary statistics: \textit{r}\textsuperscript{2}, Tajima's $\pi$, and Tajima's D. Additionally, we 
+show runtimes for generating each simulation replicate. Data were generated by simulating 100 replicates 
+of human chromosome 22 under the AncientEurasia\_9K19 model \citep{kamm2019efficiently} and using a uniform 
+rate of recombination across the chromosome. From top to bottom we show results 
+using population-size rescaling for three scaling factors: Q=1, Q=10, and Q=100.
+}
+\label{fig:slim_val_nomap}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19_census_vs_IRC.pdf}
+\caption{\textbf{Comparing census population sizes and inverse coalescence rates in humans}.
+Data are shown from human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model \citep{ragsdale2019models} 
+and using the HapMapII\_GRCh37 genetic map \citep{international2007second}. 
+From left to right we show sizes for each of the three populations in the model: YRI, CEU, and CHB.
+We plot the census sizes for each population in black, and in red we plot 
+inverse coalescence rates as calculated from the demographic model used for simulation (see text).
+}
+\label{fig:census_vs_IRC}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.8\linewidth]{display_items/HomSap_OutOfAfrica_3G09.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
@@
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
-inferred in \cite{international2007second}. From top to bottom we show estimates for each
-of the three populations in the model: YRI, CEU, and CHB. In shades of blue we show the estimated
+replicate human genomes under the OutOfAfrica\_3G09 model \citep{gutenkunst2009inferring} 
+and using the HapMapII\_GRCh37 genetic map \citep{international2007second}. 
+From top to bottom we show estimates for each of the three populations in the model: 
+YRI, CEU, and CHB. In shades of blue we show the estimated
 $N(t)$ trajectories for each replicate.
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.png}
+\includegraphics[width=0.8\linewidth]{display_items/HomSap_constant_size.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
@@
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate human genomes under a constant sized population model with $N=10^4$ and using the
-HapMapII genetic map~\citep{international2007second}.
+replicate human genomes under a constant sized population model with $N=10^4$ 
+and using the HapMapII\_GRCh37 genetic map \citep{international2007second}.
 As a proxy for the ``truth'', in black we show inverse coalescence rates
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{display_items/AraTha_African2Epoch_1H18.png}
+\includegraphics[width=0.8\linewidth]{display_items/AraTha_African2Epoch_1H18.pdf}
 \caption{\textbf{Comparing estimates of $N(t)$ in \textit{A.~thaliana}}. Here we show estimates of population
@@
 \MSMC with $n=2$ and $n=8$. Data were generated by simulating
-replicate \textit{A.~thaliana} genomes under the African2Epoch\_1H18 model and using
-the genetic map of \cite{salome2011recombination}.
+replicate \textit{A.~thaliana} genomes under the African2Epoch\_1H18 model \citep{durvasula2017african}
+and using the SalomeAveraged\_TAIR7 genetic map \citep{salome2011recombination}.
 As a proxy for the ``truth'', in black we show inverse coalescence rates

### 2020-04-27 popSimManu.tex
added text about pop-size scaling

@@
 the distribution of fitness effects \citep{Boyko:2008cr,kim2017inference,tataru2017inference,Fortier703918,Huang2019,Vecchyo770966},
-recombination rates \citep{chan2012genome,lin2013fast,Adrion662247,Barroso2019},
+recombination rates \citep{chan2012genome,lin2013fast,adrion2020predicting,Barroso2019},
 and the extent of positive selection in genome sequence data
@@
 as the default simulation engine.
-We have implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim}
-as an alternative backend,
-to allow simulation of processes that cannot be modeled under the coalescent. We validated our
-implementation of the \texttt{SLiM} engine by comparing estimates of several population
-genetic summary statistics for simulations generated by both engines 
-(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
+
+Coalescent simulations, while highly efficient, are limited in their ability to model
+complex selection scenarios, such as recurrent sweeps and background selection. For these reasons
+we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim}, 
+as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under 
+the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
+population sizes can be highly demanding of computational resources. One solution to this obstacle is to 
+rescale the population size, \textit{N}, such that population-dependent parameters (e.g. \textit{N$\mu$}, \textit{Nr},
+and \textit{Ns}) remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
+
+We validated our implementation of the \texttt{SLiM} engine by comparing 
+estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
+and \texttt{msprime} under different factors of population-size rescaling 
+(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}). We show that rescaling by up to a factor
+of $50$ ($Q=50$ in \texttt{SLiM}) produces patterns of both diversity and linkage disequilibrium that are 
+qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map 
+(Figure \ref{fig:slim_val_map}). It is important to note that rescaling \textit{N} is not, in actuality, the equivalent to simulating 
+all individuals in the population, and that differences between the two methods may be subtle and difficult to interpret.
+To emphasize this point, we show deviations between patterns of diversity generated by \texttt{SLiM} and those generated by 
+\texttt{msprime} engines at $Q=50$, when simulating under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}). 
+The causes of these differences remain unclear. 
+While an in-depth investigation of the effects of population-size rescaling under selection 
+scenarios is outside the scope of this report, we are enthusiastic about the potential for \texttt{stdpopsim} and the 
+\texttt{SLiM} engine to facilitate addressing such pertinent questions in the near future.
 
@@
 and downstream aspects of data quality (genotyping and mapping error).
+Moreover, an in-depth investigation into the effects of population-size rescaling under many of the
+above scenarios is warented, given our preliminary findings using neutral simulations 
+(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
 These plans would not be possible without recent important work on SLiM \citep{haller2019slim},

### 2020-04-28 popSimManu.tex

@@
 the degree of sophistication of the questions, models, data, and computational approaches
-used have all increased over the past two decades. Currently there exist myriad computational methods
+used have all increased over the past two decades. Currently there exist a myriad of computational methods
 that can infer the histories of populations
@@
 the distribution of fitness effects \citep{Boyko:2008cr,kim2017inference,tataru2017inference,Fortier703918,Huang2019,Vecchyo770966},
-recombination rates \citep{chan2012genome,lin2013fast,adrion2020predicting,Barroso2019},
+recombination rates \citep{mcvean2004ldhat,chan2012genome,lin2013fast,adrion2020predicting,Barroso2019},
 and the extent of positive selection in genome sequence data
-\citep{eyre2009estimating, alachiotis2012omegaplus,degiorgio2016sweepfinder2,kern2018diplos,sugden2018localization}.
+\citep{kim2002detecting,eyre2009estimating, alachiotis2012omegaplus,garud2015sweeps,degiorgio2016sweepfinder2,kern2018diplos,sugden2018localization}.
 While these methods have undoubtedly increased our understanding of
 genetic and evolutionary processes, very little has been done to systematically
-benchmark the quality of these inferences.
-% or their robustness to deviations from the assumptions.
+benchmark the quality of these inferences or their robustness to deviations from the assumptions.
 As large databases of population genetic variation begin to be used to inform public health procedures,
@@
 including identifying appropriate demographic models from the literature,
-translating them into input for a simulator, obtaining appropriate genetic
-maps, and choosing appropriate values for key population genetic parameters.
+translating them into input for a simulator, and choosing appropriate values for key population genetic parameters,
+such as the mutation and recombination rates.
 
@@
 correctly implemented the models as described in their original publication
-and have documented methods for others to contribute new modules.
+and have documented methods for others to contribute new models.
 We invite new collaborators to join our community:
@@
 on a catalog of species (Fig.~\ref{fig:cartoon}A), initially consisting of humans,
-\textit{D.~melanogaster}, and \textit{A.~thaliana}.
+\textit{D.~melanogaster}, \textit{A.~thaliana},
+\textit{Escherichia coli}, and \textit{Canis familiaris}.
 A species definition consists of two key elements.
@@
 \citeauthor{gutenkunst2009inferring} model which does not include Asian populations, but more explicitly models
-recent rapid human population growth (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
+recent rapid human population growth in Europe (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
 for American populations which specifies ancestral African, European, and Asian population
@@
 ``true'' simulated population size due to model misspecification.
-Instead, many methods work by matching coalescence time distributions,
-and a single-population model with varying population size can match any coalescence time distribution
-(in which case coalescence rate is the inverse of the effective size).
+Instead, many methods work by matching coalescence time distributions, and a single-population model 
+with varying population size can be fit to data generated from a more complex model including migration. 
+In this case, the coalescence time distributions
+will exactly match, and the coalescence rate is the inverse of the effective population size.
 For this reason,
@@
 simulated under the OutOfAfrica\_3G09 model. Our results highlight many
-of the strengths and weaknesses of the different types of methods we used.
+of the strengths and weaknesses of the different types of methods.
 For instance, the SFS-based approaches where we fit simple IM models do not capture
@@
 Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
-but ignore the brief population bottleneck in Europe. In addition, \dadi
+but but do not detect the brief bottleneck in Europe, due to inference model
+not allowing for population size changes after the population split. In addition, \dadi
 and \fastsimcoal both do reasonably well at correctly inferring the absence of
@@
 are roughly consistent with the hypothesized scenario.
-Currently, many studies would benefit from simulation-based checks such as these,
-but this presents a major barrier,
+Currently, many studies would benefit from such simulation-based checks.
+However,  there are major barriers to implementation,
 since individual research groups must reimplement complex, previously published demographic models,
@@
 Dmitri Petrov, and Sohini Ramachandran.
-In addition we would like to thank NESCENT and Matt Hahn, Victoria Sork, and Michael Whitlock for organizing a 2014
-catalysis meeting in which many of the goals of this effort were first laid out. 
 CCK and KEL were funded under NIH Award R35GM119856.

### 2020-04-29 response-to-reviewers.md
IG edits
@grahamgower edits

.

@@
 
-Issue #150 (jadrion) will add the figures we need and some explanation;
-we need to explain here (and maybe in the text) that selection is beyond the scope of this paper,
-and instead we are focusing in the paper on validating scaling.
-This will also be covered by issue #164.
+We would like to see this also and agree that it would be of great interest.
+However, we felt that performing adequate simulations with selection would require
+a substantial amount of additional work and would significantly expand the scope of
+the manuscript.
+Furthermore, discussing selection would distract from the current relatively simple focus
+on inference of demographic history.
+We definitely intend to include selection (and expand the available generic models) in future work,
+as we now discuss in greater detail in the Discussion ("next steps").
+
+Consequently, we decided to add an example for possible use of SLiM in neutral simulations (see "simulation engines" section in the results and Supplementary Figures S1 and S2).
+Our focus there was to show that under neutral simulations, SLiM produced data that were consistent with the coalescent simulations of msprime.
+To expand on this fundamental point, we also examined the influence of population size down-scaling, which is a common practice used in forward simulations.
+While this is by no means a comprehensive investigation into this issue, we do believe that it demonstrates the use of SLiM in future applications that will also simulate selection.
 
@@
 
-We have added an additional paragraph to the discussion exploring some of the additional factors that may affect choice of inference method, beyond results like those presented here.
-These include data required, type of model, and software engineering.
-A comprehensive investigation of these issues is beyond the scope of the current manuscript, which focuses on the resource itself, rather than its application.
+We have added an additional paragraph to the discussion exploring various factors that may affect choice of inference method.
+These include the data required by the method, the type of model, and software engineering.
+We also provide a qualitative comparison between the performance of methods, as shown in our limited analysis.
+We believe that a more comprehensive comparison between methods is beyond the scope of the current manuscript, which focuses on the resource itself, rather than its application.
 
@@
 We made the suggested change:
-"Importantly, we have rigorous quality control methods to ensure that we have
+"Importantly, we developed rigorous quality control methods to ensure that we have
 correctly implemented the models as described in their original publication
-and have documented methods for others to contribute new modules."
+and provide documented methods for others to contribute new models."
 
@@
 
-We would like to see this also!
-But, we felt that the amount of work it took to get to this point
-made this a good point to release the publication.
-Furthermore, discussing selection would distract from the current relatively simple focus
+As mentioned in our response to the editor's comments, we agree that this is an important avenue to pursue in the near future.
+However, performing adequate simulations with selection would require
+a substantial amount of additional work and would likely distract from the current relatively simple focus
 on inference of demographic history.
 We definitely intend to include selection (and expand the available generic models) in future work,
-however, as we now discuss in greater detail in the Discussion ("next steps").
+as we now discuss in greater detail in the Discussion ("next steps").
 
@@
 
-Maybe so, but beyond the scope?
-Issue #166 (andrewkern)
+Our initial motivation for masking was to reduce the overrepresentation of marginal trees with little to no recombination from biasing patterns of diversity in such a way that demographic inference methods would be misled.
+While we agree with the reviewer that different masking approaches might better reflect the masking done on real genomic data, at this time the best way to mask remains an open question, and we feel that a nuanced analysis of masking is outside the scope of this paper.
+For these reasons, we feel that masking using a simple recombination rate threshold is the most appropriate approach.
+We agree that there may indeed be better ways to mask, which is why we allow users to mask their simulations as they see fit.
 
@@
 
-Each software package requires its own kind of masking - eg smc++ requires a bedfile.
-Issue #166
+Mask files are not currently a component of `stdpopsim` proper, rather they were implemented separately from running `stdpopsim`, for the sole purpose of comparing demographic inference methods.
+Users who download `stdpopsim` will always be simulating raw and unmasked tree sequence files, to which they can apply any variety of masks *post hoc*, if they so choose.
 
@@
 
-We agree that adding both census size and inverse coalescence rates to these plots would instructive.
-Thus, we have added census size to all of the single population analysis plots (Figs 2, 3, S1, S2, and S3) in addition to ICR.
-However, after some discussion, we have decided that adding ICR to the two population analysis plots (Figs 4, S5, and S7, which currently only have census size) might not be appropriate since these multi-population analyses do to some extent account for the effects of migration that are ignored in the single-population case, and should therefore be able to recover the true simulated population sizes.
-
-Finally, we note that we have decided to instead describe the true simulated population sizes as the "simulated population size" rather than "census size" as we realized the use of "census" here could be misleading (see below for further discussion).
-
-FIXME: add the census sizes to Figs 2, 3, S1, S2, and S3 and edit figure captions
-Issue #160 (jradrion)
+First, we note that we decided to avoid using "census size", and simply refer to it as the "simulated population size", as we realized the use of "census" here could be misleading (see below for further discussion).
+Regarding the main point raised by the reviewer, we agree that in the single-population analyses, it would be instructive to show both the simulated (census) population size and the inverse coalescence rates.
+We thus plotted the simulated population sizes against the inverse coalescence rate for the model used in Figure 2, and found that they were near identical.
+As a result, we decided to show this comparison in Supplementary Figure S1 and discuss this issue in the text without adding the simulated population sizes to the main figures (as they didn't appear to add much).
+Note that in the multi-population analyses, the inverse coalescence rate is less appropriate, because some of the inference methods used in these analyses model gene flow, and should therefore be able to recover the true simulated population sizes.
+We now clarify this in the text as well.
 
@@
 In addition, we had previously plotted divergence times here as vertical lines, however we felt that plotting it this way made the figure hard to read since there was a great deal of overlap between these lines and the N(t) lines on the figure, which was was especially difficult to read when including multiple replicates.
-Thus, we prefer to keep the lower divergence times panel for improved readability.
+Thus, we decided to depict the divergence times (simulated and inferred) in a separate panel in the bottom of the figure, to improve readability.
 
@@
 
-FIXME add description of what we added to the manuscript here.
+As mentioned in our response to the editor's comment, we added a demonstration of the use of SLiM to the results (see "Simulation engines" section and Supplementary Figures S1 and S2).
+This new analysis uses SLiM in neutral simulations to validate that it produces data that is consistent with data simulated under the coalescent.
+Furthermore, we provide a rudamentary examination of the effects of population size down-scaling, which is a common practice used in forward simulations such as SLiM.
 
@@
 
-FIXME: fill this in.
-Issue #150 and #164, as above.
+As mentioned in our response to the editor's comment, we added a demonstration of the use of SLiM to the results (see "Simulation engines" section and Supplementary Figures S1 and S2).
+See also our response to related comments made by Reviewers #1 and #2.
 
@@
 
-FIXME: maybe we should do this?
-
-Note that zigzag is in human because the parameterization is human-ish.
+This is an interesting idea. We did discuss adding a suite of generic models to stdpopsim that do not necessarily capture demographic parameters from a certain species,
+but do reflect general trends observed in inference. The "zigzag" model could definitely be one of these.
+This did not make it to this release, but we will note this for the next one.
+the reason why the zigzag model is currently defined as a human model is that its
+effective population size values are taken from (or at least inspired by) values inferred from human genomes.
 
@@
 
-"However, because model misspecification
-here is less severe  compared to the single-population case, we compare
-our inferred population sizes only to the simulated population sizes (black line in
+"However, since the model we use for inference also
+allows gene flow between populations, we directly compare estimated effective population sizes
+to the values used in simulations (black line in
 Figure 4C) and not the inverse coalescence rates."
@@
 
-Really, we just mean that the inferred history has more population size changes
+We meant to say that the inferred history has more population size changes
 than the truth did;

### 2020-04-29 popSimManu.tex
IG edits
@grahamgower edits

.

@@
 to continue its growth and maintain its quality.
-Importantly, we have rigorous quality control methods to ensure that we have
+Importantly, we developed rigorous quality control methods to ensure that we have
 correctly implemented the models as described in their original publication
-and have documented methods for others to contribute new models.
+and provide documented methods for others to contribute new models.
 We invite new collaborators to join our community:
@@
 
-Given the genome data and simulation model descriptions defined within the
-library, it is then straightforward to run standardized simulations
-across a range of organisms. \Stdpopsim has a Python API and a user-friendly
-command line interface, allowing users with minimal experience direct access to
-state-of-the-art simulations. Simulations are output in the ``tree sequence''
-format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
-contains complete genealogical information about the simulated samples, is
-extremely compact, and can be processed efficiently using the \tskit library
-\citep{kelleher2016efficient,kelleher2018efficient}.
-The tree sequence format could also be converted
-to other formats (e.g., VCF) by the user if desired.
-Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
-as the default simulation engine.
-
-Coalescent simulations, while highly efficient, are limited in their ability to model
-complex selection scenarios, such as recurrent sweeps and background selection. For these reasons
-we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
-as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under
-the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
-population sizes can be highly demanding of computational resources. One solution to this obstacle is to
-rescale the population size, \textit{N}, such that population-dependent parameters (e.g. \textit{N$\mu$}, \textit{Nr},
-and \textit{Ns}) remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
-
-We validated our implementation of the \texttt{SLiM} engine by comparing
-estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
-and \texttt{msprime} under different factors of population-size rescaling
-(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}). We show that rescaling by up to a factor
-of $50$ ($Q=50$ in \texttt{SLiM}) produces patterns of both diversity and linkage disequilibrium that are
-qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
-(Figure \ref{fig:slim_val_map}). It is important to note that rescaling \textit{N} is not, in actuality, the equivalent to simulating
-all individuals in the population, and that differences between the two methods may be subtle and difficult to interpret.
-To emphasize this point, we show deviations between patterns of diversity generated by \texttt{SLiM} and those generated by
-\texttt{msprime} engines at $Q=50$, when simulating under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
-The causes of these differences remain unclear.
-While an in-depth investigation of the effects of population-size rescaling under selection
-scenarios is outside the scope of this report, we are enthusiastic about the potential for \texttt{stdpopsim} and the
-\texttt{SLiM} engine to facilitate addressing such pertinent questions in the near future.
-
-The \stdpopsim command line interface, by default, outputs citation information
-for the models, genetic maps and simulation engines used in any particular run.
-We hope that this will encourage users to appropriately acknowledge the
-resources used in published work, and encourage authors
-publishing demographic models to contribute to our ongoing community-driven development process.
-Together with the \stdpopsim version number and the long-term stable identifiers
-for population models and genetic maps,
-this citation information will result in well-documented and reproducible
-simulation workflows. The individual tree sequence files produced by
-\stdpopsim also contain complete provenance information including the command
-line arguments, operating system environment and versions of key libraries
-used.
-
-\subsection*{The Species Catalog}
+\subsection*{The species Catalog}
 The central feature of \stdpopsim is the species catalog, a systematic organization
@@
 
+ \subsection*{Simulation engines}
+
+Given the genome data and simulation model descriptions defined within the
+library, it is then straightforward to run standardized simulations
+across a range of organisms. \Stdpopsim has a Python API and a user-friendly
+command line interface, allowing users with minimal experience direct access to
+state-of-the-art simulations. Simulations are output in the ``tree sequence''
+format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
+contains complete genealogical information about the simulated samples, is
+extremely compact, and can be processed efficiently using the \tskit library
+\citep{kelleher2016efficient,kelleher2018efficient}.
+The tree sequence format could also be converted
+to other formats (e.g., VCF) by the user if desired.
+Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
+as the default simulation engine.
+
+Coalescent simulations, while highly efficient, are limited in their ability to model
+complex selection scenarios, such as recurrent sweeps and background selection. For these reasons
+we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
+as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under
+the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
+population sizes can be highly demanding of computational resources. One common practice used to address this challenge is to
+scale down the population size used in the simulations (\textit{N}),
+and then scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
+rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr},
+and \textit{Ns}, remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
+Such down-scaled simulations are not completely equivalent to simulating
+all individuals in the population, and may lead to subtle differences.
+However, since many sequence-based measures of population diversity remain unchanged when rescaling in this fashion,
+this practice is very common when using forward-time simulators, such as \texttt{SLiM}.
+
+We validated our implementation of the \texttt{SLiM} engine by comparing
+estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
+and \texttt{msprime}.
+Examples of this validation for different down-scaling factors are shown in
+Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scaling down the simulated
+population size by up to a factor
+of $50$ ($Q=50$ in \texttt{SLiM}) produces patterns of both diversity and linkage disequilibrium that are
+qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
+(Figure \ref{fig:slim_val_map}).
+We do observe a slight departure in values of Tajima's D when scaling down by a factor of $50$,
+likely due to the subtle differences implied by down-scaling.
+Interestingly, this effect of down-scaling is more prominent and extends to other sequence-based statistcs
+when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
+Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied,
+and they suggest that down-scaling population sizes can significantly reduce running time at a fairly modest cost to accuracy.
+However, since the effects of rescaling are not fully understood, scaling factors should be used as conservatively as practicable.
+The framework provided by \texttt{stdpopsim} should enable a deeper inspection of this issue, which we expect will contribute to
+best practices used when applying forward-time simulators, such as \texttt{SLiM}.
+
+%Ilan: added this short subsection since this paragraph didn't quite fit any of the other subsections
+\subsection*{Documentation and reproducability}
+
+The \stdpopsim command line interface, by default, outputs citation information
+for the models, genetic maps and simulation engines used in any particular run.
+We hope that this will encourage users to appropriately acknowledge the
+resources used in published work, and encourage authors
+publishing demographic models to contribute to our ongoing community-driven development process.
+Together with the \stdpopsim version number and the long-term stable identifiers
+for population models and genetic maps,
+this citation information will result in well-documented and reproducible
+simulation workflows. The individual tree sequence files produced by
+\stdpopsim also contain complete provenance information including the command
+line arguments, operating system environment and versions of key libraries
+used.
+
 \subsection*{Use case: comparing methods of demographic inference}
@@
 that allow efficient computing in multicore or cluster environments.
-
-For single-population population size histories, we compared
-\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
-\stairwayplot~\citep{liu2015exploring}
- on simulated genomes sampled from a single population,
-in a number of the demographic models described above.
 Our workflow generates $R$ replicates of $C$ chromosomes,
@@
 Each file is then converted into an input file appropriate for each inference method
-(so, all inference methods run on the same simulation replicates).
+(such that all inference methods run on the same simulation replicates).
 Each of the inference programs are then run in parallel, followed by plotting of
@@
 
+\subsubsection*{ Single-population demographic models.}
+For single-population demographic models, we compared
+\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
+\stairwayplot~\citep{liu2015exploring}
+ on simulated genomes sampled from a single population,
+in a number of the demographic models described above.
+When fitting single-population models to data simulated for multiple
+populations with migration, we do not necessarily expect the inference methods to recover the
+actual simulated population size due to model misspecification.
+Indeed, the best one may expect in such a scenario is to infer a model which accurately reflects the
+coalescence time distribution of the simulated model. While in the inferred single-population model,
+this distribution is only affected by the ancestral effective population sizes, in the simulated multi-population
+model it is also influenced by migration between the target population and populations not analyzed in inference.
+Since ancestral population sizes correspond to the inverse of the coalescence rates, by computing the
+inverse coalescence rates from the simulated model, we obtain the best estimates
+one may hope to infer by analyzing a single population. We thus analytically compute inverse coalescence
+rates in \texttt{msprime} for each simulated model, and use them as a benchmark of the ``true'' effective population sizes.
+See the Appendix for a precise definition and description of the inverse coalescence rate computation.
+
 Figure \ref{fig:n_t_ragsdale} presents the results
@@
 In each row we show comparisons among the methods (including two sample sizes for \MSMC).
-Blue lines show estimates from each of three replicate whole genome simulations.
-When fitting single-population models to data simulated for multiple
-populations with migration, however, we do not necessarily expect the inference methods to recover the
-``true'' simulated population size due to model misspecification.
+Blue lines show estimates from each of three replicate whole genome simulations,
+and black lines indicate the ``true" values depicted by the inverse coalescence rates.
+The latter are included as an appropriate reference because
+when fitting single-population models to data simulated for multiple
+populations with migration, we do not necessarily expect the inference methods to recover the
+actual sizes of the populations used in simulation due to model misspecification.
 Instead, many methods work by matching coalescence time distributions, and a single-population model
@@
 will exactly match, and the coalescence rate is the inverse of the effective population size.
-For this reason,
-we plot the inverse coalescence rates (black lines) calculated analytically in \texttt{msprime}
-as a benchmark of the ``truth'' in the single-population inference setting. However,
-in this case inverse coalescence rates are very close to census population sizes
+However, in this specific model,
+the inverse coalescence rates are very close to the simulated population sizes
 (Figure \ref{fig:census_vs_IRC}).
-See the Appendix for a precise definition and description of the inverse coalescence rate computation.
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 and in black simulated population sizes for the respective population.
-The population split date, $T_{DIV}$, is shown at
-the bottom, with a common $x$-axis to the population size panels.}
+The population split time, $T_{DIV}$, is shown at
+the bottom (simulated value in black and inferred values in blue),
+with a common $x$-axis to the population size panels.}
 \label{fig:IM_popn_human}
@@
 
-\paragraph*{Multi-population demographic models.}
+\subsubsection*{Multi-population demographic models.}
 As \stdpopsim implements multi-population demographic models, we also
@@
 not match the model from which the data were generated (Figures
-\ref{fig:IM_popn_human}A and B). However, because model misspecification
-here is less severe compared to the single-population case, we compare
-our inferred population sizes only to the simulated population sizes (black line in
-Figure \ref{fig:IM_popn_human}C) and not the inverse coalescence rates.
+\ref{fig:IM_popn_human}A and B). However, since the model we use for inference also
+allows gene flow between populations, we directly compare estimated effective population sizes
+to the values used in simulations (black line in Figure \ref{fig:IM_popn_human}C)
+and not the inverse coalescence rates.
 
@@
 Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
-but but do not detect the brief bottleneck in Europe, due to inference model
+but they do not detect the brief bottleneck in Europe, due to inference model
 not allowing for population size changes after the population split. In addition, \dadi
@@
 Moreover, an in-depth investigation into the effects of population-size rescaling under many of the
-above scenarios is warented, given our preliminary findings using neutral simulations
+above scenarios is warranted, given our preliminary findings using neutral simulations
 (Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
@@
 will be easily incorporated, once they are published.
+Although the models in the catalog are state of the art, it is
+important to acknowledge that these are current best estimates of parameter values
+which will change over time. Additional complications such as changing
+genome builds as assemblies improve must also incorporated.
+As the selection of models and species grows, we hope to develop an improved
+documentation system that incorporates crowd-sourced community input
+to highlight the most appropriate models and genome details.
 
@@
 using a variety of models (see Table 1): OutOfAfricaArchaicAdmixture\_5R19, OutOfAfrica\_3G09,
-a constant-sized generic model, and a two-epoch generic model where the
-population size instantaneously decreased from $N=10^4$ to $N=10^3$
-five hundred generations before the present. Each simulation
+and a constant-sized generic model.
+% ILAN REMOVED THIS APR 29, 2020 - NOT USED ANYMORE
+%, and a two-epoch generic model where the population size instantaneously decreased from $N=10^4$ to $N=10^3$
+% five hundred generations before the present.
+Each simulation
 used the HapmapII\_GRCh37 genetic map. For \textit{D.~melanogaster}
@@
 Specifically we masked all regions of \SI{1}{cM} or greater in the lowest 5th percentile of the empirical
-distribution of recombination, regions which are nearly uniformly absent for
-empirical analysis.
+distribution of recombination, regions which are nearly uniformly absent for empirical analysis.
+This approach to masking was chosen to prevent marginal trees with low or no recombination 
+from biasing the comparisons of demographic inference methods. It should be noted that masking is not implemented 
+within \texttt{stdpopsim} proper; tree sequences generated by \texttt{stdpopsim} are always raw and unmasked. 
+This allows users the flexibility to implement masking approaches that are specific to their needs 
+for downstream analysis.
 
@@
 population that was ignored by our inference.
-To ground-truth our inference approach, we also conducted
+To validate our inference approach, we also conducted
 inference on a generic IM model that was identical to the model used for inference (Figure \ref{fig:generic_IM}).
@@
 HapMapII\_GRCh37 genetic map \citep{international2007second}. From top to bottom we show results
-using population-size rescaling for three scaling factors: Q=1, Q=10, and Q=100.
+using three scaling factors for the population sizes: Q=1, Q=10, and Q=50.
 }
@@
 rate of recombination across the chromosome. From top to bottom we show results
-using population-size rescaling for three scaling factors: Q=1, Q=10, and Q=100.
+using three scaling factors for the population sizes: Q=1, Q=10, and Q=50.
 }
@@
 \includegraphics[width=0.8\linewidth]{display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19_census_vs_IRC.pdf}
-\caption{\textbf{Comparing census population sizes and inverse coalescence rates in humans}.
+\caption{\textbf{Comparing simulated population sizes and inverse coalescence rates in humans}.
 Data are shown from human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model \citep{ragsdale2019models}
@@
 From left to right we show sizes for each of the three populations in the model: YRI, CEU, and CHB.
-We plot the census sizes for each population in black, and in red we plot
+We plot the simulated sizes for each population in black, and in red we plot
 inverse coalescence rates as calculated from the demographic model used for simulation (see text).
+In this specific model, these two measures are near identical, but in other models with higher migration rates
+we expect to see a larger departure between the two.
 }
@@
 OutOfAfrica\_3G09 demographic model, with a HapmapII\_GRCh37 genetic map.
-Note that resulting data for input to all inference methods \smcpp, \MSMC, and \stairwayplot,
+Note that the data used as input by all inference methods \smcpp, \MSMC, and \stairwayplot,
 come from the same set of simulations.}
@@
 An analogous observation holds for populations of changing size,
-if we define the "coalescence rate" of a given demographic model
+if we define the ``coalescence rate" of a given demographic model
 at a particular point back in time
 to be the rate of coalescence of remaining lineages
-and define the "coalescence effective size" at that time,
+and define the ``coalescence effective size" at that time,
 denoted $N_e(t)$, so that the coalescence rate at time $t$ in the past

### 2020-04-29 popSimManu.tex
added text about masking

@@
 distribution of recombination, regions which are nearly uniformly absent for
-empirical analysis.
+empirical analysis. This approach to masking was chosen to prevent marginal trees with low or no recombination 
+from biasing the comparisons of demographic inference methods. It should be noted that masking is not implemented 
+within \texttt{stdpopsim} proper; tree sequences generated by \texttt{stdpopsim} are always raw and unmasked. 
+This allows users the flexibility to implement masking approaches that are specific to their needs 
+for downstream analysis.
 

### 2020-04-30 popSimManu.tex
manuscript edits from dan

--- 2020-04-29 popSimManu.tex
@@
 These available models and parameters are meant to represent the field's current understanding,
-and we intend for this resource to evolve as new results become available.
+and we intend for this resource to evolve as new results become available, and relevant
+existing models are added to \stdpopsim by the community.
 Through either a command line interface or a simple Python API, users can specify which
@@
 isolation with migration (IM) model which can be used with any genome and genetic map.
+%DRS: do we want to include these generic models in Table 1?
 
@@
 scale down the population size used in the simulations (\textit{N}),
-and then scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
+scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
 rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr},
@@
 (Figure \ref{fig:slim_val_map}).
-We do observe a slight departure in values of Tajima's D when scaling down by a factor of $50$,
-likely due to the subtle differences implied by down-scaling.
+We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$,
+likely due to the subtle differences produced by down-scaling.
 Interestingly, this effect of down-scaling is more prominent and extends to other sequence-based statistcs
@@
 The framework provided by \texttt{stdpopsim} should enable a deeper inspection of this issue, which we expect will contribute to
-best practices used when applying forward-time simulators, such as \texttt{SLiM}.
+best practices used when applying forward-time simulators such as \texttt{SLiM}.
 
@@
 The \stdpopsim command line interface, by default, outputs citation information
-for the models, genetic maps and simulation engines used in any particular run.
+for the models, genetic maps, and simulation engines used in any particular run.
 We hope that this will encourage users to appropriately acknowledge the
@@
 methods, our aim at this stage is not to provide an exhaustive
-evaluation or ranking of these methods. Our hope is instead that future work built upon this resource
-will enable more detailed exploration of the strengths and weaknesses of the numerous
+evaluation or ranking of these methods. Our hope is instead to demonstrate how \stdpopsim
+will facilitate more detailed future explorations of the strengths and weaknesses of the numerous
 inference methods that are available to the population genetics community
@@
 inverse coalescence rates from the simulated model, we obtain the best estimates
-one may hope to infer by analyzing a single population. We thus analytically compute inverse coalescence
+one may hope to infer by analyzing a single population with a method that seeks to match coalescence time distributions.
+. We thus analytically compute inverse coalescence
 rates in \texttt{msprime} for each simulated model, and use them as a benchmark of the ``true'' effective population sizes.
@@
 Blue lines show estimates from each of three replicate whole genome simulations,
-and black lines indicate the ``true" values depicted by the inverse coalescence rates.
-The latter are included as an appropriate reference because
-when fitting single-population models to data simulated for multiple
-populations with migration, we do not necessarily expect the inference methods to recover the
-actual sizes of the populations used in simulation due to model misspecification.
-Instead, many methods work by matching coalescence time distributions, and a single-population model
-with varying population size can be fit to data generated from a more complex model including migration.
-In this case, the coalescence time distributions
-will exactly match, and the coalescence rate is the inverse of the effective population size.
-However, in this specific model,
-the inverse coalescence rates are very close to the simulated population sizes
-(Figure \ref{fig:census_vs_IRC}).
+and black lines indicate the ``true" values depicted by the inverse coalescence rates (although in this specific model
+the inverse coalescence rates are very close to the simulated population sizes;
+Figure \ref{fig:census_vs_IRC}).
+%DRS: trying to cut down on the redundancy between the coal rate stuff here and the previous paragraph
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 For instance, \stdpopsim could be used in power analyses to determine adequate sample sizes,
-or in sanity checks, to see if observed data
+or in sanity checks to see if observed data
 (e.g., levels of divergence or the allele frequency spectrum)
@@
 the site frequency spectrum-based method (\stairwayplot),
-which tends to over-estimate the sizes of oscillations.
+which tends to over-estimate the sizes of oscillations
 (Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}).
@@
 In simulations of two human populations
-(Figure~\ref{fig:IM_popn_human}), most methods do reasonably well at
+(Figure~\ref{fig:IM_popn_human}), \dadi and \fastsimcoal do reasonably well at
 reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
 history, in large part because the methods assume constant population sizes.
-An exception is \smcpp, which does not have the same
+On the other hand, \smcpp does not have the same
 restrictions on its inferred history, and as a result does somewhat better
@@
 \dadi and \fastsimcoal require a pre-specified parametric model, unlike \MSMC, \smcpp, and \stairwayplot.
-Using a parametric approach yields less noisy results, but a model that is too simple may not capture important demographic events (Figures~\ref{fig:IM_popn_human} and~\ref{fig:two_popn_fly}).
+Using a parametric approach yields less noisy results, but a model that is too simple may not capture important demographic events (Figures~\ref{fig:IM_popn_human} and~\ref{fig:two_popn_fly}),
+and other forms of model misspeficification may also produce undesireable behavior.
 From a software engineering perspective, methods also differ in their ease of installation and use.
@@
 Some of these additions that we are enthusiastic to add are:
-selection (including distributions of fitness effects, maps of functional elements, and single sweeps),
+selection (including distributions of fitness effects, maps of functional elements, and both single and recurrent hitchhiking events),
 gene conversion, mutation models (rate heterogeneity),
@@
 % note: this is addressing reviewer comments about model uncertainty
-In particular, although the included demographic models represent the current state of the art,
+In particular, although the included demographic histories are some of the most widely used models for our current set of species,
 we anticipate the set of available models to expand
@@
 will be easily incorporated, once they are published.
-Although the models in the catalog are state of the art, it is
-important to acknowledge that these are current best estimates of parameter values
-which will change over time. Additional complications such as changing
+%DRS: trying to make this part a little less subjective. We were kind of painting the picture that
+%we have all the best/useful models and everything currently published but not included is not useful,
+%and I don't think this is justified.
+Our current set may also omit other widely used models that can be readily incorporated into \stdpopsim.
+Additional complications such as changing
 genome builds as assemblies improve must also incorporated.
-As the selection of models and species grows, we hope to develop an improved
-documentation system that incorporates crowd-sourced community input
-to highlight the most appropriate models and genome details.
+As the selection of models and species grows, we hope to
+incorporate crowd-sourced community input
+to highlight the most appropriate models and genome annotations.
+%DRS: not in love with the word "annotations" here as some might think we are talking gene annotations (which we will be eventually but not yet), but "details" is too vague.
 

### 2020-04-30 popSimManu.tex

@@
 
+The possibility for error and the importance of careful qualty control
+was inadvertantly illustrated during our own development process:
+during the final revisions of this paper, we noticed that the 
+OutOfAfrica\_3G09 model \citep{gutenkunst2009inferring}
+had not in fact gone through the QC process.
+Then, the QC process revealed that not only was our implementation
+slightly wrong---migration rates had not been set to zero to the European population in the longest-ago time period
+when there should have only been a single population.
+This error was in fact propagated from the \texttt{msprime} documentation.
+
 \subsection*{Workflow for analysis of simulated data}

### 2020-05-01 popSimManu.tex
small style edits, and reordered some paragraphs in results to avoid redundancy

@@
 (perhaps provided as long command-line calls in supplementary material).
-Often these studies focus more on showcasing a novel method than on
-rigorously comparing it with competing methods.
+Often these simulations are more useful to showcase a novel method than to
+rigorously compare it with competing methods.
 Moreover, this situation results in a great deal of duplicated effort,
@@
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five organisms: humans,
+The \stdpopsim catalog currently contains five species: humans,
 \textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
 \textit{Escherichia coli}, and \textit{Canis familiaris}.
-For each organism, the catalog contains information on our current understanding of
-the physical organization (e.g., chromosome structure) of its genome,
+For each species, the catalog contains information on our current understanding of
+the physical organization of its genome (e.g., chromosome structure),
 one or more inferred genetic maps (except for \textit{E.~coli}),
-default population-level parameters (mutation rate and generation time estimates).
+and default population-level parameters (mutation rate and generation time estimates).
 The catalog also contains numerous published demographic models for the first three species.
@@
 representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig.~\ref{fig:cartoon}A), initially consisting of humans,
+on a catalog of species (Fig.~\ref{fig:cartoon}A), 
+and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).
+
+\subsection*{The species Catalog}
+
+The central feature of \stdpopsim is the species catalog, a systematic organization
+of the key quantitative data needed to simulate a given species. 
+Data is currently available for humans,
 \textit{D.~melanogaster}, \textit{A.~thaliana},
@@
 
-\subsection*{The species Catalog}
-The central feature of \stdpopsim is the species catalog, a systematic organization
-of the key quantitative data needed to simulate a given species. These include a description
-of the genome assembly, information about mutation rate estimates, inferred recombination rate(s), and generation time
-in addition to a series of demographic models that have been estimated for that organism.
-
 The current demographic models in the \stdpopsim catalog are shown in \autoref{tab:catalog}.
@@
 admixture \cite[e.g.,][]{ragsdale2019models}.
-In addition to those models shown, at time of writing
-the PopSim Consortium has demographic models in development for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli}.
+In addition to models from Table \label{tab:catalog}, 
+the PopSim Consortium has demographic models for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli} in development at time of writing.
 
@@
 and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).
-Together these models
-contain features believed to have widespread impacts in real data (e.g., bottlenecks, population growth,
-admixture) and are therefore highly pertinent in the context of method development.
 
@@
 %For \emph{E. coli} we have a simple, large constant population size model implemented.
+
 In addition to organism-specific models, \stdpopsim also includes a generic piecewise constant size model and
 isolation with migration (IM) model which can be used with any genome and genetic map. 
+Together these models
+contain many features believed to affect observed patterns of polymorphism (e.g., bottlenecks, population growth,
+admixture) and therefore provide useful benchmarks for method development.
+
 %DRS: do we want to include these generic models in Table 1?
@@
 %Ilan: added this short subsection since this paragraph didn't quite fit any of the other subsections
-\subsection*{Documentation and reproducability}
+\subsection*{Documentation and reproducibility}
 

### 2020-05-01 popSimManu.tex
minor adjustment

@@
 admixture \cite[e.g.,][]{ragsdale2019models}.
-In addition to models from Table \label{tab:catalog}, 
+In addition to models from Table \autoref{tab:catalog}, 
 the PopSim Consortium has demographic models for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli} in development at time of writing.
@@
 had not in fact gone through the QC process.
-Then, the QC process revealed that not only was our implementation
+The subsequent QC process revealed that in fact our implementation
 slightly wrong---migration rates had not been set to zero to the European population in the longest-ago time period
 when there should have only been a single population.
-This error was in fact propagated from the \texttt{msprime} documentation.
+This error was propagated from the \texttt{msprime} documentation,
+code which has been used in a number of other publications.
 

### 2020-05-01 popSimManu.tex
adjustment to intro para

@@
 
-\paragraph{The \stdpopsim library.}
-The first contribution of the PopSim consortium is \stdpopsim, a
-community-maintained library of empirical genome data and population genetics simulation
-models. Figure \ref{fig:cartoon} shows a graphical
-representation of the structure of \stdpopsim. The package centers
-on a catalog of species (Fig.~\ref{fig:cartoon}A), 
+The first contribution of the PopSim consortium is \stdpopsim,
+a community-maintained library of empirical genome data and population genetics simulation models.
+Figure \ref{fig:cartoon} shows a graphical representation of the structure of \stdpopsim.
+The package centers on a catalog of genomic information and demographic models
+for a growing list of taxa (Fig.~\ref{fig:cartoon}A), 
 and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).

### 2020-05-01 popSimManu.tex
species

@@
 The package centers on a catalog of genomic information and demographic models
-for a growing list of taxa (Fig.~\ref{fig:cartoon}A), 
+for a growing list of species (Fig.~\ref{fig:cartoon}A), 
 and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).

### 2020-05-01 popSimManu.tex
deleted stray line

@@
 leading to inconsistency and duplication of effort.
-This presents a major barrier to empirical researchers seeking
+This situation presents a major barrier to empirical researchers seeking
 to use simulations for power analyses of upcoming studies
@@
 can be used to systematically compare demographic inference methods,
-and we encourage an even broader
+and we encourage a broader
 community of developers to contribute to this growing resource.
@@
 genetic and evolutionary processes, very little has been done to systematically
-benchmark the quality of these inferences or their robustness to deviations from the assumptions.
+benchmark the quality of these inferences or their robustness to deviations from their underlying assumptions.
 As large databases of population genetic variation begin to be used to inform public health procedures,
@@
 operandi to date has been for individual groups to validate their own
-methods using simulations coded from scratch
-(perhaps provided as long command-line calls in supplementary material).
+methods using simulations coded from scratch.
+%ACS: this level of detail seems a bit superfluous here
+%(perhaps provided as long command-line calls in supplementary material).
 Often these simulations are more useful to showcase a novel method than to
@@
 
+%ACS: This paragraph seems to me to need a more explicit connection to 
+%the enterprise of providing standardized simulations.  It feels a bit tangential
+%as written.  I have tried to edit it accordingly.
 A related issue is that it has been challenging to assess the degree to which modeling assumptions
 and choices of data summaries can affect population genetic inferences.  
-However, there are clear examples of different methods yielding fundamentally
+Standardized simulations would enable these questions to be systematically examined.
+Importantly, there are clear examples of different methods yielding fundamentally
 different conclusions. For example, Markovian coalescent methods applied to human genomes have
@@
 researchers lack principled guidelines for selecting the best method for addressing
-their particular questions. The need for empirical
-guidance will only increase as researchers
+their particular questions. The need for 
+guidance from simulated data will only increase as researchers
 seek to apply population genetic methods to a growing collection of non-model taxa.
@@
 correctly implemented the models as described in their original publication
-and provide documented methods for others to contribute new models.
+and provided documented methods for others to contribute new models.
 We invite new collaborators to join our community:
@@
 
-The first contribution of the PopSim consortium is \stdpopsim,
-a community-maintained library of empirical genome data and population genetics simulation models.
+%ACS: the consortium really isn't the point; let's get right to describing what stdpopsim is
+%The first contribution of the PopSim consortium is 
+The \stdpopsim library is a
+a community-maintained collection of empirical genome data and population genetics simulation models.
 Figure \ref{fig:cartoon} shows a graphical representation of the structure of \stdpopsim.
@@
 
-\subsection*{The species Catalog}
+\subsection*{The species catalog}
 
@@
 \caption{\label{tab:catalog}
-Initial set of demographic models in the Catalog and simple benchmarks.
-For each model we report the CPU time, maximum memory usage and the
+Initial set of demographic models in the catalog and summary of computing resources needed for simulation.
+For each model, we report the CPU time, maximum memory usage and the
 size of the output \tskit file, as simulated using the \texttt{msprime} simulation engine.
-In each case we simulate 100 samples
+In each case, we simulate 100 samples
 drawn from the first population, for the shortest chromosome of that species
@@
 model with only the African population specified (expansion from the ancestral
-population and recent growth; Africa\_1T12), the three-population model of \cite{gutenkunst2009inferring}
+population and recent growth; Africa\_1T12); the three-population model of \cite{gutenkunst2009inferring},
 which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
-the European and Asian populations (OutOfAfrica\_3G09), the \cite{tennessen2012evolution} two-population variant of the
-\citeauthor{gutenkunst2009inferring} model which does not include Asian populations, but more explicitly models
-recent rapid human population growth in Europe (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
-for American populations which specifies ancestral African, European, and Asian population
-components (AmericanAdmixture\_4B11), a three-population out-of-Africa model from \cite{ragsdale2019models}
-which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19),
-a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19),
+the European and Asian populations (OutOfAfrica\_3G09); the \cite{tennessen2012evolution} two-population variant of the
+\citeauthor{gutenkunst2009inferring} model, which does not include Asian populations but more explicitly models
+recent rapid human population growth in Europe (OutOfAfrica\_2T12); the \cite{browning2018ancestry} admixture model
+for American populations, which specifies ancestral African, European, and Asian population
+components (AmericanAdmixture\_4B11); a three-population out-of-Africa model from \cite{ragsdale2019models},
+which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19);
+a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19);
 and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).
 
-For \textit{D.~melanogaster} we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
+For \textit{D.~melanogaster}, we have implemented the three-epoch model estimated by \cite{sheehan2016deep} from
 an African sample (African3Epoch\_1S16), as well as the out-of-Africa divergence

### 2020-05-01 popSimManu.tex
Wordsmithing/polishing of Results
More minor changes, just polishing

@@
 Given the genome data and simulation model descriptions defined within the
-library, it is then straightforward to run standardized simulations
+library, it is straightforward to run standardized simulations
 across a range of organisms. \Stdpopsim has a Python API and a user-friendly
@@
 Coalescent simulations, while highly efficient, are limited in their ability to model
-complex selection scenarios, such as recurrent sweeps and background selection. For these reasons
+complex selection scenarios, such as recurrent sweeps and background selection. For these reasons,
 we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
@@
 However, since many sequence-based measures of population diversity remain unchanged when rescaling in this fashion,
-this practice is very common when using forward-time simulators, such as \texttt{SLiM}.
+this practice is effective for many purposes and widely employed.
 
@@
 Examples of this validation for different down-scaling factors are shown in
-Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scaling down the simulated
-population size by up to a factor
-of $50$ ($Q=50$ in \texttt{SLiM}) produces patterns of both diversity and linkage disequilibrium that are
+Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scale factors of up to
+$50$ ($Q=50$ in \texttt{SLiM}) produce patterns of both diversity and linkage disequilibrium that are
 qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
 (Figure \ref{fig:slim_val_map}).
-We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$,
-likely due to the subtle differences produced by down-scaling.
-Interestingly, this effect of down-scaling is more prominent and extends to other sequence-based statistcs
+We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$.
+%likely due to the subtle differences produced by down-scaling.
+Interestingly, the effects of down-scaling are more prominent and extend to other sequence-based statistics
 when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
@@
 and they suggest that down-scaling population sizes can significantly reduce running time at a fairly modest cost to accuracy.
+% ACS: We might consider deleting the next two sentences.  This feels like a lot on down-scaling, and it's a bit verbose as written.  Maybe 
+% better to come back to it in the discussion if we want to say more.
 However, since the effects of rescaling are not fully understood, scaling factors should be used as conservatively as practicable.
@@
 
-The \stdpopsim command line interface, by default, outputs citation information
+The \stdpopsim command-line interface, by default, outputs citation information
 for the models, genetic maps, and simulation engines used in any particular run.
-We hope that this will encourage users to appropriately acknowledge the
+We hope that this feature will encourage users to appropriately acknowledge the
 resources used in published work, and encourage authors
@@
 
-As an example of the utility of \stdpopsim, we demonstrate how it can be used
-to easily and fairly compare popular demographic inference methods.
+As an example of the utility of \stdpopsim, we demonstrate how it can be easily used
+to perform a fair comparison of popular demographic inference methods.
 Although we present comparison of results from several
@@
 Our workflow generates $R$ replicates of $C$ chromosomes,
-producing $n$ samples in each of a total of $R \times C$ simulations for each demographic model.
+% ACS: "sample" could be confusing because we are both sampling 
+% by simulation and sampling from a population.  Does "population samples" (below) help?
+producing $n$ population samples in each of a total of $R \times C$ simulations for each demographic model.
 After simulation,
@@
 (such that all inference methods run on the same simulation replicates).
-Each of the inference programs are then run in parallel, followed by plotting of
-estimates of population size history from each program.
-
+Each of the inference programs are then run in parallel, and finally,
+estimates of population size history from each program are plotted.
+
+% ACS: do we need the "Homo sapiens" at top left in the figure?  It breaks the visual symmetry among YRI, CEU, and CHB, and 
+% I think it is unnecessary given the population labels and the description in the legend
+% Also: should we include "N(t)" after "Population size" in the y-axis label, to drive home its meaning (used in the text and legend)?
 \begin{figure}
@@
 of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
-$N(t)$ trajectories for each replicate.
+$N(t)$ trajectories for each of three replicates.
 As a proxy for the ``truth'', in black we show inverse coalescence rates
+% perhaps add: "(expressed in numbers of individuals)"    Seems useful to make the units clear
 as calculated from the demographic model used for simulation (see text).
@@
 
+% ACS: there was indeed a lot of redundancy here.  DRS improved it a lot, but I've tried to do a bit more
 \subsubsection*{ Single-population demographic models.}
@@
  on simulated genomes sampled from a single population,
-in a number of the demographic models described above.
-When fitting single-population models to data simulated for multiple
-populations with migration, we do not necessarily expect the inference methods to recover the
-actual simulated population size due to model misspecification.
-Indeed, the best one may expect in such a scenario is to infer a model which accurately reflects the
-coalescence time distribution of the simulated model. While in the inferred single-population model,
-this distribution is only affected by the ancestral effective population sizes, in the simulated multi-population
-model it is also influenced by migration between the target population and populations not analyzed in inference.
-Since ancestral population sizes correspond to the inverse of the coalescence rates, by computing the
-inverse coalescence rates from the simulated model, we obtain the best estimates
-one may hope to infer by analyzing a single population with a method that seeks to match coalescence time distributions.
-. We thus analytically compute inverse coalescence
-rates in \texttt{msprime} for each simulated model, and use them as a benchmark of the ``true'' effective population sizes.
+under several of the demographic models described above.
+However, these experiments raise the question of what to use as the ``true" population sizes in the case of multi-population models with migration.
+In particular, when simple single-population models are fit to data simulated under such complex models, 
+they are not expected to recover the actual simulated population sizes owing to model misspecification.
+Instead, we argue that the best one may expect in such a scenario is to infer a model that accurately reflects the
+coalescence time distribution of the simulated model.  In turn, the inverses of these coalescence times (expressed in appropriate units) 
+correspond to population sizes.
+Under a multi-population model, the coalescence time distribution is influenced by migration between the target population and populations not analyzed in inference,
+as well as by the ancestral effective population sizes.
+We thus analytically computed inverse coalescence
+rates in \texttt{msprime} for each simulated model, and used them as benchmarks for the ``true'' effective population sizes.
 See the Appendix for a precise definition and description of the inverse coalescence rate computation.
@@
 While there is variation in accuracy among methods, populations, and individual replicates,
-the methods generally infer back a good estimate
-of the true effective population sizes of the simulations --
-inferred values are mostly within a factor of two of the truth,
-and many of the methods infer a bottleneck at the correct time.
+the methods generally produce a good estimate
+of the true effective population sizes of the simulations, with
+inferred values mostly within a factor of two of the truth,
+and most methods inferring a bottleneck at approximately the correct time.
 
@@
 All methods recover population size within a factor of two of the simulated values, however
-SMC-based methods, perhaps due to their regularization, tend to infer sinusoidal
+SMC-based methods
+% ACS: suggest eliminating this vague conjecture.  Which regularization exactly is meant?
+%perhaps due to their regularization, 
+tend to infer sinusoidal
 patterns of population size even though no such change is present.
@@
 For human models with more than two populations \citep[e.g.,][]{gutenkunst2009inferring}
-this means that we are inferring parameters for a model that does
+this limitation means that users are inferring parameters for a model that does
 not match the model from which the data were generated (Figures
-\ref{fig:IM_popn_human}A and B). However, since the model we use for inference also
+\ref{fig:IM_popn_human}A and B). However, since the model used for inference also
 allows gene flow between populations, we directly compare estimated effective population sizes
@@
 simulated under the OutOfAfrica\_3G09 model. Our results highlight many
-of the strengths and weaknesses of the different types of methods.
-For instance, the SFS-based approaches where we fit simple IM models do not capture
+of the strengths and weaknesses of the different methods.
+For instance, the SFS-based approaches with simple IM models do not capture
 recent exponential growth in the CEU population, but do consistently recover the
-simulated YRI population size history. Moreover, these approaches allow for estimating
-migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
-of divergence times. However, these migration rate estimates are somewhat biased,
-likely due to model misspecification.
+simulated YRI population size history. Moreover, these approaches allow 
+migration rates to be estimated (Figure \ref{fig:homsap_mig_rates}), and lead to more accurate inferences
+of divergence times. However, these migration rate estimates are somewhat biased.
+%ACS: seems unnecessary and redundant
+%likely due to model misspecification.
 In contrast, \smcpp is much better at capturing the recent exponential
@@
 
-Again, we can compare between species and look at the performance of these methods
+Again, we can extend this analysis to other taxa and examine the performance of these methods
 for a two-population model of \textit{D.~melanogaster}. Figure
@@
 Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
-but they do not detect the brief bottleneck in Europe, due to inference model
+but they do not detect the brief bottleneck in Europe, due to the inference model
 not allowing for population size changes after the population split. In addition, \dadi

### 2020-05-04 popSimManu.tex

@@
 
-\paragraph{Benchmarking population size inference}
+\paragraph{Benchmarking population size inference.}
 We have illustrated in this paper how \stdpopsim can be used for direct
@@
 In addition, the ability of \stdpopsim to generate data with and without significant features, such
-as a genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
+as a genetic map or population-size changes (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
 investigation of the failure modes of popular methods.
@@
 
-\paragraph{Next steps}
+\paragraph{Next steps.}
 \Stdpopsim is intended to be a fully open, community-developed project.
@@
 there are many biological processes that we do not model.
-Some of these additions that we are enthusiastic to add are:
+Some of the additions that we are enthusiastic to add are:
 selection (including distributions of fitness effects, maps of functional elements, and both single and recurrent hitchhiking events),
@@
 Additional complications such as changing
-genome builds as assemblies improve must also incorporated.
-As the selection of models and species grows, we hope to
+genome builds as assemblies improve must also be incorporated.
+As the collection of models and species grows, we hope to
 incorporate crowd-sourced community input

### 2020-05-05 popSimManu.tex
updated figures and src

@@
 existing models are added to \stdpopsim by the community.
-Through either a command line interface or a simple Python API, users can specify which
-organism, genetic map, chromosome, and demographic history they are interested in simulating, and the
-simulation output from their chosen model is returned.
+We have implemented both a command line interface and a simple Python API
+that can be used to simulate genomic data
+from a choice of organism, genetic map, chromosome, and demographic history.
 In this way, \stdpopsim will lower the barrier to high-quality simulation for exploratory analyses,
@@
 The \stdpopsim library is a
-a community-maintained collection of empirical genome data and population genetics simulation models.
-Figure \ref{fig:cartoon} shows a graphical representation of the structure of \stdpopsim.
+a community-maintained collection of empirical genome data and population genetics simulation models,
+with a structure depicted in Figure \ref{fig:cartoon}.
 The package centers on a catalog of genomic information and demographic models
@@
 and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).
+
+Given the genome data and simulation model descriptions defined within the
+library, it is straightforward to run standardized simulations
+across a range of organisms. \Stdpopsim has a Python API and a user-friendly
+command line interface, allowing users with minimal experience direct access to
+state-of-the-art simulations. Simulations are output in the ``tree sequence''
+format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
+contains complete genealogical information about the simulated samples, is
+extremely compact, and can be processed efficiently using the \tskit library
+\citep{kelleher2016efficient,kelleher2018efficient}.
+The tree sequence format could also be converted
+to other formats (e.g., VCF) by the user if desired.
 
@@
 
-Given the genome data and simulation model descriptions defined within the
-library, it is straightforward to run standardized simulations
-across a range of organisms. \Stdpopsim has a Python API and a user-friendly
-command line interface, allowing users with minimal experience direct access to
-state-of-the-art simulations. Simulations are output in the ``tree sequence''
-format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
-contains complete genealogical information about the simulated samples, is
-extremely compact, and can be processed efficiently using the \tskit library
-\citep{kelleher2016efficient,kelleher2018efficient}.
-The tree sequence format could also be converted
-to other formats (e.g., VCF) by the user if desired.
 Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
 as the default simulation engine.
-
 Coalescent simulations, while highly efficient, are limited in their ability to model
-complex selection scenarios, such as recurrent sweeps and background selection. For these reasons,
+continuous geography or 
+complex selection scenarios, such as recurrent sweeps and background selection.
+For these reasons,
 we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
@@
 population sizes can be highly demanding of computational resources. One common practice used to address this challenge is to
-scale down the population size used in the simulations (\textit{N}),
-scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
-rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr},
-and \textit{Ns}, remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
+simulate a \emph{smaller} population, but to rescale resulting times,
+mutation rates, recombination rates, and selection coefficients
+so that the intensity of mutation, recombination, and allele frequency change due to selection
+per unit time remains the same
+%% PLR: the following is not worded quite right, or at least confusingly, because if you scale time
+%% then this scales rates also; I've opted for a less precise but more descriptive formulation?
+% scale down the population size used in the simulations (\textit{N}),
+% scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
+% rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr}, and \textit{Ns}, remain unchanged
+\citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
+Our implementation of the \texttt{SLiM} backend allows easy use of this \emph{rescaling}
+through a single ``scaling factor'' argument.
 Such down-scaled simulations are not completely equivalent to simulating
-all individuals in the population, and may lead to subtle differences.
-However, since many sequence-based measures of population diversity remain unchanged when rescaling in this fashion,
+all individuals in the population, and may lead to subtle differences,
+especially in the presence of selection.
+However, since many sequence-based measures of population diversity remain nearly unchanged when rescaling in this fashion,
 this practice is effective for many purposes and widely employed.
@@
 and \texttt{msprime}.
-Examples of this validation for different down-scaling factors are shown in
-Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scale factors of up to
-$50$ ($Q=50$ in \texttt{SLiM}) produce patterns of both diversity and linkage disequilibrium that are
-qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
-(Figure \ref{fig:slim_val_map}).
-We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$.
+Examples of this validation for the AncientEurasia\_9K19 model \citep{AncientEurasia 9K19}
+are shown in Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}.
+For this model,
+down-scaling factors of up to $50$ produce patterns of both diversity and linkage disequilibrium that are
+very similar to those observed under the coalescent (i.e., \texttt{msprime}), when both are simulated using the HapMapII\_GRCh37 recombination map
+(Figure \ref{fig:slim_val_map}),
+although the distribution of Tajima's $D$ differs slightly when scaling down by a factor of $50$.
 %likely due to the subtle differences produced by down-scaling.
@@
 when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
-Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied,
-and they suggest that down-scaling population sizes can significantly reduce running time at a fairly modest cost to accuracy.
-% ACS: We might consider deleting the next two sentences.  This feels like a lot on down-scaling, and it's a bit verbose as written.  Maybe 
-% better to come back to it in the discussion if we want to say more.
-However, since the effects of rescaling are not fully understood, scaling factors should be used as conservatively as practicable.
-The framework provided by \texttt{stdpopsim} should enable a deeper inspection of this issue, which we expect will contribute to
-best practices used when applying forward-time simulators such as \texttt{SLiM}.
-
-%Ilan: added this short subsection since this paragraph didn't quite fit any of the other subsections
+Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied.
+The results are also optimistic about the rescaling strategy to reduce computational burden,
+but the possible effects are not well-understood,
+so results relying on rescaled simulations should be carefully validated.
+
 \subsection*{Documentation and reproducibility}
@@
 However, these experiments raise the question of what to use as the ``true" population sizes in the case of multi-population models with migration.
-In particular, when simple single-population models are fit to data simulated under such complex models, 
-they are not expected to recover the actual simulated population sizes owing to model misspecification.
+In particular, a simple single-population model that is fit to data simulated
+under a multi-population model, 
+is not expected to recover the actual simulated population sizes,
+because of model misspecification.
 Instead, we argue that the best one may expect in such a scenario is to infer a model that accurately reflects the
-coalescence time distribution of the simulated model.  In turn, the inverses of these coalescence times (expressed in appropriate units) 
-correspond to population sizes.
+coalescence time distribution of the simulated model.
 Under a multi-population model, the coalescence time distribution is influenced by migration between the target population and populations not analyzed in inference,
 as well as by the ancestral effective population sizes.
+The inverse coalescence rate is commonly interpreted as the effective population size,
+since these are equal in a single-population model with random mating.
 We thus analytically computed inverse coalescence
@@
 of divergence times. However, these migration rate estimates are somewhat biased.
-%ACS: seems unnecessary and redundant
-%likely due to model misspecification.
 In contrast, \smcpp is much better at capturing the recent exponential
-growth in the CEU population, though the inferred population sizes are generally noisier.
-In addition, \smcpp consistently underestimates divergence times
+growth in the CEU population, though it consistently underestimates divergence times
 because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).
@@
 (Figure~\ref{fig:IM_popn_human}), \dadi and \fastsimcoal do reasonably well at
-reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
+reconstructing the simulated YRI history and estimating divergence times, but struggle with the more complex simulated CEU
 history, in large part because the methods assume constant population sizes.
 On the other hand, \smcpp does not have the same
-restrictions on its inferred history, and as a result does somewhat better
-with the CEU history but tends to overcomplicate the simple YRI history.
-The results for
-the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly})
-are generally similar.
-In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, differing only in how they calculate model expectations and optimize parameters.
+restrictions on its inferred history, and as a result does much better
+with the CEU history but tends to underestimate divergence times
+due to the assumption of no migration.
+The results for the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly}) are generally similar.
+In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, 
+differing only in how they calculate model expectations and optimize parameters.
 

### 2020-05-05 src/format_plot_data_dromel_two_pop.R

@@ -1,0 +1,347 @@
+
+
+setwd("~/Desktop/PopSim/analysis/two_population_analysis/dmel_n50_LSTP_joint_3Mb_no_mask/")
+
+
+######## Read in dadi and fsc files ##########
+simulated_genome_length <- 3000000 # genome length for 3 Mb dmel sims
+dadi <- read.table("Results/best_dadi_runs.txt", header=T)
+dadi$Na <- dadi$theta/(4*8.4e-9*simulated_genome_length)
+dadi$N1 <- dadi$nu1*dadi$Na
+dadi$N2 <- dadi$nu2*dadi$Na
+dadi$TDIV <- dadi$T*dadi$Na*0.1*2
+dadi$MIG12 <- dadi$m1/(dadi$Na*2)
+dadi$MIG21 <- dadi$m2/(dadi$Na*2)
+
+fsc <- read.table("Results/best_fsc_runs.txt", header=T)
+fsc$Na = fsc$ANCSIZE/2
+fsc$N1 = fsc$NPOP2/2
+fsc$N2 = fsc$NPOP1/2
+fsc$TDIV = fsc$TDIV*0.1
+
+
+
+
+######### create dataframes in the format for plotting ########
+
+### this script is hard coded to plot three replicates - could change to up to 10
+num_reps <- 3
+
+plot_df_af <- data.frame(matrix(ncol = 4, nrow = 0))
+colnames(plot_df_af) <- c("x", "y", "method", "rep")
+
+plot_df_eu <- data.frame(matrix(ncol = 4, nrow = 0))
+colnames(plot_df_eu) <- c("x", "y", "method", "rep")
+
+step_size <- -1000
+
+# need to set some arbitrary time (years ago) to begin plotting the dadi and fsc trajectories 
+t_start <- 500000
+
+#### Add dadi output from Africa pop
+
+for(i in 1:num_reps){
+  
+  dadi_x_pre_split <- seq(from=t_start, to=dadi$TDIV[i], by=step_size)
+  dadi_x <- c(dadi_x_pre_split, seq(from=dadi$TDIV[i], to = 0, by = step_size))
+  dadi_y_Na <- rep(x=dadi$Na[i], times=length(dadi_x_pre_split) )
+  dadi_y_N1 <- c(dadi_y_Na, rep(x=dadi$N1[i], times=length(dadi_x)-length(dadi_x_pre_split)))
+  
[…]

### 2020-05-05 src/plot_IM_two_pop.R

@@ -6,7 +6,7 @@ library(RColorBrewer)
 
 
 ## read in data
-setwd("../data/")
+setwd("~/Desktop/PopSim/figures/data/")
 popsize_1 <- read.csv("IM_popsize_pop1.csv")
 popsize_2 <- read.csv("IM_popsize_pop2.csv")
 tdiv <- read.csv("IM_tdiv.csv")
@@ -46,20 +46,18 @@ p2C_data = subset(tdiv, method == "smcpp")
 
 
 #PLOT
-species<-"Generic IM model"
-population<-"pop 1"
-speciesPop <- bquote(italic(.(species)) ~ (.(population)))
+speciesPop <- "Population 1"
 p0A<-ggplot(p0A_data, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(25,xMax),ylim=c(y0Min,y0Max)) +
-  geom_step(data=p0Z_data,aes(x=x, y=y),size=2.5,alpha=1.0, col=coal_col) +
-  geom_step(size=1.2, alpha=0.9, show.legend = FALSE) +
+  geom_step(data=p0Z_data,aes(x=x, y=y),size=3.3,alpha=1.0, col=coal_col) +
+  geom_step(size=2, alpha=0.9, show.legend = FALSE) +
   scale_x_continuous(trans='log10', breaks=c(1e2,1e3,1e4,1e5,1e6)) +
   scale_y_continuous(trans='log10', breaks=c(5e3,1e4,2e4), labels=scales::scientific) +
   annotation_logticks(sides="b") +
   annotation_logticks(sides="l") +
   scale_color_manual(values=cPal_rep) +
   guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
-  theme_classic(base_size = 20) +
+  theme_classic(base_size = 24) +
   labs(x="", y="", title=speciesPop, subtitle="dadi", colour="", fill="")+
   theme(axis.title.x=element_blank(),
         axis.text.x=element_blank(),
@@ -68,15 +66,15 @@ p0A<-ggplot(p0A_data, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
 
 p0B<-ggplot(p0B_data, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(25,xMax),ylim=c(y0Min,y0Max)) +
[…]

### 2020-05-05 src/plot_mig_rates_homsap.py

@@ -7,8 +7,8 @@ import subprocess
 
 
 
-dadi_infile = "homo_sapiens_mask_full_genome/Results/best_dadi_runs.txt"
-fsc_infile = "homo_sapiens_mask_full_genome/Results/best_fsc_runs.txt"
+dadi_infile = "homo_sapiens_full_genome/Results/best_dadi_runs.txt"
+fsc_infile = "homo_sapiens_full_genome/Results/best_fsc_runs.txt"
 simulated_genome_length = 2681441344.0
 
 outfile = "homsap_migration_rates.pdf"

### 2020-05-05 src/plot_AraTha_African2Epoch_1H18.R
updated figures & added asiepel's suggestions

@@ -2,10 +2,8 @@ library(ggplot2)
 library(cowplot)
 
 ## define variables
-infile="../data/AraTha_African2Epoch_1H18.txt"
+infile="../data/AraTha_African2Epoch_1H18_pop0_sizes.txt"
 outfile="../display_items/AraTha_African2Epoch_1H18.pdf"
-species="Arabidopsis thaliana"
-pop0_name="Africa"
 method1_name="SMC++"
 method2_name="stairway plot"
 method3_name="MSMC (n = 2)"
@@ -36,7 +34,6 @@ tab0D = subset(tab0,method == method4_id)
 tab0Y = subset(tab0,method == "census")
 tab0Z = subset(tab0,method == "coal")
 
-speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
 p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
   geom_step(data=tab0Z,aes(x=x, y=y),size=3.3,alpha=1.0, col=truth_color) +
@@ -48,9 +45,9 @@ p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   scale_color_manual(values=replicate_colors) +
   guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
   theme_classic(base_size = 24) +
-  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
-  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
-
+  labs(x="", y="", subtitle=method1_name, colour="", fill="")+
+  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5),
+        plot.subtitle = element_text(size = 28)) 
 
 p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
@@ -63,11 +60,11 @@ p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   scale_color_manual(values=replicate_colors) +
   guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
   theme_classic(base_size = 24) +
[…]

### 2020-05-05 src/plot_HomSap_OutOfAfricaArchaicAdmixture_5R19.R

@@ -6,7 +6,6 @@ infile_pop0="../data/HomSap_OutOfAfricaArchaicAdmixture_5R19_pop0_sizes.txt"
 infile_pop1="../data/HomSap_OutOfAfricaArchaicAdmixture_5R19_pop1_sizes.txt"
 infile_pop2="../data/HomSap_OutOfAfricaArchaicAdmixture_5R19_pop2_sizes.txt"
 outfile="../display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19.pdf"
-species="Homo sapiens"
 pop0_name="YRI"
 pop1_name="CEU"
 pop2_name="CHB"
@@ -62,7 +61,6 @@ tab2D = subset(tab2,method == method4_id)
 tab2Y = subset(tab2,method == "census")
 tab2Z = subset(tab2,method == "coal")
 
-speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
 p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(y0Min,y0Max)) +
   geom_step(data=tab0Z,aes(x=x, y=y),size=3.3,alpha=1.0, col=truth_color) +
@@ -77,7 +75,8 @@ p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
   theme(axis.title.x=element_blank(),
         axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
+        axis.ticks.x=element_blank(),
+        plot.subtitle = element_text(size = 28))
 
 p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(y0Min,y0Max)) +
@@ -96,7 +95,8 @@ p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
         axis.ticks.x=element_blank()) +
   theme(axis.title.y=element_blank(),
         axis.text.y=element_blank(),
-        axis.ticks.y=element_blank())
+        axis.ticks.y=element_blank(),
+        plot.subtitle = element_text(size = 28))
 
[…]

### 2020-05-05 src/plot_HomSap_constant_size.R

@@ -4,8 +4,6 @@ library(cowplot)
 ## define variables
 infile="../data/HomSap_constant_size.txt"
 outfile="../display_items/HomSap_constant_size.pdf"
-species="Homo sapiens"
-pop0_name="Constant size"
 method1_name="SMC++"
 method2_name="stairway plot"
 method3_name="MSMC (n = 2)"
@@ -33,10 +31,9 @@ tab0A = subset(tab0,method == method1_id)
 tab0B = subset(tab0,method == method2_id)
 tab0C = subset(tab0,method == method3_id)
 tab0D = subset(tab0,method == method4_id)
-#tab0Y = subset(tab0,method == "census")
+tab0Y = subset(tab0,method == "census")
 tab0Z = subset(tab0,method == "coal")
 
-speciesPop <- bquote(italic(.(species)) ~ (.(pop0_name)))
 p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
   geom_step(data=tab0Z,aes(x=x, y=y),size=3.3,alpha=1.0, col=truth_color) +
@@ -48,9 +45,9 @@ p0A<-ggplot(tab0A, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   scale_color_manual(values=replicate_colors) +
   guides(colour=guide_legend(keywidth=0.3,keyheight=0.21,default.unit="inch")) +
   theme_classic(base_size = 24) +
-  labs(x="", y="", title="", subtitle=method1_name, colour="", fill="")+
-  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5)) 
-
+  labs(x="", y="", subtitle=method1_name, colour="", fill="")+
+  theme(axis.text.x = element_text(angle = 45,hjust = 0.5,vjust = 0.5),
+        plot.subtitle = element_text(size = 28)) 
 
 p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   coord_cartesian(xlim=c(xMin,xMax),ylim=c(yMin,yMax)) +
@@ -63,11 +60,11 @@ p0B<-ggplot(tab0B, aes(x=x, y=y, group=interaction(rep), colour=rep)) +
   scale_color_manual(values=replicate_colors) +
[…]

### 2020-05-07 popSimManu.tex

@@
 of divergence times. However, these migration rate estimates are somewhat biased.
+%ACS: seems unnecessary and redundant
+%likely due to model misspecification.
 In contrast, \smcpp is much better at capturing the recent exponential
-growth in the CEU population, though it consistently underestimates divergence times
+growth in the CEU population, though the inferred population sizes are generally noisier.
+In addition, \smcpp consistently underestimates divergence times
 because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).
@@
 (Figure~\ref{fig:IM_popn_human}), \dadi and \fastsimcoal do reasonably well at
-reconstructing the simulated YRI history and estimating divergence times, but struggle with the more complex simulated CEU
+reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
 history, in large part because the methods assume constant population sizes.
 On the other hand, \smcpp does not have the same
-restrictions on its inferred history, and as a result does much better
-with the CEU history but tends to underestimate divergence times
-due to the assumption of no migration.
-The results for the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly}) are generally similar.
-In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, 
-differing only in how they calculate model expectations and optimize parameters.
+restrictions on its inferred history, and as a result does somewhat better
+with the CEU history but tends to overcomplicate the simple YRI history.
+The results for
+the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly})
+are generally similar.
+In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, differing only in how they calculate model expectations and optimize parameters.
 

### 2020-05-12 src/simulation_benchmark.py
Updated simulation benchmark results.
Based on commit d8ef28850eea778d418e1972990a3763963866d5 from stdpopsim.

@@ -49,7 +49,7 @@ def main():
     seed = 42
 
     data = {"species": [], "model": [], "cpu_time": [], "ram": [], "file_size": []}
-    for species_id in ["PonPyg", "HomSap", "DroMel", "AraTha"]:
+    for species_id in ["PonAbe", "HomSap", "DroMel", "AraTha"]:
         species = stdpopsim.get_species(species_id)
         # Get the shortest chromosome
         chrom = sorted(species.genome.chromosomes, key=lambda x: x.length)[0]

### 2020-05-12 popSimManu.tex
Update model table with results and fix references.

--- 2020-05-05 popSimManu.tex
@@
 and a constant chromosome-specific recombination rate.
-The times reported are for a single run on an Intel i5-7600 CPU.
+The times reported are for a single run on an Intel i5-7600K CPU.
 Computing resources required will vary widely depending on sample sizes, chromosome length,
@@
 and \texttt{msprime}.
-Examples of this validation for the AncientEurasia\_9K19 model \citep{AncientEurasia 9K19}
+Examples of this validation for the AncientEurasia\_9K19 model
+\citep{kamm2019efficiently}
 are shown in Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}.

### 2020-05-12 src/catalog_table.py

@@ -38,7 +38,7 @@ def main():
             if j < 2:
                 print("&")
         print("\\\\")
-        for species_id in ["HomSap", "DroMel", "AraTha", "PonPyg"]:
+        for species_id in ["HomSap", "DroMel", "AraTha", "PonAbe"]:
             species = stdpopsim.get_species(species_id)
             escaped_id = species.id.replace("_", "\\_")
             species_header = f"{escaped_id} (\\emph{{{species.name}}})"
@@ -53,7 +53,6 @@ def main():
                 file_size = float(row.file_size) / megabyte
                 bibtex = model.citations[0].fetch_bibtex()
                 bibtex_key = get_bibtex_key(bibtex, model_refs)
-                # bibtex_key = "ragsdale2019models"
                 escaped_id = model.id.replace("_", "\\_")
                 print("&")
                 print(escaped_id, end="& ")

### 2020-05-12 popSimManu.tex

@@
 This error was propagated from the \texttt{msprime} documentation,
-code which has been used in a number of other publications.
+where the model was presented as an illustrative example.
+A number of studies have been published using copies of
+this example code.
 

### 2020-05-12 popSimManu.tex
Fixup table refs and editing pass.

@@
 choosing pre-implemented models from a community-maintained catalog.
-The \stdpopsim catalog currently contains five species: humans,
+The \stdpopsim catalog currently contains six species: humans,
+\textit{Pongo abelii},
 \textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
 \textit{Escherichia coli}, and \textit{Canis familiaris}.
-For each species, the catalog contains information on our current understanding of
-the physical organization of its genome (e.g., chromosome structure),
-one or more inferred genetic maps (except for \textit{E.~coli}),
-and default population-level parameters (mutation rate and generation time estimates).
-The catalog also contains numerous published demographic models for the first three species.
-These available models and parameters are meant to represent the field's current understanding,
-and we intend for this resource to evolve as new results become available, and relevant
-existing models are added to \stdpopsim by the community.
+For each species, the catalog contains curated information on our current understanding of
+the physical organization of its genome,
+inferred genetic maps,
+population-level parameters (e.g., mutation rate and generation time
+estimates), and published demographic models.
+These models and parameters are meant to represent the field's current understanding,
+and we intend for this resource to evolve as new results become available, and
+other existing models are added to \stdpopsim by the community.
 We have implemented both a command line interface and a simple Python API
@@
 distributed open source model, with strong procedures in place
-to continue its growth and maintain its quality.
+to continue its growth and maintain quality.
 Importantly, we developed rigorous quality control methods to ensure that we have
@@
 %The first contribution of the PopSim consortium is
-The \stdpopsim library is a
+The \stdpopsim library is
 a community-maintained collection of empirical genome data and population genetics simulation models,
-with a structure depicted in Figure \ref{fig:cartoon}.
+illustrated in Figure \ref{fig:cartoon}.
 The package centers on a catalog of genomic information and demographic models
@@
 command line interface, allowing users with minimal experience direct access to
-state-of-the-art simulations. Simulations are output in the ``tree sequence''
+state-of-the-art simulations. Simulations are output in the ``succinct tree sequence''
 format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
@@
 Data is currently available for humans,
+\textit{P.~abelii},
 \textit{D.~melanogaster}, \textit{A.~thaliana},
-\textit{Escherichia coli}, and \textit{Canis familiaris}.
+\textit{E.~coli}, and \textit{C.~familiaris}.
 A species definition consists of two key elements.
@@
 deCODE~\citep{kong2010fine} genetic maps for humans;
+the \cite{nater2017morphometric} maps for \textit{P.~abelii};
 the \cite{salome2011recombination} map for \textit{A.~thaliana};
@@
 development and quality-control process for these models.)
-
-The current demographic models in the \stdpopsim catalog are shown in \autoref{tab:catalog}.
-These range from
-simple, single population histories \cite[e.g.,][]{sheehan2016deep},
-to complex models which include population splitting, migration, and archaic
-admixture \cite[e.g.,][]{ragsdale2019models}.
-In addition to models from Table \autoref{tab:catalog},
-the PopSim Consortium has demographic models for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli} in development at time of writing.
-
 \renewcommand{\arraystretch}{1.2}
@@
 For each model, we report the CPU time, maximum memory usage and the
-size of the output \tskit file, as simulated using the \texttt{msprime} simulation engine.
+size of the output \tskit file, as simulated using the \texttt{msprime}
+simulation engine (version 0.7.4).
 In each case, we simulate 100 samples
@@
 
-Currently, {\em Homo sapiens} has the largest number of population models in
-\stdpopsim (see Table \ref{tab:catalog}).
-These models include: a simplified version of the \cite{tennessen2012evolution}
+The current demographic models in the \stdpopsim catalog are shown in \autoref{tab:catalog}.
+% These range from
+% simple, single population histories \cite[e.g.,][]{sheehan2016deep},
+% to complex models which include population splitting, migration, and archaic
+% admixture \cite[e.g.,][]{ragsdale2019models}.
+{\em Homo sapiens} currently has the richest selection of population models.
+These include: a simplified version of the \cite{tennessen2012evolution}
 model with only the African population specified (expansion from the ancestral
@@
 (Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
-These plans would not be possible without recent important work on SLiM \citep{haller2019slim},
-as well as current work on \texttt{msprime}.
 Some other important processes are more challenging to model with current simulation software,
@@
 Further details of our QC process can be found in our developer documentation
-(\url{https://stdpopsim.readthedocs.io/en/latest/development.html#}).
+(\url{https://stdpopsim.readthedocs.io/en/latest/development.html}).
 
 The possibility for error and the importance of careful qualty control
-was inadvertantly illustrated during our own development process:
-during the final revisions of this paper, we noticed that the
+was illustrated very clearly during our own development process:
+while carrying out the final revisions of this paper, we noticed that the
 OutOfAfrica\_3G09 model \citep{gutenkunst2009inferring}
-had not in fact gone through the QC process.
-The subsequent QC process revealed that in fact our implementation
+had not gone through our QC process.
+The subsequent QC revealed that our implementation was in fact
 slightly wrong---migration rates had not been set to zero to the European population in the longest-ago time period
@@
 A number of studies have been published using copies of
-this example code.
+this erroneous example code.
 
@@
 \section*{Resource availability}
-The version 0.1 release of \stdpopsim is available for download on Github:
-\url{https://github.com/popsim-consortium/stdpopsim/releases}.
+The \stdpopsim package is available for download on the Python Package Index:
+\url{https://pypi.org/project/stdpopsim/}.
 Documentation for the project can be found here:

### 2020-05-13 response-to-reviewers.md
edited responses document

--- 2020-04-29 response-to-reviewers.md
@@
-Dear Graham,
+Dear Graham and the Editorial Board of *eLife*,
 
-[Introductory text]
+Please find attached our revised manuscript entitled, 
+"A community maintained standard library of population genetic models"
+for consideration for publication.
 
-Please see below a point-by-point response to the comments.
+We thank you and the reviewers of our manuscript for your enthusiasm for our project
+and the constructive criticisms that you have raised. As you will find below, we have tried
+our level-best to incorporate all of the changes suggested in the reviews. 
+While that is so, in a few places we have decided that suggested changes were beyond
+the scope of the current manuscript and leave them for future work.
+
+What follows below is a point-by-point response to the comments.
 
@@
 We fully agree that the language in the original manuscript was not careful enough in conveying uncertainty about the models and parameters we discuss.
-To address this, we have made edited the text throughout the manuscript trying to better convey this uncertainty.
+We have made edited the text throughout the manuscript to address this.
 See our response to Reviewer 1 below for examples.
@@
 
-Consequently, we decided to add an example for possible use of SLiM in neutral simulations (see "simulation engines" section in the results and Supplementary Figures S1 and S2).
+Instead, we have decided to add an a section validating the use of SLiM in stdpopsim neutral simulations (see "simulation engines" section in the results and Supplementary Figures S1 and S2).
 Our focus there was to show that under neutral simulations, SLiM produced data that were consistent with the coalescent simulations of msprime.
-To expand on this fundamental point, we also examined the influence of population size down-scaling, which is a common practice used in forward simulations.
+To expand on this fundamental point, we also examined the influence of population size down-scaling, which is a common practice used in forward simulations, and indeed often crucial for tractable compute times.
 While this is by no means a comprehensive investigation into this issue, we do believe that it demonstrates the use of SLiM in future applications that will also simulate selection.
@@
 
-Good idea -
+Thank you for this good idea -
 we have expanded the section of the discussion ("next steps")
 with more detail on our future plans.
-We're trying to walk the line between being forward-looking and promising too much.
+While that is so we are wary of promising too much at this point.
 
@@
 We have added an additional paragraph to the discussion exploring various factors that may affect choice of inference method.
-These include the data required by the method, the type of model, and software engineering.
+These include the data required by the method, the type of model, and the implementation.
 We also provide a qualitative comparison between the performance of methods, as shown in our limited analysis.
@@
 
-(The reviewer's suggestion is good, but we think that "current understanding"
+<!-- (The reviewer's suggestion is good, but we think that "current understanding"
 conveys roughly the same meaning as "the latest reference genome" in less words,
 and without making promises about genome assemblies.)
+
+ADK commenting out-- don't think we need this
+-->
 
@@
 
+Moreover we believe that the issues feature on Github provides us with the current utility we need along these lines.
+
 > 3.
@@
 For these reasons, we feel that masking using a simple recombination rate threshold is the most appropriate approach.
+We note that their is however a substantial correlation between recombination rate and mapability (by any of the measures mentioned above) wherein the notoriously difficult portions of genomes to assemble are generally in regions of low recombination.
  We agree that there may indeed be better ways to mask, which is why we allow users to mask their simulations as they see fit.
@@
 Mask files are not currently a component of `stdpopsim` proper, rather they were implemented separately from running `stdpopsim`, for the sole purpose of comparing demographic inference methods.
+All of the masks that we have used are available on the `Analysis` repo that is associated with this manuscript.
 Users who download `stdpopsim` will always be simulating raw and unmasked tree sequence files, to which they can apply any variety of masks *post hoc*, if they so choose.
@@
 
-True, but the details are rather pedantic, and we provide a link
-to our documentation where we describe the system in detail.
+True, but the details are rather pedantic and we feel not material for the  manuscript.
+We provide a link to our extensive documentation where we describe the system in detail.
 
@@
 
-We've replaced "bespoke" with "coded from scratch", which we think is more descriptive,
-and added a parenthetical note about command-line calls in supplementary material (which is a good point).
+We've replaced "bespoke" with "coded from scratch", which we think is more descriptive.
 We hope that the other points are made elsewhere.
@@
 
-We agree that the orientation of this figure is confusing, and so have changed it as suggested.
+We agree that the orientation of this figure was confusing, and so have changed it as suggested.
 In addition, we had previously plotted divergence times here as vertical lines, however we felt that plotting it this way made the figure hard to read since there was a great deal of overlap between these lines and the N(t) lines on the figure, which was was especially difficult to read when including multiple replicates.
@@
 There is not a command-line flag for obtaining the coalescence rates,
-but this is easily obtainable in python.
+but this is easily obtainable through the python API.
 We've added a paragraph to the supplement explaining what methods to use.
@@
 
-Thank YOU! We hope you are right.
+Thank you! We hope you are right.
 
@@
 This new analysis uses SLiM in neutral simulations to validate that it produces data that is consistent with data simulated under the coalescent.
-Furthermore, we provide a rudamentary examination of the effects of population size down-scaling, which is a common practice used in forward simulations such as SLiM.
+Furthermore, we provide a rudimentary examination of the effects of population size down-scaling, which is a common practice used in forward simulations such as SLiM.
 
@@
 
-This is an interesting idea. We did discuss adding a suite of generic models to stdpopsim that do not necessarily capture demographic parameters from a certain species,
-but do reflect general trends observed in inference. The "zigzag" model could definitely be one of these.
-This did not make it to this release, but we will note this for the next one.
-the reason why the zigzag model is currently defined as a human model is that its
+This is an interesting idea. We have added suite of generic models to stdpopsim that do not necessarily capture demographic parameters from a certain species,
+but do reflect general trends observed in inference.
+The "zigzag" model could definitely be one of these, and indeed was initially implemented as a generic.
+However the reason we settled on the zigzag model is being defined as a human model is that its
 effective population size values are taken from (or at least inspired by) values inferred from human genomes.
+To see our internal discussion of these issues from September 2019 please see this [closed issue](https://github.com/popsim-consortium/stdpopsim/issues/106).
 
@@
 
-We see why this is confusing.
+This is a good catch! We see why this is confusing.
 We introduced this notation (R, C, and n) to make it completely clear
@@
 We made the following change:  
-"Accuracy is mixed among methods when doing inference on simulated data from these D. melanogaster
-and A. thaliana models, and generally worse than what we
+"Accuracy is mixed among methods when doing inference on simulated data from these _D. melanogaster_
+and _A. thaliana_ models, and generally worse than what we
 observe for simulations of the human genome."

### 2020-05-14 popSimManu.tex
ADK final(?) pass on the manuscript

@@
 The \stdpopsim catalog currently contains six species: humans,
-\textit{Pongo abelii},
+\textit{Pongo abelii}, \textit{Canis familiaris},
 \textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
-\textit{Escherichia coli}, and \textit{Canis familiaris}.
+and \textit{Escherichia coli}.
 For each species, the catalog contains curated information on our current understanding of
@@
 Data is currently available for humans,
-\textit{P.~abelii},
+\textit{P.~abelii}, \textit{C.~familiaris},
 \textit{D.~melanogaster}, \textit{A.~thaliana},
-\textit{E.~coli}, and \textit{C.~familiaris}.
+and \textit{E.~coli}.
 A species definition consists of two key elements.
@@
 the \cite{nater2017morphometric} maps for \textit{P.~abelii};
+the \cite{campbell2016pedigree} map for \textit{C.~familiaris};
 the \cite{salome2011recombination} map for \textit{A.~thaliana};
-the \cite{comeron2012many} map for \textit{D.~melanogaster},
-and the \cite{campbell2016pedigree} map for \textit{C.~familiaris}.
+and the \cite{comeron2012many} map for \textit{D.~melanogaster}. 
 Adding further maps to the library is straightforward.
@@
 %DRS: do we want to include these generic models in Table 1?
+%ADK: not a bad idea...
 
@@
 this practice is effective for many purposes and widely employed.
+%ADK: I think we need a citaiton above for scaling. Maybe https://doi.org/10.1534/genetics.113.156935
 
@@
 
-% ACS: there was indeed a lot of redundancy here.  DRS improved it a lot, but I've tried to do a bit more
 \subsubsection*{ Single-population demographic models.}
@@
 under a multi-population model,
-is not expected to recover the actual simulated population sizes,
+is not expected to recover the actual simulated population sizes
 because of model misspecification.
@@
 Figure \ref{fig:census_vs_IRC}).
-%DRS: trying to cut down on the redundancy between the coal rate stuff here and the previous paragraph
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 As most method development for population genetics has been focused on human
-data, it is of consequence to ask how such methods might perform in non-human
+data, it is important to ask how such methods might perform in non-human
 genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from the African3Epoch\_1S16
@@
 Although these results do not represent an exhaustive benchmarking,
-we have highlighted some of the strengths and weaknesses of these methods.
+we have begun to highlight some of the strengths and weaknesses of these methods.
 Future work should build on these results and undertake more in-depth comparisons
@@
 Using a parametric approach yields less noisy results, but a model that is too simple may not capture important demographic events (Figures~\ref{fig:IM_popn_human} and~\ref{fig:two_popn_fly}),
-and other forms of model misspeficification may also produce undesireable behavior.
+and other forms of model misspecification may also produce undesirable behavior.
 From a software engineering perspective, methods also differ in their ease of installation and use.
@@
 Some of the additions that we are enthusiastic to add are:
-selection (including distributions of fitness effects, maps of functional elements, and both single and recurrent hitchhiking events),
+selection (including distributions of fitness effects, maps of functional elements, 
+both single and recurrent hitchhiking events, and selection on polygenic traits),
 gene conversion, mutation models (rate heterogeneity),
@@
 % note: this is addressing reviewer comments about model uncertainty
-In particular, although the included demographic histories are some of the most widely used models for our current set of species,
+We wish to emphasize that although the included demographic histories are some of the
+most widely used models for our current set of species,
 we anticipate the set of available models to expand
@@
 \Stdpopsim's framework is sufficiently general that more realistic population models
-will be easily incorporated, once they are published.
-%DRS: trying to make this part a little less subjective. We were kind of painting the picture that
-%we have all the best/useful models and everything currently published but not included is not useful,
-%and I don't think this is justified.
-Our current set may also omit other widely used models that can be readily incorporated into \stdpopsim.
-Additional complications such as changing
-genome builds as assemblies improve must also be incorporated.
-As the collection of models and species grows, we hope to
-incorporate crowd-sourced community input
-to highlight the most appropriate models and genome annotations.
+will be easily incorporated, as they are published.
+
+%ADK: I cut this off here-- I think thais is a fine ending
+
+%Our current set of models is not exhaustive at this point however,
+%and othercan be readily incorporated into \stdpopsim.
+%Additional complications such as changing
+%genome builds as assemblies improve must also be incorporated.
+%As the collection of models and species grows, we hope to
+%incorporate crowd-sourced community input
+%to highlight the most appropriate models and genome annotations.
 %DRS: not in love with the word "annotations" here as some might think we are talking gene annotations (which we will be eventually but not yet), but "details" is too vague.
@@
 The subsequent QC revealed that our implementation was in fact
-slightly wrong---migration rates had not been set to zero to the European population in the longest-ago time period
+slightly wrong---migration rates had not been set to zero to the European population in the most ancient time period
 when there should have only been a single population.
@@
 Dmitri Petrov, and Sohini Ramachandran.
+In addition we would like to thank NESCENT and Matt Hahn, Victoria Sork, and Michael Whitlock for organizing a 2014
+catalysis meeting in which many of the goals of this effort were first laid out. 
 CCK and KEL were funded under NIH Award R35GM119856.

### 2020-05-14 response-to-reviewers.md
plr adjustments

@@
 the scope of the current manuscript and leave them for future work.
+
+An additional change that has occurred since the first submission of the manuscript
+is that the Gutenkunst 2009 model hadn't been through QC, and had a bug.
+This has been corrected, resulting in mostly cosmetic changes to the figures and text
+(inferred population histories are slightly better than before; see Figure 4);
+but we have also explained the situation,
+as an example of the utility of QC and the resource more generally.
 
@@
 The first quoted passage now reads:
-"For each organism, the catalog contains information on our current understanding of
-the physical organization (e.g., chromosome structure) of its genome,
-one or more inferred genetic maps (except for \textit{E.~coli}),
-default population-level parameters (mutation rate and generation time estimates).
-The catalog also contains numerous published demographic models for the first three species.
-These available models and parameters are meant to represent the field's current understanding,
-and we intend for this resource to evolve as new results become available."
+"For each species, the catalog contains curated information on our current understanding of
+the physical organization of its genome, inferred genetic maps,
+population-level parameters (e.g., mutation rate and generation time estimates),
+and published demographic models.
+These models and parameters are meant to represent the field's current understanding,
+and we intend for this resource to evolve as new results become available, and
+other existing models are added to stdpopsim by the community."
 
 The second passage now reads:
-" Firstly, the library defines some basic information about our current understanding of each 
+"Firstly, the library defines some basic information about our current understanding of each
 species' genome, including information about chromosome
@@
 which model observed heterogeneity in recombination rate along chromosomes."
-
-<!-- (The reviewer's suggestion is good, but we think that "current understanding"
-conveys roughly the same meaning as "the latest reference genome" in less words,
-and without making promises about genome assemblies.)
-
-ADK commenting out-- don't think we need this
--->
 
@@
 correctly implemented the models as described in their original publication
-and provide documented methods for others to contribute new models."
+and provided documented methods for others to contribute new models."
 
@@
 > a.
-Shouldn't the approach instead be to drop "low complexity" regions (e.g., as defined by an excessive number of "N"'s in the reference, low mapability scores, or via tools like RepeatMasker?).
+Shouldn't the approach instead be to drop "low complexity" regions (e.g., as defined by an excessive number of "N"'s in the reference, low mappability scores, or via tools like RepeatMasker?).
 This part of the pipeline seems open to refinement.
@@
 While we agree with the reviewer that different masking approaches might better reflect the masking done on real genomic data, at this time the best way to mask remains an open question, and we feel that a nuanced analysis of masking is outside the scope of this paper.
-For these reasons, we feel that masking using a simple recombination rate threshold is the most appropriate approach.
-We note that their is however a substantial correlation between recombination rate and mapability (by any of the measures mentioned above) wherein the notoriously difficult portions of genomes to assemble are generally in regions of low recombination.
+For these reasons, we feel that masking using a simple recombination rate threshold is a reasonable approach.
+We note that there is however a substantial correlation between recombination rate and mappability (by any of the measures mentioned above) wherein the notoriously difficult portions of genomes to assemble are generally in regions of low recombination.
  We agree that there may indeed be better ways to mask, which is why we allow users to mask their simulations as they see fit.
@@
 Mask files are not currently a component of `stdpopsim` proper, rather they were implemented separately from running `stdpopsim`, for the sole purpose of comparing demographic inference methods.
-All of the masks that we have used are available on the `Analysis` repo that is associated with this manuscript.
+All of the masks that we have used are available on the `analysis` repository that is associated with this manuscript.
 Users who download `stdpopsim` will always be simulating raw and unmasked tree sequence files, to which they can apply any variety of masks *post hoc*, if they so choose.
+We are currently making plans, however, to incorporate some version of this process into a future
+`stdpopsim` release.
 
@@
 
-Good point - we've clarified the nature of the resource,
+Good point - in the revision we've clarified the nature of the resource,
 but haven't included the URL since googling "stdpopsim" features the correct page as the top hit.
@@
 We've replaced "bespoke" with "coded from scratch", which we think is more descriptive.
-We hope that the other points are made elsewhere.
+We think that the other points are made elsewhere.
 
@@
 
-This is an interesting idea. We have added suite of generic models to stdpopsim that do not necessarily capture demographic parameters from a certain species,
-but do reflect general trends observed in inference.
-The "zigzag" model could definitely be one of these, and indeed was initially implemented as a generic.
-However the reason we settled on the zigzag model is being defined as a human model is that its
-effective population size values are taken from (or at least inspired by) values inferred from human genomes.
-To see our internal discussion of these issues from September 2019 please see this [closed issue](https://github.com/popsim-consortium/stdpopsim/issues/106).
+We considered both of these suggestions, and here's why we've left it the way it is.
+Most of the columns of the table don't apply to generic models,
+so it seems strange to include them there.
+The "zigzag" model could definitely be a generic model,
+and indeed was initially implemented as such.
+However, the reason we settled on the zigzag model is being defined as a human model is that its
+effective population size values are taken from (or at least inspired by) values inferred from human genomes: as implemented, it is not as "generic" as it might seem.
 
->
 > 3) On page 8 the authors set up notation for the number of replicates (R), number of chromosomes (C), and sample size (n), but don't seem to use it afterward (or use it inconsistently).
@@
 "While there is variation in accuracy among methods, populations, and individual replicates,
-the methods generally infer back a good estimate
-of the true effective population sizes of the simulations --
-inferred values are mostly within a factor of two of the truth,
-and many of the methods infer a bottleneck at the correct time."
+the methods generally produce a good estimate of the true effective population sizes of the
+simulations, with inferred values mostly within a factor of two of the truth, and most
+methods inferring a bottleneck at approximately the correct time."
 
@@
 
-We meant to say that the inferred history has more population size changes
-than the truth did;
-we've change this to "overcomplicate the simple YRI history".
+We meant to say that the inferred history has more population size changes than the truth did;
+but this phrase has been removed.
 

### 2020-05-14 response-to-reviewers.md
put back in github issue link

@@
 and indeed was initially implemented as such.
-However, the reason we settled on the zigzag model is being defined as a human model is that its
-effective population size values are taken from (or at least inspired by) values inferred from human genomes: as implemented, it is not as "generic" as it might seem.
+However, the reason we settled on the zigzag model is being defined as a human model 
+(see discussion at https://github.com/popsim-consortium/stdpopsim/issues/106)
+is that its effective population size values are taken from (or at least inspired by) values inferred from human genomes: as implemented, it is not as "generic" as it might seem.
 

### 2020-05-15 response-to-reviewers.md
closes #203

@@
 
-Instead, we have decided to add an a section validating the use of SLiM in stdpopsim neutral simulations (see "simulation engines" section in the results and Supplementary Figures S1 and S2).
+Instead, we have decided to add a section validating the use of SLiM in stdpopsim neutral simulations (see "simulation engines" section in the results and Supplementary Figures S1 and S2).
 Our focus there was to show that under neutral simulations, SLiM produced data that were consistent with the coalescent simulations of msprime.
@@
 We thus plotted the simulated population sizes against the inverse coalescence rate for the model used in Figure 2, and found that they were near identical.
-As a result, we decided to show this comparison in Supplementary Figure S1 and discuss this issue in the text without adding the simulated population sizes to the main figures (as they didn't appear to add much).
+As a result, we decided to show this comparison in Supplementary Figure S3 and discuss this issue in the text without adding the simulated population sizes to the main figures (as they didn't appear to add much).
 Note that in the multi-population analyses, the inverse coalescence rate is less appropriate, because some of the inference methods used in these analyses model gene flow, and should therefore be able to recover the true simulated population sizes.

### 2020-05-15 popSimManu.tex
cleaned up single data is instance
closes #203

@@
 and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).
-
 Given the genome data and simulation model descriptions defined within the
@@
 of the key quantitative data needed to simulate a given species.
-Data is currently available for humans,
+Data are currently available for humans,
 \textit{P.~abelii}, \textit{C.~familiaris},
@@
 
-%DRS: do we want to include these generic models in Table 1?
-%ADK: not a bad idea...
-
 To guarantee reproducibility,
@@
 on simulations of \textit{D.~melanogaster} or \textit{A.~thaliana} populations,
-in which linkage disequilibrium is reduced (Figures~\ref{fig:n_t_sheehan} \& \ref{fig:n_t_AraTha}).
+in which linkage disequilibrium is reduced (Figures~\ref{fig:n_t_sheehan} and \ref{fig:n_t_AraTha}).
 In simulations of two human populations
@@
 will be easily incorporated, as they are published.
-
-%ADK: I cut this off here-- I think thais is a fine ending
-
-%Our current set of models is not exhaustive at this point however,
-%and othercan be readily incorporated into \stdpopsim.
-%Additional complications such as changing
-%genome builds as assemblies improve must also be incorporated.
+Additional aspects of the framework,
+such as genome builds, will also continue to change
+as improvements are made to our understanding of genome structure.
+
 %As the collection of models and species grows, we hope to

### 2020-05-15 popSimManu.tex

@@
 For this model,
-down-scaling factors of up to $50$ produce patterns of both diversity and linkage disequilibrium that are
-very similar to those observed under the coalescent (i.e., \texttt{msprime}), when both are simulated using the HapMapII\_GRCh37 recombination map
-(Figure \ref{fig:slim_val_map}),
-although the distribution of Tajima's $D$ differs slightly when scaling down by a factor of $50$.
-%likely due to the subtle differences produced by down-scaling.
-Interestingly, the effects of down-scaling are more prominent and extend to other sequence-based statistics
-when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
+down-scaling factors of up to $10$ produce patterns of both diversity and linkage disequilibrium that are
+indistinguishable from those observed under the coalescent (i.e., \texttt{msprime}).
+Scaling down by a factor of $50$ does appear to modify the distribution of these sequence statistics.
+Interestingly, the apparent difference between distributions is somewhat larger when simulating using 
+a uniform recombination rate (Figure \ref{fig:slim_val_nomap}), likely due to the lower variation in
+the values of these statistics.
 Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied.
@@
 show runtimes for generating each simulation replicate. Data were generated by simulating 100 replicates
-of human chromosome 22 under the AncientEurasia\_9K19 model \citep{kamm2019efficiently} and using the
-HapMapII\_GRCh37 genetic map \citep{international2007second}. From top to bottom we show results
+of human chromosome 22 under the AncientEurasia\_9K19 model \citep{kamm2019efficiently} using the
+HapMapII\_GRCh37 genetic map \citep{international2007second}. 12 samples were drawn from each population 
+(excluding basal Eurasians). From top to bottom we show results
 using three scaling factors for the population sizes: Q=1, Q=10, and Q=50.
+Kolmogorov-Smirnov 2-sample test statistics (\textit{D}) and p-values are shown, testing the null hypothesis that
+the quantiles were drawn from the same continuous distribution.
 }
@@
 show runtimes for generating each simulation replicate. Data were generated by simulating 100 replicates
-of human chromosome 22 under the AncientEurasia\_9K19 model \citep{kamm2019efficiently} and using a uniform
-rate of recombination across the chromosome. From top to bottom we show results
+of human chromosome 22 under the AncientEurasia\_9K19 model \citep{kamm2019efficiently} using a 
+uniform rate of recombination across the chromosome. 12 samples were drawn from each population 
+(excluding basal Eurasians). From top to bottom we show results
 using three scaling factors for the population sizes: Q=1, Q=10, and Q=50.
+Kolmogorov-Smirnov 2-sample test statistics (\textit{D}) and p-values are shown, testing the null hypothesis that
+the quantiles were drawn from the same continuous distribution.
 }

### 2020-06-19 popSimManu.tex
formatting changes requested by eLife

@@
 \usepackage{booktabs}
+\usepackage{caption}
 \usepackage[binary-units]{siunitx}
@@
         \setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}%
-        \renewcommand{\theHfigure}{S\thefigure}
+        \captionsetup[figure]{name=Appendix Figure}
+        \renewcommand{\figurename}{Appendix Figure} 
+        %\renewcommand{\thefigure}{S\arabic{figure}}%
+        %\renewcommand{\theHfigure}{S\thefigure}
      }
@@
         \setcounter{figure}{0}
+        \captionsetup[figure]{name=Figure}
         \renewcommand{\thefigure}{\arabic{figure}}%
@@
 \citep{kamm2019efficiently}
-are shown in Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}.
+are shown in Appendix Figure \ref{fig:slim_val_map} and Appendix Figure \ref{fig:slim_val_nomap}.
 For this model,
@@
 Interestingly, the apparent difference between distributions is somewhat larger when simulating using 
-a uniform recombination rate (Figure \ref{fig:slim_val_nomap}), likely due to the lower variation in
+a uniform recombination rate (Appendix Figure~\ref{fig:slim_val_nomap}), likely due to the lower variation in
 the values of these statistics.
@@
 the inverse coalescence rates are very close to the simulated population sizes;
-Figure \ref{fig:census_vs_IRC}).
+Appendix Figure \ref{fig:census_vs_IRC}).
 While there is variation in accuracy among methods, populations, and individual replicates,
@@
 Using \stdpopsim, we can readily compare performance on this benchmark to
-that based on a different model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
+that based on a different model of human history. In Appendix Figure \ref{fig:n_t_gutenkunst} we show estimates of
 $N(t)$ from simulations using the same physical and genetic maps, but from the OutOfAfrica\_3G09
@@
 We show results of such an
-experiment in Figure \ref{fig:n_t_HomSap_constant}.
+experiment in Appendix Figure \ref{fig:n_t_HomSap_constant}.
 All methods recover population size within a factor of two of the simulated values, however
@@
 model, originally estimated from an African sample of \textit{D.~melanogaster} \citep{sheehan2016deep},
-and Figure \ref{fig:n_t_AraTha} shows estimates from simulations of \textit{A.~thaliana}
+and Appendix Figure \ref{fig:n_t_AraTha} shows estimates from simulations of \textit{A.~thaliana}
 under the African2Epoch\_1H18 model originally inferred by \cite{huber2018gene}.
@@
 simulated YRI population size history. Moreover, these approaches allow
-migration rates to be estimated (Figure \ref{fig:homsap_mig_rates}), and lead to more accurate inferences
+migration rates to be estimated (Appendix Figure \ref{fig:homsap_mig_rates}), and lead to more accurate inferences
 of divergence times. However, these migration rate estimates are somewhat biased.
@@
 Again, we can extend this analysis to other taxa and examine the performance of these methods
-for a two-population model of \textit{D.~melanogaster}. Figure
+for a two-population model of \textit{D.~melanogaster}. Appendix Figure
 \ref{fig:two_popn_fly} shows inference results using data simulated under
@@
 Africa from which a European population splits off following a bottleneck, with no
-post-divergence gene flow between the African and European population (Figure \ref{fig:two_popn_fly}A).
+post-divergence gene flow between the African and European population (Appendix Figure \ref{fig:two_popn_fly}A).
 Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
@@
 and \fastsimcoal both do reasonably well at correctly inferring the absence of
-migration (Figure \ref{fig:dromel_mig_rates}). In contrast, the inferred
+migration (Appendix Figure \ref{fig:dromel_mig_rates}). In contrast, the inferred
 demographic parameters from \smcpp are more noisy, though in some cases better
@@
 which tends to over-estimate the sizes of oscillations
-(Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}).
+(Figure~\ref{fig:n_t_ragsdale} and Appendix Figure~\ref{fig:n_t_gutenkunst}).
 In contrast, \stairwayplot outperforms the sequence-based methods
 on simulations of \textit{D.~melanogaster} or \textit{A.~thaliana} populations,
-in which linkage disequilibrium is reduced (Figures~\ref{fig:n_t_sheehan} and \ref{fig:n_t_AraTha}).
+in which linkage disequilibrium is reduced (Figure~\ref{fig:n_t_sheehan} and Appendix Figure~\ref{fig:n_t_AraTha}).
 In simulations of two human populations
@@
 due to the assumption of no migration.
-The results for the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly}) are generally similar.
+The results for the two-population \textit{D.~melanogaster} model (Appendix Figure~\ref{fig:two_popn_fly}) are generally similar.
 In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data,
@@
 \dadi and \fastsimcoal require a pre-specified parametric model, unlike \MSMC, \smcpp, and \stairwayplot.
-Using a parametric approach yields less noisy results, but a model that is too simple may not capture important demographic events (Figures~\ref{fig:IM_popn_human} and~\ref{fig:two_popn_fly}),
+Using a parametric approach yields less noisy results, but a model that is too simple may not capture important demographic events (Figure~\ref{fig:IM_popn_human} and Appendix Figure~\ref{fig:two_popn_fly}),
 and other forms of model misspecification may also produce undesirable behavior.
@@
 In addition, the ability of \stdpopsim to generate data with and without significant features, such
-as a genetic map or population-size changes (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
+as a genetic map or population-size changes (e.g., Appendix Figure~\ref{fig:n_t_HomSap_constant}), allows
 investigation of the failure modes of popular methods.
@@
 above scenarios is warranted, given our preliminary findings using neutral simulations
-(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
+(Appendix Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
 Some other important processes are more challenging to model with current simulation software,
@@
 workflows \citep{koster2012snakemake} that perform demographic inference on
-tree sequence output from our package using a few common software packages (see Figure \ref{fig:n_t_dag} for an example workflow).
+tree sequence output from our package using a few common software packages (see Appendix Figure \ref{fig:n_t_dag} for an example workflow).
 Our choice of \texttt{Snakemake} allows complete reproducibility of the
@@
 DroMel model is a two population model (Africa and Europe) with no post-divergence
-migration and constant population sizes (Figure \ref{fig:two_popn_fly}).
+migration and constant population sizes (Appendix Figure \ref{fig:two_popn_fly}).
 
@@
 Importantly, this meant that for both species, the
-fitted model did not match the simulated model (Figures \ref{fig:IM_popn_human} and \ref{fig:two_popn_fly}).
+fitted model did not match the simulated model (Figure \ref{fig:IM_popn_human} and Appendix Figure \ref{fig:two_popn_fly}).
 In the HomSap case, we therefore performed inference solely on the Africa
@@
 To validate our inference approach, we also conducted
-inference on a generic IM model that was identical to the model used for inference (Figure \ref{fig:generic_IM}).
+inference on a generic IM model that was identical to the model used for inference (Appendix Figure \ref{fig:generic_IM}).
 
@@
 Estimates from the highest log-likelihood (out of 10 runs) for each simulation replicate
-are shown in Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C.
+are shown in Figure~\ref{fig:IM_popn_human}C and Appendix Figure~\ref{fig:two_popn_fly}C.
 
@@
 for each simulation replicate alongside the simulated population sizes
-(Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C).
+(Figure~\ref{fig:IM_popn_human}C and Appendix Figure~\ref{fig:two_popn_fly}C).
 Here, unlike the single-population workflow,
@@
 \beginsupplement
-\section*{Supplemental Figures}
+\section*{Appendix Figures}
 \begin{figure}
